### 2014-10-22 manuscript/cordiv.tex
updated figures + r4sJC + manuscript

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{footnote}
\usepackage{rotating}
\usepackage{slashbox}
\usepackage{afterpage}
\usepackage{float}
\citestyle{apj}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%force setting figures at the top of the page:
    \makeatletter
    \setlength{\@fptop}{0pt}
    \makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  personal abbreviations and macros
%    the following package contains macros used in this document:
%\input econfmacros.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textwidth=6.9in  \textheight=9.6in
%%  Adjust these for your printer:
\leftmargin=-5in   \topmargin=-0.8in
\hoffset = -0.95in  \voffset = 0pt
\marginparwidth = 0pt
%% preprint number data:
%\newcommand\pubnumber{Article 14 in eConf C1304143}
\newcommand\pubnumber{}
\newcommand\pubdate{\today}

%%  address and funding acknowledgement data:
\def\affiliation{Department of Physics \\
The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
%\def\support{\footnote{Work supported by the Office of Science,Kingdom of the Two Sicilies, under contract OSS--32456.}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   document style macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gtrsim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
\newcommand{\liso}{L_{iso}}
\newcommand{\eiso}{E_{iso}}
\newcommand{\epkz}{E_{p,z}}
\newcommand{\durz}{T_{90,z}}
\newcommand{\pbol}{P_{bol}}
\newcommand{\sbol}{S_{bol}}
\newcommand{\epk}{E_{p}}
\newcommand{\dur}{T_{90}}
\newcommand{\pph}{P_{50-300}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\Title#1{\begin{center} {\Large #1 } \end{center}}
\def\Author#1{\begin{center}{ \sc #1} \end{center}}
\def\Address#1{\begin{center}{ \it #1} \end{center}}
\def\andauth{\begin{center}{and} \end{center}}
\def\submit#1{\begin{center}Submitted to {\sl #1} \end{center}}
\newcommand\pubblock{\rightline{\begin{tabular}{l} \pubnumber\\
         \pubdate  \end{tabular}}}
\newenvironment{Abstract}{\begin{quotation}  }{\end{quotation}}
%\newenvironment{Presented}{\begin{quotation} \begin{center}
%             PRESENTED AT\end{center}\bigskip
%      \begin{center}\begin{large}}{\end{large}\end{center} \end{quotation}}
\def\Acknowledgements{\bigskip  \bigskip \begin{center} \begin{large}
             \bf ACKNOWLEDGEMENTS \end{large}\end{center}}

\graphicspath{{../analysis/figures/}}

\begin{document}
\begin{titlepage}
\pubblock

\vfill
\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
\vfill
\Author{Amir Shahmoradi, Claus Wilke, Eleisha Jackson, ...}%\support}
\Address{\affiliation}
\vfill
\begin{Abstract}
    Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations. The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with correlations ranging from $0.1$ to $0.8$. In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes. We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

\section{Introduction}
\label{sec:intro}

    Patterns of amino acid sequence variation are known to be influenced by the function of proteins (xx). The general consensus, based on the flurry of research done over the past several decades, is that the amino-acid sequence determines the 3D structure of proteins, known as the native conformation. This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein. According to stability threshold model of proteins (xx), substitutions at specific sites may be tolerated more than other sites in the protein sequence, if the substitution does not significantly change the energy landscape and therefore the functional conformation of the protein. Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins (xx). As one of the earliest discovered examples, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
    \\

    Other structural properties have also been identified recently and proposed to influence or explain the site-specific evolutionary variations of proteins. Among the simplest properties is a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest, named residue {\it contact number (CN)}(xx). Variants of this quantity that eliminate the free-parameter (i.e., radius) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx). In a recent work Echave et al. xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins ($\Delta\Delta G$) to the rate at which mutations accumulate at the corresponding sites over evolutionary time. They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
    \\

    Not all proteins exhibit the same correlation strength and association between sequence variation and structural properties. By analyzing a data set of $216$ monomeric enzymes, Yeh et al. xx find a wide range of $\rho\sim0.1-0.8$ for the distributions of the correlations of sequence variability with two site-specific properties: contact number and residue solvent accessibility. Similarly, Echave et al. xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by $\delta\delta G$ with evolutionary rates. Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}. This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
    \\

    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence evolution, implies the existence of one or more structural or evolutionary characteristics of protein that modulate sequence-structure correlations in all proteins. Motivated by this observation, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide-range variations seen in correlation strengths of sequence evolutionary rates with different structural properties.  We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations.
    \\

    \begin{figure}[tbh]
        \begin{center}
        \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
        \end{center}
        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
        \label{fig:seqent_structure_cors}
    \end{figure}

\section{Materials and Methods}
\label{sec:mam}

    \subsection*{Sequence Data, Alignments and Evolutionary Rates}

    The results presented in this work are based on data set of 213 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al. 2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al. 1995). To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al. 2009; Ashkenazy et al. 2010). Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al. 2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

    \begin{equation}
        \label{eqn:shannon}
        H_i = -\sum_j P_{ij}\ln P_{ij}
    \end{equation}

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software rate4site (xx). To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity (Stamatakis 2014). For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) (Mayrose et al. 2004).

    \subsection*{Structural Properties}

    The goal of the presented work is to identify the prominent structural or evolutionary properties of proteins that modulate sequence-structure correlations. These potential modulators represent a unique characteristics of the protein as a whole.  In general, the structural and evolutionary properties fall into two major categories. 1. Residue-level properties: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence. Prominent examples of the former include site-specific Relative Solvent Accessibility (RSA) (Tien et al. 2012 xx), Weighted Contact Number (WCN) (shih? xx).   2. PDB-level properties: structural or evolutionary characteristics that are representative of the protein as a whole. Examples include pdb Contact Order (CO) as defined by xx, protein size, protein compactness.  In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein. A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.

    \subsection*{Eliminating Degeneracy in Structural Property Definitions}

    In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s). There are however degeneracies in the definition of the some site-specific variables. For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins. There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins. The same definition degeneracy also exists for the set of atomic Bfactors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor (e.g., Halle 2001 xx).
    \\

    Similar definition degeneracy also exists for the set of coordinates that can be used for Voronoi tessellation of proteins. A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other. These regions are called Voronoi cells. The structure of proteins can be considered as a set of 3D coordinates representing individual sites. Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
    \\

    Thus, for the sake of comprehensiveness and in order to identify the best definitions of structural properties such as WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of properties depending on the choice of the representative set of atomic coordinates used. These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$). In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains. We also calculate a representative coordinate for each site that is an average over all heavy atom coordinates in the side chain and backbone of the amino acid. In rare cases where the side chain atoms are not resolved in the PDB file or the amino acid lacks the heavy atom needed (e.g., $C_\beta$ for Glycine). The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
    \\

    We calculate the relevant Voronoi cell properties of all sites in all proteins using the software VORO++ (xx), and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. In addition to ASA values, we also extract from DSSP information about the secondary structure of proteins, such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns. To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structural properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
    \\

    All data including a list of 213 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.

\section{Results}
\label{sec:results}

    \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}

        As explained in the previous section, there is a high level of redundancy in the initial set of collected protein properties. In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties. This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.  Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used. 
        \\

        For the measure of local packing density in proteins (the Weighted Contact Number) we find that the average of coordinates of all heavy atoms for each residue results in WCN values that show the best correlation with other structural properties.

    \begin{figure}[tbh]
        \begin{center}
        \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
        \end{center}
        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
        \label{fig:seqent_structure_cors}
    \end{figure}

\section{Discussion and Concluding Remarks}
\label{sec:dcr}

\Acknowledgements

....

\bibliographystyle{apj}
\bibliography{proc}

\end{document}

### 2014-10-22 analysis/src/best_bfactor.r

@@ -19,19 +19,31 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+npdbs = 209         # number of pdb structures in the dataset
+
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
 res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
+res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
+res_prop_hps         = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
 res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
+res_prop_dssp        = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
 res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
+res_prop_wcn_bf      = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
 res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
 res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
+res_prop_voroAA      = res_prop_voroAA[!(res_prop_voroAA$pdb %in% excluded_pdbs),]
 res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
[…]

### 2014-10-22 analysis/src/best_bfactor_selected_variables.r

@@ -20,38 +20,58 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+npdbs = 209         # number of pdb structures in the dataset
+
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
 res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
+res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
+res_prop_hps         = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
 res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
+res_prop_dssp        = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
 res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
+res_prop_wcn_bf      = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
 res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
 res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
+res_prop_voroAA      = res_prop_voroAA[!(res_prop_voroAA$pdb %in% excluded_pdbs),]
 res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
-res_prop_voroAA      = cbind(res_prop_voroAA, VAAmodified_volume = res_prop_voroAA$VAAvolume)
[…]

### 2014-10-22 analysis/src/best_vorvol_selected_variables.r

@@ -18,42 +18,62 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+npdbs = 209         # number of pdb structures in the dataset
+
+res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
 res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
+res_prop_hps         = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
 res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
+res_prop_dssp        = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
 res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
+res_prop_wcn_bf      = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
 res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
 res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
+res_prop_voroAA      = res_prop_voroAA[!(res_prop_voroAA$pdb %in% excluded_pdbs),]
 res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
-res_prop_voroAA      = cbind(res_prop_voroAA, VAAmodified_volume = res_prop_voroAA$VAAvolume)
[…]

### 2014-10-22 analysis/src/best_wcn_selected_variables.r

@@ -18,38 +18,58 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+npdbs = 209         # number of pdb structures in the dataset
+
+res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
 res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
+res_prop_hps         = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
 res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
+res_prop_dssp        = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
 res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
+res_prop_wcn_bf      = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
 res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
 res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
+res_prop_voroAA      = res_prop_voroAA[!(res_prop_voroAA$pdb %in% excluded_pdbs),]
 res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
-res_prop_voroAA      = cbind(res_prop_voroAA, VAAmodified_volume = res_prop_voroAA$VAAvolume)
[…]

### 2014-10-22 manuscript/cordiv.tex
MN revisions

@@
 
-        For the measure of local packing density in proteins (the Weighted Contact Number) we find that the average of coordinates of all heavy atoms for each residue results in WCN values that show the best correlation with other structural properties.
+        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms for each each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_wcn}
+
+        For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
+Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively.
+
+        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells.
 
@@
         \begin{center}
-        \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
+        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf} % Fig. 1
         \end{center}
@@
 For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-        \label{fig:seqent_structure_cors}
+        \label{fig:best_wcn}
+    \end{figure}
+
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf} % Fig. 1
+        \end{center}
+        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
+correlations of other structural properties with evolutionary rates.
+Detailed description of the structural properties is given in Section \label{sec:mam}.
+The red lines in each plot represent equality line.
+It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
+For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
+        \label{fig:best_wcn}
+    \end{figure}
+
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf} % Fig. 1
+        \end{center}
+        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
+correlations of other structural properties with evolutionary rates.
+Detailed description of the structural properties is given in Section \label{sec:mam}.
+The red lines in each plot represent equality line.
+It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
+For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
+        \label{fig:best_wcn}
     \end{figure}

### 2014-10-24 manuscript/cordiv.tex
regularized_regression tables and figures added.

@@
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
-        \end{center}
-        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
-correlations of other structural properties with evolutionary rates.
-Detailed description of the structural properties is given in Section \label{sec:mam}.
-The red lines in each plot represent equality line.
-It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
-For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-        \label{fig:seqent_structure_cors}
-    \end{figure}
-
 \section{Materials and Methods}
@@
         For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms for each each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_wcn}
+Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_wcn}).
+        \\
 
         For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively.
+Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_bf}).
+        \\
 
-        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells.
+        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
+        \\
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf} % Fig. 1
-        \end{center}
-        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
-correlations of other structural properties with evolutionary rates.
-Detailed description of the structural properties is given in Section \label{sec:mam}.
-The red lines in each plot represent equality line.
-It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
-For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-        \label{fig:best_wcn}
-    \end{figure}
+        All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites.
+In particular, the strength of structure and sequence correlations decrease when moving from side chain to backbone atoms.
+An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
+        \\
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf} % Fig. 1
-        \end{center}
-        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
-correlations of other structural properties with evolutionary rates.
-Detailed description of the structural properties is given in Section \label{sec:mam}.
-The red lines in each plot represent equality line.
-It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
-For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-        \label{fig:best_wcn}
-    \end{figure}
+        Based on the observation that side chain structural properties have the best correlations with other structure and sequence characteristics of individual sites, we keep only variables measured based on average side chain properties and coordinates throughout the rest of the analysis and omit all other possible measures that show only weaker correlations with other site-specific characteristics.
+        \\
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf} % Fig. 1
-        \end{center}
-        \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
-correlations of other structural properties with evolutionary rates.
-Detailed description of the structural properties is given in Section \label{sec:mam}.
-The red lines in each plot represent equality line.
-It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
-For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-        \label{fig:best_wcn}
-    \end{figure}
+    \subsection*{Identifying Determinants of Sequence-Structure Correlation Strengths}
+
+        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence between variables in a collection of observations, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
+The choice of Spearman vs.
+the popular Pearson's correlation measure is to minimize the effects of any potential nonlinear relationships between the set of variables on the measures of associations between the variables.
+
+        In the following lines we summarize the primary factors affecting the strength of individual correlations of interest.
+
+        -- most importantly WCN, RSA $\Delta\Delta G$ Entropy, and Voronoi cell volume and surface area --
 
@@
 
+        \begin{figure}[tbh]
+            \begin{center}
+            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
+            \end{center}
+            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
+correlations of other structural properties with evolutionary rates.
+Detailed description of the structural properties is given in Section \label{sec:mam}.
+The red lines in each plot represent equality line.
+It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
+For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
+            \label{fig:seqent_structure_cors}
+        \end{figure}
+
+        \begin{figure}[tbh]
+            \begin{center}
+            \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf} % Fig. 1
+            \end{center}
+            \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
+The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
+The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+            \label{fig:best_wcn}
+        \end{figure}
+
+        \begin{figure}[tbh]
+            \begin{center}
+            \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf} % Fig. 1
+            \end{center}
+            \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset.
+In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other.
+The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors.
+The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+            \label{fig:best_bf}
+        \end{figure}
+
+        \begin{figure}[tbh]
+            \begin{center}
+            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf} % Fig. 1
+            \end{center}
+            \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
+            \label{fig:best_voronoi}
+        \end{figure}
+
 \bibliographystyle{apj}

### 2014-10-24 analysis/src/glm_reg.r

@@ -1,7 +1,7 @@
 # This R code performs regularized regression on pdb properties data.
 # Amir Shahmoradi, Sunday 4:47 AM, Oct 19 2014, Wilke Lab, ICMB, UT Austin
 
-install.packages("glmnet", repos = "http://cran.us.r-project.org")
+# install.packages("glmnet", repos = "http://cran.us.r-project.org")
 library(glmnet)
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
@@ -12,10 +12,56 @@ seqent_scors = grep('^r.+seqent',colnames(all_pdb_prop_subset),value = TRUE)  # 
 r4sJC_scors = grep('^r.+r4sJC',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
 not_in_pcr = c(seqent_scors,r4sJC_scors)
 
-fit = glmnet(x = as.matrix(all_pdb_prop_subset[,-which(names(all_pdb_prop_subset) %in% not_in_pcr)]), y = as.vector(all_pdb_prop_subset$r.seqent.wcnSC))
+x = as.matrix(all_pdb_prop_subset[,-which(names(all_pdb_prop_subset) %in% not_in_pcr)])   # predictor variables
+
+for (correlation in not_in_pcr)
+{
+  cat (paste(' Now fitting for correlation:', correlation, '\n'))
+  #fit = glmnet(x = x,
+  #             y = as.vector(all_pdb_prop_subset[[correlation]]),
+  #             alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
+  #             )
+
+  cvfit = cv.glmnet(x = x,
+                    y = as.vector(all_pdb_prop_subset[[correlation]]),
+                    alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
+                    )
+  
+  lambda.min.coef = data.frame(as.matrix(coef(cvfit, s = 'lambda.min')))
+  lambda.min.coef = data.frame(variable = row.names(lambda.min.coef),
+                               coef     = lambda.min.coef$X1,
+                               abs.coef = abs(lambda.min.coef$X1)
+                               )
+  lambda.min.coef = lambda.min.coef[with(lambda.min.coef, order(-abs.coef)),]
+
[…]

### 2014-10-30 manuscript/cordiv.tex
some new analysis and data

@@
 
-    We calculate the relevant Voronoi cell properties of all sites in all proteins using the software VORO++ (xx), and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-In addition to ASA values, we also extract from DSSP information about the secondary structure of proteins, such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
-To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structural properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
+    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
+To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
     \\
@@
 
-        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms for each each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_wcn}).
+        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates ($r4sJC$) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
         \\
@@
         For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, Voronoi cell volumes, sequence entropy and evolutionary rates by a median Spearman correlation difference of $00$, $00$, $00$ \& $00$, respectively (Figure \ref{fig:best_bf}).
+Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
         \\
@@
 
-        Based on the observation that side chain structural properties have the best correlations with other structure and sequence characteristics of individual sites, we keep only variables measured based on average side chain properties and coordinates throughout the rest of the analysis and omit all other possible measures that show only weaker correlations with other site-specific characteristics.
+        Based on the observations described in the previous paragraphs, we keep only variables measured based on average side chain properties and coordinates throughout the rest of the analysis and omit all other possible measures that show only weaker correlations with other site-specific characteristics.
         \\
@@
 the popular Pearson's correlation measure is to minimize the effects of any potential nonlinear relationships between the set of variables on the measures of associations between the variables.
+
+        The exclusion of alternative measures of structural quantities as described in the previous section, results in a significant reduction in the number of pdb-level variables to be further analysed, from $$ to $$, without compromising generality and comprehensiveness of the analysis.
+
+        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest using regularized regression methods (xx) on the original dataset of pdb-level properties.
+, also on its rank transformation to ensure non-linearities
+
+        first remove all other sequence-structure correlations from the list of potential contributing factors. As explained in the previous sections, Then we apply regularized regression methods
 

### 2014-10-30 analysis/src/get_res_data.r

@@ -7,39 +7,42 @@ setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
 
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
-res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 res_prop_elj = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
+res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
 res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
-res_prop_jec$pdb     = factor(res_prop_jec$pdb)
 res_prop_jec = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
 
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
-res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 res_prop_hps = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
+res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
-res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 res_prop_dssp = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
+res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
-res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 res_prop_wcn_bf = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
+res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
-#res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
-#res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
[…]

### 2014-10-30 analysis/src/regularized_regresison_rank_data.r

@@ -47,6 +47,12 @@ for (correlation in not_in_pcr)
   filename = paste0('../tables/regularized_regression_rank_data/',correlation,'_l1se_rank.csv')
   write.csv(lambda.1se.coef, filename, row.names = F)
   
+  
+  
+  lfit = lm()
+  
+  
+  
   # Now generate the error-lambda plots:
   
   # filename = paste0('../figures/regularized_regression_rank_data/',correlation,'.pdf')
@@ -65,7 +71,7 @@ predictor_names[103]
 plot(fit, label = TRUE)
 plot(fit, xvar = 'dev', label = TRUE,
      xlim = c(0.,0.1),
-     xlim = c(-0.1,0.1)
+     ylim = c(-0.05,0.05)
      )
 print(fit)
 coef(fit,s=0.1)

### 2014-10-30 analysis/src/res_prop_correlations.r

@@ -13,132 +13,112 @@
 # Last updated by Amir Shahmoradi, Tuesday 7:41 PM, Sep 30 2014, Wilke Lab, ICMB, UT Austin
 
 #install.packages("reshape2")
-#library("reshape2")
+library("reshape2")
 #library('corrplot')
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+
 res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+res_prop_elj = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
 res_prop_elj$pdb     = factor(res_prop_elj$pdb)
 
+res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+res_prop_jec = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
+res_prop_hps = res_prop_hps[!(res_prop_hps$pdb %in% excluded_pdbs),]
 res_prop_hps$pdb     = factor(res_prop_hps$pdb)
 
 res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
+res_prop_dssp = res_prop_dssp[!(res_prop_dssp$pdb %in% excluded_pdbs),]
 res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
 
 res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
+res_prop_wcn_bf = res_prop_wcn_bf[!(res_prop_wcn_bf$pdb %in% excluded_pdbs),]
 res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
 
 res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
+res_prop_voroAA      = cbind(res_prop_voroAA, VAAsphericity = 4.8359758620494089221509005399179*(res_prop_voroAA$VAAvolume^(2./3.))/res_prop_voroAA$VAAarea)
[…]

### 2014-11-03 manuscript/cordiv.tex
new data, analysis and figures added for proportion of variance explained.

@@
 \def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
-\newcommand{\liso}{L_{iso}}
-\newcommand{\eiso}{E_{iso}}
-\newcommand{\epkz}{E_{p,z}}
-\newcommand{\durz}{T_{90,z}}
-\newcommand{\pbol}{P_{bol}}
-\newcommand{\sbol}{S_{bol}}
-\newcommand{\epk}{E_{p}}
-\newcommand{\dur}{T_{90}}
-\newcommand{\pph}{P_{50-300}}
+\newcommand{\ddg}{$\Delta\Delta G~$}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In a recent work Echave et al.
-xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins ($\Delta\Delta G$) to the rate at which mutations accumulate at the corresponding sites over evolutionary time.
+xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins (\ddg) to the rate at which mutations accumulate at the corresponding sites over evolutionary time.
 They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
@@
 Similarly, Echave et al.
-xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by $\delta\delta G$ with evolutionary rates.
+xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg with evolutionary rates.
 Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
@@
 
-    The results presented in this work are based on data set of 213 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al.
+    The results presented in this work are based on data set of 209 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al.
 2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
@@
 
-    All data including a list of 213 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
+    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
 
@@
 
-        As explained in the previous section, there is a high level of redundancy in the initial set of collected protein properties.
+        As explained in previous section, there is a high level of redundancy in the initial set of collected protein properties.
 In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties.
@@
         For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates ($r4sJC$) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
+Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
         \\
@@
         For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, $\Delta\Delta$ entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
+Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
         \\
@@
 
-        Based on the observations described in the previous paragraphs, we keep only variables measured based on average side chain properties and coordinates throughout the rest of the analysis and omit all other possible measures that show only weaker correlations with other site-specific characteristics.
+        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
+The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
         \\
 
-    \subsection*{Identifying Determinants of Sequence-Structure Correlation Strengths}
+    \subsection*{Sequence divergence as the main Determinant of Sequence-Structure Relation}
 
-        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence between variables in a collection of observations, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
-The choice of Spearman vs.
-the popular Pearson's correlation measure is to minimize the effects of any potential nonlinear relationships between the set of variables on the measures of associations between the variables.
+        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
+The choice of Spearman versus the popular Pearson's correlation measure is to minimize the effects of any nonlinear relationships between the variables on the strengths of the correlations.
+As summary of the most influential factors to structure-sequence relations is illustrated in Figure \ref{fig:}.
 
-        The exclusion of alternative measures of structural quantities as described in the previous section, results in a significant reduction in the number of pdb-level variables to be further analysed, from $$ to $$, without compromising generality and comprehensiveness of the analysis.
-
-        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest using regularized regression methods (xx) on the original dataset of pdb-level properties.
+        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest, using regularized regression methods (xx) on the original dataset of pdb-level properties.
 , also on its rank transformation to ensure non-linearities
@@
 
-        -- most importantly WCN, RSA $\Delta\Delta G$ Entropy, and Voronoi cell volume and surface area --
+        -- most importantly WCN, RSA \ddg Entropy, and Voronoi cell volume and surface area --
 
@@
             \begin{center}
-            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf} % Fig. 1
+            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
             \end{center}
@@
             \begin{center}
-            \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf} % Fig. 1
+            \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
             \end{center}
@@
             \begin{center}
-            \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf} % Fig. 1
+            \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
             \end{center}
@@
             \begin{center}
-            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf} % Fig. 1
+            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf}
             \end{center}
@@
 
+        \begin{figure*}[tbh]
+            \centering
+            \begin{tabular}{cc}
+            \includegraphics[width=3.4in]{/modulators_similarity_Sr4sJC_scors.pdf} & \includegraphics[width=3.4in]{/modulators_similarity_Sseqent_scors.pdf} \\
+            \end{tabular}
+            \caption{Proportion of explained variance of structure-sequence correlations by 182 pdb-level properties considered individually.}
+            \label{fig:modsim}
+        \end{figure*}
+
 \bibliographystyle{apj}

### 2014-11-03 analysis/src/cordiv_modulators.r

@@ -1,0 +1,58 @@
+# This R scripts generates a plot of the most influential pdb-level variables on the structure-sequence correlations.
+
+
+index = names(result) %in% c("r.rsa_avg_md", "r.wcn_avg_md", "r.chi1_var_md", "r.rmsf_avg_md", "r.bfca") #, "avg_cr_rho") # columns we want to plot
+index.p = names(result) %in% c("p.rsa_avg_md", "p.wcn_avg_md", "p.chi1_var_md", "p.rmsf_avg_md", "p.bfca") # columns that store significance
+
+colors = c('red', 'blue', 'green', 'purple', 'orange3', 'darkgreen', 'black', 'gray', 'cyan2') #, 'darkred', 'darkgreen', 'bisque2')
+###proteins = c('1RD8_AB', '2FP7_B', '2JLY_A', '2Z83_A', '3GOL_A', '3LYF_A', '4AQF_B', '4GHA_A', '4IRY_A') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
+proteins = c('1RD8_AB', '2FP7_B', '2Z83_A', '3LYF_A', '4AQF_B', '4GHA_A', '4IRY_A') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
+###labels = c('1RD8', '2FP7', '2JLY', '2Z83', '3GOL', '3LYF', '4AQF', '4GHA', '4IRY') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
+labels = c('1RD8', '2FP7', '2Z83', '3LYF', '4AQF', '4GHA', '4IRY') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
+
+pdf( "correlation_analysis/figures/cor_r4s_all.pdf", width=7, height=4, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.0, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
[…]

### 2014-11-03 analysis/src/cormat_cormat_ss_scors.r

@@ -1,0 +1,38 @@
+# This code searches for the similarity of the of variables that describe the correlations coefficients.
+# This is complicated, ask Amir to explain for you what this code does.
+# Amir Shahmoradi, Sunday 1:40 PM, November 2 2014, Wilke Lab, ICMB, UT Austin
+
+install.packages('corrplot')
+library('corrplot')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+
+
+# first Structure_r4sJC correlations:
+
+all_pdb_prop_cormat_SS_scors = read.csv('../tables/cormat_all_pdb_prop_cormat_Sr4s_scors.csv', header=T)
+
+pdf( "../figures/modulators_similarity_Sr4sJC_scors.pdf", width=4.5, height=4, useDingbats=FALSE )
+cormat_all_pdb_prop_cormat_SS_scors = cor(subset(all_pdb_prop_cormat_SS_scors, select = -c(variable,vcompactness,vvolume)), method = 'spearman')
+corrplot.mixed(cormat_all_pdb_prop_cormat_SS_scors, lower = "ellipse", upper = "number", order = 'alphabet')
+dev.off()
+
+# Now Structure_seqent correlations:
+
+all_pdb_prop_cormat_SS_scors = read.csv('../tables/cormat_all_pdb_prop_cormat_Sseqent_scors.csv', header=T)
+
+pdf( "../figures/modulators_similarity_Sseqent_scors.pdf", width=4.5, height=4, useDingbats=FALSE )
+cormat_all_pdb_prop_cormat_SS_scors = cor(subset(all_pdb_prop_cormat_SS_scors, select = -c(variable,vcompactness,vvolume)), method = 'spearman')
+corrplot.mixed(cormat_all_pdb_prop_cormat_SS_scors, lower = "ellipse", upper = "number", order = 'alphabet')
+dev.off()
+
+
+# corrplot(cormat_all_pdb_prop_cormat_SS_scors, order = "hclust")
+# cormat_all_pdb_prop_cormat_SS_scors = cor(abs(subset(all_pdb_prop_cormat_SS_scors, select = -c(variable))), method = 'spearman')
+# corrplot(cormat_all_pdb_prop_cormat_SS_scors, order = "FPC")
+# 
[…]

### 2014-11-03 analysis/src/lm_PEV.r

@@ -1,0 +1,68 @@
+# This R script performs linear regression of sequence-structure on the set of variables that have the strongest explanatory power according to regularized regression results.
+# Last updated by Amir Shahmoradi, Sunday 6:33 PM, November 2 2014, Wilke Lab, ICMB, UT Austin
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+all_pdb_prop_select_wide = read.csv("../tables/all_pdb_prop_select_wide.csv",header=T)
+all_pdb_prop_subset = subset(all_pdb_prop_select_wide, select = -c(pdb,sum.nssb,mean.nssb))  # These two columns are all zero for all proteins and do not carry any valuable infromation.
+all_pdb_prop_cormat = read.csv("../tables/all_pdb_prop_cormat.csv", header=T )
+all_pdb_prop_cormat_pvalues = read.csv("../tables/all_pdb_prop_cormat_pvalues.csv", header=T )
+#plot(log10(all_pdb_prop_cormat_pvalues$contact_orderSC),abs(all_pdb_prop_cormat$contact_orderSC))
+
+seqent_scors = grep('^r.+seqent',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
+r4sJC_scors = grep('^r.+r4sJC',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
+ssscors = c(seqent_scors,r4sJC_scors)   # structure-sequence spearman correlation names
+'%ni%' = Negate('%in%')  # not in
+
+#all_pdb_prop_cormat = subset(all_pdb_prop_cormat, select = names(all_pdb_prop_cormat) %ni% ssscors)
+all_pdb_prop_cormat = all_pdb_prop_cormat[!(all_pdb_prop_cormat$X %in% ssscors),]
+
+#all_pdb_prop_cormat_pvalues = subset(all_pdb_prop_cormat_pvalues, select = names(all_pdb_prop_cormat_pvalues) %ni% ssscors)
+all_pdb_prop_cormat_pvalues = all_pdb_prop_cormat_pvalues[!(all_pdb_prop_cormat_pvalues$X %in% ssscors),]
+
[…]

### 2014-11-03 analysis/src/regularized_regresison_rank_data.r

@@ -1,78 +1,0 @@
-# This R code performs regularized regression on pdb properties RANK data.
-# Amir Shahmoradi, Friday 8:09 PM, Oct 24 2014, Wilke Lab, ICMB, UT Austin
-
-# install.packages("glmnet", repos = "http://cran.us.r-project.org")
-library(glmnet)
-
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
-
-all_pdb_prop_select_wide_rank = read.csv("../tables/all_pdb_prop_select_wide_rank.csv",header=T)
-all_pdb_prop_subset = subset(all_pdb_prop_select_wide_rank, select = -c(pdb)) # ,sum.nssb,mean.nssb))  # These two columns are all zero for all proteins and do not carry any valuable infromation.
-seqent_scors = grep('^r.+seqent',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
-r4sJC_scors = grep('^r.+r4sJC',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
-not_in_pcr = c(seqent_scors,r4sJC_scors)
-
-x = as.matrix(all_pdb_prop_subset[,-which(names(all_pdb_prop_subset) %in% not_in_pcr)])   # predictor variables
-
-for (correlation in not_in_pcr)
-{
-  cat (paste(' Fitting for correlation:', correlation, '\n'))
-  fit = glmnet(x = x,
-               y = as.vector(all_pdb_prop_subset[[correlation]]),
-               alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
-               )
-
-  cvfit = cv.glmnet(x = x,
-                    y = as.vector(all_pdb_prop_subset[[correlation]]),
-                    alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
-                    )
-  
-  lambda.min.coef = data.frame(as.matrix(coef(cvfit, s = 'lambda.min')))
-  lambda.min.coef = data.frame(variable = row.names(lambda.min.coef),
-                               coef     = lambda.min.coef$X1,
-                               abs.coef = abs(lambda.min.coef$X1)
-                               )
[…]

### 2014-11-03 analysis/src/regularized_regression_rank_data.r

@@ -1,0 +1,132 @@
+# This R code performs regularized regression on pdb properties RANK data.
+# Amir Shahmoradi, Friday 8:09 PM, Oct 24 2014, Wilke Lab, ICMB, UT Austin
+
+# install.packages("glmnet", repos = "http://cran.us.r-project.org")
+library(glmnet)
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+all_pdb_prop_select_wide_rank = read.csv("../tables/all_pdb_prop_select_wide_rank.csv",header=T)
+all_pdb_prop_subset = subset(all_pdb_prop_select_wide_rank, select = -c(pdb)) # ,sum.nssb,mean.nssb))  # These two columns are all zero for all proteins and do not carry any valuable infromation.
+seqent_scors = grep('^r.+seqent',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
+r4sJC_scors = grep('^r.+r4sJC',colnames(all_pdb_prop_subset),value = TRUE)  # Select column names that correspond to seqent-variable correlations
+not_in_pcr = c(seqent_scors,r4sJC_scors)
+
+x = as.matrix(all_pdb_prop_subset[,-which(names(all_pdb_prop_subset) %in% not_in_pcr)])   # predictor variables
+
+for (correlation in not_in_pcr)
+{
+  cat (paste(' Fitting for correlation:', correlation, '\n'))
+  fit = glmnet(x = x,
+               y = as.vector(all_pdb_prop_subset[[correlation]]),
+               alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
+               )
+
+  cvfit = cv.glmnet(x = x,
+                    y = as.vector(all_pdb_prop_subset[[correlation]]),
+                    alpha = 0.5   # elastic-net mixing parameter, with a range from 0 (ridge regression) to 1 (lasso regression).
+                    )
+  
+  lambda.min.coef = data.frame(as.matrix(coef(cvfit, s = 'lambda.min')))
+  lambda.min.coef = data.frame(variable = row.names(lambda.min.coef),
[…]

### 2014-11-03 analysis/src/summarize_pdb_data.r

@@ -1,4 +1,5 @@
 # Amir Shahmoradi, Thursday 4:20 PM, Aug 21 2014, Wilke Lab, ICMB, UT Austin
+# This code summarizes all PDB-level properties of all proteins in the dataset in a long-table format.
 
 #install.packages("reshape2")
 #library("reshape2")

### 2014-11-03 analysis/src/summarize_pdb_data_select_variables.r

@@ -3,6 +3,8 @@
 
 #install.packages("reshape2")
 library("reshape2")
+#install.packages("Hmisc")
+#library("Hmisc")
 #library('corrplot')
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
@@ -59,18 +61,35 @@ pdbs = data.frame(pdb = all_pdb_prop_select_wide$pdb)
 all_pdb_prop_select_wide = cbind(pdbs, all_pdb_prop_select_wide[,as.vector(colnames_all_pdb_prop_select_wide$colnames)])
 write.csv(all_pdb_prop_select_wide, "../tables/all_pdb_prop_select_wide.csv", row.names=F )
 
+all_pdb_prop_select_wide = read.csv("../tables/all_pdb_prop_select_wide.csv", header = T)
 all_pdb_prop_subset = subset(all_pdb_prop_select_wide, select = -c(pdb,sum.nssb,mean.nssb))
 all_pdb_prop_cormat = as.data.frame(cor(all_pdb_prop_subset, method='spearman'))
 write.csv(all_pdb_prop_cormat, "../tables/all_pdb_prop_cormat.csv", row.names=T )
 write.csv(abs(all_pdb_prop_cormat), "../tables/all_pdb_prop_cormat_abs_values.csv", row.names=T )
+
 cor(all_pdb_prop_cormat$r.rsa.seqent,all_pdb_prop_cormat$sd.vsphericitym, method='sp')
 cor(all_pdb_prop_cormat$r.seqent.wcnSC,all_pdb_prop_cormat$r.seqent.vsphericitym, method='sp')
+
 cormat_all_pdb_prop_cormat = as.data.frame(cor(all_pdb_prop_cormat, method='spearman'))
 write.csv( cormat_all_pdb_prop_cormat, "../tables/cormat_all_pdb_prop_cormat.csv", row.names=T )
 write.csv( abs(cormat_all_pdb_prop_cormat), "../tables/cormat_all_pdb_prop_cormat_abs_values.csv", row.names=T )
 
+# Now calculate the corresponding P-value matrix
 
+# install.packages('psych')
+library('psych')
 
+all_pdb_prop_select_wide = read.csv("../tables/all_pdb_prop_select_wide.csv", header = T)
+all_pdb_prop_subset = subset(all_pdb_prop_select_wide, select = -c(pdb,sum.nssb,mean.nssb))
[…]

### 2014-11-11 manuscript/cordiv.tex
New analysis and figures for cross validation added.

@@
 \newcommand{\ddg}{$\Delta\Delta G~$}
+%\newcommand{\se}{\it seqent}
+%\newcommand{\rs}{\it r4sJC}
 
@@
         In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
-The choice of Spearman versus the popular Pearson's correlation measure is to minimize the effects of any nonlinear relationships between the variables on the strengths of the correlations.
-As summary of the most influential factors to structure-sequence relations is illustrated in Figure \ref{fig:}.
+The choice of Spearman versus the popular Pearson's correlation measure is made in order to minimize the effects of any nonlinear variable relationships on the strengths of the correlations.
+The resulting correlation matrix reveals a myriad of pdb-level properties each having a small but nonzero contribution to the strength of the structure-sequence correlations.
+        \\
 
-        Now with the reduced data set at hand, we attempt to identify the main determinants of each sequence-structure relation of interest, using regularized regression methods (xx) on the original dataset of pdb-level properties.
-, also on its rank transformation to ensure non-linearities
+        A hierarchical clustering of the correlation matrix however, reveals two main independent factors that have the strongest influence on the strengths of sequence-structure correlations: 1.
+The sequence divergence as measured by the standard deviation of sequence entropy and evolutionary rates (denoted by {\it sd.seqent} \& {\it sd.r4sJC}) among all sites in each protein structure.
+2.
+The homogeneity of the hydrogen bond strengths among the back bone atoms of each protein structure, as measured by the standard deviation of hydrogen bond energies (denoted by {\it sd.hbe}) among all pdb sites.
+A reduced-size of the Spearman correlation matrix for the most influential factors on the two strongest sequence--structure (seqent$/$r4s -- wcnSC$/$varea) relations is illustrated in Figure \ref{fig:mainmods}.
+        \\
 
-        first remove all other sequence-structure correlations from the list of potential contributing factors. As explained in the previous sections, Then we apply regularized regression methods
+        For the other weaker sequence--structure relations, i.e.
+the correlations of seqent$/$r4s with RSA, \ddg entropy (ddgent), and Bfactor (bfSC) we find other pdb-level properties that also contribute to the correlation strengths, comparable to or even stronger than in sequence divergence and hydrogen bond homogeneity.
+In general, we observe that for the weaker the sequence-structure correlations, factors that determine the accuracy of the measured residue properties become more influential on the strength of the correlations.
+In particular, the X-ray crystallographic resolution of the structure and the definition of the \ddg entropy play dominant roles, with Spearman correlation coefficients of $\rho~0.3$, on the strengths of the corresponding sequence-structure relations.
 
-        In the following lines we summarize the primary factors affecting the strength of individual correlations of interest.
+        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also apply multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the models, and the pdb-level properties as the explanatory variables.
+Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression ( reference to R package xx) on the entire dataset of pdb-level properties, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
+Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
 
-        -- most importantly WCN, RSA \ddg Entropy, and Voronoi cell volume and surface area --
+        In addition to regularized regression, we have also employed Principal Component Regression on the original dataset and its rank transformation, all of which point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
 
@@
 \label{sec:dcr}
+
+        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work.
+Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., wcnSC), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
+In addition, we have derived new site-specific properties based on Voronoi Tessellation of protein 3D structures that are comparable to or better than several previously known structural properties in explaining site specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
+Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,
+
+        \begin{equation}
+        \label{eqn:sphericity}
+        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
+        \end{equation}
+
+        in which $V$ \& $A$ represent {\it vvolume} \& {\it varea} respectively.
+We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
+Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates $0.12$ in terms of Spearman correlation strength.
 
@@
 
+        \begin{figure}[tbh]
+            \centering
+            \includegraphics[width=7.2in]{/main_modulators_cormat.pdf} \\
+            \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
+The prefixes {\it r, sd \& mn}}
+            \label{fig:mainmods}
+        \end{figure}
+
         \begin{figure*}[tbh]
@@
 
+        \begin{figure*}[tbh]
+            \centering
+            \begin{tabular}{cc}
+            \includegraphics[width=3.4in]{/validation_sdseqent.pdf} & \includegraphics[width=3.4in]{/validation_sdhbe.pdf} \\
+            \end{tabular}
+            \caption{Proportion of explained variance of structure-sequence correlations by 182 pdb-level properties considered individually.}
+            \label{fig:validation}
+        \end{figure*}
+
 \bibliographystyle{apj}

### 2014-11-11 analysis/src/cordiv_modulators.r

@@ -1,13 +1,11 @@
 # This R scripts generates a plot of the most influential pdb-level variables on the structure-sequence correlations.
-
+# Amir Shahmoradi, Sunday 4:39 PM, November 9 2014, Wilke Lab, ICMB, UT Austin
 
 index = names(result) %in% c("r.rsa_avg_md", "r.wcn_avg_md", "r.chi1_var_md", "r.rmsf_avg_md", "r.bfca") #, "avg_cr_rho") # columns we want to plot
 index.p = names(result) %in% c("p.rsa_avg_md", "p.wcn_avg_md", "p.chi1_var_md", "p.rmsf_avg_md", "p.bfca") # columns that store significance
 
 colors = c('red', 'blue', 'green', 'purple', 'orange3', 'darkgreen', 'black', 'gray', 'cyan2') #, 'darkred', 'darkgreen', 'bisque2')
-###proteins = c('1RD8_AB', '2FP7_B', '2JLY_A', '2Z83_A', '3GOL_A', '3LYF_A', '4AQF_B', '4GHA_A', '4IRY_A') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
 proteins = c('1RD8_AB', '2FP7_B', '2Z83_A', '3LYF_A', '4AQF_B', '4GHA_A', '4IRY_A') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
-###labels = c('1RD8', '2FP7', '2JLY', '2Z83', '3GOL', '3LYF', '4AQF', '4GHA', '4IRY') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
 labels = c('1RD8', '2FP7', '2Z83', '3LYF', '4AQF', '4GHA', '4IRY') #, '3GSZ_A', '3I5K_A', '2JLY_A_temp_50', '2JLY_A_temp_100', '2JLY_A_temp_200', '2JLY_A_temp_450')
 
 pdf( "correlation_analysis/figures/cor_r4s_all.pdf", width=7, height=4, useDingbats=FALSE )

### 2014-11-11 analysis/src/cormat_cormat_ss_scors.r

@@ -28,10 +28,47 @@ corrplot.mixed(cormat_all_pdb_prop_cormat_SS_scors, lower = "ellipse", upper = "
 dev.off()
 
 
+# Now create the cor matrix of the main modulators of correlations of wcnSC and varea with r4sJC and seqent:
+
+all_pdb_prop_select_wide_rank = read.csv('../tables/all_pdb_prop_select_wide_rank.csv',header=T)
+# selection = subset( all_pdb_prop_select_wide_rank
+#                   , select = c(r.r4sJC.wcnSC,r.r4sJC.varea,r.seqent.wcnSC,r.seqent.varea
+#                               ,r.r4sJC.seqent,sd.seqent,sd.r4sJC
+#                               ,sd.hbe,mean.helix,mean.betas))
+
+selection = data.frame( r.r4s.wcn      = all_pdb_prop_select_wide_rank$r.r4sJC.wcnSC
+                      , r.r4s.varea    = all_pdb_prop_select_wide_rank$r.r4sJC.varea
+                      , r.se.wcn       = all_pdb_prop_select_wide_rank$r.seqent.wcnSC
+                      , r.se.varea     = all_pdb_prop_select_wide_rank$r.seqent.varea
+                      , r.r4s.se       = all_pdb_prop_select_wide_rank$r.r4sJC.seqent
+                      , sd.se          = all_pdb_prop_select_wide_rank$sd.seqent
+                      , sd.r4s         = all_pdb_prop_select_wide_rank$sd.r4sJC
+                      , sd.hbe         = all_pdb_prop_select_wide_rank$sd.hbe
+                      , mn.helix       = all_pdb_prop_select_wide_rank$mean.helix
+                      , mn.betas       = all_pdb_prop_select_wide_rank$mean.betas
+                      )
+
+pdf( "../figures/main_modulators_cormat.pdf", width=9, height=8, useDingbats=FALSE )
+cormat_select = cor(subset(selection), method = 'spearman')
+corrplot.mixed(cormat_select, lower = "ellipse", upper = "number"
+               #, tl.pos = 'lt'
+               , tl.pos = 'd'
+               , tl.cex = 0.95
+               #, tl.srt = 45
+               #, order = 'hclust'
[…]

### 2014-11-11 analysis/src/get_res_data.r

@@ -46,3 +46,33 @@ res_prop_voroSC$VSCmodified_sphericity = res_prop_voroSC$VSCsphericity
 res_prop_voroSC$VSCmodified_sphericity[res_prop_voroSC$VSCvolume_change_diff != 0] = -res_prop_voroSC$VSCsphericity[res_prop_voroSC$VSCvolume_change_diff != 0]
 res_prop_voroSC = res_prop_voroSC[!(res_prop_voroSC$pdb %in% excluded_pdbs),]
 res_prop_voroSC$pdb  = factor(res_prop_voroSC$pdb)
+
+scaling_data = res_prop_voroSC[res_prop_voroSC$VSCvolume_change_diff == 0,]
+#plot(1.4*(log10(scaling_data$VSCarea)-log10(4.8359758620494089221509005399179)),log10(scaling_data$VSCvolume))
+plot(log10(scaling_data$VSCarea),log10(scaling_data$VSCvolume/scaling_data$VSCarea^1.25))
+cor(log10(scaling_data$VSCarea),log10(scaling_data$VSCvolume/scaling_data$VSCarea^1.25),method='sp')
+#x = 1.4*(log10(scaling_data$VSCarea)-log10(4.8359758620494089221509005399179))
+x = log10(scaling_data$VSCarea)
+summary(lm(log10(scaling_data$VSCvolume)~ x))
+#plot(lm(log10(scaling_data$VSCvolume)~ x))
+lx = c(1,40)
+ly = 0.8807092*lx + 0.3023839
+lines(lx,ly,col='red')
+
+#H-clustering of correlation matrix:
+
+all_pdb_prop_cormat = read.csv("../tables/all_pdb_prop_cormat.csv", header=T )
+row.names(all_pdb_prop_cormat) = all_pdb_prop_cormat$X
+all_pdb_prop_cormat = subset(all_pdb_prop_cormat, select = -c(X))
+pdf( "../figures/cormat_hcluster.pdf", width=36, height=16, useDingbats=FALSE )
+plot( hclust(dist(abs(all_pdb_prop_cormat)))
+    , xlab = 'Hierarchical Clustering of the Spearman Correlation Matrix')
+dev.off()
+
+all_pdb_prop_cormat = read.csv("../tables/all_pdb_prop_cormat.csv", header=T )
+row.names(all_pdb_prop_cormat) = all_pdb_prop_cormat$X
[…]

### 2014-11-11 analysis/src/input_ASAP_data.r

@@ -17,15 +17,15 @@ data_4IRY_A  = read.csv( "correlation_analysis/combined_data/data_4IRY_A.csv" )
 #data_3GSZ_A  = read.csv( "correlation_analysis/combined_data/data_3GSZ_A.csv" )
 #data_3I5K_A  = read.csv( "correlation_analysis/combined_data/data_3I5K_A.csv" )
 
-ASAP_res_data = rbind(data_1RD8_AB, data_2FP7_B, data_2Z83_A, data_2JLY_A, data_3GOL_A, data_3LYF_A, data_4AQF_B, data_4GHA_A, data_4IRY_A)
-ASAP_res_data$protein = factor(ASAP_res_data$protein)
-ASAP_pdb_data = data.frame()
+ASAP_res_prop = rbind(data_1RD8_AB, data_2FP7_B, data_2Z83_A, data_2JLY_A, data_3GOL_A, data_3LYF_A, data_4AQF_B, data_4GHA_A, data_4IRY_A)
+ASAP_res_prop$protein = factor(ASAP_res_prop$protein)
+ASAP_pdb_prop = data.frame()
 counter = 0
-for (pdb in levels(ASAP_res_data$protein))
+for (pdb in levels(ASAP_res_prop$protein))
 {
   counter = counter + 1
   cat (counter[[1]][1], ' ', pdb, '\n')
-  pdb_data = ASAP_res_data[ASAP_res_data$protein == pdb,]
+  pdb_data = ASAP_res_prop[ASAP_res_prop$protein == pdb,]
   x = cor.test(pdb_data$entropy, pdb_data$rsa_avg_md, method = 'spearman')
   r.seqent.rsa = x$estimate
   x = cor.test(pdb_data$entropy, pdb_data$wcn_avg_md, method = 'spearman')
@@ -48,11 +48,11 @@ for (pdb in levels(ASAP_res_data$protein))
                     median.rsa     = median(pdb_data$rsa_avg_md)
                     )
   
-  ASAP_pdb_data = rbind(ASAP_pdb_data,row)
+  ASAP_pdb_prop = rbind(ASAP_pdb_prop,row)
 }
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-write.csv(ASAP_pdb_data, "../tables/ASAP_pdb_prop.csv", row.names = F)
+write.csv(ASAP_pdb_prop, "../tables/ASAP_pdb_prop.csv", row.names = F)
 
 

### 2014-11-11 analysis/src/lm_PEV.r

@@ -37,32 +37,95 @@ lfit = lm(all_pdb_prop_subset$r.seqent.wcnSC ~ ., regressor)
 
 
 
-# Now generate a single file containing all coefficients for the 5 most important structure-sequence correlations:
+# Now generate a single file containing all coefficients for the 4 most important structure-sequence correlations:
 
+# Fisrt seqent-structure correlations
 all_ssscors = abs(subset(all_pdb_prop_cormat, select = names(all_pdb_prop_cormat) %in% cornames_seqent))
 all_ssscors = data.frame(variable = all_pdb_prop_cormat$X,all_ssscors)
-#all_ssscors = all_ssscors[with(all_ssscors, order(-r.seqent.wcnSC)),]
-all_ssscors = all_ssscors[with(all_ssscors, order(-r.rsa.seqent)),]
+all_ssscors = all_ssscors[with(all_ssscors, order(-r.seqent.wcnSC)),]
+#all_ssscors = all_ssscors[with(all_ssscors, order(-r.rsa.seqent)),]
 
 #plot( log10(order(-all_ssscors$r.seqent.wcnSC))
-plot( log10(order(-all_ssscors$r.rsa.seqent))
-    , all_ssscors$r.rsa.seqent
+pdf( "../figures/PEV_structure_seqent_cors_rank_data_wcnSC_ordered.pdf", width=4.5, height=4, useDingbats=FALSE )
+#pdf( "../figures/PEV_structure_seqent_cors_rank_data_RSA_ordered.pdf", width=4.5, height=4, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(
+    log10(order(-all_ssscors$r.seqent.wcnSC))
+    #log10(order(-all_ssscors$r.rsa.seqent))
+    , all_ssscors$r.seqent.wcnSC^2
+    #, all_ssscors$r.rsa.seqent^2
     , type = 'l'
     #, xlim = c(0,3)
-    , ylim = c(0,0.50)
+    , ylim = c(0,0.3)
+    , xlab = "Log10 (Order of Explanatory Variables)"
+    , ylab = "Proportion of Explained Variance (Rank Data)"
     )
[…]

### 2014-11-11 analysis/src/pdb_prop_figures.r

@@ -57,17 +57,17 @@ dev.off()
 # Now generate a similar screen plot of the correlation of r4sJC-wcnSC vs. r4sJC-other
 filename = paste0('../figures/cordiv_similarities_r4sJC.pdf')
 pdf( filename, width=4.5*n.screen.cols, height=4*n.screen.rows, useDingbats=FALSE )
-column_list = c('r.r4sJC.rsa'
-                ,'r.ddgent.r4sJC'
-                ,'r.bfSC.r4sJC'
-                ,'r.r4sJC.varea'
-                ,'r.r4sJC.vsphericity'
-                ,'r.r4sJC.veccentricity'
-                )
-name_list = c('( Evol. Rates - RSA )'
+column_list = c('r.r4sJC.varea'
+               ,'r.r4sJC.rsa'
+               ,'r.ddgent.r4sJC'
+               ,'r.bfSC.r4sJC'
+               ,'r.r4sJC.vsphericity'
+               ,'r.r4sJC.veccentricity'
+               )
+name_list = c('( Evol. Rates - Voronoi Cell Surface Area )'
+              ,'( Evol. Rates - RSA )'
               ,'( Evol. Rates - ddG Entropy )'
               ,'( Evol. Rates - Mean Residue Bfactor )'
-              ,'( Evol. Rates - Voronoi Cell Surface Area )'
               ,'( Evol. Rates - Voronoi Cell Compactness )'
               ,'( Evol. Rates - Voronoi Cell Symmetry )'
               )
@@ -80,7 +80,7 @@ for (counter in seq(1,n.screen.rows*n.screen.cols))
   plot(abs(all_pdb_prop_select_wide[[column_list[counter]]]),
        abs(all_pdb_prop_select_wide$r.r4sJC.wcnSC),
        xlab = bquote('abs.' ~ rho ~ .(name_list[counter])),
-       ylab = expression(paste('abs. ', rho,' ( Seq. Entropy - Side Chain WCN )')),
+       ylab = expression(paste('abs. ', rho,' ( Evol. Rates - Side Chain WCN )')),
        xlim = c(0.,0.9),
        ylim = c(0.,0.9),
        #cex.axis = 1.4,
[…]

### 2014-11-11 manuscript/cordiv.tex
SAP pdb data added

@@
 
-        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also apply multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the models, and the pdb-level properties as the explanatory variables.
-Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression ( reference to R package xx) on the entire dataset of pdb-level properties, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
+        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the models, and the pdb-level properties as the explanatory variables.
+Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset of pdb-level properties, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
 Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
-
 In addition to regularized regression, we have also employed Principal Component Regression on the original dataset and its rank transformation, all of which point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
@@
             \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
-The prefixes {\it r, sd \& mn}}
+The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), the variance of hydrogen bond energies ({\it var.hbe}), and the fraction of helical \& $\beta-$sheet secondary structures in the proteins ({\it mean.helix} \& {\it mean.betas} respectively).}
             \label{fig:mainmods}

### 2014-11-12 manuscript/cordiv.tex
some minor changes in seach for structural differences between viral and cellular proteins.

@@
     Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations.
-The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with correlations ranging from $0.1$ to $0.8$.
+The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
 In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes.
-We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships.
+We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths.
+In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations.
+In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
 \end{Abstract}

### 2014-11-12 analysis/src/compare_cellular_viral_pdb_prop.r

@@ -1,0 +1,23 @@
+# I am writing this R script in an attempt to find potential structural differences between viral (ASAP) and cellular (Echave) proteins.
+# Last updated by Amir Shahmoradi, Wednesday 12:23 PM, Nov 12 2014, Wilke Lab, ICMB, UT Austin
+
+# input files:  
+#               ../../elj_pdb_entropies.in
+#               ../../properties/res_prop_HPS_asap.out
+#               ../../properties/res_prop_dssp_asap.out
+#               ../../properties/res_prop_wcn_bf_asap.out
+
+
+#install.packages("reshape2")
+#library("reshape2")
+#library('corrplot')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+pdb_prop_from_residue_prop = read.csv( "../tables/pdb_prop_from_residue_prop.csv", header=T )
+pdb_prop_from_residue_prop_asap = read.csv( "../tables/pdb_prop_from_residue_prop_asap.csv", header=T )
+
+for ()
+{
+    
+}

### 2014-11-12 analysis/src/summarize_res_data_asap.r

@@ -1,0 +1,210 @@
+# This R function takes in the residue properties of all ASAP proteins from DSSP, and PDB files and and combines them into a single dataframe for further analysis at pdb level.
+
+# input files:  
+#               ../../elj_pdb_entropies.in
+#               ../../properties/res_prop_HPS_asap.out
+#               ../../properties/res_prop_dssp_asap.out
+#               ../../properties/res_prop_wcn_bf_asap.out
+
+# Last updated by Amir Shahmoradi, Thursday 3:40 PM, Aug 28 2014, Wilke Lab, ICMB, UT Austin
+
+#install.packages("reshape2")
+#library("reshape2")
+#library('corrplot')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+#res_prop_elj            = read.table('../../elj_pdb_entropies.in', header=T)
+#res_prop_elj$pdb        = factor(res_prop_elj$pdb)
+                         
+#res_prop_jec            = read.csv('../../jec_pdb_r4s.csv', header=T)
+#res_prop_jec$pdb        = factor(res_prop_jec$pdb)
+                         
+res_prop_hps_asap        = read.table('../../properties/res_prop_HPS_asap.out', header=T)
+res_prop_hps_asap$pdb    = factor(res_prop_hps_asap$pdb)
+                         
+res_prop_dssp_asap       = read.table('../../properties/res_prop_dssp_asap.out', header=T)
+res_prop_dssp_asap$pdb   = factor(res_prop_dssp_asap$pdb)
+
+res_prop_wcn_bf_asap     = read.table('../../properties/res_prop_wcn_bf_asap.out', header=T)
+res_prop_wcn_bf_asap$pdb = factor(res_prop_wcn_bf_asap$pdb)
+
+res_prop_voroSC_asap     = read.table('../../properties/res_prop_voronoiSC_asap.out', header=T)
+res_prop_voroSC_asap$pdb = factor(res_prop_voroSC_asap$pdb)
+res_prop_voroSC_asap     = cbind(res_prop_voroSC_asap, VSCmodified_volume = res_prop_voroSC_asap$VSCvolume)
+maxval = max(res_prop_voroSC_asap$VSCvolume)
[…]

### 2014-11-12 properties/src/get_res_prop_dssp.py

@@ -52,7 +52,7 @@ Usage:'''
        sum_out_file = open(sum_out,'a')
     else:
        sum_out_file = open(sum_out,'w')
-       sum_out_file.write('pdb' + '\t' + 'resnam' + '\t' + 'resnum' + '\t' + 'asa' + '\t' + 'rsa' + '\t' + 'hbe_mean' + '\t' + 'rss' + '\n')
+       sum_out_file.write('pdb' + '\t' + 'resnam' + '\t' + 'resnum' + '\t' + 'asa' + '\t' + 'rsa' + '\t' + 'hbe' + '\t' + 'rss' + '\n')
     
     ######################################################################################################################################
     # Now calculate the mean of all 4 hydrogen bonds energies and residue RSAs for each individual residue:

### 2014-11-12 properties/src/get_pdb_prop_CO.f90
Contact order for viral proteins added. structural properties for thermophilic proteins also added.

@@ -96,10 +96,11 @@ implicit none
 ! FIRST DETERMINE THE FOUR-LETTER NAME OF THE PDB STRUCTURE:
   pdb_in = adjustr(pdb_in)
   i = len(adjustr(pdb_in))
-  if (pdb_in(i-3:i) /= '.pdb' .or. pdb_in(i-10:i-10) /= '_') then
+  if (pdb_in(i-3:i) /= '.pdb') then ! .or. pdb_in(i-10:i-10) /= '_') then
     write(*,*) 'WARNING: pdb filename does not seem to be correct'
     write(*,*) 'pdb filename: ', pdb_in(i-9:i)
-    read(*,*)
+    write(*,*) 'The last 10 letters of pdb filename must be of the form: ***_*.pdb'
+    STOP  ! read(*,*)
   else
     pdb_name = pdb_in(i-9:i-4)
     write(*,*); write(*,*) 'pdb name: ', pdb_name
@@ -133,7 +134,7 @@ implicit none
       exit
     else
       if (trim(adjustl(record(1)))=='ATOM') natoms = natoms + 1
-      if (trim(adjustl(atom_name(1))) == 'CA') then
+      if (trim(adjustl(record(1)))=='ATOM' .and. trim(adjustl(atom_name(1))) == 'CA') then
         if (nres == 0) then
           nres = nres + 1
         elseif (res_num_old == res_num(1) .and. res_nam_old == res_name(1) .and. res_cod_old == res_code(1)) then
@@ -304,7 +305,7 @@ implicit none
   allocate (no_residue_contact(nres,nres))
   no_residue_contact = .TRUE.     ! Initially assume no contact between any two atoms of a pair of residues.
   ncontacts          = 0
-  do i = 1,natoms-1
+  do i = 1,natoms-1 
     do j = i+1,natoms
       not_same_residue = res_num(i) /= res_num(j) .or. res_name(i) /= res_name(j) .or. res_code(i) /= res_code(j)
       ! safety check point
@@ -313,6 +314,7 @@ implicit none
           stop
         end if
       if ( not_same_residue .and. no_residue_contact(res_num_renumbered(i),res_num_renumbered(j)) ) then
+        !write(*,*) 'Calculating CA contact order for residue number ', res_num(i)
[…]

### 2014-11-12 rate_calculations/entropy/prelim_entropies/calculate_entropies.py
Adding the prelim entropies.

@@ -1,0 +1,100 @@
+#!/usr/bin/python
+import cor_div_helper as ch
+import numpy as np
+import subprocess
+import sys
+
+'''
+Last Updated: November 12, 2014
+Description: This is a script that calculates entropy values for the viral sequence alignments
+'''
+
+aa_list = ['G', 'A', 'L', 'V', 'I', 'P', 'R', 'T', 'S', 'C', 'M', 'K', 'E', 'Q', 'D', 'N', 'W', 'Y', 'F', 'H'] #List of amino acids
+
+def calculate_aa_frequencies(list_of_aligned_seqs): 
+	'''
+	This function calculates the amino acid frequencies for each site in an alignment
+	Args: 
+		list_of_aligned_seqs: A list of sequences that have been aligned
+	
+	Returns:
+		all_freqs: A list of lists where each list are the amino acid frequencies for a site
+	
+	'''
+	all_freqs = []
+	aligned_seqs = []
+	for seq in list_of_aligned_seqs:
+		aligned_seqs.append(ch.get_sequence_array(seq)) #Turn the lists of sequences into a list of arrays (where each array represents a sequence 
+	seq_alignment = np.array(aligned_seqs)
+	rows, cols = seq_alignment.shape
+
+	for i in xrange(0, cols): #For each column in the alignment numpy array (each site in the alignment)
+		#print "site: ", i
+		site_freqs = []
+		site = seq_alignment[:,i].tolist()
+		num_aas_at_site = len(site) - site.count('-') #Count the number of amino acids at the site
+		for aa in aa_list: #For each of the 20 amino acid types
+			aa_freq = float(site.count(aa))/float(num_aas_at_site) #Calculate the AA's frequences
+			site_freqs.append(aa_freq) #Append it to the list of frequencies for that site
+		all_freqs.append(site_freqs) #Append it to the list of lists for the frequencies
+	return all_freqs
+
+def calculate_entropy(data_array):
+	'''
+	Calculates the entropy of list of lists where each list represents a site
+	
+	Args:
+		data_array: A 2D numpy array where the rows are sites and the columns represent a value for each of the 20 amino acids
+			(The values could be its frequency at that site or another meaningful calculated quantity)
+	Returns:
+		entropy_values: The entropy value for each site	
+	'''
+	
[…]

### 2014-11-12 analysis/src/summarize_pdb_data_select_variables_asap.r
naming convention fro vairal proteins modified

@@ -1,0 +1,62 @@
+# Amir Shahmoradi, Wednesday 3:31 PM, Nov 12 2014, Wilke Lab, ICMB, UT Austin
+# This code is very similar to 'get_pdb_data.r' with the sole difference that here only select important and best representative variables are used.
+
+#install.packages("reshape2")
+library("reshape2")
+#install.packages("Hmisc")
+#library("Hmisc")
+#library('corrplot')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+nonviral_pdbs = c('1AJ8_A','1AOR_A','1CTS_A','1MP9_A','3GSZ_A','3I5K_A')   # These are thermophilic proteins, in addition to the 2 viral PDBs that are from the same family as 3GOL_A and therefore redundant.
+
+# pdb_general = read.csv('../../properties/pdb_prop_general.csv',header=T)
+# pdb_general = data.frame(pdb  = paste0(pdb_general$Protein,'_',pdb_general$chain),
+#                          sres = pdb_general$Resolution,    # structure resolution
+#                          nseq = pdb_general$Number.of.Sequences   # number of sequences in the alignment
+#                          )
+# pdb_general = pdb_general[!(pdb_general$pdb %in% nonviral_pdbs),]
+# pdb_general$pdb = factor(pdb_general$pdb)
+
+pdb_CO_asap = read.table('../../properties/pdb_prop_CO_asap.out',header=T)
+pdb_CO_asap = pdb_CO_asap[!(pdb_CO_asap$pdb %in% nonviral_pdbs),]
+
+pdb_prop_dssp_asap = read.table('../../properties/pdb_prop_dssp_asap.out',header=T)
+pdb_prop_dssp_asap = cbind( pdb_prop_dssp_asap,
+                       data.frame(sum.nhbpa.dif = pdb_prop_dssp_asap$sum.nhbps - pdb_prop_dssp_asap$sum.nhbas),
+                       data.frame(mean.nhbpa.dif = pdb_prop_dssp_asap$mean.nhbps - pdb_prop_dssp_asap$mean.nhbas)
+                       )
+pdb_prop_dssp_asap = pdb_prop_dssp_asap[!(pdb_prop_dssp_asap$pdb %in% nonviral_pdbs),]
+pdb_prop_dssp_asap$pdb = factor(pdb_prop_dssp_asap$pdb)
+
[…]

### 2014-11-12 analysis/src/summarize_res_data_asap.r

@@ -20,7 +20,7 @@ setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 #res_prop_jec            = read.csv('../../jec_pdb_r4s.csv', header=T)
 #res_prop_jec$pdb        = factor(res_prop_jec$pdb)
                          
-res_prop_hps_asap        = read.table('../../properties/res_prop_HPS_asap.out', header=T)
+res_prop_hps_asap        = read.table('../../properties/res_prop_hps_asap.out', header=T)
 res_prop_hps_asap$pdb    = factor(res_prop_hps_asap$pdb)
                          
 res_prop_dssp_asap       = read.table('../../properties/res_prop_dssp_asap.out', header=T)
@@ -36,13 +36,13 @@ maxval = max(res_prop_voroSC_asap$VSCvolume)
 res_prop_voroSC_asap$VSCmodified_volume[res_prop_voroSC_asap$VSCvolume_change_ratio != 0] = maxval
 res_prop_voroSC_asap$VSCmodified_volume = res_prop_voroSC_asap$VSCmodified_volume * res_prop_voroSC_asap$VSCvolume_change_ratio
 
-excluded_pdbs = c('3GSZ_A','3I5K_A')   # These are the 2 PDBs that are from the same family as 3GOL_A and therefore redundant.
+nonviral_pdbs = c('1AJ8_A','1AOR_A','1CTS_A','1MP9_A','3GSZ_A','3I5K_A')   # These are thermophilic proteins, in addition to the 2 viral PDBs that are from the same family as 3GOL_A and therefore redundant.
 pdb_prop_from_residue_prop_asap = data.frame()    # This dataframe will contain the mean median and variance of sequqence entropy and ddG entropy for each pdb file.
 counter = 0
 
 for(pdb in levels(res_prop_dssp_asap$pdb))
 {
-  if (!(pdb %in% excluded_pdbs))
+  if (!(pdb %in% nonviral_pdbs))
   {
     counter = counter + 1
     cat( paste(str(counter),pdb) )

### 2014-11-12 analysis/src/summarize_res_data_select_variables_asap.r

@@ -1,14 +1,14 @@
-# This R function takes in all the the residue properties of all proteins from all sorts of analysis, and then selects the  most relevant quantites that will be later used in the analyses. Unlike its ancestor code 'get_res_data.r', this code only selects variables that performed the best among other similar variables definitions. For example, among all Bfactor definitions, only the average over SideChain atoms (bfSC) will be taken and used. This is because apparently, as far as I have searched, residue properties that are based on side chain atoms, correlate best with other residue properties. The only exception to this is the H-bond energy (hbe) of residues which correlates best with CA atom properties (such as bfCA and wcnCA).
-# Also, among the three hydrophobicity scales, I am only picking hpshh, which seems to correlate best with other residue properties.
+# This R function takes in the residue properties of all ASAP proteins from DSSP, and PDB files and and combines them into a single dataframe for further analysis at pdb level.
+
+# Last updated by Amir Shahmoradi, Wednesday 3:02 PM, Nov 12 2014, Wilke Lab, ICMB, UT Austin
 
 # input files:  
 #               ../../elj_pdb_entropies.in
-#               ../../properties/res_prop_hps.out
-#               ../../properties/res_prop_dssp.out
-#               ../../properties/res_prop_wcn_bf.out
-#               ../../properties/res_prop_voroSC.out
+#               ../../properties/res_prop_hps_asap.out
+#               ../../properties/res_prop_dssp_asap.out
+#               ../../properties/res_prop_wcn_bf_asap.out
+#               ../../properties/res_prop_voroSC_asap.out
 
-# Last updated by Amir Shahmoradi, Thursday 3:40 PM, Aug 28 2014, Wilke Lab, ICMB, UT Austin
 
 #install.packages("reshape2")
 #library("reshape2")
@@ -16,59 +16,59 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
[…]

### 2014-11-12 rate_calculations/entropy/prelim_entropies/calculate_entropies.py
Adding entropy script

@@ -69,15 +69,14 @@ def calculate_entropy(data_array):
 
 def main(argv = sys.argv):
 	process = subprocess.call("mkdir entropies", shell = "True")
-	alignments = ["aln_aa_H1N1_HA", "aln_aa_H1N1_NP", "aln_aa_HCV_NS5B", "aln_aa_JEV", "aln_aa_crimean_congo.Nucleoprotein", 
-	"aln_aa_dengue_ns3", "aln_aa_marburg.vp35", "aln_aa_riftvalley", "aln_aa_westNile_ns2b", "aln_aa_westnile_chainb"]
-	maps = ["HP_1RD8_AB", "INP_4IRY_A", "HCP_3GOL_A", "JEHN_2Z83_A", "CCHFN_4AQF_B", "DPH_2JLY_A", "MRNABD_4GHA_A", "RVFVNP_3LYF_A", "MRNABD_4GHA_A", "WNPB_2FP7_B" ]
-	map_counter = 0
-	for a in alignments: #For each viral alignment
+	alignments = ["CCHFN", "DPH", "HCP", "HCP", "HCP", "HCP", "HP", "INP", "JEHN","JEHN", "MRNABD", "RVFVNP", "WNPB"]
+	maps = ["CCHFN_4AQF_B", "DPH_2JLY_A", "HCP_3GOL_A", "HCP_3GSZ_A", "HCP_3I5K_A", "HP_1RD8_AB", "INP_4IRY_A", "JEHN_2JLY_A", "JEHN_2Z83_A", "MRNABD_4GHA_A", "RVFVNP_3LYF_A", "WNPB_2FP7_B" ]
+	alignment_counter = 0
+	for m in maps: #For each viral alignment
 		freq_list = []
-		alignment_file = "../alignments/" + a + ".fasta"
-		map_file = "../../pdb_maps/" + maps[map_counter] + ".map"
-		entropy_file = open("entropies/" + a + "_entropies.csv", "w")
+		alignment_file = "../alignments/" + alignments[alignment_counter] + ".fasta"
+		map_file = "../../pdb_maps/" + m + ".map"
+		entropy_file = open("entropies/" + m + "_entropies.csv", "w")
 		entropy_file.write("alignment_pos,pdb_pos,entropy\n")
 		aligned_seqs, headers = ch.get_sequences(alignment_file)
 		aa_freqs = calculate_aa_frequencies(aligned_seqs)
@@ -95,6 +94,6 @@ def main(argv = sys.argv):
[…]

### 2014-11-13 analysis/src/compare_cellular_viral_pdb_prop.r
structural difference between viral proteins and Echave dataset added.

@@ -14,10 +14,53 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-pdb_prop_from_residue_prop = read.csv( "../tables/pdb_prop_from_residue_prop.csv", header=T )
-pdb_prop_from_residue_prop_asap = read.csv( "../tables/pdb_prop_from_residue_prop_asap.csv", header=T )
+all_pdb_prop_select = read.csv( "../tables/all_pdb_prop_select.csv", header=T )
+all_pdb_prop_select$variable = factor(all_pdb_prop_select$variable)
+all_pdb_prop_select_asap = read.csv( "../tables/all_pdb_prop_select_asap.csv", header=T )
+all_pdb_prop_select_asap$variable = factor(all_pdb_prop_select_asap$variable)
 
-for ()
+cellular_viral_pdb_differences = data.frame()
+
+for (vvar in levels(all_pdb_prop_select_asap$variable))
 {
+  if (vvar %in% levels(all_pdb_prop_select$variable))
+  {
+    # Echave data (213 cellular pdbs)
+    cdata = all_pdb_prop_select[all_pdb_prop_select$variable == vvar,]
+    x = quantile(cdata$value, probs = c(0,0.25,0.5,0.75, 1.))
+    crow = data.frame(c.var      = vvar,
+                      c.mean     = mean(cdata$value),
+                      c.median   = x[3],
+                      c.sd       = sd(cdata$value),
+                      c.min      = x[1],
+                      c.q1       = x[2],
+                      c.q3       = x[4],
+                      c.max      = x[5]
+                      )
     
+    # Viral data (9 pdbs)
+    vdata = all_pdb_prop_select_asap[all_pdb_prop_select_asap$variable == vvar,]
+    x = quantile(vdata$value, probs = c(0,0.25,0.5,0.75, 1.))
+    vrow = data.frame(v.var      = vvar,
+                      v.mean     = mean(vdata$value),
+                      v.median   = x[3],
+                      v.sd       = sd(vdata$value),
+                      v.min      = x[1],
+                      v.q1       = x[2],
+                      v.q3       = x[4],
+                      v.max      = x[5]
+                      )
+                      
+    # difference row
+    drow = data.frame(c.v.d.mean   = crow$c.mean - vrow$v.mean,
+                      c.v.d.median = crow$c.median - vrow$v.median,
[…]

### 2014-11-13 analysis/src/summarize_res_data_select_variables_asap.r

@@ -16,11 +16,9 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-# res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
-# res_prop_elj$pdb     = factor(res_prop_elj$pdb)
+# First get ASAP entropies:
 
-# res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
-# res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+
 
 res_prop_hps_asap        = read.table('../../properties/res_prop_hps_asap.out', header=T)
 res_prop_hps_asap$pdb    = factor(res_prop_hps_asap$pdb)

### 2014-11-16 manuscript/cordiv.tex
New figures for validation added.

@@
 In particular, the X-ray crystallographic resolution of the structure and the definition of the \ddg entropy play dominant roles, with Spearman correlation coefficients of $\rho~0.3$, on the strengths of the corresponding sequence-structure relations.
+        \\
 
-        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the models, and the pdb-level properties as the explanatory variables.
-Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset of pdb-level properties, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
+        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.
+Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
 Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
-In addition to regularized regression, we have also employed Principal Component Regression on the original dataset and its rank transformation, all of which point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
+In addition to regularized regression, we have also employed Principal Component Regression (PCR) on the original dataset and its rank transformation.
+Both regression methods, PCR \& regularized, point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
 
@@
             \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
-The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), the variance of hydrogen bond energies ({\it var.hbe}), and the fraction of helical \& $\beta-$sheet secondary structures in the proteins ({\it mean.helix} \& {\it mean.betas} respectively).}
+The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
             \label{fig:mainmods}
@@
             \end{tabular}
-            \caption{Proportion of explained variance of structure-sequence correlations by 182 pdb-level properties considered individually.}
-            \label{fig:modsim}
+            \caption{{\bf Spearman correlation matrices illustrating the similarity of the set of structure and sequence properties that contribute to the strength of sequence--structure relations}.
+{\bf Left:} The correlation matrix for the relation of evolutionary rates ({\it r4sJC}) with structural quantities on the diagonal of the matrix.
+{\bf Right:} The correlation matrix for the relation of sequence entropy ({\it seqent}) with structural quantities on the diagonal of the matrix.}
+            \label{fig:simcormats}
         \end{figure*}
@@
             \end{tabular}
-            \caption{Proportion of explained variance of structure-sequence correlations by 182 pdb-level properties considered individually.}
+            \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
+The black circles represent $209$ proteins used in this work.
+For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}.
             \label{fig:validation}

### 2014-11-16 analysis/src/pdb_prop_figures.r

@@ -125,44 +125,66 @@ dev.off()
 
 # Now plot data for validation
 
-source('input_ASAP_data.r')
+all_pdb_prop_select_wide = read.csv('../tables/all_pdb_prop_select_wide.csv', header = T)
+all_pdb_prop_select_wide_asap = read.csv('../tables/all_pdb_prop_select_wide_asap.csv', header = T)
 
-filename = paste0('../figures/validation_sdseqent.pdf')
-pdf( filename, width=4.5, height=4, useDingbats=FALSE )
+filename = paste0('../figures/validation_var_seqent.pdf')
+pdf( filename, width=6, height=5, useDingbats=FALSE )
 par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
 
-plot( all_pdb_prop_select_wide$sd.seqent
+plot( all_pdb_prop_select_wide$sd.seqent^2
     , all_pdb_prop_select_wide$r.seqent.wcnSC
     , pch = 19
-    , xlab = 'St. Dev. of Seq. Entropy ( sd.seqent )'
-    , ylab = bquote('Spearman' ~ rho ~ 'seqent - wcnSC')
-    , xlim = c(-0.05,0.9)
-    , ylim = c(-0.8,0.1)
+    , xlab = 'Sequence Divergence:  Variance ( Seq. Entropy )'
+    , ylab = bquote('Spearman' ~ rho ~ ':  Seq. Entropy - wcnSC')
+    , xlim = c(-0.1,0.85)
+    , ylim = c(0.2,-0.8)
     )
-points( ASAP_pdb_prop$sd.seqent
-      , ASAP_pdb_prop$r.seqent.wcnCA
+points( all_pdb_prop_select_wide_asap$sd.seqent
+      , all_pdb_prop_select_wide_asap$r.seqent.wcnSC
       , col = 'red'
       , pch = 19
       )
 dev.off()
 
-filename = paste0('../figures/validation_sdhbe.pdf')
-pdf( filename, width=4.5, height=4, useDingbats=FALSE )
+
+filename = paste0('../figures/validation_var_hbe.pdf')
+pdf( filename, width=6, height=5, useDingbats=FALSE )
 par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
 
[…]

### 2014-11-16 analysis/src/summarize_res_data_select_variables_asap.r

@@ -18,7 +18,43 @@ setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
 # First get ASAP entropies:
 
+seqent_1RD8_X = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/1RD8_X_HP_entropies.csv')
+seqent_1RD8_X = data.frame(pdb = '1RD8_X',seqent_1RD8_X)
+seqent_2FP7_B = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/2FP7_B_WNPB_entropies.csv')
+seqent_2FP7_B = data.frame(pdb = '2FP7_B',seqent_2FP7_B)
+seqent_2JLY_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/2JLY_A_DPH_entropies.csv')
+seqent_2JLY_A = data.frame(pdb = '2JLY_A',seqent_2JLY_A)
+seqent_2Z83_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/2Z83_A_JEHN_entropies.csv')
+seqent_2Z83_A = data.frame(pdb = '2Z83_A',seqent_2Z83_A)
+seqent_3GOL_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/3GOL_A_HCP_entropies.csv')
+seqent_3GOL_A = data.frame(pdb = '3GOL_A',seqent_3GOL_A)
+seqent_3GSZ_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/3GSZ_A_HCP_entropies.csv')
+seqent_3GSZ_A = data.frame(pdb = '3GSZ_A',seqent_3GSZ_A)
+seqent_3I5K_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/3I5K_A_HCP_entropies.csv')
+seqent_3I5K_A = data.frame(pdb = '3I5K_A',seqent_3I5K_A)
+seqent_3LYF_A = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/3LYF_A_RVFVNP_entropies.csv')
+seqent_3LYF_A = data.frame(pdb = '3LYF_A',seqent_3LYF_A)
+seqent_4AQF_B = read.csv('../../rate_calculations/entropy/prelim_entropies/entropies/4AQF_B_CCHFN_entropies.csv')
[…]

### 2014-11-21 analysis/src/best_wcn_selected_variables.r
New progress on all aspects, in particular figures

@@ -285,7 +285,43 @@ for (variable in variable_names)
           ylab = paste0('Spearman cor. with ',variable_list[[counter]][1]),
           cex.axis = 1.3,
           cex.lab = 1.3
+          #, ylim = c(0.05,-1.)
           )
 }
 graphics.off()
 
+
+# Now create two box plots in one figure for correlation of r4sJC and seqent with wcnSC.
+
+wcn_scors_all_pdbs = read.csv( "../tables/best_wcn/selected_variables/wcn_scors_all_pdbs.csv", header = TRUE )
+wcn_scors_all_pdbs$variable = factor(wcn_scors_all_pdbs$variable)
+wcn_scors_all_pdbs$wcn = factor(wcn_scors_all_pdbs$wcn)
+
+#variable_list = c('Seq. Entropy','ddG Entropy','RSA','ASA','Hydrophobicity','H-bond Energy')
+#variable_names = c('seqent','ddgent','rsa','asa','hpshh','hbe_mean')
+variable_list = c('Evol. Rates','Seq. Entropy','ddG Entropy','RSA','Hydrophobicity','H-bond Energy')
+variable_names = c('r4s_JC','seqent','ddgent','rsa','hpshh','hbe_mean')
+wcn_list = c('wcnSC','wcnAA','wcnCB','wcnCA','wcnN','wcnC','wcnO')
+counter = 0
+filename = paste0('../figures/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf')
+pdf( filename, width=15, height=4, useDingbats=FALSE )
+split.screen(c(1,2))
+for (variable in variable_names[1:2])
+{ 
+  counter = counter + 1
+  screen(counter)
+  temp_data_variable_long = wcn_scors_all_pdbs[wcn_scors_all_pdbs$variable == variable,c('pdb','wcn','value')]
+  temp_data_variable = reshape(temp_data_variable_long, timevar = 'wcn', idvar = 'pdb', direction = 'wide')
+  temp_data_variable = subset (temp_data_variable, select = -c(pdb))
+  colnames(temp_data_variable) = levels(wcn_scors_all_pdbs$wcn)
+  temp_data_variable = temp_data_variable[wcn_list]
+  par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+  boxplot( temp_data_variable
+         , xlab = 'Representative Weighted Contact Number (wcn)'
[…]

### 2014-11-21 analysis/src/input_ASAP_data.r

@@ -1,3 +1,4 @@
+# ATTN, Wednesday 5:59 PM, Nov 19 2014, : This is an old R code that extracts data from the ASAP original dataset. I have now written a new set of codes that regrenerate all ASAP pdb properties from scratch. 
 # This file reads all data for the ASAP project, in order to compare the correlation strengths of seqent-StructuralVariables with the variance of seqent for each pdb.
 # Amir Shahmoradi, Thursday 2:08 PM, Oct 1, 2014, ICMB, UT Austin
 

### 2014-11-21 rate_calculations/entropy/scripts/calculate_sequence_entropy.py
Adding new files

@@ -5,8 +5,8 @@ import subprocess
 import sys
 
 '''
-Last Updated: November 12, 2014
-Description: This is a script that calculates entropy values for the viral sequence alignments
+Last Updated: November 21, 2014
+Description: This is a script that calculates entropy values for an alignment
 '''
 
 aa_list = ['G', 'A', 'L', 'V', 'I', 'P', 'R', 'T', 'S', 'C', 'M', 'K', 'E', 'Q', 'D', 'N', 'W', 'Y', 'F', 'H'] #List of amino acids
@@ -60,7 +60,6 @@ def calculate_entropy(data_array):
 		for j in xrange(0,len(probs_values)): #Calculate the entropy (Can look up this formula, just the native entropy)
 			value = (float(probs_values[j])*np.log(float(probs_values[j])))
 			entropy_number = entropy_number + value
-			#print -entropy_number
 		if (entropy_number == 0.0):
 			entropy_values.append(0.0) #Append entropy value to entropy at sites. 
 		else:
@@ -68,32 +67,29 @@ def calculate_entropy(data_array):
 	return entropy_values
 
 def main(argv = sys.argv):
-	process = subprocess.call("mkdir entropies", shell = "True")
-	alignments = ["CCHFN", "DPH", "HCP", "HCP", "HCP", "HCP", "HP", "INP", "JEHN","JEHN", "MRNABD", "RVFVNP", "WNPB"]
-	maps = ["CCHFN_4AQF_B", "DPH_2JLY_A", "HCP_3GOL_A", "HCP_3GSZ_A", "HCP_3I5K_A", "HP_1RD8_AB", "INP_4IRY_A", "JEHN_2JLY_A", "JEHN_2Z83_A", "MRNABD_4GHA_A", "RVFVNP_3LYF_A", "WNPB_2FP7_B" ]
 	alignment_counter = 0
-	for m in maps: #For each viral alignment
+	if len( sys.argv ) != 3:
+		print 
+		print "Usage:     ", sys.argv[0], "<input alignment>", "<output entropy file>"
+	else:
+		alignment_file = sys.argv[1]
+		outfile = sys.argv[2]
+		
 		freq_list = []
-		alignment_file = "../alignments/" + alignments[alignment_counter] + ".fasta"
-		map_file = "../../pdb_maps/" + m + ".map"
-		entropy_file = open("entropies/" + m + "_entropies.csv", "w")
[…]

### 2014-11-21 rate_calculations/entropy/scripts/create_pdb_to_alignment_map.py

@@ -1,0 +1,84 @@
+#!/usr/bin/python
+import cor_div_helper as ch
+import subprocess, os, sys, re
+
+'''
+Last Edited By: Eleisha Jackson (November 21, 2014)
+Description: This script create a map called *_map.txt. This map maps a position in the alignment to each position in the pdb sequence. Must have MAFFT installed to use this script.
+'''
+#Not really used but just a dictionary of the pdbs and there associated alignments for a references
+pdb_dict = {"1RD8_X": "HP", "2FP7_B" : "WNPB" , "2JLY_A": "DPH" , "2Z83_A" : "JEHN" ,
+			"3GOL_A" : "HCP" , "3GSZ_A": "HCP" , "3I5K_A": "HCP" , "3LYF_A": "RVFVNP" , 
+			"4AQF_B": "CCHFN" , "4GHA_A": "MRNABD" , "4IRY_A": "INP" }
+	
+def create_pdb_alignment_map(seqs, pdb_res_nums, map_filename):
+	'''
+	Creates a file that maps a pdb seqeunce onto a given alignments. 
+	Args: 
+		seqs: A list of aligned sequences (
+		pdb_res_nums: A list of residue numbers from the pdb file
+		map_filename: The file name to write the map to
+	
+	Returns:
+		Does not return anything but creates a file with a map
+	'''
+
+	
+	alignment_length = len(seqs[0]) 
+	aligned_pdb_seq = seqs[0]
+	#Make map list
+	map_outfile = open(map_filename, "w")
+	map_outfile.write("align_pos\tpdb_pos\tpdb_aa\n")
+	res_counter = 0
+	for j in xrange(0, alignment_length):
+		map_outfile.write(str(j+1) + "\t")
+		if(aligned_pdb_seq[j]!= "-"):
+			map_outfile.write(str(pdb_res_nums[res_counter]) + "\t" + aligned_pdb_seq[j])
+			res_counter = res_counter + 1
+		else:
+			map_outfile.write("NA" + "\t" + "-")
+		map_outfile.write("\n")
+
+def main():
+	if len( sys.argv ) != 4:
+		print '''      Usage:      '''
+		print "     ", sys.argv[0], "<input alignment>", "<pdb file> " "<output entropy file>"
+	else:
+		align_file = sys.argv[1]
+		pdb_file = sys.argv[2]
+		map_filename = sys.argv[3]
+		
+		fileparts = re.split("/", pdb_file)
+		parts = fileparts[-1]
[…]

### 2014-11-21 rate_calculations/entropy_test/my_sequence_entropy.py

@@ -1,0 +1,100 @@
+#!/usr/bin/python
+import cor_div_helper as ch
+import numpy as np
+import subprocess
+import sys
+
+'''
+Last Updated: November 12, 2014
+Description: This is a script that calculates entropy values for the viral sequence alignments
+'''
+
+aa_list = ['G', 'A', 'L', 'V', 'I', 'P', 'R', 'T', 'S', 'C', 'M', 'K', 'E', 'Q', 'D', 'N', 'W', 'Y', 'F', 'H'] #List of amino acids
+
+def calculate_aa_frequencies(list_of_aligned_seqs): 
+	'''
+	This function calculates the amino acid frequencies for each site in an alignment
+	Args: 
+		list_of_aligned_seqs: A list of sequences that have been aligned
+	
+	Returns:
+		all_freqs: A list of lists where each list are the amino acid frequencies for a site
+	
+	'''
+	all_freqs = []
+	aligned_seqs = []
+	for seq in list_of_aligned_seqs:
+		aligned_seqs.append(ch.get_sequence_array(seq)) #Turn the lists of sequences into a list of arrays (where each array represents a sequence 
+	seq_alignment = np.array(aligned_seqs)
+	rows, cols = seq_alignment.shape
+
+	for i in xrange(0, cols): #For each column in the alignment numpy array (each site in the alignment)
+		#print "site: ", i
+		site_freqs = []
+		site = seq_alignment[:,i].tolist()
+		num_aas_at_site = len(site) - site.count('-') #Count the number of amino acids at the site
+		for aa in aa_list: #For each of the 20 amino acid types
+			aa_freq = float(site.count(aa))/float(num_aas_at_site) #Calculate the AA's frequences
+			site_freqs.append(aa_freq) #Append it to the list of frequencies for that site
+		all_freqs.append(site_freqs) #Append it to the list of lists for the frequencies
+	return all_freqs
+
+def calculate_entropy(data_array):
+	'''
+	Calculates the entropy of list of lists where each list represents a site
+	
+	Args:
+		data_array: A 2D numpy array where the rows are sites and the columns represent a value for each of the 20 amino acids
+			(The values could be its frequency at that site or another meaningful calculated quantity)
+	Returns:
+		entropy_values: The entropy value for each site	
+	'''
+	
+	entropy_values = []
+	num_residues,num_AA = data_array.shape  
[…]

### 2014-11-21 rate_calculations/entropy_test/sequence_entropy.py

@@ -1,0 +1,74 @@
+#!/usr/bin/python
+# This is a novice python code that reads in a pdb file with a given pdb id and the corresponding fasta sequence alignment file.
+import sys, subprocess, os
+
+from Bio import SeqIO
+from Bio.Seq import Seq
+from Bio.SeqRecord import SeqRecord
+from Bio.PDB.Polypeptide import *
+from Bio.PDB import PDBParser
+from Bio.PDB import PDBIO
+from Bio.PDB import Dice
+import math
+import numpy
+
+
+def readAlignment(filename):
+    align = {}
+    for seq_record in SeqIO.parse(filename,"fasta"):
+        if seq_record.id in align:
+            print "Warning: skipping duplicated entry", seq_record.id
+        else:
+            align[seq_record.id] = seq_record.seq
+    return align
+
+def seqent(align):
+    seqs = align.values()
+    L = len(seqs[0])	# Length of each sequence
+    counts = [{} for i in xrange(L)]
+    entropy_list = []
+    for seq in seqs:
+        for i in xrange(L):
+            c = seq[i]
+            if c not in "ACDEFGHIKLMNPQRSTVWYacdefghiklmnpqrstvwy":
+				continue
+            if c in counts[i]:
+                counts[i][c] += 1
+            else:
+                counts[i][c] = 1
+	entropy = 0.
+    for i in xrange(L):
+		sum_freq = sum(counts[i].values())
+		entropy = 0.
+		for n in counts[i].values():
+			entropy += n*(math.log(float(n)/sum_freq))/sum_freq
+		entropy = -entropy
+		if entropy < 0.:
+			print entropy
+			raw_input ('Are you joking? Entropy is Negative?')
+			print ('This code will terminate itself (equivalent to suicide in human society).')
+			stop
+		print entropy
+		entropy_list.append(entropy/sum_freq)
+    return entropy_list
+
+def main():
+    if len( sys.argv ) != 3:
+        print '''
+  
+  Usage:'''
+        print "     ", sys.argv[0], "<input alignment>", "<output entropy file>"
+    else:
+        filename = sys.argv[1]
+        outfilename = sys.argv[2]
+	align = readAlignment(filename)
+	entropy = seqent(align)
+	outputfile=open(outfilename,'w')
+	i = 0
+	outputfile.write("res_num entropy\n")
+	for item in entropy:
+		i += 1
+		outputfile.write(str(i) + " " + str(item) + "\n")
+	return
+
+main()

### 2014-11-21 rate_calculations/pdb_maps/src/map_alignment_to_structure.py

@@ -1,244 +1,0 @@
-#!/usr/bin/python
-import sys, subprocess, os
-
-from Bio import SeqIO
-from Bio.Seq import Seq
-from Bio.SeqRecord import SeqRecord
-from Bio.PDB.Polypeptide import *
-from Bio.PDB import PDBParser
-from Bio.PDB import PDBIO
-from Bio.PDB import Dice
-
-def parsePDBStructure( pdb_fname ):
-    parser = PDBParser()
-    structure = parser.get_structure('test', pdb_fname)
-    return structure
-
-
-def renumberResidues( structure, new_pdb ):
-    model = structure[0]
-
-    i = 1
-    for chain in model:
-        for residue in chain:
-            residue.id = (' ', i, ' ')
-            i += 1
-
-    w = PDBIO()
-    w.set_structure(structure)
-    w.save(new_pdb)
-
-    return structure
-
-
-def getPDBSequence( pdb_fname, chain_id ):
-    # first we create a renumbered structure of just the chain we're interested in
-    structure = parsePDBStructure( pdb_fname )
-    tmp_file = "tmp_sdielccskeoksmndivls.pdb"
-    Dice.extract( structure, chain_id, 0, 100000, tmp_file )
-    
-    pdb_basename = os.path.basename( os.path.splitext( pdb_fname )[0] )
-    new_pdb = pdb_basename + ".renumbered.pdb"
-    structure = parsePDBStructure( tmp_file )
-    renumbered_structure = renumberResidues( structure, new_pdb )
-    
-    polypeptides = ""
-    ppb=PPBuilder()
-    for pp in ppb.build_peptides(structure):
-        polypeptides += pp.get_sequence()
-    os.unlink( tmp_file )
-    return polypeptides
-
-
-def readAlignment( filename ):
-    align = {}
-    for seq_record in SeqIO.parse( filename, "fasta"):
-        if seq_record.id in align:
-            print "Warning: skipping duplicated entry", seq_record.id
-        else:
-            align[seq_record.id] = seq_record.seq
-    return align
-
-
-def createMafftInput( align_fname, pdb_record, mafft_in_fname ):
-    # first write out pdb sequence
-    out_handle = open( mafft_in_fname, "w" )
-    SeqIO.write( [pdb_record], out_handle, "fasta" )
-
-    # now just copy over the input alignment
-    in_handle = open( align_fname, "r" )
-    for line in in_handle:
[…]

### 2014-11-21 manuscript/manuscript_cordiv.tex

--- 2014-11-16 manuscript/cordiv.tex
@@
 The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
-%\def\support{\footnote{Work supported by the Office of Science,Kingdom of the Two Sicilies, under contract OSS--32456.}}
+%\def\support{\footnote{Work supported by the BEACON-NSF center for the study of evolution in action, under contract xxxxxx.}}
 
@@
 \vfill
-\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
+%\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
+\Title{Dissecting the relationship between protein structure and sequence variation}
 \vfill
-\Author{Amir Shahmoradi, Claus Wilke, Eleisha Jackson, ...}%\support}
+\Author{Amir Shahmoradi, Eleisha Jackson, Claus Wilke}%\support}
 \Address{\affiliation}
@@
 \begin{Abstract}
-    Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations.
+    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution.
+The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
+Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes.
+We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations.
+In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations.
+In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relation.
+
+    %Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations.
 The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
@@
         Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work.
-Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., wcnSC), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
-In addition, we have derived new site-specific properties based on Voronoi Tessellation of protein 3D structures that are comparable to or better than several previously known structural properties in explaining site specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
+Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., {\it wcnSC}), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
+In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
 Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,
@@
 We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
-Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates $0.12$ in terms of Spearman correlation strength.
+Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates by as much as $0.12$ in terms of Spearman correlation strength.
+
+        In search for the determinants of the strength of sequence-structure relations, we compiled a set of more than $200$ protein properties for a dataset of $209$ monomeric enzymes.
+By employing several independent parametric and non-parametric statistics, such as Spearman rank test, regularized regression and Principal Regression methods, we identify sequence divergence as the dominant factor in the strength of sequence-structure correlations, capable of explaining $10-30\%$ of the observed correlation strengths alone, in both the original and rank-transformed data.
 
@@
 
+%\clearpage
+%\newpage
+
+%\section*{Figures}
+
     \begin{figure}[tbh]
             \begin{center}
-            \includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
+            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
+            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
             \end{center}
@@
     \end{figure}
+
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
+        \end{center}
+        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
+The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
+The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+        \label{fig:best_wcn}
+    \end{figure}
+
+    \begin{figure}[tbh]
+        \centering
+        \includegraphics[width=3.6in]{/main_modulators_cormat_r4s_wcn.pdf} \\
+        \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
+The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
+        \label{fig:mainmods}
+    \end{figure}
+
+    \begin{figure*}[tbh]
+        \centering
+        %\begin{tabular}{cc}
+        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
+        %\end{tabular}
+        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
+The black circles represent $209$ proteins used in this work.
+For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
+        \label{fig:validation}
+    \end{figure*}
+
+%\newpage
+
+%\section*{Supplementary Figures}
+
+    %\centerline{\includegraphics[width=6.5in]{PC_screen_entropy_CSrmsf.pdf}}
+    %\noindent \textbf{Fig.
+S1} Principal Component (PC) Regression of sequence entropy against the structural variables, including CS RMSF.
+{\bf (A)} Variance in entropy explained by each principal component.
+For most proteins, PC1 and either PC2 or PC3 show the strongest correlations with sequence entropy.
+Significant correlations ($P<0.05$) are shown as filled symbols, and insignificant correlations ($P\geq0.05$) are shown as open symbols.
+{\bf (B)} and {\bf (C)} Composition of the three leading components.
+Red arrows represent the loadings of each of the structural variables on the principal components; black dots represent the amino acid sites in the PC coordinate system.
+    %\customlabel{fig:cor_entropy_PC_screen_CSrmsf}{S1}
 
@@
         \begin{center}
-            \includegraphics[width=6.9in]{/best_vorvol/selected_variables/boxplot_vorvol_all_in_one.pdf}
+        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
         \end{center}
@@
         \label{fig:best_voronoi}
-        \end{figure}
-
-        \begin{figure}[tbh]
-            \centering
-            \includegraphics[width=7.2in]{/main_modulators_cormat.pdf} \\
-            \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
-The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
-            \label{fig:mainmods}
     \end{figure}
@@
         \centering
-            \begin{tabular}{cc}
-            \includegraphics[width=3.4in]{/validation_sdseqent.pdf} & \includegraphics[width=3.4in]{/validation_sdhbe.pdf} \\
-            \end{tabular}
+        %\begin{tabular}{cc}
+        \includegraphics[width=3.4in]{/validation_r_seqent_wcn_var_seqent.pdf}
+        %\end{tabular}
         \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
 The black circles represent $209$ proteins used in this work.
-For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}.
+For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
         \label{fig:validation}

### 2015-01-29 manuscript/cordiv/manuscript_cordiv.tex
manuscript now divided into two separate articles

--- 2014-11-21 manuscript/manuscript_cordiv.tex
@@
 
-\graphicspath{{../analysis/figures/}}
+\graphicspath{{../../analysis/figures/}}
 
@@
 In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations.
-In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relation.
+In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relations.
 
@@
 This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein.
-According to stability threshold model of proteins (xx), substitutions at specific sites may be tolerated more than other sites in the protein sequence, if the substitution does not significantly change the energy landscape and therefore the functional conformation of the protein.
+According to stability threshold model of proteins (xx), some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation.
 Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins (xx).
-As one of the earliest discovered examples, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
+One of the earliest discovered examples of such relations, is the correlation of the site-specific Relative Solvent Accessibility (RSA) with sequence variation, that is, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
     \\
 
-    Other structural properties have also been identified recently and proposed to influence or explain the site-specific evolutionary variations of proteins.
-Among the simplest properties is a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest, named residue {\it contact number (CN)}(xx).
-Variants of this quantity that eliminate the free-parameter (i.e., radius) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx).
+    Other structural properties have also been recently identified and proposed to influence or explain the site-specific evolutionary variations of proteins.
+Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest (xx).
+Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx).
 In a recent work Echave et al.
-xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins (\ddg) to the rate at which mutations accumulate at the corresponding sites over evolutionary time.
+xx presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time.
 They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
@@
 
-    Not all proteins exhibit the same correlation strength and association between sequence variation and structural properties.
+    Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures.
 By analyzing a data set of $216$ monomeric enzymes, Yeh et al.
-xx find a wide range of $\rho\sim0.1-0.8$ for the distributions of the correlations of sequence variability with two site-specific properties: contact number and residue solvent accessibility.
+xx found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA).
 Similarly, Echave et al.
-xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg with evolutionary rates.
-Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
+xx recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
+It appears that sequence-structure correlations tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
 This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
@@
 
-    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence evolution, implies the existence of one or more structural or evolutionary characteristics of protein that modulate sequence-structure correlations in all proteins.
-Motivated by this observation, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide-range variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
+    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability, implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins.
+Motivated by these observations, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
 We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations.
+In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.
+In the following sections, we present evidence in support of these findings and discuss their implications.
     \\
@@
 
-    The results presented in this work are based on data set of 209 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
+    The results presented in this work are based on a dataset of $209$ monomeric enzymes (Echave papers, Wilke ddg paper xx) randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
+2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
 1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
 2009; Ashkenazy et al.
@@
 1.
-Residue-level properties: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the former include site-specific Relative Solvent Accessibility (RSA) (Tien et al.
-2012 xx), Weighted Contact Number (WCN) (shih?
+{\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
+Prominent examples of the site-specific structural properties include RSA (Tien et al.
+2012 xx), WCN (shih?
 xx).
 2.
-PDB-level properties: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include pdb Contact Order (CO) as defined by xx, protein size, protein compactness.
+{\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
+Examples include pdb Contact Order (CO) as defined by xx, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
 In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
+Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.
 A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.
@@
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+%    \begin{figure}[tbh]
+%        \begin{center}
+%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
+%        \end{center}
+%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
 In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
@@
 The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-        \label{fig:best_wcn}
-    \end{figure}
+%        \label{fig:best_wcn}
+%    \end{figure}
 
@@
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+%    \begin{figure}[tbh]
+%        \begin{center}
+%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
+%        \end{center}
+%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
 In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
@@
 The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-        \label{fig:best_wcn}
-    \end{figure}
-
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset.
+%        \label{fig:best_wcn}
+%    \end{figure}
+%
+%    \begin{figure}[tbh]
+%        \begin{center}
+%        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
+%        \end{center}
+%        \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset.
 In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other.
@@
 The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-        \label{fig:best_bf}
-    \end{figure}
-
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
-        \end{center}
-        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
-        \label{fig:best_voronoi}
-    \end{figure}
+%        \label{fig:best_bf}
+%    \end{figure}
+%
+%    \begin{figure}[tbh]
+%        \begin{center}
+%        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
+%        \end{center}
+%        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
+%        \label{fig:best_voronoi}
+%    \end{figure}
 

### 2015-01-29 manuscript/voronoi/manuscript_voronoi.tex
manuscript now divided into two separate articles

--- 2014-11-21 manuscript/manuscript_cordiv.tex
@@
 %%  address and funding acknowledgement data:
-\def\affiliation{Department of Physics \\
-The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
+\def\affiliation{$^1$ Department of Physics, The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu \\
+                 $^2$ Department of Integrative Biology, The University of Texas at Austin, TX 78712, USA; wilke@austin.utexas.edu
+                 }
 %\def\support{\footnote{Work supported by the BEACON-NSF center for the study of evolution in action, under contract xxxxxx.}}
@@
 \vfill
-%\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
-\Title{Dissecting the relationship between protein structure and sequence variation}
+\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
 \vfill
-\Author{Amir Shahmoradi, Eleisha Jackson, Claus Wilke}%\support}
+\Author{Amir Shahmoradi$^{1}$, Claus Wilke$^2$}
 \Address{\affiliation}
@@
 \begin{Abstract}
-    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution.
-The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
-Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes.
-We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations.
-In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations.
-In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relation.
-
-    %Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations.
-The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
-In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes.
-We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths.
-In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations.
-In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
+    A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence.
+Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset.
+We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.
+Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates.
+This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability.
+Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins.
+On average, the choice of geometrical centers of side-chains vs.
+the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths.
+We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
+By contrast, other definitions perform equally well or better.
+This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
 \end{Abstract}
@@
 
-    Patterns of amino acid sequence variation are known to be influenced by the function of proteins (xx).
-The general consensus, based on the flurry of research done over the past several decades, is that the amino-acid sequence determines the 3D structure of proteins, known as the native conformation.
-This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein.
-According to stability threshold model of proteins (xx), substitutions at specific sites may be tolerated more than other sites in the protein sequence, if the substitution does not significantly change the energy landscape and therefore the functional conformation of the protein.
-Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins (xx).
-As one of the earliest discovered examples, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
+    Over the past decade, extensive attempt has been made to predict protein sequence evolution from structural properties.
+A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins (xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
 \\
 
-    Other structural properties have also been identified recently and proposed to influence or explain the site-specific evolutionary variations of proteins.
-Among the simplest properties is a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest, named residue {\it contact number (CN)}(xx).
-Variants of this quantity that eliminate the free-parameter (i.e., radius) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx).
-In a recent work Echave et al.
-xx presented a biophysical model that links the thermodynamic stability changes due to mutations at sites in proteins (\ddg) to the rate at which mutations accumulate at the corresponding sites over evolutionary time.
-They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
+    Although structural characteristics have been individually extensively studied and explored, it is yet unknown whether these seemingly independent quantities are merely different manifestations of an underlying more fundamental characteristics of individual sites in proteins.
+It is perceivable that quantities such as B factor, RSA, and CN can all serve as {\it approximate} measures of local packing density of individual  sites in proteins, or the local flexibility of individual amino acids.
+Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, in contrast to local flexibility.
+Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
+A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure.
+An estimate of the accessible volume to each site can be obtained through Contact Number.
+The definition of this CN however, depends on an arbitrary parameter: the radius of neighborhood definition around each site.
+An alternative more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure.
 \\
 
-    Not all proteins exhibit the same correlation strength and association between sequence variation and structural properties.
-By analyzing a data set of $216$ monomeric enzymes, Yeh et al.
-xx find a wide range of $\rho\sim0.1-0.8$ for the distributions of the correlations of sequence variability with two site-specific properties: contact number and residue solvent accessibility.
-Similarly, Echave et al.
-xx find a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg with evolutionary rates.
-Interestingly however, it appears that sequence-structure correlations also tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
-This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
+    Motivated by this gap in the current understanding of the role of flexibility and other structural properties on sequence-structure relation in globular proteins, here we employ tessellation methods from the field of computational geometry to calculate several new site-specific quantities for globular proteins, including new measures of site-specific flexibility.
+Contrary to what is currently perceived about the role of flexibility in sequence variability, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
+For structural properties that are calculated based on a set of representative site coordinates, particularly Contact Number, we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
+We also show that the original kernel proposed for the definition of Weighted Contact Number by (xx) and supported further by (xx) and extensively used in other studies, has no significant advantage whatsoever in predicting B factors or the sequence variability, when compared to other possible types of kernels.
+A discussion of the methodology used in this work, the results and implications of our findings on the energy landscape of proteins and sequence-structure relations will be presented in the following sections.
 \\
 
-    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence evolution, implies the existence of one or more structural or evolutionary characteristics of protein that modulate sequence-structure correlations in all proteins.
-Motivated by this observation, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide-range variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
-We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations.
+\section{Methods}
+\label{sec:methods}
+
+    \subsection*{Voronoi Tessellation}
+
+    There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials.
+For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points.
 \\
 
-\section{Materials and Methods}
-\label{sec:mam}
+    In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of centroid points to partition protein 3D structure according to Voronoi tessellation.
+The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure (e.g., xx), energy landscape (xx) or protein--protein interactions (xx).
+\\
 
-    \subsection*{Sequence Data, Alignments and Evolutionary Rates}
+    Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
+The $209$ proteins (Echave papers, Wilke ddg paper xx) were randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
+2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
+1995).
 
-    The results presented in this work are based on data set of 209 monomeric enzymes randomly picked from the Catalytic Site Atlas 2.2.11 (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$ and including representatives of all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
+    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
+
+     Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific variables.
+For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins.
+There is however no reason to believe this set of atomic coordinates represent individual sites in proteins the best.
+Indeed, some earlier works have already suggested the use of center-of-mass of side chains coordinates to represent the 3D structure of protein (xx).
+More recently, (Echave 2015) have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence entropy.
+The same definition degeneracy also exists for the set of atomic B factors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic B factor (e.g., Halle 2001 xx).
+Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
+\\
+
+    In order to identify which set of atomic coordinates result in the highest prediction power for WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of these variables based on different choices of the representative set of atomic coordinates used.
+These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
+In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
+We also calculate a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
+In rare cases where the side chain atoms had not been not resolved in the PDB file or the amino acid lacks the heavy atom needed (e.g., $C_\beta$ for Glycine).
+The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
+\\
+
+    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+\\
+
+    \subsection*{Contact Number Definition}
+
+    In its simplest mathematical form, the Contact Number (CN) for a given site in protein is defined as the number of amino acids within a fixed radius of neighborhood around it (xx).
+Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms in the calculation of CN.
+A major problem with the traditional definition of contact number, is the existence of an arbitrary parameter -- the radius of neighborhood --  in the definition of CN.
+There is no consensus on the optimal value of this cutoff distance, typically ranging from $7\AA$ to $13\AA$ (e.g., lin, franzosa).
+
+    Several 
+    The statistical kernel often used in the definition of Weighted Contact Number is generally the square of the reciprocal distance between the
+    contacting pair of sites in the protein.
+
+    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
+
+    \subsection*{Dataset}
+
+    The results presented in this work are based on a dataset of $209$ monomeric enzymes (Echave papers, Wilke ddg paper xx) randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
+2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
 1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
 2009; Ashkenazy et al.
@@
 1.
-Residue-level properties: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the former include site-specific Relative Solvent Accessibility (RSA) (Tien et al.
-2012 xx), Weighted Contact Number (WCN) (shih?
+{\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
+Prominent examples of the site-specific structural properties include RSA (Tien et al.
+2012 xx), WCN (shih?
 xx).
 2.
-PDB-level properties: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include pdb Contact Order (CO) as defined by xx, protein size, protein compactness.
+{\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
+Examples include pdb Contact Order (CO) as defined by xx, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
 In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
+Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.
 A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.
-
-    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
-
-    In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s).
-There are however degeneracies in the definition of the some site-specific variables.
-For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins.
-There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-The same definition degeneracy also exists for the set of atomic Bfactors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor (e.g., Halle 2001 xx).
-    \\
-
-    Similar definition degeneracy also exists for the set of coordinates that can be used for Voronoi tessellation of proteins.
-A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other.
-These regions are called Voronoi cells.
-The structure of proteins can be considered as a set of 3D coordinates representing individual sites.
-Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
-    \\
-
-    Thus, for the sake of comprehensiveness and in order to identify the best definitions of structural properties such as WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of properties depending on the choice of the representative set of atomic coordinates used.
-These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
-We also calculate a representative coordinate for each site that is an average over all heavy atom coordinates in the side chain and backbone of the amino acid.
-In rare cases where the side chain atoms are not resolved in the PDB file or the amino acid lacks the heavy atom needed (e.g., $C_\beta$ for Glycine).
-The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
-    \\
-
-    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
-To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
-    \\
-
-    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
 
@@
         \begin{center}
-            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
-            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
-            \end{center}
-            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
-correlations of other structural properties with evolutionary rates.
-Detailed description of the structural properties is given in Section \label{sec:mam}.
-The red lines in each plot represent equality line.
-It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures.
-For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-            \label{fig:seqent_structure_cors}
-    \end{figure}
-
-    \begin{figure}[tbh]
-        \begin{center}
         \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
@@
     \end{figure}
-
-    \begin{figure}[tbh]
-        \centering
-        \includegraphics[width=3.6in]{/main_modulators_cormat_r4s_wcn.pdf} \\
-        \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
-The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
-        \label{fig:mainmods}
-    \end{figure}
-
-    \begin{figure*}[tbh]
-        \centering
-        %\begin{tabular}{cc}
-        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
-        %\end{tabular}
-        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
-The black circles represent $209$ proteins used in this work.
-For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
-        \label{fig:validation}
-    \end{figure*}
 
@@
 
-    \begin{figure*}[tbh]
-        \centering
-        \begin{tabular}{cc}
-        \includegraphics[width=3.4in]{/modulators_similarity_Sr4sJC_scors.pdf} & \includegraphics[width=3.4in]{/modulators_similarity_Sseqent_scors.pdf} \\
-        \end{tabular}
-        \caption{{\bf Spearman correlation matrices illustrating the similarity of the set of structure and sequence properties that contribute to the strength of sequence--structure relations}.
-{\bf Left:} The correlation matrix for the relation of evolutionary rates ({\it r4sJC}) with structural quantities on the diagonal of the matrix.
-{\bf Right:} The correlation matrix for the relation of sequence entropy ({\it seqent}) with structural quantities on the diagonal of the matrix.}
-        \label{fig:simcormats}
-    \end{figure*}
-
-    \begin{figure*}[tbh]
-        \centering
-        %\begin{tabular}{cc}
-        \includegraphics[width=3.4in]{/validation_r_seqent_wcn_var_seqent.pdf}
-        %\end{tabular}
-        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
-The black circles represent $209$ proteins used in this work.
-For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
-        \label{fig:validation}
-    \end{figure*}
-
 \bibliographystyle{apj}

### 2015-02-09 manuscript/voronoi/manuscript_voronoi.tex
significant rearrangement and cleanup of the repository currently being performed

@@
 \begin{Abstract}
-    A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence.
+    A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for potential structural determinants of the site-specific evolution of protein sequence.
+Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset.
+We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.
+Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates.
+This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability.
+Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins.
+On average, the choice of geometrical centers of side-chains vs.
+the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths.
+We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
+By contrast, other definitions perform equally well or better.
+This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
+
+    % OLD abstract:
+    %A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence.
 Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset.
@@
 
+    \subsection*{Protein Dataset}
+
+        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes taken from xx Echave papers, Wilke ddg paper xx randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
+2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
+1995).
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
+2009; Ashkenazy et al.
+2010).
+Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al.
+2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
+The alignments were then used to calculate the site-specific sequence variability for each individual protein in dataset.
+To do so, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+
+        \begin{equation}
+            \label{eqn:shannon}
+            H_i = -\sum_j P_{ij}\ln P_{ij}
+        \end{equation}
+
+        where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
+We use DSSP software (xx) for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+\\
+
+        All data including a list of $209$ proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
+
     \subsection*{Voronoi Tessellation}
@@
 
-    In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of centroid points to partition protein 3D structure according to Voronoi tessellation.
+        In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation.
 The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure (e.g., xx), energy landscape (xx) or protein--protein interactions (xx).
@@
         Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
-The $209$ proteins (Echave papers, Wilke ddg paper xx) were randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
-1995).
-
-    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
-
-     Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific variables.
-For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins.
-There is however no reason to believe this set of atomic coordinates represent individual sites in proteins the best.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chains coordinates to represent the 3D structure of protein (xx).
-More recently, (Echave 2015) have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence entropy.
-The same definition degeneracy also exists for the set of atomic B factors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic B factor (e.g., Halle 2001 xx).
-Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
-\\
-
-    In order to identify which set of atomic coordinates result in the highest prediction power for WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of these variables based on different choices of the representative set of atomic coordinates used.
-These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
-We also calculate a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
-In rare cases where the side chain atoms had not been not resolved in the PDB file or the amino acid lacks the heavy atom needed (e.g., $C_\beta$ for Glycine).
-The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
-\\
-
-    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-\\
+We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
 
@@
 A major problem with the traditional definition of contact number, is the existence of an arbitrary parameter -- the radius of neighborhood --  in the definition of CN.
-There is no consensus on the optimal value of this cutoff distance, typically ranging from $7\AA$ to $13\AA$ (e.g., lin, franzosa).
+There is no consensus on the optimal value of this cutoff distance, but is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx), and sometimes up to $18\AA$ (e.g., xx).
+\\
 
-    Several 
-    The statistical kernel often used in the definition of Weighted Contact Number is generally the square of the reciprocal distance between the
-    contacting pair of sites in the protein.
-
-    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
-
-    \subsection*{Dataset}
-
-    The results presented in this work are based on a dataset of $209$ monomeric enzymes (Echave papers, Wilke ddg paper xx) randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
-1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
-2009; Ashkenazy et al.
-2010).
-Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al.
-2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
-The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset.
-To do so, we relied on two independent methods of measuring sequence variability measure.
-First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+        In an attempt to provide a more general definition of CN, some studies have already suggested an alternative, known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
 
         \begin{equation}
-        \label{eqn:shannon}
-        H_i = -\sum_j P_{ij}\ln P_{ij}
+            \label{eqn:wcn_pwrl}
+            WCN_{i,pwrl} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}},
         \end{equation}
 
-    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
-Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software rate4site (xx).
-To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity (Stamatakis 2014).
-For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) (Mayrose et al.
-2004).
+        in which the use of subscript {\it pwrl} is to emphasize on the power-law kernel used in the definition of WCN.
+Although WCN in general correlates more strongly with $C_\alpha$ atomic B factor and sequence entropy, the proposed definition of WCN still involves an adjustable exponent, which has been traditionally fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl}.
+So far, no physical model has been proposed to support the proposed power-law definition of WCN and the specific value of exponent used.
+\\
 
-    \subsection*{Structural Properties}
+        In order to find the best definition for Contact Number among all possibilities, we first consider the simplest definition of CN described above and calculate CN for all sites in all proteins using $C_\alpha$ atoms as reference coordinates, for a wide range of neighborhood radius $r\in[0.1\AA,50\AA]$ with strides of $\Delta r=0.1\AA$.
+We then calculate the correlation of CN with $C_\alpha$ atomic B factors and sequence entropy.
 
-    The goal of the presented work is to identify the prominent structural or evolutionary properties of proteins that modulate sequence-structure correlations.
-These potential modulators represent a unique characteristics of the protein as a whole.
-In general, the structural and evolutionary properties fall into two major categories.
-1.
-{\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the site-specific structural properties include RSA (Tien et al.
-2012 xx), WCN (shih?
-xx).
-2.
-{\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include pdb Contact Order (CO) as defined by xx, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
-In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
-Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.
-A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.
+        Next, we consider the traditional definition of WCN with power-law kernel.
+In order to find the optimal value of exponent in the power-law definition of WCN we first calculate the weighted contact number for all sites in all $209$ proteins of our dataset using $C_\alpha$ atoms as reference coordinates, for a wide range of power-law exponents $-30\leq\alpha\leq30$ with strides of $0.02$.
+We then calculate the correlation of WCN with the corresponding $C_\alpha$ atomic B factors in each protein, also with site-specific sequence entropy.
+This is done for the entire range of the exponent considered above.
+\\
+
+        We also consider alternative definitions of WCN with kernels other than power-law. These include WCN with a Gaussian kernel defined for the site $i$ in protein as,
+
+        \begin{equation}
+            \label{eqn:wcn_gaus}
+            WCN_{i,Gaus} = \sum^N_{j\neq i} \exp\bigg(-\frac{r_{ij}^2}{\sigma}\bigg),
+        \end{equation}
+
+        with $\sigma\in[0.2\AA,50\AA]$ with strides of $0.2$ acting as the free parameter of the model, and WCN with an exponential kernel defined for the site $i$ in protein as,
+
+        \begin{equation}
+            \label{eqn:wcn_exp}
+            WCN_{i,exp} = \sum^N_{j\neq i} \exp\bigg(-\frac{r_{ij}}{\lambda}\bigg),
+        \end{equation}
+
+        with $\lambda\in[0.2\AA,50\AA]$ with strides of $0.2$ acting as the free parameter of the model.
+
+    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
+
+        Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific variables.
+For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
+The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein (xx).
+More recently, (Echave 2015) have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
+Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility (e.g., Halle 2001 xx), same definition degeneracy also exists on choice of atomic B factors (xx) that are used to represent site-specific flexibility.
+In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
+\\
+
+        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates used.
+These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
+In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
+We also calculate a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
+In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid.
+\\
 
@@
 
-        As explained in previous section, there is a high level of redundancy in the initial set of collected protein properties.
-In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties.
+        As explained in previous section, In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties.
 This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.
@@
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
-The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-        \label{fig:best_wcn}
-    \end{figure}
-
 %\newpage
@@
 
-    %\centerline{\includegraphics[width=6.5in]{PC_screen_entropy_CSrmsf.pdf}}
-    %\noindent \textbf{Fig.
-S1} Principal Component (PC) Regression of sequence entropy against the structural variables, including CS RMSF.
-{\bf (A)} Variance in entropy explained by each principal component.
-For most proteins, PC1 and either PC2 or PC3 show the strongest correlations with sequence entropy.
-Significant correlations ($P<0.05$) are shown as filled symbols, and insignificant correlations ($P\geq0.05$) are shown as open symbols.
-{\bf (B)} and {\bf (C)} Composition of the three leading components.
-Red arrows represent the loadings of each of the structural variables on the principal components; black dots represent the amino acid sites in the PC coordinate system.
-    %\customlabel{fig:cor_entropy_PC_screen_CSrmsf}{S1}
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
+        \end{center}
+        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
+        \label{fig:best_voronoi}
+    \end{figure}
 
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
+        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
         \end{center}
-        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
-        \label{fig:best_voronoi}
+        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
+The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
+The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+        \label{fig:best_wcn}
     \end{figure}

### 2015-02-09 analysis/src/best_varea_select_variables.r

@@ -1,6 +1,6 @@
 
 # This code is aimed at finding which definition of Voronoi cell volume (vvol) and area (varea) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
-# "selected_variables" in the name of this code refers to the fact that I am comparing the performance of different vorvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and varea based on the set of atom coordinates used.
+# "select_variables" in the name of this code refers to the fact that I am comparing the performance of different vorvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and varea based on the set of atom coordinates used.
 
 # Last updated by Amir Shahmoradi, Wednesday 7:24 PM, November 19 2014, Wilke Lab, ICMB, UT Austin
 
@@ -17,63 +17,66 @@
 #library("reshape2")
 #library('corrplot')
 
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# 
+# excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+# npdbs = 209         # number of pdb structures in the dataset
+# 
+# res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+# res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+# res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+# 
+# res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
[…]

### 2015-02-09 analysis/src/best_vorvol_selected_variables.r

@@ -1,297 +1,0 @@
-# This code is aimed at finding which definition of Voronoi cell volumes (vorvol) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
-# "selected_variables" in the name of this code refers to the fact that I am comparing the performance of different vorvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and WCN based on the set of atom coordinates used.
-
-# Last updated by Amir Shahmoradi, Friday 1:23 PM, October 4 2014, Wilke Lab, ICMB, UT Austin
-
-# Input data:
-#               ../../elj_pdb_entropies.in
-#               ../../properties/res_prop_hps.out
-#               ../../properties/res_prop_dssp.out
-#               ../../properties/res_prop_wcn_bf.out
-#               ../../properties/res_prop_voronoiAA.out
-#               ../../properties/res_prop_voronoiCA.out
-#               ../../properties/res_prop_voronoiSC.out
-
-#install.packages("reshape2")
-#library("reshape2")
-#library('corrplot')
-
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
-
-excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
-npdbs = 209         # number of pdb structures in the dataset
-
-res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
-res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
-res_prop_jec$pdb     = factor(res_prop_jec$pdb)
-
-res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
-res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
-res_prop_elj$pdb     = factor(res_prop_elj$pdb)
-
[…]

### 2015-02-09 analysis/src/best_vvol_select_variables.r

@@ -1,0 +1,297 @@
+# This code is aimed at finding which definition of Voronoi cell volumes (vvol) can best represent a residue. This is done by comparing the correlations of different vvols with other residue variables.
+# "select_variables" in the name of this code refers to the fact that I am comparing the performance of different vvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and WCN based on the set of atom coordinates used.
+
+# Last updated by Amir Shahmoradi, Friday 1:23 PM, October 4 2014, Wilke Lab, ICMB, UT Austin
+
+# Input data:
+#               ../../elj_pdb_entropies.in
+#               ../../properties/res_prop_hps.out
+#               ../../properties/res_prop_dssp.out
+#               ../../properties/res_prop_wcn_bf.out
+#               ../../properties/res_prop_voronoiAA.out
+#               ../../properties/res_prop_voronoiCA.out
+#               ../../properties/res_prop_voronoiSC.out
+
+#install.packages("reshape2")
+#library("reshape2")
+#library('corrplot')
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# 
+# excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+# npdbs = 209         # number of pdb structures in the dataset
+# 
+# res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
+# res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
+# res_prop_jec$pdb     = factor(res_prop_jec$pdb)
+# 
+# res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
+# res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
[…]

### 2015-02-09 analysis/src/get_res_data.r

@@ -1,33 +1,52 @@
 # Last updated by Amir Shahmoradi, Monday 1:51 PM, Feb 2 2015, Wilke Lab, ICMB, UT Austin
 
 # This R script first reads in all residue properties for the entire dataset of 209 proteins from different files into R dataframes.
-
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# input files:  
+#               ../../elj_pdb_entropies.in
+#               ../../jec_pdb_r4s.csv
+#               ../../properties/res_prop_hps.out
+#               ../../properties/res_prop_dssp.out
+#               ../../properties/res_prop_wcn_bf.out
+#               ../../properties/res_prop_voronoiSC.out
+#               ../../properties/res_prop_voronoiAA.out
+#               ../../properties/res_prop_voronoiCA.out
+#               ../../properties/res_prop_voronoiCB.out
+#               ../../properties/res_prop_voronoiN.out
+#               ../../properties/res_prop_voronoiC.out
+#               ../../properties/res_prop_voronoiO.out
 
 # First read all residue-level data from different tables into R:
 
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
   excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
   
+  # site-specific entropies
   res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
   res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
   res_prop_elj$pdb     = factor(res_prop_elj$pdb)
   
+  # Rate4Site site-specific evolutionary rates
   res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
   res_prop_jec         = res_prop_jec[!(res_prop_jec$pdb %in% excluded_pdbs),]
   res_prop_jec$pdb     = factor(res_prop_jec$pdb)
   
+  # site-specific Hydrophobicity scales
[…]

### 2015-02-09 analysis/src/main.r

@@ -1,0 +1,14 @@
+# This is the main R script that runs all R scripts in the proper order, to collect, generate, and analyze all data and output results, tables and figures:
+# Last updated by Amir Shahmoradi, Sunday 7:54 PM, Feb 8 2015, Wilke Lab, ICMB, UT Austin
+
+#install.packages("reshape2")
+#library("reshape2")
+#library('corrplot')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+# First collect all residue-level data, that is, all site specific properties.
+source('get_res_data.r')
+
+# Now make plots of correlation of Voronoi cell area with other site-specific variables for all proteins in the form of box plots.
+ 

### 2015-02-09 analysis/src/summarize_res_data.r

@@ -1,46 +1,10 @@
 # This R function takes in the residue properties of all proteins from DSSP, and PDB files and and combines them into a single dataframe for further analysis at pdb level.
 
-# input files:  
-#               ../../elj_pdb_entropies.in
-#               ../../properties/res_prop_hps.out
-#               ../../properties/res_prop_dssp.out
-#               ../../properties/res_prop_wcn_bf.out
-
 # Last updated by Amir Shahmoradi, Thursday 3:40 PM, Aug 28 2014, Wilke Lab, ICMB, UT Austin
 
 #install.packages("reshape2")
 #library("reshape2")
 #library('corrplot')
-
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
-
-res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
-res_prop_elj$pdb     = factor(res_prop_elj$pdb)
-
-res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
-res_prop_jec$pdb     = factor(res_prop_jec$pdb)
-
-res_prop_hps         = read.table('../../properties/res_prop_hps.out', header=T)
-res_prop_hps$pdb     = factor(res_prop_hps$pdb)
-
-res_prop_dssp        = read.table('../../properties/res_prop_dssp.out', header=T)
-res_prop_dssp$pdb    = factor(res_prop_dssp$pdb)
-
-res_prop_wcn_bf      = read.table('../../properties/res_prop_wcn_bf.out', header=T)
-res_prop_wcn_bf$pdb  = factor(res_prop_wcn_bf$pdb)
-
-#res_prop_voroAA      = read.table('../../properties/res_prop_voronoiAA.out', header=T)
-#res_prop_voroAA$pdb  = factor(res_prop_voroAA$pdb)
-
-#res_prop_voroCA      = read.table('../../properties/res_prop_voronoiCA.out', header=T)
-#res_prop_voroCA$pdb  = factor(res_prop_voroCA$pdb)
-
-res_prop_voroSC      = read.table('../../properties/res_prop_voronoiSC.out', header=T)
-res_prop_voroSC$pdb  = factor(res_prop_voroSC$pdb)
-res_prop_voroSC      = cbind(res_prop_voroSC, VSCmodified_volume = res_prop_voroSC$VSCvolume)
-maxval = max(res_prop_voroSC$VSCvolume)
[…]

### 2015-02-10 properties/src/get_res_prop_voronoi_all_atoms.py
Variations in Voronoi Cell areas upon change of box size are now calculated and added to output files.

@@ -246,14 +246,22 @@ Usage:'''
             else:
                  for k,record in enumerate(filecontent):
                      residue_extended_volume = float(record.split()[-4])
+                     residue_extended_area = float(record.split()[-5])
                      free_volume = residue_extended_volume - resvol_dict[resnam[k]]
                      if free_volume < 0.0:
                          print 'Negative Voronoi free volume detected!', pdb_name, resnam[k], str(resnum[k]), str(free_volume)
                      pdb_voro_data[k].append(str(free_volume))
+                     # Distinguish the buried from exposed residues, via changes in cell volumes upon extending the box size
                      volume_change_diff  = residue_extended_volume - float(pdb_voro_data[k][-3])
                      volume_change_ratio = residue_extended_volume / float(pdb_voro_data[k][-3])
+                     # Distinguish the buried from exposed residues, via changes in cell areas upon extending the box size
+                     area_change_diff  = residue_extended_area - float(pdb_voro_data[k][-4])
+                     area_change_ratio = residue_extended_area / float(pdb_voro_data[k][-4])
+                     # Now append the volume and area changes in the cell number k:
                      pdb_voro_data[k].append(str(volume_change_diff))
                      pdb_voro_data[k].append(str(volume_change_ratio))
+                     pdb_voro_data[k].append(str(area_change_diff))
+                     pdb_voro_data[k].append(str(area_change_ratio))
 
         # Now write out data in the output files. First check if the output file currently exists. If not, then create the output and add the file header.
         if os.path.isfile(summary_file[i]):
@@ -264,37 +272,44 @@ Usage:'''
                 output_file.write( 'pdb' + '\t' + 'resnam' + '\t' + 'resnum' + '\t' + 'sizeSC' + '\t' + 'sizeAA' + '\t' + 'resvol' + '\t' \
                                  + 'VSCnvertices' + '\t' + 'VSCnedges' + '\t' + 'VSCedge_length_total' + '\t' + 'VSCnfaces' + '\t' \
[…]

### 2015-02-19 manuscript/voronoi/manuscript_voronoi.tex
New firgures and the corresponding source codes added.

@@
 \begin{Abstract}
-    A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for potential structural determinants of the site-specific evolution of protein sequence.
-Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset.
-We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.
-Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates.
-This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability.
-Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins.
-On average, the choice of geometrical centers of side-chains vs.
-the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths.
+    What are the best structural predictors of protein's sequence evolution?
+A number of site-specific structural properties have been proposed over the past decade to answer this question.
+The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition.
+A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.
+Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution.
+We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometrical average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
 We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
@@
 This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
-
     % OLD abstract:
@@
 
-        As explained in previous section, In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties.
+        As explained in previous section, In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, B factor and Voronoi cell properties.
 This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
+        \includegraphics[width=6.9in]{/best_varea/select_variables/boxplot_varea_all_in_one.pdf}
         \end{center}
-        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
+        \caption{A comparison of the correlation strength of 6 different measures of Voronoi cell areas with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
+The Voronoi cells are generated using 6 sets of atomic coordinates: {\it SC, AA, CB, CA, N, C, O}, used as different representations of individual sites in proteins.
+The two labels {\it SC} \& {\it AA} stand respectively for the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
+        %about different structure or sequence properties for 209 proteins in dataset.
+In each plot, the Spearman correlation strengths of a given structure/sequence property on the vertical axis with different measures of cell areas with different atomic Voronoi seeds on the horizontal axis are compared against each other.
+The capital letters on the horizontal axis denote the set of atomic coordinates used in Voronoi tessellation of protein structures.
+}
         \label{fig:best_voronoi}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
+        \includegraphics[width=6.9in]{/best_wcn/select_variables/boxplot_wcn_all_in_one.pdf}
         \end{center}
-        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
-The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+        \caption{A comparison of the correlation strength of 6 different measures of Weighted Contact Number (WCN) with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
+The contact numbers, WCN, are calculated using 6 sets of atomic coordinates: {\it SC, AA, CB, CA, N, C, O}, used as different representations of individual sites in proteins.
+The two labels {\it SC} \& {\it AA} stand respectively for the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
         \label{fig:best_wcn}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
+        \includegraphics[width=6.9in]{/best_bf/select_variables/boxplot_bf_all_in_one.pdf}
         \end{center}
-        \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors.
-The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
+        \caption{A comparison of the correlation strength of 6 different measures of B factor with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
+Shown on the horizontal axes, are the 6 representative atomic B factors: {\it SC, AA, CB, CA, N, C, O} used as flexibility measures of individual sites in proteins.
+The two variables {\it SC} \& {\it AA} stand respectively for the average B factor of all Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
         \label{fig:best_bf}
@@
 
-    \begin{figure}[tbh]
-        \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
+%    \begin{figure}[tbh]
+%        \begin{center}
+%        \includegraphics[width=6.9in]{/best_wcn/select_variables/boxplot_wcn_two_in_one.pdf}
+%        \end{center}
+%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
 In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.

### 2015-02-19 analysis/src/best_bfactor.r

@@ -85,7 +85,7 @@ for(pdb in levels(res_prop_elj$pdb))
   pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('r4sJC')]
   pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
   pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,]  # c('hpskd','hpsww','hpshh')]
-  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
   pdb_bf     = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, bf_list]
   pdb_wcn    = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb,]
   pdb_voroAA = res_prop_voroAA[res_prop_voroAA$pdb==pdb, ]
@@ -95,8 +95,8 @@ for(pdb in levels(res_prop_elj$pdb))
   pdb_temp = cbind( subset(pdb_jec, select = c(r4s_JC)),
                     subset(pdb_elj, select = c(seqent,ddgent)),
                     subset(pdb_hps, select = c(hpshh)),
-                    #subset(pdb_dssp, select = c(asa,rsa,hbe_mean)),
-                    subset(pdb_dssp, select = c(rsa,hbe_mean)),
+                    #subset(pdb_dssp, select = c(asa,rsa,hbe)),
+                    subset(pdb_dssp, select = c(rsa,hbe)),
                     subset(pdb_wcn, select = -c(pdb,resnam,resnum,bfSC,bfAA,bfN,bfCA,bfC,bfO,bfCB)),
                     subset(pdb_voroAA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,VAAnvertices,VAAnedges,VAAvolume_change_ratio)),
                     subset(pdb_voroCA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,resvol,VCAnvertices,VCAnedges,VCAvolume_change_ratio)),

### 2015-02-19 analysis/src/best_bfactor_select_variables.r

@@ -1,5 +1,5 @@
 # This code is aimed at finding which definition of B factor can best represent a residue. This is done by comparing the correlations of different atomic B factors or defenitions of B factor (SC and AA) with other residue variables.
-# "selected_variables" in the name of this code refers to the fact that I am comparing the performance of different B factors to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of WCN and voronoi Volumes and areas based on the set of atom coordinates used.
+# "select_variables" in the name of this code refers to the fact that I am comparing the performance of different B factors to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of WCN and voronoi Volumes and areas based on the set of atom coordinates used.
 
 # Last updated by Amir Shahmoradi, Wednesday 5:06 PM, October 1 2014, Wilke Lab, ICMB, UT Austin
 
@@ -20,61 +20,62 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
 npdbs = 209         # number of pdb structures in the dataset
-
-res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)
-res_prop_elj         = res_prop_elj[!(res_prop_elj$pdb %in% excluded_pdbs),]
-res_prop_elj$pdb     = factor(res_prop_elj$pdb)
-
-res_prop_jec         = read.csv('../../jec_pdb_r4s.csv', header=T)
[…]

### 2015-02-19 analysis/src/best_structural_predictors_of_ER.r

@@ -1,0 +1,156 @@
+# This code is aimed at finding which structural property best predicts  and area (varea) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
+
+# Last updated by Amir Shahmoradi, Friday 4:25 PM, February 13 2015, Wilke Lab, ICMB, UT Austin
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+all_scors_select_variables = data.frame()    # This dataframe will contain the mean median and variance of sequqence entropy and ddG entropy for each pdb file.
+counter = 0
+
+for(pdb in levels(res_prop_elj$pdb))
+{
+  if (!(pdb %in% excluded_pdbs))
+  {
+    counter = counter + 1
+    cat( paste(str(counter),pdb) )
+    
+    pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
+    pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
+    pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
+    pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+    pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe_mean','rss')] )
+    
+    pdb_temp = cbind( subset(pdb_elj, select = c(seqent,ddgent)),
+                      subset(pdb_jec, select = c(zr4s_JC)),
+                      subset(pdb_hps, select = c(hpskd,hpsww,hpshh)),
+                      subset(pdb_dssp, select = c(asa,rsa,hbe_mean)),
+                      subset(pdb_wcn_bf, select = -c(pdb,resnam,resnum))
+    )
+    
+    pdb_long = reshape(pdb_temp, ids = rownames(pdb_temp), varying = colnames(pdb_temp), v.names = 'value', timevar = 'variable', times = colnames(pdb_temp), direction = 'long')
[…]

### 2015-02-19 analysis/src/best_varea_select_variables.r

@@ -1,4 +1,3 @@
-
 # This code is aimed at finding which definition of Voronoi cell volume (vvol) and area (varea) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
 # "select_variables" in the name of this code refers to the fact that I am comparing the performance of different vorvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and varea based on the set of atom coordinates used.
 
@@ -76,7 +75,7 @@ varea_scors_all_pdbs = data.frame()
 # The above dataframe will contain the mean median and variance of sequqence entropy and ddG entropy for each pdb file.
 # Here scor stands for Sperman CORrelation.
 varea_list = c('vareaSC','vareaAA','vareaCA','vareaCB','vareaN','vareaC','vareaO') #,'wcnCA','wcnN','wcnC','wcnO')
-crd_list = c('SC','AA','CB','CA','N','C','O') #,'wcnCA','wcnN','wcnC','wcnO') # list of representative coordinates used to generate Voronoi cells.
+crd_list = c('SC','AA','CB','CA','N','C','O') # list of representative coordinates used to generate Voronoi cells.
 counter = 0
 
 for(pdb in levels(res_prop_elj$pdb))
@@ -138,8 +137,6 @@ for(pdb in levels(res_prop_elj$pdb))
   }
 }
 write.csv( varea_scors_all_pdbs, "../tables/best_varea/select_variables/varea_scors_all_pdbs.csv", row.names=F )
-
-
 
 # Now summarize all Spearman correlations over the entire dataset
 
@@ -208,103 +205,104 @@ for (i in 1:(length(varea_list)-1))
 
 # Make plots of ABS(varea) vs. ABS(varea) for comparison:
 
-x = -2:2
-colors = c('red', 'blue', 'green', 'purple', 'orange3', 'darkgreen', 'black', 'gray', 'cyan2') #, 'darkred', 'darkgreen', 'bisque2')
-#labels = c('ASA', 'ddG Entropy', 'H-bond energy', 'Hydrophobicity', 'Residue Volume', 'RSA', 'Seq. Entropy')
[…]

### 2015-02-19 analysis/src/best_vspherecity_select_variables.r

@@ -1,0 +1,257 @@
+
+# This code is aimed at finding which definition of Voronoi cell volume (vvol) and area (varea) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
+# "select_variables" in the name of this code refers to the fact that I am comparing the performance of different vorvols to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and varea based on the set of atom coordinates used.
+
+# Last updated by Amir Shahmoradi, Friday 4:04 PM, February 13 2015, Wilke Lab, ICMB, UT Austin
+
+# Input data:
+#               ../../elj_pdb_entropies.in
+#               ../../properties/res_prop_hps.out
+#               ../../properties/res_prop_dssp.out
+#               ../../properties/res_prop_wcn_bf.out
+#               ../../properties/res_prop_voronoiAA.out
+#               ../../properties/res_prop_voronoiCA.out
+#               ../../properties/res_prop_voronoiSC.out
+
+#install.packages("reshape2")
+#library("reshape2")
+#library('corrplot')
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+varea_scors_all_pdbs = data.frame()
+# The above dataframe will contain the mean median and variance of sequqence entropy and ddG entropy for each pdb file.
+# Here scor stands for Sperman CORrelation.
+varea_list = c('vareaSC','vareaAA','vareaCA','vareaCB','vareaN','vareaC','vareaO') #,'wcnCA','wcnN','wcnC','wcnO')
+crd_list = c('SC','AA','CB','CA','N','C','O') #,'wcnCA','wcnN','wcnC','wcnO') # list of representative coordinates used to generate Voronoi cells.
+counter = 0
+
+for(pdb in levels(res_prop_elj$pdb))
+{
+  counter = counter + 1
+  cat( paste(str(counter),pdb) )
+  
+  pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('r4sJC')]
+  pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
[…]

### 2015-02-19 analysis/src/best_wcn_select_variables.r

@@ -1,7 +1,7 @@
 # This code is aimed at finding which definition of WCN can best represent a residue. This is done by comparing the correlations of different atomic WCNs or definitions of WCN (SC and AA) with other residue variables.
-# "selected_variables" in the name of this code refers to the fact that I am comparing the performance of different WCN to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and voronoi Volumes and areas based on the set of atom coordinates used.
+# "select_variables" in the name of this code refers to the fact that I am comparing the performance of different WCN to only a select number of important residue properties, such as ASA, RSA, seqent, ddGent, and I am ignoring the rest which are generally definition dependent, such as different definitions of B factors and voronoi Volumes and areas based on the set of atom coordinates used.
 
-# Last updated by Amir Shahmoradi, Thursday 9:27 PM, October 2 2014, Wilke Lab, ICMB, UT Austin
+# Last updated by Amir Shahmoradi, Wednesday 2:38 PM, Feb 18 2015, Wilke Lab, ICMB, UT Austin
 
 # Input data:
 #               ../../elj_pdb_entropies.in
@@ -16,63 +16,64 @@
 #library("reshape2")
 #library('corrplot')
 
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
 npdbs = 209         # number of pdb structures in the dataset
-
[…]

### 2015-02-19 analysis/src/main.r

@@ -1,14 +1,31 @@
 # This is the main R script that runs all R scripts in the proper order, to collect, generate, and analyze all data and output results, tables and figures:
 # Last updated by Amir Shahmoradi, Sunday 7:54 PM, Feb 8 2015, Wilke Lab, ICMB, UT Austin
 
-#install.packages("reshape2")
-#library("reshape2")
-#library('corrplot')
+# install.packages("reshape2")
+# install.packages("corrplot")
+# install.packages('deldir')  # used by voronoi_diagram.r
+# install.packages('Rpdb')  # used by getcrd.r in voronoi_diagram.r
+# library("reshape2")
+# library('corrplot')
+# library('deldir')
+# library('Rpdb')
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+npdbs = 209 # number of pdbs in dataset. Will be used by other subroutines below
 
 # First collect all residue-level data, that is, all site specific properties.
 source('get_res_data.r')
 
-# Now make plots of correlation of Voronoi cell area with other site-specific variables for all proteins in the form of box plots.
+# Generate data for the best definition of Voronoi cell areas.
+# Make plots of correlation of Voronoi cell area with other site-specific variables for all proteins in the form of box plots.
+source('best_varea_select_variables.r')
+
+# Generate data for the best definition of B factors:
+# Make plots of correlation of B factor with other site-specific variables for all proteins in the form of box plots.
+source('best_bfactor_select_variables.r')
+
+# Generate data for the best definition of WCN:
+# Make plots of correlation of WCN with other site-specific variables for all proteins in the form of box plots.
+source('best_wcn_select_variables.r')
  

### 2015-02-19 analysis/src/voronoi_diagram.r

@@ -1,0 +1,31 @@
+# This R script uses the package Rpdb to extract the CA coordinates of a given pdb file which will be then used to generate a two dimensional Voronoi diagram.
+# Last updated by Amir Shhamoradi, 10:15 AM, Thursday Feb 19 2015, WilkeLab, ICMB, UT Austin
+
+# USES library Rpdb for reading parsing pdb file
+# USES library deldir for generating 2D Voronoi diagram
+
+getcrd = function(file,atom,crd)
+{
+  pdb <- read.pdb(file)
+  crd = pdb$atoms[pdb$atoms$elename == atom, c('x1','x2','x3')]
+  return(crd)
+}
+
+cacrd = getcrd(file='../../structures/RepairPDB_1LBA_A.pdb',atom='CA')
+
+vt = deldir(cacrd$x1,cacrd$x2)  # generate Voronoi tessellationa dn Delaunay trangulation
+filename = '../figures/voronoi_diagram.pdf'
+pdf( filename, width=4.5, height=4, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot( vt
+    , wlines='tess'
+    , pch=19:1
+    , cex=0.3
+    , lty = 1
+    , col = c(34,1)
+    , xlab = 'X coordinate [ Angstroms ]'
+    , ylab = 'Y coordinate [ Angstroms ]'
+    , showrect = TRUE
+    )
+graphics.off()
+

### 2015-02-20 manuscript/voronoi/manuscript_voronoi.tex
minor changes

@@
         \begin{center}
+        \includegraphics[width=6.9in]{/voronoi_diagram.pdf}
+        \end{center}
+        \caption{Example 2-dimensional Voronoi diagram for bacteriophage T7 lysozyme (Protein Data Bank ID `1LBA').
+The red dots represent the backbone $C_\alpha$ atoms projected on the X--Y plane, used as cell seeds in Voronoi tesellation.}
+        \label{fig:voronoi}
+    \end{figure}
+
+    \begin{figure}[tbh]
+        \begin{center}
         \includegraphics[width=6.9in]{/best_varea/select_variables/boxplot_varea_all_in_one.pdf}
@@
 
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{/best_structural_predictors_of_ER.pdf}
+        \end{center}
+        \caption{A comparison of the prediction power of four structural variables about site-specific evolutionary rates.
+Example 2-dimensional Voronoi diagram for bacteriophage T7 lysozyme (Protein Data Bank ID `1LBA').
+The red dots represent the backbone $C_\alpha$ atoms projected on the X--Y plane, used as cell seeds in Voronoi tesellation.}
+        \label{fig:best_predictor}
+    \end{figure}
+
 %    \begin{figure}[tbh]

### 2015-02-20 analysis/src/best_structural_predictors_of_ER.r

@@ -1,156 +1,101 @@
-# This code is aimed at finding which structural property best predicts  and area (varea) can best represent a residue. This is done by comparing the correlations of different vorvols with other residue variables.
+# This code is aimed at finding which structural property best predicts the Evolutionary Rates.
 
 # Last updated by Amir Shahmoradi, Friday 4:25 PM, February 13 2015, Wilke Lab, ICMB, UT Austin
 
 # setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
-all_scors_select_variables = data.frame()    # This dataframe will contain the mean median and variance of sequqence entropy and ddG entropy for each pdb file.
+# excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+pdb_temp = data.frame()
+ERscors = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
 counter = 0
 
 for(pdb in levels(res_prop_elj$pdb))
 {
-  if (!(pdb %in% excluded_pdbs))
-  {
-    counter = counter + 1
-    cat( paste(str(counter),pdb) )
-    
-    pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
-    pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
-    pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
-    pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
-    pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe_mean','rss')] )
-    
[…]

### 2015-02-20 analysis/src/best_structural_predictors_of_SE.r

@@ -1,0 +1,101 @@
+# This code is aimed at finding which structural property best predicts the Sequence Entropy.
+
+# Last updated by Amir Shahmoradi, Thursday 5:48 PM, February 19 2015, Wilke Lab, ICMB, UT Austin
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+# excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+pdb_temp = data.frame()
+SEscors = data.frame()  # This data frame will contain correlations of select structural variables with Sequence Entropy, for each pdb file on eaxch row
+counter = 0
+
+for(pdb in levels(res_prop_elj$pdb))
+{
+  counter = counter + 1
+  cat( paste(str(counter),pdb) )
+  
+  pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
+  pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
+  pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
+  pdb_voro   = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
+  
+  pdb_temp = cbind( subset(pdb_elj, select = c(pdb,seqent,ddgent))
+                  , subset(pdb_jec, select = c(zr4s_JC))
+                  , subset(pdb_hps, select = c(hpshh))
+                  , subset(pdb_dssp, select = c(rsa,hbe))
+                  , subset(pdb_wcn_bf, select = c(wcnSC,wcnCA,bfSC))
+                  , subset(pdb_voro, select = c(VSCarea))
+                  )
+  r.rsa.seqent = cor(pdb_temp$seqent,pdb_temp$rsa,method='sp')
[…]

### 2015-02-20 analysis/src/main.r

@@ -28,4 +28,10 @@ source('best_bfactor_select_variables.r')
 # Generate data for the best definition of WCN:
 # Make plots of correlation of WCN with other site-specific variables for all proteins in the form of box plots.
 source('best_wcn_select_variables.r')
+
+# Generate an example Voronoi diagram
+source('voronoi_diagram.r')
+
+# Generate density plots comparing the predictive power of Voronoi cell area, Weighted Contact Number, and RSA about sequence evolution:
+
  

### 2015-02-20 analysis/src/voronoi_diagram.r

@@ -11,7 +11,7 @@ getcrd = function(file,atom,crd)
   return(crd)
 }
 
-cacrd = getcrd(file='../../structures/RepairPDB_1LBA_A.pdb',atom='CA')
+cacrd = getcrd(file='../../structures/RepairPDB_1A2T_A.pdb',atom='CA')
 
 vt = deldir(cacrd$x1,cacrd$x2)  # generate Voronoi tessellationa dn Delaunay trangulation
 filename = '../figures/voronoi_diagram.pdf'
@@ -20,7 +20,7 @@ par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
 plot( vt
     , wlines='tess'
     , pch=19:1
-    , cex=0.3
+    , cex=0.35
     , lty = 1
     , col = c(34,1)
     , xlab = 'X coordinate [ Angstroms ]'

### 2015-02-24 manuscript/voronoi/manuscript_voronoi.tex
minor changes to mn

@@
 Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution.
-We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometrical average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
-We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
+We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
+    % We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
 By contrast, other definitions perform equally well or better.
@@
     Over the past decade, extensive attempt has been made to predict protein sequence evolution from structural properties.
-A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins (xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
+A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
 \\
 
-    Although structural characteristics have been individually extensively studied and explored, it is yet unknown whether these seemingly independent quantities are merely different manifestations of an underlying more fundamental characteristics of individual sites in proteins.
+    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
 It is perceivable that quantities such as B factor, RSA, and CN can all serve as {\it approximate} measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, in contrast to local flexibility.
-Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
+Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, as opposed to local flexibility.
+Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily a bais-free measure of the amino acid flexibility as a whole in a given site in protein.
 A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure.
-An estimate of the accessible volume to each site can be obtained through Contact Number.
-The definition of this CN however, depends on an arbitrary parameter: the radius of neighborhood definition around each site.
+An estimate of the accessible volume can be obtained through a quantity widely known as Contact Number (xx).
+For a given $C_\alpha$ backbone atom representing an amino acid in protein, the contact number is defined as the number of $C_\alpha$ atoms within a radius ($r$) of the reference $C_\alpha$ atom.
+This definition of CN however, depends on an arbitrary parameter: the radius of neighborhood around the reference site.
 An alternative more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure.

### 2015-02-25 manuscript/voronoi/manuscript_voronoi.tex
minor edit. .gitignore also corrected

@@
         \begin{center}
-        \includegraphics[width=6.9in]{/voronoi_diagram.pdf}
+        \includegraphics[width=6.9in]{voronoi_diagram.pdf}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_varea/select_variables/boxplot_varea_all_in_one.pdf}
+        \includegraphics[width=6.9in]{best_varea/select_variables/boxplot_varea_all_in_one.pdf}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_wcn/select_variables/boxplot_wcn_all_in_one.pdf}
+        \includegraphics[width=6.9in]{best_wcn/select_variables/boxplot_wcn_all_in_one.pdf}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_bf/select_variables/boxplot_bf_all_in_one.pdf}
+        \includegraphics[width=6.9in]{best_bf/select_variables/boxplot_bf_all_in_one.pdf}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=6.9in]{/best_structural_predictors_of_ER.pdf}
+        \includegraphics[width=6.9in]{best_structural_predictors_of_ER.pdf}
         \end{center}
@@
 %        \begin{center}
-%        \includegraphics[width=6.9in]{/best_wcn/select_variables/boxplot_wcn_two_in_one.pdf}
+%        \includegraphics[width=6.9in]{best_wcn/select_variables/boxplot_wcn_two_in_one.pdf}
 %        \end{center}

### 2015-02-25 manuscript/voronoi/manuscript_voronoi.tex
minor changes. finally gitignore works!
Used this http://stackoverflow.com/questions/927358/how-to-undo-the-last-commit
to fix the gitignore problem globally.
minor mn modification

@@
 
-    Over the past decade, extensive attempt has been made to predict protein sequence evolution from structural properties.
-A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
+    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
 \\
@@
     Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
-It is perceivable that quantities such as B factor, RSA, and CN can all serve as {\it approximate} measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, as opposed to local flexibility.
-Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily a bais-free measure of the amino acid flexibility as a whole in a given site in protein.
+It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
+Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the dominant factor in sequence variability patterns, as opposed to local flexibility.
+It is notable that site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
 A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure.
 An estimate of the accessible volume can be obtained through a quantity widely known as Contact Number (xx).
-For a given $C_\alpha$ backbone atom representing an amino acid in protein, the contact number is defined as the number of $C_\alpha$ atoms within a radius ($r$) of the reference $C_\alpha$ atom.
+For a given $C_\alpha$ backbone atom representing an amino acid in protein, the contact number is defined as the number of $C_\alpha$ atoms within a radius ($r$) from the reference $C_\alpha$ atom.
 This definition of CN however, depends on an arbitrary parameter: the radius of neighborhood around the reference site.
-An alternative more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure.
+An alternative definition has been proposed in recent years, the Weighted Contact Number (WCN), which has been circumvents the free-parameter $r$ in the definition of CN by more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure.
 \\

### 2015-02-25 manuscript/voronoi/manuscript_voronoi.tex
minor edits to MN

@@
 It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the dominant factor in sequence variability patterns, as opposed to local flexibility.
+Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to local flexibility.
 It is notable that site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
-A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure.
-An estimate of the accessible volume can be obtained through a quantity widely known as Contact Number (xx).
-For a given $C_\alpha$ backbone atom representing an amino acid in protein, the contact number is defined as the number of $C_\alpha$ atoms within a radius ($r$) from the reference $C_\alpha$ atom.
-This definition of CN however, depends on an arbitrary parameter: the radius of neighborhood around the reference site.
-An alternative definition has been proposed in recent years, the Weighted Contact Number (WCN), which has been circumvents the free-parameter $r$ in the definition of CN by more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure.
 \\
 
-    Motivated by this gap in the current understanding of the role of flexibility and other structural properties on sequence-structure relation in globular proteins, here we employ tessellation methods from the field of computational geometry to calculate several new site-specific quantities for globular proteins, including new measures of site-specific flexibility.
+    A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
+An estimate of the accessible volume for each site in protein is generally obtained through a quantity widely known as Contact Number (xx).
+In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it (xx).
+Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN.
+A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN.
+There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx), and sometimes up to $18\AA$ (e.g., xx).
+\\
+
+	In an attempt to provide a more general definition of CN, some studies have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
+
+        \begin{equation}
+            \label{eqn:wcn_pwrl}
+            WCN_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}},
+        \end{equation}
+
+	Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel, which has been traditionally fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl}.
+Moreover, no physical model has been so far proposed to support the power-law kernel used in the definition of WCN and the specific value of exponent often used.
+\\
+
+    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which unlike CN and WCN do not involve any adjustable free parameter in their definitions, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
+This is done by employing tessellation methods from the field of computational geometry to calculate several new site-specific quantities for proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
 Contrary to what is currently perceived about the role of flexibility in sequence variability, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
-For structural properties that are calculated based on a set of representative site coordinates, particularly Contact Number, we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
+\\
+
+    Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
 We also show that the original kernel proposed for the definition of Weighted Contact Number by (xx) and supported further by (xx) and extensively used in other studies, has no significant advantage whatsoever in predicting B factors or the sequence variability, when compared to other possible types of kernels.
@@
     \subsection*{Contact Number Definition}
-
-        In its simplest mathematical form, the Contact Number (CN) for a given site in protein is defined as the number of amino acids within a fixed radius of neighborhood around it (xx).
-Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms in the calculation of CN.
-A major problem with the traditional definition of contact number, is the existence of an arbitrary parameter -- the radius of neighborhood --  in the definition of CN.
-There is no consensus on the optimal value of this cutoff distance, but is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx), and sometimes up to $18\AA$ (e.g., xx).
-\\
-
-        In an attempt to provide a more general definition of CN, some studies have already suggested an alternative, known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
-
-        \begin{equation}
-            \label{eqn:wcn_pwrl}
-            WCN_{i,pwrl} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}},
-        \end{equation}
-
-        in which the use of subscript {\it pwrl} is to emphasize on the power-law kernel used in the definition of WCN.
-Although WCN in general correlates more strongly with $C_\alpha$ atomic B factor and sequence entropy, the proposed definition of WCN still involves an adjustable exponent, which has been traditionally fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl}.
-So far, no physical model has been proposed to support the proposed power-law definition of WCN and the specific value of exponent used.
-\\
 

### 2015-02-25 manuscript/voronoi/manuscript_voronoi.tex
MN update

@@
 
-    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which unlike CN and WCN do not involve any adjustable free parameter in their definitions, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
+    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, are not defined with any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
 This is done by employing tessellation methods from the field of computational geometry to calculate several new site-specific quantities for proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
@@
     Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
-We also show that the original kernel proposed for the definition of Weighted Contact Number by (xx) and supported further by (xx) and extensively used in other studies, has no significant advantage whatsoever in predicting B factors or the sequence variability, when compared to other possible types of kernels.
+Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility.
+
+    %We also show that the original kernel proposed for the definition of Weighted Contact Number by (xx) and supported further by (xx) and extensively used in other studies, has no significant advantage whatsoever in predicting B factors or the sequence variability, when compared to other possible types of kernels.
 A discussion of the methodology used in this work, the results and implications of our findings on the energy landscape of proteins and sequence-structure relations will be presented in the following sections.

### 2015-02-25 analysis/src/adjacent_averaging_screen_voro_log.r
adjacent averaging figures updated

@@ -1,37 +1,46 @@
-# Amir Shahmoradi, Sunday 7:42 PM, Oct 5 2014, Wilke Lab, ICMB, UT Austin
+# Amir Shahmoradi, Wednesday 3:07 PM, Feb 25 2015, Wilke Lab, ICMB, UT Austin
 # This code is very similar to the original code adjacent_averaging.r with the only difference being that here the figures are generated as multiple plots for the a single residue variable versus all others, in a single pdf figure.
 
-# install.packages('zoo')
+#install.packages('zoo')
+#install.packages('ggplot2')
+library('ggplot2')
 library('zoo')
 
-setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-source('get_res_data.r')
+# source('get_res_data.r')
 
 # The following is short version of the most useful residue properties that I have found so far.
 # Now since the three Voronoi quantities vnvortices, vnedges and vnfaces happen to be exactly the same as seen in the cormat calculated above, I am going to remove them from the data set, in addition to resvol which is not as informative.
-res_prop_concise = data.frame(zr4s_JC       = res_prop_jec$zr4s_JC,
-                              seqent        = res_prop_elj$seqent,
-                              ddgent        = res_prop_elj$ddgent,
-                              rsa           = res_prop_dssp$rsa,
-                              hbe           = res_prop_dssp$hbe_mean,
-                              hpshh         = res_prop_hps$hpshh,
-                              wcnSC         = res_prop_wcn_bf$wcnSC,
-                              bfSC          = res_prop_wcn_bf$bfSC,
-                              #resvol        = res_prop_voroSC$resvol,
-                              #vnvertices    = res_prop_voroSC$VSCnvertices,
-                              #vnedges       = res_prop_voroSC$VSCnedges,
-                              vnfaces       = res_prop_voroSC$VSCnfaces,
-                              #vedge         = log10(res_prop_voroSC$VSCedge_length_total),
-                              varea         = log10(res_prop_voroSC$VSCarea),
-                              vvolume       = log10(res_prop_voroSC$VSCvolume),
-                              veccentricity = res_prop_voroSC$VSCeccentricity,
-                              vsphericity   = res_prop_voroSC$VSCsphericity
[…]

### 2015-02-26 analysis/src/adjacent_averaging.r
new figures for voronoi cell correlations added.

@@ -13,7 +13,7 @@ res_prop_concise = data.frame(zr4s_JC       = res_prop_jec$zr4s_JC,
                               seqent        = res_prop_elj$seqent,
                               ddgent        = res_prop_elj$ddgent,
                               rsa           = res_prop_dssp$rsa,
-                              hbe           = res_prop_dssp$hbe_mean,
+                              hbe           = res_prop_dssp$hbe,
                               hpshh         = res_prop_hps$hpshh,
                               wcnSC         = res_prop_wcn_bf$wcnSC,
                               bfSC          = res_prop_wcn_bf$bfSC,
@@ -144,6 +144,22 @@ for (i in 1:length(varnames_short))
     }
   }
 }
+
+# Update: Feb 25 2015, Amir
+# Now let's make a plot of r4sJC vs. inverse of wcnSC, that is 1/wcnSC, to see if it is linear or not.
+
+res_prop_ordered = res_prop_concise[with(res_prop_concise, order(res_prop_concise$wcnSC)),]
+temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
+temp = data.frame(temp)
+filename = paste0('../figures/adjacent_averaging/zr4sJC_wcnSCinv_all_cells.pdf')
+pdf( filename, width=5.625, height=5, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(1./temp$wcnSC,
+     temp$zr4s_JC,
+     xlab = "1 / wcnSC",
+     ylab = "Evolutionary Rates (zr4sJC)",
+     type='l')
+graphics.off()
 
 #temp_quantile = rollapply(cbind(res_prop_all_ordered$wcnSC,res_prop_all_ordered$volume, res_prop_all_ordered$rsa), width = 1000, FUN = quantile)
 #temp = data.frame(temp_quantile)

### 2015-02-26 analysis/src/best_voronoi_predictors_of_ER.r

@@ -1,0 +1,198 @@
+# This code aims at finding which Voronoi cell property best predicts the Evolutionary Rates.
+
+# Last updated by Amir Shahmoradi, Thursday 9:38 AM, February 26 2015, Wilke Lab, ICMB, UT Austin
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+# excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+library(ppcor)
+pdb_temp = data.frame()
+ERscors = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
+counter = 0
+
+for(pdb in levels(res_prop_elj$pdb))
+{
+  counter = counter + 1
+  cat( paste(str(counter),pdb) )
+  
+  #pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
+  pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
+  #pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
+  #pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  #pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
+  pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
+  #pdb_voroCA = res_prop_voroCA[res_prop_voroCA$pdb==pdb, ]
+  
+  pdb_temp = cbind( subset(pdb_jec, select = c(zr4s_JC))
+                  #, subset(pdb_elj, select = c(pdb,seqent,ddgent))
+                  #, subset(pdb_hps, select = c(hpshh))
+                  #, subset(pdb_dssp, select = c(rsa,hbe))
+                  #, subset(pdb_wcn_bf, select = c(wcnSC,bfSC))
[…]

### 2015-02-26 analysis/src/main.r

@@ -33,5 +33,10 @@ source('best_wcn_select_variables.r')
 source('voronoi_diagram.r')
 
 # Generate density plots comparing the predictive power of Voronoi cell area, Weighted Contact Number, and RSA about sequence evolution:
+source('best_structural_predictors_of_ER.r')    # for Evolutionary Rates: r4sJC
+source('best_structural_predictors_of_SE.r')    # for Sequence Entropy: seqent
+
+# Generate density plots comparing the predictive power of different properties of Voronoi cell:
+source('best_voronoi_predictors_of_ER.r')
 
  

### 2015-02-26 analysis/src/best_structural_predictors_of_ER.r
minor edits to figures

@@ -6,7 +6,7 @@
 
 # excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
 pdb_temp = data.frame()
-ERscors = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
+best_structural_predictors_of_ER = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
 counter = 0
 
 for(pdb in levels(res_prop_elj$pdb))
@@ -39,16 +39,17 @@ for(pdb in levels(res_prop_elj$pdb))
   
   #row = data.frame( pdb = pdb, rsa = r.rsa.r4sJC, wcn = r.wcn.r4sJC, vareaSC = r.vareaSC.r4sJC, ddgent = r.ddgent.r4sJC )
   row = data.frame( pdb, r.wcnCA.r4sJC, r.wcnSC.r4sJC, r.rsa.r4sJC, r.vareaSC.r4sJC, r.ddgent.r4sJC , r.vareaCA.r4sJC )
-  ERscors = rbind( ERscors, row )
+  best_structural_predictors_of_ER = rbind( best_structural_predictors_of_ER, row )
 }
 
+write.csv(best_voronoi_predictors_of_ER, file = "../tables/best_structural_predictors_of_ER.csv", row.names=F )
 
 # NOW GENERATE CORRELATIONS HISTOGRAM DATA:
-hist.rsa = density(ERscors$r.rsa.r4sJC)
-hist.wcnSC = density(ERscors$r.wcnSC.r4sJC)
-hist.wcnCA = density(ERscors$r.wcnCA.r4sJC)
-hist.vareaSC = density(ERscors$r.vareaSC.r4sJC)
-hist.ddgent = density(ERscors$r.ddgent.r4sJC)
+hist.rsa = density(best_structural_predictors_of_ER$r.rsa.r4sJC)
+hist.wcnSC = density(best_structural_predictors_of_ER$r.wcnSC.r4sJC)
+hist.wcnCA = density(best_structural_predictors_of_ER$r.wcnCA.r4sJC)
+hist.vareaSC = density(best_structural_predictors_of_ER$r.vareaSC.r4sJC)
[…]

### 2015-02-26 analysis/src/best_voronoi_predictors_of_ER.r

@@ -7,7 +7,7 @@
 # excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
 library(ppcor)
 pdb_temp = data.frame()
-ERscors = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
+best_voronoi_predictors_of_ER = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
 counter = 0
 
 for(pdb in levels(res_prop_elj$pdb))
@@ -53,10 +53,10 @@ for(pdb in levels(res_prop_elj$pdb))
   #row = data.frame( pdb, r.nvertices.r4sJC, r.nedges.r4sJC, r.nfaces.r4sJC, r.edgelength.r4sJC, r.area.r4sJC , r.volume.r4sJC, r.eccentricity.r4sJC, r.sphericity.r4sJC )
   row = data.frame( pdb, edge, area , volume, eccentricity, sphericity
                   , r.edg_given_area, r.vol_given_area, r.ecc_given_area, r.sph_given_area )
-  ERscors = rbind( ERscors, row )
+  best_voronoi_predictors_of_ER = rbind( best_voronoi_predictors_of_ER, row )
 }
 
-write.csv(ERscors, file = "../tables/best_voronoi_predictors_of_ER.csv", row.names=F )
+write.csv(best_voronoi_predictors_of_ER, file = "../tables/best_voronoi_predictors_of_ER.csv", row.names=F )
 
 #GGPLOT is a waste of time!
 ### library(reshape2)
@@ -68,23 +68,23 @@ cbbPalette <- c("#009E73", "#0072B2", "#D55E00", "#CC79A7", "#000000", "#F0E442"
 # To use for line and point colors, add
 #scale_colour_manual(values=cbPalette)
 
-### ERscors_long = melt(ERscors, id.vars = "pdb")
+### best_voronoi_predictors_of_ER_long = melt(best_voronoi_predictors_of_ER, id.vars = "pdb")
 ### pdf( "../figures/best_voronoi_predictors_of_ER.pdf", width=4.5, height=4, useDingbats=FALSE )
[…]

### 2015-02-26 manuscript/voronoi/manuscript_voronoi.tex
MN updated

@@
 \vfill
-\Author{Amir Shahmoradi$^{1}$, Claus Wilke$^2$}
+\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
 \Address{\affiliation}
@@
 It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to local flexibility.
+Franzosa and Xia (xx) use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
+Recently however, Huang et al 2014 (xx) have argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to RSA and local flexibility measures.
+
     It is notable that site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
-\\
-
 A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
-An estimate of the accessible volume for each site in protein is generally obtained through a quantity widely known as Contact Number (xx).
+An estimate of the accessible volume for each site in protein can be generally obtained through a quantity widely known as Contact Number originally introduced by (xx).
 In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it (xx).
@@
 
-	In an attempt to provide a more general definition of CN, some studies have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
+	In an attempt to provide a more general definition of CN, some studies (xx) have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
 
@@
 The alignments were then used to calculate the site-specific sequence variability for each individual protein in dataset.
-To do so, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (Mayrose et al.
+2004), hereafter abbreviated as {\it r4sJC}.
+        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
 We use DSSP software (xx) for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-\\
+The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program (c.f., Echave 2014 for the details of the methodology employed xx).
+In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
+Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
 
@@
 
-        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence.
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials.
-For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points.
+        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence (Richards 1974, Gerstein 1994 xx).
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials (xx).
+For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
 \\
@@
         In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation.
-The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure (e.g., xx), energy landscape (xx) or protein--protein interactions (xx).
-\\
-
+The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions.
 Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
-We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
-
-    \subsection*{Contact Number Definition}
-
-        In order to find the best definition for Contact Number among all possibilities, we first consider the simplest definition of CN described above and calculate CN for all sites in all proteins using $C_\alpha$ atoms as reference coordinates, for a wide range of neighborhood radius $r\in[0.1\AA,50\AA]$ with strides of $\Delta r=0.1\AA$.
-We then calculate the correlation of CN with $C_\alpha$ atomic B factors and sequence entropy.
-
-        Next, we consider the traditional definition of WCN with power-law kernel.
-In order to find the optimal value of exponent in the power-law definition of WCN we first calculate the weighted contact number for all sites in all $209$ proteins of our dataset using $C_\alpha$ atoms as reference coordinates, for a wide range of power-law exponents $-30\leq\alpha\leq30$ with strides of $0.02$.
-We then calculate the correlation of WCN with the corresponding $C_\alpha$ atomic B factors in each protein, also with site-specific sequence entropy.
-This is done for the entire range of the exponent considered above.
-\\
-
-        We also consider alternative definitions of WCN with kernels other than power-law. These include WCN with a Gaussian kernel defined for the site $i$ in protein as,
+We use VORO++ software (Rycroft2009xx) to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
+Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
+In addition, for each cell we also calculate {\it sphericity}, a measure of the cell {\it compactness} defined as,
 
         \begin{equation}
-            \label{eqn:wcn_gaus}
-            WCN_{i,Gaus} = \sum^N_{j\neq i} \exp\bigg(-\frac{r_{ij}^2}{\sigma}\bigg),
+        \label{eqn:sphericity}
+        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
         \end{equation}
 
-        with $\sigma\in[0.2\AA,50\AA]$ with strides of $0.2$ acting as the free parameter of the model, and WCN with an exponential kernel defined for the site $i$ in protein as,
-
-        \begin{equation}
-            \label{eqn:wcn_exp}
-            WCN_{i,exp} = \sum^N_{j\neq i} \exp\bigg(-\frac{r_{ij}}{\lambda}\bigg),
-        \end{equation}
-
-        with $\lambda\in[0.2\AA,50\AA]$ with strides of $0.2$ acting as the free parameter of the model.
+        in which $V$ \& $A$ stand for the volume \& area of the cell respectively.
+For a perfectly spherical cell, $\Psi=1$, while it becomes zero for a 2-dimensional object that has no volume but only surface area.
 
@@
 
-        Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific variables.
+        Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific structural variables.
 For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
 The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein (xx).
-More recently, (Echave 2015) have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
-Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility (e.g., Halle 2001 xx), same definition degeneracy also exists on choice of atomic B factors (xx) that are used to represent site-specific flexibility.
-In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein (soyer2000xx).
+More recently, Marcos2015 have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
+
+        Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility (e.g., Halle 2002), same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
+In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra.
 \\
 
-        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates used.
+        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates.
 These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
@@
 
+    \subsection*{Voronoi Cell Area and Volume as a Measure of Local Packing Density in Proteins}
+
+    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra.
+Figure \ref{fig:best_voronoi} (left plot) depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
+It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
+In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
+The cell properties are also highly associated with each other.
+Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in the left plot of Figure \ref{fig:best_voronoi}.
+On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
+In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell.
+Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates.
+These cell characteristics are however, discrete (integer) quantities and in general have a limited range.
+
+    Not shown here for brevity, we also obtain identical results to the above if we use sequence entropy as defined by Eqn.
+\ref{eqn:shannon} instead of sequence evolutionary rates, although sequence entropy generally results in weaker correlation strengths due to the discreteness and limited range inherent in its definition.
+
     \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}
 
-        As explained in previous section, In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, B factor and Voronoi cell properties.
-This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.
+        Depending on the set of atomic coordinates that represent the protein structure, there are $7$ different measures for site-specific structural properties such as Weighted Contact Number, B factor and Voronoi cell properties.
 Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
@@
 
-        For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
+        For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
+Specifically, bfSC outperforms the commonly used $C_\alpha$ B factor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
         \\
 
-        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
+        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
         \\
+
+\section{Discussion and Concluding Remarks}
+\label{sec:dcr}
+
+		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced, using Voronoi tessellation from the field of computational geometry, that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al.
+(2014xx), site-specific measures of flexibility, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
+It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.
 
@@
 
-    \subsection*{Sequence divergence as the main Determinant of Sequence-Structure Relation}
-
-        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
-The choice of Spearman versus the popular Pearson's correlation measure is made in order to minimize the effects of any nonlinear variable relationships on the strengths of the correlations.
-The resulting correlation matrix reveals a myriad of pdb-level properties each having a small but nonzero contribution to the strength of the structure-sequence correlations.
-        \\
-
-        A hierarchical clustering of the correlation matrix however, reveals two main independent factors that have the strongest influence on the strengths of sequence-structure correlations: 1.
-The sequence divergence as measured by the standard deviation of sequence entropy and evolutionary rates (denoted by {\it sd.seqent} \& {\it sd.r4sJC}) among all sites in each protein structure.
-2.
-The homogeneity of the hydrogen bond strengths among the back bone atoms of each protein structure, as measured by the standard deviation of hydrogen bond energies (denoted by {\it sd.hbe}) among all pdb sites.
-A reduced-size of the Spearman correlation matrix for the most influential factors on the two strongest sequence--structure (seqent$/$r4s -- wcnSC$/$varea) relations is illustrated in Figure \ref{fig:mainmods}.
-        \\
-
-        For the other weaker sequence--structure relations, i.e.
-the correlations of seqent$/$r4s with RSA, \ddg entropy (ddgent), and Bfactor (bfSC) we find other pdb-level properties that also contribute to the correlation strengths, comparable to or even stronger than in sequence divergence and hydrogen bond homogeneity.
-In general, we observe that for the weaker the sequence-structure correlations, factors that determine the accuracy of the measured residue properties become more influential on the strength of the correlations.
-In particular, the X-ray crystallographic resolution of the structure and the definition of the \ddg entropy play dominant roles, with Spearman correlation coefficients of $\rho~0.3$, on the strengths of the corresponding sequence-structure relations.
-        \\
-
-        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.
-Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
-Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
-In addition to regularized regression, we have also employed Principal Component Regression (PCR) on the original dataset and its rank transformation.
-Both regression methods, PCR \& regularized, point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
-
-\section{Discussion and Concluding Remarks}
-\label{sec:dcr}
+		One potential caveat with Voronoi tessellation is the edge effects, that is, Voronoi cells that are on the surface of the protein are open. To ensure that
 
@@
 In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
-Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,
+Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and
 
-        \begin{equation}
-        \label{eqn:sphericity}
-        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
-        \end{equation}
-
-        in which $V$ \& $A$ represent {\it vvolume} \& {\it varea} respectively.
          We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
@@
 
-....
+We thank Austin G. Meyer, Stephanie Spielman and Eleisha Jackson at UT Austin for helpful discussions.
 
@@
     \end{figure}
+
+    \begin{figure*}[tbh]
+        \begin{center}
+        \begin{tabular}{cc}  
+        \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER.pdf} & \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER_given_area.pdf}
+        \end{tabular}
+        \end{center}
+        \caption{{\bf Right:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER).
+Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER.
+{\bf Left:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
+        \label{fig:voronoi_ER}
+    \end{figure*}
 
@@
         \end{center}
-        \caption{A comparison of the prediction power of four structural variables about site-specific evolutionary rates.
-Example 2-dimensional Voronoi diagram for bacteriophage T7 lysozyme (Protein Data Bank ID `1LBA').
-The red dots represent the backbone $C_\alpha$ atoms projected on the X--Y plane, used as cell seeds in Voronoi tesellation.}
+        \caption{A comparison of the prediction power of four structural variables about site-specific evolutionary rates (ER).
+All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively.
+For better illustration however, the absolute Spearman's correlation of WCN with ER are shown in the Figure.}
         \label{fig:best_predictor}

### 2015-02-27 analysis/src/adjacent_averaging_screen_voro_nonlog.r
new analysis added.

@@ -14,7 +14,7 @@ res_prop_concise = data.frame(zr4s_JC       = res_prop_jec$zr4s_JC,
                               seqent        = res_prop_elj$seqent,
                               ddgent        = res_prop_elj$ddgent,
                               rsa           = res_prop_dssp$rsa,
-                              hbe           = res_prop_dssp$hbe_mean,
+                              hbe           = res_prop_dssp$hbe,
                               hpshh         = res_prop_hps$hpshh,
                               wcnSC         = res_prop_wcn_bf$wcnSC,
                               bfSC          = res_prop_wcn_bf$bfSC,

### 2015-02-27 analysis/src/best_hps.r

@@ -62,15 +62,15 @@ for(pdb in levels(res_prop_elj$pdb))
   
   pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
   pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,c('hpskd','hpsww','hpshh')]
-  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
-  pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe_mean','rss')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
+  pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe','rss')] )
   pdb_voroAA = res_prop_voroAA[res_prop_voroAA$pdb==pdb, ]
   pdb_voroCA = res_prop_voroCA[res_prop_voroCA$pdb==pdb, ]
   pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
   
   pdb_temp = cbind( subset(pdb_elj, select = c(seqent,ddgent)),
                     #subset(pdb_hps, select = c(hpskd,hpsww,hpshh)),
-                    subset(pdb_dssp, select = c(asa,rsa,hbe_mean)),
+                    subset(pdb_dssp, select = c(asa,rsa,hbe)),
                     subset(pdb_wcn_bf, select = -c(pdb,resnam,resnum)),
                     subset(pdb_voroAA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,VAAnvertices,VAAnedges,VAAvolume_change)),
                     subset(pdb_voroCA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,resvol,VCAnvertices,VCAnedges,VCAvolume_change)),

### 2015-02-27 analysis/src/convert_res_prop_2_rank_data.r

@@ -1,0 +1,89 @@
+# This code generates a ranked version of the dataframe containing all residue PDB data.
+# Amir Shahmoradi, Thursday 6:26 PM, Feb 26 2014, Wilke Lab, ICMB, UT Austin 
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+pdb_temp = data.frame()
+best_structural_predictors_of_ER = data.frame()  # This data frame will contain correlations of select structural variables with r4sJC evolutionary rates, for each pdb file on eaxch row
+counter = 0
+
+for(pdb in levels(res_prop_elj$pdb))
+{
+  counter = counter + 1
+  cat( paste(str(counter),pdb) )
+  
+  pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
+  pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
+  pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
+  pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
+  pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
+  pdb_voroCA = res_prop_voroCA[res_prop_voroCA$pdb==pdb, ]
+  
+  pdb_temp = cbind( subset(pdb_elj, select = c(pdb,seqent,ddgent))
+                    , subset(pdb_jec, select = c(zr4s_JC))
+                    , subset(pdb_hps, select = c(hpshh))
+                    , subset(pdb_dssp, select = c(rsa,hbe))
+                    , subset(pdb_wcn_bf, select = c(wcnSC,wcnCA,bfSC))
+                    , subset(pdb_voroSC, select = c(VSCarea))
+                    , subset(pdb_voroCA, select = c(VCAarea))
+  )
+  r.rsa.r4sJC = cor(pdb_temp$zr4s_JC,pdb_temp$rsa,method='sp')
[…]

### 2015-02-27 manuscript/voronoi/manuscript_voronoi.tex
MN update

@@
     In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra.
-Figure \ref{fig:best_voronoi} (left plot) depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
+Plot {\it A} of Figure \ref{fig:best_voronoi} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
 It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
@@
 The cell properties are also highly associated with each other.
-Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in the left plot of Figure \ref{fig:best_voronoi}.
+Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}.
 On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
@@
 
-		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced, using Voronoi tessellation from the field of computational geometry, that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al.
-(2014xx), site-specific measures of flexibility, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
+		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al.
+(2014xx), site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
 It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.
@@
 
-    \begin{figure*}[tbh]
+    \begin{figure}[tbh]
         \begin{center}
-        \begin{tabular}{cc}  
-        \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER.pdf} & \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER_given_area.pdf}
-        \end{tabular}
+        \includegraphics[width=7in]{best_voronoi_predictors_of_ER_screen.pdf}
         \end{center}
-        \caption{{\bf Right:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER).
+        \caption{{\bf A:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER).
 Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER.
-{\bf Left:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
+{\bf B:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
         \label{fig:voronoi_ER}
-    \end{figure*}
+    \end{figure}
 

### 2015-02-27 analysis/src/best_structural_predictors_of_ER.r

@@ -17,7 +17,7 @@ for(pdb in levels(res_prop_elj$pdb))
   pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
   pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
   pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
-  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
   pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
   pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
   pdb_voroCA = res_prop_voroCA[res_prop_voroCA$pdb==pdb, ]
@@ -36,9 +36,11 @@ for(pdb in levels(res_prop_elj$pdb))
   r.vareaSC.r4sJC = cor(pdb_temp$zr4s_JC,pdb_temp$VSCarea,method='sp')
   r.ddgent.r4sJC = cor(pdb_temp$zr4s_JC,pdb_temp$ddgent,method='sp')
   r.vareaCA.r4sJC = cor(pdb_temp$zr4s_JC,pdb_temp$VCAarea,method='sp')
+  r.bfSC.r4sJC = cor(pdb_temp$zr4s_JC,pdb_temp$bfSC,method='sp')
+  r.hbe.r4sJC = cor(pdb_temp$zr4s_JC,abs(pdb_temp$hbe),method='sp')
   
   #row = data.frame( pdb = pdb, rsa = r.rsa.r4sJC, wcn = r.wcn.r4sJC, vareaSC = r.vareaSC.r4sJC, ddgent = r.ddgent.r4sJC )
-  row = data.frame( pdb, r.wcnCA.r4sJC, r.wcnSC.r4sJC, r.rsa.r4sJC, r.vareaSC.r4sJC, r.ddgent.r4sJC , r.vareaCA.r4sJC )
[…]

### 2015-02-27 analysis/src/best_voronoi_predictors_of_ER.r

@@ -18,7 +18,7 @@ for(pdb in levels(res_prop_elj$pdb))
   #pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
   pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
   #pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
-  #pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  #pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
   #pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
   pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
   #pdb_voroCA = res_prop_voroCA[res_prop_voroCA$pdb==pdb, ]
@@ -196,3 +196,134 @@ legend( 'topleft'
       )
 
 graphics.off()
+
+
+
+
+#####################################################
+#####################################################
+#####################################################
+#####################################################
+#####################################################
+
+
+
+# Now combine both histograms in one screen figure:
+
+# Now plot histograms in a single plot
+#colors = c('green', 'blue', 'red', 'black', 'gray', 'cyan2')
+pdf( "../figures/best_voronoi_predictors_of_ER_screen.pdf", width=9, height=4, useDingbats=FALSE )
+
+split.screen(c(1,2))
+
+  screen(1)
+  par( mai=c(0.65, 0.65, 0.2, 0.2), mgp=c(2, 0.5, 0), tck=-0.03 )
+  plot( hist.edge$x
+    , hist.edge$y
+    #, col = 'blue'
+    ,  col = cbbPalette[[1]]
+    ,   xlim = c(0.1,0.85)
+    ,   ylim = c(0,5.5)
+    #,   col=colors[1]
+    #,   ylim=c(0,7)
+    #,   border = colors[1]
+    #,   lty = 0
+    ,   type = 'l'
+    ,   lwd  = 2 
+    #,   main = 'Correlations with Evolutionary Rates'
+    #,   xlab = expression(paste('Absolute Spearman Cor. with Evolutionary Rates ',rho))
[…]

### 2015-02-27 analysis/src/cordiv_pearson.r

@@ -37,7 +37,7 @@ for(pdb in levels(res_prop_elj$pdb))
   r.seqent_wcnca = x$estimate
   p.seqent_wcnca = x$p.value
   
-  x = cor.test( pdb_elj$entropy_from_alignments, pdb_dssp$hbe_mean, method='pearson', na.action="na.omit" )
+  x = cor.test( pdb_elj$entropy_from_alignments, pdb_dssp$hbe, method='pearson', na.action="na.omit" )
   r.seqent_hbe = x$estimate
   p.seqent_hbe = x$p.value
   
@@ -53,7 +53,7 @@ for(pdb in levels(res_prop_elj$pdb))
   r.ddgent_wcnca = x$estimate
   p.ddgent_wcnca = x$p.value
   
-  x = cor.test( pdb_elj$entropy_from_ddGs, pdb_dssp$hbe_mean, method='pearson', na.action="na.omit" )
+  x = cor.test( pdb_elj$entropy_from_ddGs, pdb_dssp$hbe, method='pearson', na.action="na.omit" )
   r.ddgent_hbe = x$estimate
   p.ddgent_hbe = x$p.value
   
@@ -65,7 +65,7 @@ for(pdb in levels(res_prop_elj$pdb))
   r.asa_wcnca = x$estimate
   p.asa_wcnca = x$p.value
   
-  x = cor.test( pdb_dssp$asa, pdb_dssp$hbe_mean, method='pearson', na.action="na.omit" )
+  x = cor.test( pdb_dssp$asa, pdb_dssp$hbe, method='pearson', na.action="na.omit" )
   r.asa_hbe = x$estimate
   p.asa_hbe = x$p.value
   
@@ -73,11 +73,11 @@ for(pdb in levels(res_prop_elj$pdb))
   r.rsa_wcnca = x$estimate
   p.rsa_wcnca = x$p.value
   
-  x = cor.test( pdb_dssp$rsa, pdb_dssp$hbe_mean, method='pearson', na.action="na.omit" )
+  x = cor.test( pdb_dssp$rsa, pdb_dssp$hbe, method='pearson', na.action="na.omit" )
   r.rsa_hbe = x$estimate
   p.rsa_hbe = x$p.value
   
-  x = cor.test( pdb_dssp$wcn_ca, pdb_dssp$hbe_mean, method='pearson', na.action="na.omit" )
[…]

### 2015-02-27 analysis/src/cordiv_spearman.r

@@ -49,7 +49,7 @@ for(pdb in levels(res_prop_elj$pdb))
   #r.seqent_wcnca = x$estimate
   #p.seqent_wcnca = x$p.value
   
-  x = cor.test( pdb_elj$entropy_from_alignments, pdb_dssp$hbe_mean, method='spearman', na.action="na.omit" )
+  x = cor.test( pdb_elj$entropy_from_alignments, pdb_dssp$hbe, method='spearman', na.action="na.omit" )
   r.seqent_hbe = x$estimate
   p.seqent_hbe = x$p.value
   
@@ -65,7 +65,7 @@ for(pdb in levels(res_prop_elj$pdb))
   #r.ddgent_wcnca = x$estimate
   #p.ddgent_wcnca = x$p.value
   
-  x = cor.test( pdb_elj$entropy_from_ddGs, pdb_dssp$hbe_mean, method='spearman', na.action="na.omit" )
+  x = cor.test( pdb_elj$entropy_from_ddGs, pdb_dssp$hbe, method='spearman', na.action="na.omit" )
   r.ddgent_hbe = x$estimate
   p.ddgent_hbe = x$p.value
   
@@ -77,7 +77,7 @@ for(pdb in levels(res_prop_elj$pdb))
   #r.asa_wcnca = x$estimate
   #p.asa_wcnca = x$p.value
   
-  x = cor.test( pdb_dssp$asa, pdb_dssp$hbe_mean, method='spearman', na.action="na.omit" )
+  x = cor.test( pdb_dssp$asa, pdb_dssp$hbe, method='spearman', na.action="na.omit" )
   r.asa_hbe = x$estimate
   p.asa_hbe = x$p.value
   
@@ -85,11 +85,11 @@ for(pdb in levels(res_prop_elj$pdb))
   #r.rsa_wcnca = x$estimate
   #p.rsa_wcnca = x$p.value
   
-  x = cor.test( pdb_dssp$rsa, pdb_dssp$hbe_mean, method='spearman', na.action="na.omit" )
+  x = cor.test( pdb_dssp$rsa, pdb_dssp$hbe, method='spearman', na.action="na.omit" )
   r.rsa_hbe = x$estimate
   p.rsa_hbe = x$p.value
   
-  #x = cor.test( pdb_dssp$wcn_ca, pdb_dssp$hbe_mean, method='spearman', na.action="na.omit" )
[…]

### 2015-02-27 analysis/src/res_prop_correlations_selected_variables.r

@@ -64,13 +64,13 @@ for(pdb in levels(res_prop_elj$pdb))
   
   pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
   pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
-  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
   pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ]
   pdb_voroSC = res_prop_voroSC[res_prop_voroSC$pdb==pdb, ]
   
   pdb_temp = cbind( subset(pdb_elj, select = c(seqent,ddgent)),
                     subset(pdb_hps, select = c(hpshh)),
-                    subset(pdb_dssp, select = c(rsa,hbe_mean)),
+                    subset(pdb_dssp, select = c(rsa,hbe)),
                     subset(pdb_wcn_bf, select = c(wcnSC,bfSC)),
                     #subset(pdb_voroAA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,VAAnvertices,VAAnedges,VAAvolume_change)),
                     #subset(pdb_voroCA, select = -c(pdb,resnam,resnum,sizeSC,sizeAA,resvol,VCAnvertices,VCAnedges,VCAvolume_change)),

### 2015-02-27 analysis/src/summarize_res_data.r

@@ -20,13 +20,13 @@ for(pdb in levels(res_prop_elj$pdb))
     pdb_elj    = res_prop_elj[res_prop_elj$pdb==pdb,] # c('seqent','ddgent')]
     pdb_jec    = res_prop_jec[res_prop_jec$pdb==pdb,] # c('zr4s_JC')]
     pdb_hps    = res_prop_hps[res_prop_hps$pdb==pdb,] # c('hpskd','hpsww','hpshh')] )
-    pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
-    pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe_mean','rss')] )
+    pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
+    pdb_wcn_bf = res_prop_wcn_bf[res_prop_wcn_bf$pdb==pdb, ] #c('asa','rsa','hbe','rss')] )
     
     pdb_temp = cbind( subset(pdb_elj, select = c(seqent,ddgent)),
                       subset(pdb_jec, select = c(zr4s_JC)),
                       subset(pdb_hps, select = c(hpskd,hpsww,hpshh)),
-                      subset(pdb_dssp, select = c(asa,rsa,hbe_mean)),
+                      subset(pdb_dssp, select = c(asa,rsa,hbe)),
                       subset(pdb_wcn_bf, select = -c(pdb,resnam,resnum))
     )
     
@@ -77,7 +77,7 @@ for(pdb in levels(res_prop_dssp$pdb))
   counter = counter + 1
   cat( paste(str(counter),pdb) )
   
-  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe_mean','rss')] )
+  pdb_dssp   = res_prop_dssp[res_prop_dssp$pdb==pdb,] # c('asa','rsa','hbe','rss')] )
 
   sum.GSS    = length(which(pdb_dssp$rss == 'G'))
   sum.HSS    = length(which(pdb_dssp$rss == 'H'))

### 2015-02-28 analysis/src/adjacent_averaging.r
MN / Figure update

@@ -151,7 +151,7 @@ for (i in 1:length(varnames_short))
 res_prop_ordered = res_prop_concise[with(res_prop_concise, order(res_prop_concise$wcnSC)),]
 temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
 temp = data.frame(temp)
-filename = paste0('../figures/adjacent_averaging/zr4sJC_wcnSCinv_all_cells.pdf')
+filename = paste0('../figures/adjacent_averaging/wcnSCinv_zr4sJC_all_cells.pdf')
 pdf( filename, width=5.625, height=5, useDingbats=FALSE )
 par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
 plot(1./temp$wcnSC,
@@ -160,6 +160,50 @@ plot(1./temp$wcnSC,
      ylab = "Evolutionary Rates (zr4sJC)",
      type='l')
 graphics.off()
+
+# Update: Feb 26 2015, Amir
+# Now let's make a plot of r4sJC vs. inverse of Voronoi Volume (volinv), that is 1/volume, to see if it is linear or not.
+
+res_prop_ordered = res_prop_concise[with(res_prop_concise, order(res_prop_concise$vvolume)),]
+temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
+temp = data.frame(temp)
+filename = paste0('../figures/adjacent_averaging/vvolumeSCinv_zr4sJC_all_cells.pdf')
+pdf( filename, width=5.625, height=5, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(1./temp$vvolume,
+     temp$zr4s_JC,
+     xlab = "1 / Voronoi Volume",
+     ylab = "Evolutionary Rates (zr4sJC)",
+     type='l')
+graphics.off()
+
+res_prop_concise_closed = res_prop_concise[res_prop_voroSC$VSCvolume_change_diff == 0,]
+res_prop_closed_ordered = res_prop_concise_closed[with(res_prop_concise_closed, order(res_prop_concise_closed$vvolume)),]
+temp_closed = data.frame(rollapply(res_prop_closed_ordered, width = 3000, FUN = mean))
[…]

### 2015-02-28 analysis/src/best_structural_predictors_of_ER.r

@@ -44,7 +44,7 @@ for(pdb in levels(res_prop_elj$pdb))
   best_structural_predictors_of_ER = rbind( best_structural_predictors_of_ER, row )
 }
 
-write.csv(best_voronoi_predictors_of_ER, file = "../tables/best_structural_predictors_of_ER.csv", row.names=F )
+write.csv(best_structural_predictors_of_ER, file = "../tables/best_structural_predictors_of_ER.csv", row.names=F )
 
 # NOW GENERATE CORRELATIONS HISTOGRAM DATA:
 hist.rsa = density(best_structural_predictors_of_ER$r.rsa.r4sJC)
@@ -62,7 +62,7 @@ par( mai=c(0.65, 0.65, 0.05, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
 plot(  hist.rsa$x
     ,  hist.rsa$y
     ,   col = 'blue'
-    ,   xlim = c(-0.5,0.85)
+    ,   xlim = c(0.0,0.85)
     ,   ylim = c(0,5.5)
     #,   col=colors[1]
     #,   ylim=c(0,7)
@@ -101,15 +101,16 @@ lines( hist.bfSC$x
      , col = 'cyan2'
      , lwd = 2
      )
-lines( -hist.hbe$x
-     , hist.hbe$y
-     , col = 'grey'
-     , lwd = 2
-     )
+#lines( -hist.hbe$x
+#     , hist.hbe$y
+#     , col = 'grey'
+#     , lwd = 2
+#     )
 
 legend( 'topleft'
-      , c("Voronoi Cell Area","H-bond strength", "WCN (SC)", "WCN (CA)", "ddG Rate", "Bfactor", "RSA")
-      , col = c('red','grey','black','black','green','cyan2','blue')
+      #, c("Voronoi Cell Area","H-bond strength", "WCN (SC)", "WCN (CA)", "ddG Rate", "Bfactor", "RSA")
+      , c("Voronoi Cell Area","WCN (SC)", "WCN (CA)", "ddG Rate", "Bfactor", "RSA")
+      , col = c('red','black','black','green','cyan2','blue')
       , lty = c(1,1,2,1,1)
       , lwd = 2
       , bty = 'n'

### 2015-03-16 manuscript/voronoi/manuscript_voronoi.tex
corrections to MN references added

@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) (Goldman1998, Bustamante2000, Conant2009, Franzosa2009, Ramsey2011, Scherrer2012, Yeh2014a, Meyer2013a, Meyer2013b, Yeh2014b, Shahmoradi2014, Sikosek2014), Contact Number (Rodionov1998, Hamelryck2005, Liao2005, Bloom2006, Huang2014, Marcos2014, Yeh2014b, Shahmoradi2014 ), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{} Wilke2005, Echave2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (Liao2005, Shih2012, Shahmoradi2014) or flexibility measures based elastic network models (Liu2012) and Molecular Dynamics (MD) simulations (shahmoradi2014).
 \\
@@
 It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Franzosa and Xia (xx) use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
+\citep[][]{}Franzosa and Xia (xx) use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
 Recently however, Huang et al 2014 (xx) have argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to RSA and local flexibility measures.
@@
 A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN.
-There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx), and sometimes up to $18\AA$ (e.g., xx).
+There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx).
 \\
 
-	In an attempt to provide a more general definition of CN, some studies (xx) have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
+	In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{Lin2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
 
@@
 2004), hereafter abbreviated as {\it r4sJC}.
-        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{Shenkin1991}, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
         where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
-We use DSSP software (xx) for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program (c.f., Echave 2014 for the details of the methodology employed xx).
+We use DSSP software \citep[][]{Kabsch1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{Tien2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{Echave2014}.
 In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
@@
 
-        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence (Richards 1974, Gerstein 1994 xx).
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials (xx).
+        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \citep[e.g., ][]{Richards 1974, Gerstein 1994}.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g., ][]{Zomorodian2006, Zhou2014, Zia2014}.
 For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
@@
 Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
-We use VORO++ software (Rycroft2009xx) to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
+We use VORO++ software \citep[][]{Rycroft2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
 Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
@@
 The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein (soyer2000xx).
-More recently, Marcos2015 have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \citep[e.g., ][]{soyer2000}.
+More recently, \citet{Marcos2015} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
 
@@
 
-		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al.
-(2014xx), site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
+		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by \citet{Echave2014}, site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
 It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.

### 2015-03-16 analysis/src/adjacent_averaging.r
minor changes to graphs and scripts

@@ -205,6 +205,22 @@ graphics.off()
 
 
 
+
+res_prop_ordered = res_prop_concise[with(res_prop_concise, order(res_prop_concise$vvolume)),]
+temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
+temp = data.frame(temp)
+filename = paste0('../figures/adjacent_averaging/vvolumeSC_zr4sJC_all_cells_nonlog.pdf')
+pdf( filename, width=5.625, height=5, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.1, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(10^temp$vvolume,
+     temp$zr4s_JC,
+     xlab = "Voronoi Cell Volume",
+     ylab = "Evolutionary Rates (zr4sJC)",
+     type='l')
+graphics.off()
+
+
+
 #temp_quantile = rollapply(cbind(res_prop_all_ordered$wcnSC,res_prop_all_ordered$volume, res_prop_all_ordered$rsa), width = 1000, FUN = quantile)
 #temp = data.frame(temp_quantile)
 #View(temp)

### 2015-03-16 manuscript/voronoi/mnv.tex
cript citations are now corrected

--- 2015-02-27 manuscript/voronoi/manuscript_voronoi.tex
@@
 \usepackage{float}
-\citestyle{apj}
+\citestyle{mbe}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014).
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
 \\
@@
 It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-Franzosa and Xia (xx) use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
-Recently however, Huang et al 2014 (xx) have argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to RSA and local flexibility measures.
+\citet{franzosa_structural_2009} use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
+Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to RSA and local flexibility measures.
 
@@
 A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
-An estimate of the accessible volume for each site in protein can be generally obtained through a quantity widely known as Contact Number originally introduced by (xx).
-In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it (xx).
+An estimate of the accessible volume for each site in protein can be generally obtained through a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
+In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}.
 Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN.
 A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN.
-There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ (e.g., lin, franzosa, xx), and sometimes up to $18\AA$ (e.g., xx).
+There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009}.
 \\
 
-	In an attempt to provide a more general definition of CN, some studies (xx) have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
+	In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{lin_deriving_2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
 
@@
 
-        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes taken from xx Echave papers, Wilke ddg paper xx randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
-1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
-2009; Ashkenazy et al.
-2010).
-Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al.
-2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
+        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes taken from xx Echave papers, Wilke ddg paper xx randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citet{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}.
+Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep[][]{katoh_mafft_2005}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
 The alignments were then used to calculate the site-specific sequence variability for each individual protein in dataset.
-For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (Mayrose et al.
-2004), hereafter abbreviated as {\it r4sJC}.
-        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model \citep[][]{mayrose_comparison_2004}, hereafter abbreviated as {\it r4sJC}.
+        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991}, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
         where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
-We use DSSP software (xx) for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program (c.f., Echave 2014 for the details of the methodology employed xx).
+We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
 In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
@@
 
-        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence (Richards 1974, Gerstein 1994 xx).
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials (xx).
+        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \citep[e.g., ][]{richards_interpretation_1974, gerstein_volume_1994}.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g., ][]{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
 For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
@@
 Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
-We use VORO++ software (Rycroft2009xx) to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
+We use VORO++ software \citep[][]{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
 Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
@@
 The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein (soyer2000xx).
-More recently, Marcos2015 have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \citep[e.g., ][]{soyer_voronoi_2000}.
+More recently, \citet{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
 
-        Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility (e.g., Halle 2002), same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
+        Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \citep[e.g.,][]{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
 In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra.
@@
 
-		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al.
-(2014xx), site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
+		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by \citet{echave_relationship_2014}, site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
 It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.
@@
 
-\bibliographystyle{apj}
-\bibliography{proc}
+\bibliographystyle{mbe}
+\bibliography{mnv}
 

### 2015-03-23 manuscript/voronoi/mnv.tex
minor corrections to MN

@@
 \citet{franzosa_structural_2009} use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
-Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical model, for the local packing density as the dominant factor in sequence variability patterns, as opposed to RSA and local flexibility measures.
+Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in sequence variability patterns in contrast to RSA and local flexibility measures.
 
-    It is notable that site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
+    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
 A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
@@
 
-	Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel, which has been traditionally fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl}.
+	Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel, which is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
 Moreover, no physical model has been so far proposed to support the power-law kernel used in the definition of WCN and the specific value of exponent often used.
@@
 
-    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, are not defined with any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
-This is done by employing tessellation methods from the field of computational geometry to calculate several new site-specific quantities for proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
-Contrary to what is currently perceived about the role of flexibility in sequence variability, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
+    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, their definitions does not involve any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
+This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
+Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
 \\
@@
 
-        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes taken from xx Echave papers, Wilke ddg paper xx randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}.
+        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes \citep[e.g.,][]{yeh_site-specific_2014, echave_relationship_2014} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$ amino acids, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}.
 To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citet{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}.
@@
 For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model \citep[][]{mayrose_comparison_2004}, hereafter abbreviated as {\it r4sJC}.
-        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991}, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
 We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-The {\it ddG rate} estimates are taken for all structures in the dataset were all calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
+The {\it ddG rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
 In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
@@
         There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \citep[e.g., ][]{richards_interpretation_1974, gerstein_volume_1994}.
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g., ][]{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g.,][]{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
 For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
@@
 
-        Depending on the choice of coordinates used, there exist degeneracies in the definition of the some site-specific structural variables.
+        Depending on the choice of coordinates used, there exist degeneracy in the definition of some site-specific structural variables.
 For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.

### 2015-03-23 manuscript/voronoi/mnv.tex
minor corrections to MN

@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
 \\

### 2015-03-27 manuscript/voronoi/mnv.tex
MN modified significantly, new figures in png format added

@@
 \documentclass[11pt]{article}
+\usepackage{amssymb}
 \usepackage{hyperref}
@@
 This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
-Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
+Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number (CN) and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
 \\
@@
 For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
-The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
+The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins.
 Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \citep[e.g., ][]{soyer_voronoi_2000}.
@@
 
-        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates.
+         Here we consider all possible choices of the representative set of atomic coordinates in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
 These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
@@
 
-    \subsection*{Voronoi Cell Area and Volume as a Measure of Local Packing Density in Proteins}
+    \subsection*{Voronoi Cell Area and Volume as Proxy Measures of Local Packing Density and Flexibility in Proteins}
 
-    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra.
+    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in dataset and use them as the seeds of Voronoi polyhedra.
 Plot {\it A} of Figure \ref{fig:best_voronoi} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
-It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
+It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
 In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
 The cell properties are also highly associated with each other.
-Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}.
+Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}.
 On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
@@
 
-    Not shown here for brevity, we also obtain identical results to the above if we use sequence entropy as defined by Eqn.
-\ref{eqn:shannon} instead of sequence evolutionary rates, although sequence entropy generally results in weaker correlation strengths due to the discreteness and limited range inherent in its definition.
+    Not shown here for brevity, we also obtain identical results to above if we use sequence entropy as defined by Eqn.
+\ref{eqn:shannon} instead of sequence evolutionary rates.
+The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in its definition.
 
@@
 
-        Depending on the set of atomic coordinates that represent the protein structure, there are $7$ different measures for site-specific structural properties such as Weighted Contact Number, B factor and Voronoi cell properties.
+        Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual  site-specific structural properties such as Weighted Contact Number, B factor and Voronoi cell properties.
 Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
@@
         For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
+Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
         \\
 
-        For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ B factor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
+        For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain results in the best correlations with other structural and sequence properties.
+Specifically, the average side chain B factor outperforms the commonly used $C_\alpha$ B factor in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
         \\
@@
 
-\section{Discussion and Concluding Remarks}
+\section{Discussion}
 \label{sec:dcr}
 
-		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by \citet{echave_relationship_2014}, site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
-It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.
-
-        All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites.
-In particular, the strength of structure and sequence correlations decrease when moving from side chain to backbone atoms.
-An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
+        Throughout this work, we have carried out a comprehensive analysis and comparison of the main structural determinants of sequence variability, some of which are newly reported and discussed in this work.
+Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work, and sequence entropy with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg Rate as defined by \citet{echave_relationship_2014} which is essentially a proxy measure of stability of protein's native conformation upon substitution of amino acids in individual sites in proteins.
+In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg Rate, and the traditional definitions of Contact Number (CN) and the Weighted Contact Number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figure \ref{fig:best_predictor}).
 \\
 
-        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
-The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
+        One potential caveat with Voronoi tessellation of finite structures in Euclidean space is the {\it edge effects}.
+Sites that are close to the surface of protein are often associated with Voronoi cells that are bounded by the cubic box containing the protein.
+To ensure that these edge effects do not influence the observed sequence-structure correlations, we identified open cells -- i.e., cells that are partially bounded and closed by the cubic box containing the protein -- in all proteins by examining the variations in individual cell volumes upon changing the size of the cubic box containing the protein.
+The open cells in individual proteins were then ranked by the fraction of volume changes observed upon changing the box size and then normalized to the the largest volume observed among closed cells.
 \\
 
-		One potential caveat with Voronoi tessellation is the edge effects, that is, Voronoi cells that are on the surface of the protein are open. To ensure that
+        Our conclusion is that {\it edge effects} due to Voronoi tessellation have $\lesssim0.01$ influence on the observed sequence-structure correlation strengths in the dataset of $209$ proteins considered in this work.
+Similar conclusions are reached if the open cells were alternatively ranked by the fractional changes in cell area (vs.
+cell volume) upon changing the box size.
+Our conclusion is that Voronoi cell characteristics, in particular cell volume and area can be safely used in predicting sequence variability without recourse to corrections for edge effects.
+An exception is cell sphericity as defined in Eqn.
+\ref{eqn:sphericity}, which turns out to behave differently for open and closed cells.
+This is well illustrated in the adjacent averaging plots of Figures \ref{fig:adjavg_ER} \& \ref{fig:adjavg_SE} where we plot the behavior open and closed cell characteristics averaged over all sites in all proteins in our dataset, versus sequence evolutionary rate and sequence entropy.
+\\
 
-        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work.
-Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., {\it wcnSC}), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
-In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
-Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and
+        The definitions and values of some site-specific structural quantities, such as WCN and Voronoi cell characteristics, depend on the set of atomic coordinates that are used for representing individual sites in proteins.
+In order to identify which set of atomic coordinates best represent protein 3-dimensional structure for the calculation of structural quantities, we calculated WCN and Voronoi cell properties using different sets of atomic coordinates representing individual sites in proteins.
+Moreover, there is also ambiguity as to which set of atomic B factors best represent the flexibility of individual sites in protein.
+\\
 
-         We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
-Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates by as much as $0.12$ in terms of Spearman correlation strength.
+        Our finding is that sequence variability measures correlate best with site-specific structural quantities such as WCN and Voronoi cell characteristics if the geometric center of the side-chains of amino acids are used for the calculation of these structural quantities instead of using the more commonly adopted backbone $C_\alpha$ atomic coordinates.
+In addition, the average of side-chain atomic B factors appear to correlate significantly better with sequence variability measures and site-specific structural quantities such as Relative Solvent Accessibility, \ddg Rate, and measures of amino acid hydrophobicities as illustrated in the plots of Figure \ref{fig:best_bf}.
 
-        In search for the determinants of the strength of sequence-structure relations, we compiled a set of more than $200$ protein properties for a dataset of $209$ monomeric enzymes.
-By employing several independent parametric and non-parametric statistics, such as Spearman rank test, regularized regression and Principal Regression methods, we identify sequence divergence as the dominant factor in the strength of sequence-structure correlations, capable of explaining $10-30\%$ of the observed correlation strengths alone, in both the original and rank-transformed data.
+        All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites rather than by the properties and the spatial coordinates of the backbone $C_\alpha$ atoms.
+In particular, the strength of structure-structure and sequence-structure correlations decrease monotonically when moving from side chain to backbone atoms.
+An exception to this general pattern is the correlation of the hydrogen-bond energies of individual sites with other site-specific structural properties.
+In general, average site-specific H-bond energies correlate more strongly with representative B factors, contact number, and Voronoi cell characteristics, if these quantities are calculated using the backbone Oxygen atoms as representation of individual sites, instead of using the average side chain properties and coordinates.
+This is well illustrated in the bottom-right plots of Figures \ref{fig:best_voronoi}, \ref{fig:best_wcn}, \& \ref{fig:best_bf}.
+The observed monotonic increase in the correlation strengths with H-bond energies from side chain to backbone $O$ atom can be explained away knowing that the backbone Oxygen atom is responsible for virtually all Hydrogen bonds in proteins.
+The influence of individual atoms on H-bond energies monotonically decreases from the neighbor backbone atoms $C$ \& $N$ to the farthest atoms from Oxygen in the amino acid side chains.
+\\
+
+        Regardless of the set of atomic coordinates used for representing individual sites in proteins and for calculating site-specific structural quantities it is notable that once WCN is recalculated using the same set of coordinates used for Voronoi tessellation of protein structure, the quantity WCN as defined by Eqn \ref{eqn:wcn_pwrl} outperforms all other structural quantities --  including those derived from Voronoi tessellation -- in explaining site-specific sequence variability measured, as depicted in Figure \ref{fig:best_predictor}.
+This might not be surprising, knowing that the mathematical definition of WCN includes all sites in proteins which are all likely involved in long-range correlations and electrostatic interactions.
+In contrast to WCN, measures of site flexibility, in particular, the Voronoi cell volume or its inverse as a more local measure of packing density than WCN, do not include the possible effects of long-range interactions among non-neighbor amino acids in proteins.
+In this regard, it is reasonable to expect the Voronoi cell area to perform slightly better than cell volume in explaining the general patterns of sequence evolutions, since the Voronoi cell area is primarily determined by the local packing in addition to the number of neighboring amino acids around a given site of interest in protein, whereas the cell volume is more representative of the local packing and site flexibility than the number of neighboring amino acids around the site of interest.
 
@@
 
-We thank Austin G. Meyer, Stephanie Spielman and Eleisha Jackson at UT Austin for helpful discussions.
+We thank Austin G. Meyer, Stephanie Spielman and Eleisha Jackson at UT Austin for helpful discussions and comments.
 
@@
 
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{adjacent_averaging_screen/voro_log/zr4s_JC.png}
+        \end{center}
+        \caption{General behavior of site-specific structural characteristics versus site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+For plots depicting evolutionary rates versus versus Voronoi cell characteristics, the red \& blue curves represent respectively the general behaviors of closed \& blue open Voronoi cells, in addition to the black curves which represent the behavior of all open and closed cells together.
+The amino acid hydrophobicity scales were taken from \citet{hessa_recognition_2005}.}
+        \label{fig:adjavg_ER}
+    \end{figure}
+
+    \begin{figure}[tbh]
+        \begin{center}
+        \includegraphics[width=6.9in]{adjacent_averaging_screen/voro_log/seqent.png}
+        \end{center}
+        \caption{General behavior of site-specific structural characteristics versus site-specific sequence variability, as measured by sequence entropy (Eqn.
+\ref{eqn:shannon}), among all sites in all 209 proteins in dataset.
+For plots depicting sequence entropy versus Voronoi cell characteristics, the red \& blue curves represent respectively the general behaviors of closed \& blue open Voronoi cells, in addition to the black curves which represent the behavior of all open and closed cells together.
+The amino acid hydrophobicity scales were taken from \citet{hessa_recognition_2005}.}
+        \label{fig:adjavg_SE}
+    \end{figure}
+
 %    \begin{figure}[tbh]

### 2015-03-27 analysis/src/adjacent_averaging_screen_voro_log.r

@@ -43,9 +43,9 @@ res_prop_concise_open = res_prop_concise[res_prop_voroSC$VSCvolume_change_diff !
 
 
 # The following is an ordered list, in agreement with the column names of the above data frame.
-varnames_long = c('Evolutionary Rates (r4sJC)' , 'Sequence Entropy (seqent)' , 'ddG Entropy (ddGent)' , 'Relative Solvent Accessibility (RSA)' , 'Hydrogen Bond Energy (HBE)' ,
-                  'Hydrophobicity Scale (HPS)' , 'Side-Chain Contact Number (wcnSC)' ,
-                  'Average Side-Chain Bfactor (bfSC)' , 'Voronoi Cell Faces' , 'log10 ( Voronoi Cell Surface Area )' ,
+varnames_long = c('Evolutionary Rates (r4sJC)' , 'Sequence Entropy' , 'ddG Rate (Stability upon Substitution)' , 'Relative Solvent Accessibility (RSA)' , 'Hydrogen Bond Energy (HBE)' ,
+                  'Hydrophobicity Scale' , 'Side-Chain Weighted Contact Number' ,
+                  'Average Side-Chain B factor' , 'Number of Voronoi Cell Faces' , 'log10 ( Voronoi Cell Surface Area )' ,
                   'log10 ( Voronoi Cell Volume )' , 'Voronoi Cell Eccentricity' , 'Voronoi Cell Sphericity')
 
 varnames_short = colnames(res_prop_concise)
@@ -80,8 +80,9 @@ for (i in 1:length(varnames_short))
 {
   #i = 1
   cat('generating graph # ', i, '\n')
-  filename = paste0('../figures/adjacent_averaging_screen/voro_log/',varnames_short[i],'.pdf')
-  pdf( filename, width=13.5, height=16, useDingbats=FALSE )  
+  filename = paste0('../figures/adjacent_averaging_screen/voro_log/',varnames_short[i],'.png')
+  #pdf( filename, width=13.5, height=16, useDingbats=FALSE )  
+  png( filename, width=1100, height=1200 )  
   split.screen(c(4,3))
   
   counter = 0

### 2015-04-17 manuscript/voronoi/mnv.tex
new results, figures,...MN updated

--- 2015-03-23 manuscript/voronoi/mnv.tex
@@
 \vfill
-\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
+%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
+\Title{Structural Determinants of Sequence Evolution in Enzymatic proteins}
 \vfill
@@
 A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.
-Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution.
+Expanding on these studies, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution.
 We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
-    % We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins.
-By contrast, other definitions perform equally well or better.
-This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
+Long-range interactions between individual amino acids are found to have a minor but significant contribution to the evolutionary patterns of protein sequence.
     % OLD abstract:
@@
 Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in sequence variability patterns in contrast to RSA and local flexibility measures.
+\\
 
@@
 Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility.
+\\
 
@@
 
-\section{Methods}
+\section{Protein Dataset and Structure$/$Sequence Variability Measures}
 \label{sec:methods}
-
-    \subsection*{Protein Dataset}
 
@@
 For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model \citep[][]{mayrose_comparison_2004}, hereafter abbreviated as {\it r4sJC}.
+\\
+
     In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
@@
 Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
+\\
+
+    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins can be defined and obtained following the stability threshold model of Bloom et al.
+(2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al.
+(2014) \cite{echave_relationship_2014}.
+Based upon this model which was extensively described in Chapter \ref{chap:sschar}, a quantity $\Delta\Delta G$ {\it rate} (or {it ddG rate}) was derived for each individual site in all proteins in dataset.
+A high ddG rate for the $i^{th}$ site in a protein indicates a high stability of the site and the overall conformation of the protein to perturbations caused by substitution of the amino acid residing the site.
+\\
+
+    As a measure of local flexibility or fluctuation in different parts of the protein structure the temperature, factors ({\it B factor}) for all atoms in PDB files were extracted (c.f., Chapter \ref{chap:sschar}).
+Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.
+Alternatively, the site-specific fluctuation could be calculated from MD simulations.
+This was however impossible for this study for the large dataset of $209$ proteins considered in this work.
+\\
 
@@
 
-    \subsection*{Voronoi Tessellation}
+\section{Results}
 
-        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \citep[e.g., ][]{richards_interpretation_1974, gerstein_volume_1994}.
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g.,][]{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
+    \subsection*{Voronoi Partitioning of Protein's Structure}
+    \label{sec:voro}
+
+    There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
 For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
@@
 
+    % Figure goes here xx
+
     In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation.
+An example of Voronoi tessellation of protein structure in two dimensions (PDB ID: {\it 1LBA}) is shown in Figure \ref{fig:voronoi}.
 The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions.
-Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes.
-We use VORO++ software \citep[][]{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
+\\
+
+    Here in this work, the simplest and most widely used definition of Voronoi tessellation described above is applied on a dataset of $209$  monomeric enzymes.
+We use VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
 Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
-In addition, for each cell we also calculate {\it sphericity}, a measure of the cell {\it compactness} defined as,
+A measure of the cell {\it eccentricity} can be also obtained by finding the distance between the cell seed and geometrical center of the cell.
+In addition, the cell {\it sphericity} can be calculated as a measure of the cell's {\it compactness} defined as,
 
@@
 
-        in which $V$ \& $A$ stand for the volume \& area of the cell respectively.
+    \noindent in which $V$ \& $A$ stand for the volume \& area of the cell respectively.
 For a perfectly spherical cell, $\Psi=1$, while it becomes zero for a 2-dimensional object that has no volume but only surface area.
 
-    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
+    \subsection*{Voronoi Cell Area and Volume as Proxy Measures of Local Packing Density and Flexibility in Proteins}
 
-        Depending on the choice of coordinates used, there exist degeneracy in the definition of some site-specific structural variables.
+    % Figure goes here xx
+
+    % Figure goes here xx
+
+    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
+Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
+It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
+In general, it is observed that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
+\\
+
+    The cell properties are also strongly correlated with each other.
+Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in Figure \ref{fig:voronoi_ER_cond}.
+Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
+In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell.
+Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates.
+These cell characteristics are however, discrete (integer) quantities and in general have a limited range.
+\\
+
+    Not shown here for brevity, almost identical results to the above are obtained if sequence entropy as defined by Eqn.
+\ref{eqn:shannon} were used in place of sequence evolutionary rates.
+The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy.
+\\
+
+    One potential caveat with Voronoi tessellation of finite structures in Euclidean space is the {\it edge effects}.
+Sites that are close to the surface of protein are often associated with Voronoi cells that are bounded by the cubic box containing the protein (Figure \ref{fig:voronoi}).
+Here to ensure that these edge effects do not influence the observed sequence-structure correlations, the open cells -- i.e., cells that are partially bounded and closed by the cubic box containing the protein -- are identified in all proteins by examining the variations in individual cell volumes upon changing the size of the cubic box containing the protein to a given extreme value.
+The open cells in individual proteins are then ranked by the fraction of volume changes observed upon changing the box size and then normalized to the the largest volume observed among closed cells.
+It should be noted that the specific extreme value chosen for the box sizes of the proteins or the rank ordering of the open cells does not have any influence on the resulting correlation strengths, since the Spearman's $\rho$ by its definition is a rank correlation coefficient.
+\\
+
+    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work.
+Similar conclusions are reached if the open cells were alternatively ranked by different criteria such as the fractional changes in cell area (vs.
+cell volume) upon changing the box size.
+The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for edge effects.
+An exception however is cell sphericity as defined in Eqn.
+\ref{eqn:sphericity}, which turns out to behave differently for open and closed cells.
+This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates.
+For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths.
+\\
+
+    % Figure goes here xx
+
+    % Figure goes here xx
+
+    \subsection{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
+    \label{sec:voronoi_SC}
+
+    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables.
 For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
-The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \citep[e.g., ][]{soyer_voronoi_2000}.
-More recently, \citet{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\
+The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins.
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}.
+More recently, Marcos \& Echave (2014)\cite{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
+\\
 
-        Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \citep[e.g.,][]{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
+    Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \cite{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
 In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra.
@@
 
-        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates.
+    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
+Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties.
 These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
-We also calculate a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
+In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains.
+Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
 In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid.
+The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations.
+The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}.
+Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all.
 \\
 
-\section{Results}
-\label{sec:results}
+    % Figure goes here xx
 
-    \subsection*{Voronoi Cell Area and Volume as a Measure of Local Packing Density in Proteins}
+    % Figure goes here xx
 
-    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra.
-Plot {\it A} of Figure \ref{fig:best_voronoi} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
-It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
-In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
-The cell properties are also highly associated with each other.
-Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}.
-On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
-In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell.
-Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates.
-These cell characteristics are however, discrete (integer) quantities and in general have a limited range.
-
-    Not shown here for brevity, we also obtain identical results to the above if we use sequence entropy as defined by Eqn.
-\ref{eqn:shannon} instead of sequence evolutionary rates, although sequence entropy generally results in weaker correlation strengths due to the discreteness and limited range inherent in its definition.
-
-    \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}
-
-        Depending on the set of atomic coordinates that represent the protein structure, there are $7$ different measures for site-specific structural properties such as Weighted Contact Number, B factor and Voronoi cell properties.
-Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
+    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
 \\
 
-        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
+    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells.
+Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}).
 \\
 
-        For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ B factor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
+    It is notable that the standard deviations of the difference distributions for both quantities: WCN, and Voronoi cell area, are an order of magnitude smaller than the observed differences, implying that the correlation coefficients for all proteins in dataset uniformly translate to higher values by moving from C$_\alpha$ atomic coordinates to the geometric centers of the side chains, regardless of the strength of the correlation coefficients.
 \\
 
-        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
+    \subsection{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+
+    For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain results in the best correlations with other structural and sequence properties.
+Specifically, the average side chain B factor outperforms the commonly used $C_\alpha$ B factor in predicting RSA, \ddg rate, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
 \\
 
-\section{Discussion and Concluding Remarks}
-\label{sec:dcr}
+    % Figure goes here xx
 
-		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by \citet{echave_relationship_2014}, site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates.
-It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.
-
-        All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites.
-In particular, the strength of structure and sequence correlations decrease when moving from side chain to backbone atoms.
-An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
+    The observed improvements in correlations of average side-chain B factor (vs.
+C$_\alpha$ B factor) with other structural properties also merit further attention.
+It was discussed in Section \ref{sec:voronoi_SC} and depicted in the plots of Figure \ref{fig:best_bf} that in general, as one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve.
+In particular, the use of average side-chain B factor turned out to result in the highest correlation strengths with other site-specific properties, implying that this average B factor is likely the best representation of the overall amino acid fluctuations and flexibility in a given site in protein.
+The definition of B factor and its derivation from Debye-Waller factor has been already discussed in Chapter \ref{chap:sschar}, Eqns.
+\ref{eqn:dwf_def}--\ref{eqn:bf_def}.
 \\
 
-        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
-The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
+    The mean-square-displacement $\langle u^2 \rangle$ in Eqn. \ref{eqn:bf_def} can be decomposed into four contributing components \cite{frauenfelder_temperature-dependent_1979},
+
+    \begin{equation}
+        \label{eqn:bf_comp}
+        \langle u^2 \rangle = \langle u^2 \rangle_c + \langle u^2 \rangle_d + \langle u^2 \rangle_{ld} + \langle u^2 \rangle_v,
+    \end {equation}
+
+    \noindent in which subscripts $c, d, ld, v$ refer to fluctuations due to conformational substates, diffusion, lattice disorder, and thermal vibrations respectively.
+The second term $\langle u^2 \rangle_d$ is generally negligible and can be ignored in Eqn.
+\ref{eqn:bf_comp}.
+Of particular interest to this study is the first term, which is also typically the major contributor to the overall value of the atomic B factor, specially in high-resolution X-ray crystallography of proteins.
+This term represents the positional displacements of the atom of interest together with other atoms in the amino acid between many different conformational substates of the protein, with the transition probability between the substates governed by the Boltzmann distribution.
+Compared to atomic coordinates, there are comparatively fewer restraints on the atomic B factors during X-ray crystallography refinement process, and thus in this regard B factor is generally considered as the {\it error sinks} for static and dynamic disorder and various kinds of model errors in the refinement process \cite{read_structure-factor_1990}.
+The noise and model uncertainty contributions to the atomic B factors in particular increase with decreasing the resolution of the X-ray crystallography.
+Better resolution in general corresponds to lower average B factors for the entire structure of the protein \cite{read_structure-factor_1990}.
 \\
 
-		One potential caveat with Voronoi tessellation is the edge effects, that is, Voronoi cells that are on the surface of the protein are open. To ensure that
+    Although the extraction of conformational fluctuations from noise in B factors seems a daunting task \cite{read_structure-factor_1990}, the effects of noise, model error and uncertainties due to limited X-ray crystallography resolution can be minimized by averaging B factors over the entire amino acid in a given site: To expand on this, consider the contribution of conformational fluctuations between different substates to be approximately the same for all atoms in the amino acid.
+The conformational fluctuations can be regarded as the collective motion of all atoms in the amino acid, on top of which there are noise fluctuations in each of the atoms.
+These collective motions are the type of fluctuations in B factors that are expected to reflect the biologically relevant and important factors for the proper functioning of the protein.
+The stochastic noise in the fluctuations is often assumed to have an isotropic Gaussian origin.
+Therefore, averaging over the atomic B factors in each individual amino acid essentially results in higher Signal-to-Noise Ratio (SNR) in the measurement of the amino acid conformational fluctuations.
+Figure \ref{fig:bfC_bfAA_ratio} illustrates how this averaging over all atomic B factors increases the SNR in measuring the fluctuations due to conformational substate transitions of the amino acid.
+\\
 
-        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work.
-Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., {\it wcnSC}), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
-In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
-Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and
+    % Figure goes here
 
-         We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
-Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates by as much as $0.12$ in terms of Spearman correlation strength.
+    To expand further on this, a simple argument may be given to explain the observed strongly-positive approximately-linear correlation between the two parameters in the plot of Figure \ref{fig:bfC_bfAA_ratio}.
+The contributions to the atomic B factor values of the $i^{th}$ atom in the amino acid in the $j^{th}$ site in a given protein can be assumed to originate from two major sources: conformational substates and stochastic noise due to model uncertainties in refinement process and limited resolution of the X-ray crystallography,
 
-        In search for the determinants of the strength of sequence-structure relations, we compiled a set of more than $200$ protein properties for a dataset of $209$ monomeric enzymes.
-By employing several independent parametric and non-parametric statistics, such as Spearman rank test, regularized regression and Principal Regression methods, we identify sequence divergence as the dominant factor in the strength of sequence-structure correlations, capable of explaining $10-30\%$ of the observed correlation strengths alone, in both the original and rank-transformed data.
+    \begin{equation}
+        \label{eqn:bf_comp}
+        \langle u^2 \rangle_{ij} = \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij}.
+    \end {equation}
+
+    For simplicity and without loss of generality, one can assume that the contribution of fluctuations due to conformational substate transitions is approximately the same for all atomic B factors in a given amino acid residing the $j^{th}$ site.
+In other words, the term $\langle u^2 \rangle_{\text{substates},ij}$ in the above equation has almost the same value $\langle u^2 \rangle_{\text{substates},j}$ for all atoms in the amino acid in the $j^{th}$ site in protein.
+Thus, the average B factor for the entire amino acid molecule of size $N_j$ atoms would be,
+
+    \begin{eqnarray}
+        \label{eqn:bf_avg}
+        \langle u^2 \rangle_{j} &=& \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
+                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
+                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \mu_{\text{noise},j}
+    \end {eqnarray}
+
+    \noindent in which $\mu_{\text{noise},j}$ is the average noise in the $j^{th}$ amino acid.
+The ratio of the B factor of the $ij^{th}$ atom to the average B factor of the $j^{th}$ site in protein can be approximated as,
+
+    \begin{eqnarray}
+        \label{eqn:bf_ratio}
+        \frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} &\simeq& \frac{\langle u^2 \rangle_{substates,j} + \langle u^2 \rangle_{noise,ij}}{\langle u^2 \rangle_{substates,j} + \mu_{noise,j}} \\
+                                 &=& \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \nonumber \\
+                                 &+& \bigg(\frac{\langle u^2 \rangle_{\text{noise},ij}}{\langle u^2 \rangle_{\text{substates},j}} \bigg)
+                                 \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \label{eqn:bf_ratio1} \\
+                                 &\simeq& 1 - \frac{\mu_{noise,j}}{\langle u^2 \rangle_{\text{substates},j}}, \label{eqn:bf_ratio2}
+    \end {eqnarray}
+
+    \noindent where from line \ref{eqn:bf_ratio1} to \ref{eqn:bf_ratio2}, an assumption was made that the second term in line \ref{eqn:bf_ratio1} could be neglected compared to the first term and that the noise compared to conformational fluctuation is small, that is, $\mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j} < 1$ (an error of $0.2\AA$ corresponds approximately to $1\AA$ increase in B factor \cite{read_structure-factor_1990}).
+Knowing that the average noise across different amino acids is approximately the same \cite{frauenfelder_temperature-dependent_1979}, that is $\mu_{noise,j}\sim \mu_{noise}$, and that the noise due to X-ray crystallography almost negatively linearly correlates with crystallography resolution in the range $\sim1-3~[\AA]$ \cite{read_structure-factor_1990}, that is $\mu_{noise}\propto -\text{resolution}$, a positive approximately-linear relationship between the average of the B factor ratios over the entire amino acids in the protein structure and the X-ray crystallography resolution would be obtained,
+
+    \begin{eqnarray}
+        \frac{\text{BF}_C}{\text{BF}_{AA}} &=& \frac{1}{L}\sum_{j=1}^{L}\frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} \label{eqn:avg_bf_ratio} \\
+                                           &\propto& -\mu_{noise} \sum_{j=1}^{L}\frac{1}{\langle u^2 \rangle_{substates,j}} \label{eqn:avg_bf_ratio1} \\
+                                           &\propto& \text{resolution}
+    \end {eqnarray}
+
+    \noindent in which $L$ represents the length of the protein sequence.
+The summation term in line \ref{eqn:avg_bf_ratio1} would not influence this linear relationship, causing only scatter in the relation, so long as the length of the protein not does impose limitations on the resolution of X-ray crystallography of proteins.
+In general, however this may not be the case.
+For the sample of 209 proteins considered here, there exists indeed a weak Spearman's correlation coefficient of $\rho\sim0.2$ between protein length ($L$) and resolution.
+Figure \ref{fig:bfC_bfAA_ratio} illustrates the relationship between the average B factors ratio and the resolution in the dataset, using atom $C$ in the backbone of all amino acids in proteins representing the $i^{th}$ atom in the notation of Eqn.
+\ref{eqn:avg_bf_ratio}.
+It is also notable that the the atomic fluctuations due to conformational substates may not be exactly the same for all atoms in an amino acid in a given site in protein.
+Indeed, one may expect the conformational fluctuations in the backbone atoms would be less significant compared to conformational fluctuations of side-chain atoms.
+\\
+
+    Although averaging B factor over the entire amino acid atoms would reduce the noise further than averaging over side-chain atoms, the functionally important conformational fluctuations that are better captured by the side-chain atomic B factors would compensate for the increase in the noise, such that overall, the B factors averaged over side-chain atoms results in slightly better correlations with sequence variability and other relevant structural characteristics depicted in Figure \ref{fig:best_bf}.
+\\
+
+    \subsection*{Effects of Long-Range Amino Acid Interactions on Sequence Variability}
+
+    Throughout the previous sections, a variety of site-specific structural characteristics of proteins were calculated and discussed, using different sets of atomic coordinates representing individual sites in proteins.
+These include the weighted contact number, the Voronoi cell characteristics, and representative site-specific B factor.
+To make comprehensive comparison of all site-specific structural properties of proteins, we also calculate the Relative Solvent Accessibility (RSA) and \ddg rate \citep[e.g.,][]{echave_relationship_2014} of individual amino acids in their sites in proteins.
+\\
+
+    Figure \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE} compare the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) and sequence entropy.
+It is notable that once WCN is recalculated using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific sequence variability.
+The better performance of WCN compared to local packing density as measured from Voronoi cell volume and area may not be surprising, since WCN by its definition takes into account the potential long-range interactions among amino acids in different regions of protein.
+Indeed, the fractal dimension of proteins very much resembles that of lattice percolation models \cite{stauffer_introduction_1994} and similarly the random packing of hard spheres near percolation threshold \cite{lorenz_universality_1993, liang_are_2001}.
+To expand on this, define the average maximum extent of a protein as,
+
+    % Figure goes here
+
+    % Figure goes here
+
+    \begin{equation}
+        \label{eqn:max_extent}
+        R_m = \frac{1}{2d}\sum_{i=1}^d \big( x_{i,max} - x_{i,min} \big),
+    \end{equation}
+
+    \noindent in which $d=3$ is the dimension of the Euclidean space $x_{i,max} - x_{i,min}$ is the maximum physical extent of the protein, as represented by the geometric center of the side-chain coordinates, in each of the three spatial dimensions.
+Alternatively, the radius of gyration of a protein of length $N$ can be defined as (similar to that of a finite size cluster: Egn.
+45a, Sec.
+3.2, in \cite{stauffer_introduction_1994}),
+
+    \begin{equation}
+        \label{eqn:rg}
+        R_g = \sqrt{\sum_{i=1}^N \frac{|\vec r_i-\vec r_0|^2}{N}},
+    \end{equation}
+
+    \noindent where,
+
+    \begin{equation}
+        \label{eqn:com}
+        \vec r_0 = \sum_{i=1}^N \frac{\vec r_i}{N},
+    \end{equation}
+
+    \noindent is the geometric center of the protein, and $r_i$ is the position of the geometric center of the side-chain of the $i^{th}$ amino acid in protein.
+This definition is such that the kinetic energy and the angular momentum of the protein about the $\vec r_0$ is equivalent to the kinetic energy and the angular momentum of all amino acids residing on a ring of radius $R_g$ centered at $\vec r_0$.
+Figure \ref{fig:maxextent} \& \ref{fig:rg} depict the behavior of the maximum extent ($R_m$) versus protein volume ($V$) and the radius of gyration ($R_g$) versus the protein length ($N$), respectively.
+\\
+
+    % Figure goes here xx
+
+    % Figure goes here xx
+
+    The protein volumes and surface areas calculated using {\it 3V} software \cite{voss_3v:_2010}.
+A symmetric linear fit (i.e., Deming regression) to the plot of $\log R_m$ vs.
+$\log V$ and $\log R_g$ vs.
+$\log N$ results in regression slopes of $D\simeq2.47\pm0.06$ \& $D\simeq2.60\pm0.08$ respectively.
+The observed exponents are very similar to those of the scaling relation,
+
+    \begin{equation}
+        V \propto R^{D}, ~~~~~ (p\simeq p_c)
+    \end {equation}
+
+    \noindent with $D\simeq2.51\pm0.01$ in numerical simulations of hard-sphere packing \cite{lorenz_universality_1993} in three dimensions near percolation threshold ($p\simeq p_c$) or the exponents derived from lattice models $D\simeq2.54\pm0.05$ \cite{adler_series_1990} \& $D\simeq2.5$ \cite{stauffer_introduction_1994}.
+It is notable that far from percolation threshold (i.e., $p\rightarrow 0$) in three dimensions, $D=2$ \cite{parisi_critical_1981}, while above the threshold ($p>p_c$), $D=d=3$, in 3D space.
+
+\section{Discussion}
+\label{sec:discussion}
+
+    Throughout the previous sections,  a comprehensive analysis and comparison of the main structural determinants of sequence variability was carried out, using a dataset of $209$ monomeric enzymes.
+Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg rate as defined in Chapter \ref{chap:sschar} (see also Echave et al.
+(2014) \cite{echave_relationship_2014}), which is essentially a proxy measure of the stability of protein's native conformation upon substitution of amino acids in individual sites in proteins.
+In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg rate, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figures \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE}).
 
@@
 
-We thank Austin G. Meyer, Stephanie Spielman and Eleisha Jackson at UT Austin for helpful discussions.
+We thank Austin G. Meyer, Stephanie Spielman and Eleisha Jackson at UT Austin for helpful discussions and comments.
 
-%\clearpage
-%\newpage
+\bibliographystyle{mbe}
+\bibliography{mnv}
 
-%\section*{Figures}
+\newpage
 
-%\newpage
-
-%\section*{Supplementary Figures}
-
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=6.9in]{voronoi_diagram.pdf}
+        \includegraphics[width=6in]{voronoi_diagram.pdf}
         \end{center}
-        \caption{Example 2-dimensional Voronoi diagram for bacteriophage T7 lysozyme (Protein Data Bank ID `1LBA').
-The red dots represent the backbone $C_\alpha$ atoms projected on the X--Y plane, used as cell seeds in Voronoi tesellation.}
+        \caption{An Example 2-dimensional Voronoi diagram for bacteriophage T7 lysozyme (Protein Data Bank ID `1LBA').
+The red dots represent the backbone $C_\alpha$ atoms projected on the X--Y plane, used as cell seeds in Voronoi tessellation.}
         \label{fig:voronoi}
@@
 
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=7in]{best_voronoi_predictors_of_ER_screen.pdf}
+        \includegraphics[width=5.5in]{best_voronoi_predictors_of_ER.pdf}
         \end{center}
-        \caption{{\bf A:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER).
-Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER.
-{\bf B:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
+        \caption{ A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER).
+Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER.}
         \label{fig:voronoi_ER}
@@
 
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=6.9in]{best_varea/select_variables/boxplot_varea_all_in_one.pdf}
+        \includegraphics[width=5.5in]{best_voronoi_predictors_of_ER_given_area.pdf}
         \end{center}
-        \caption{A comparison of the correlation strength of 6 different measures of Voronoi cell areas with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
-The Voronoi cells are generated using 6 sets of atomic coordinates: {\it SC, AA, CB, CA, N, C, O}, used as different representations of individual sites in proteins.
-The two labels {\it SC} \& {\it AA} stand respectively for the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
-        %about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure/sequence property on the vertical axis with different measures of cell areas with different atomic Voronoi seeds on the horizontal axis are compared against each other.
-The capital letters on the horizontal axis denote the set of atomic coordinates used in Voronoi tessellation of protein structures.
-}
-        \label{fig:best_voronoi}
+        \caption{The partial correlation strengths of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
+        \label{fig:voronoi_ER_cond}
     \end{figure}
 
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=6.9in]{best_wcn/select_variables/boxplot_wcn_all_in_one.pdf}
+        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_voro/zr4s_JC_voro.png}
+        \end{center}
+        \caption{General behavior of Voronoi cell characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The black \& orange curves represent respectively the general behaviors of closed \& open Voronoi cell characteristics.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+        \label{fig:zr4s_JC_voro}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
+        \end{center}
+        \caption{General behavior of site-specific structural characteristics versus site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+        \label{fig:zr4s_JC_nonvoro}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{best_wcn/select_variables/boxplot_wcn_all_in_one.pdf}
         \end{center}
@@
 
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=6.9in]{best_bf/select_variables/boxplot_bf_all_in_one.pdf}
+        \includegraphics[width=5.5in]{best_varea/select_variables/boxplot_varea_all_in_one.pdf}
+        \end{center}
+        \caption{A comparison of the correlation strength of 6 different measures of Voronoi cell areas with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
+The Voronoi cells are generated using 6 sets of atomic coordinates: {\it SC, AA, CB, CA, N, C, O}, used as different representations of individual sites in proteins.
+The two labels {\it SC} \& {\it AA} stand respectively for the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
+        \label{fig:best_voronoi}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{best_bf/select_variables/boxplot_bf_all_in_one.pdf}
         \end{center}
@@
 
-    \begin{figure}[tbh]
+    \begin{figure}
         \begin{center}
-        \includegraphics[width=6.9in]{best_structural_predictors_of_ER.pdf}
+        \includegraphics[width=5.5in]{mean_bfC_bfAA_ratio_resol.pdf}
         \end{center}
-        \caption{A comparison of the prediction power of four structural variables about site-specific evolutionary rates (ER).
-All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively.
-For better illustration however, the absolute Spearman's correlation of WCN with ER are shown in the Figure.}
-        \label{fig:best_predictor}
+        \caption{An illustration of the strong positive correlation of X-ray crystallography resolution with the ratio of the backbone $C$ atomic B factor to the average amino acid B factor ($\text{BF}_C/\text{BF}_{AA}$), averaged over all sites in individual proteins, highlighting the significant contributions of noise and model errors to atomic B factor values.
+The Spearman's correlation coefficient between the two quantities is $\rho\sim0.76$.
+No significant correlation would be expected in the absence of noise due to limited resolution of the X-ray crystallography of proteins.
+Each filled circle in the plot represents one protein in the dataset of 209 enzymes used in this work.}
+        \label{fig:bfC_bfAA_ratio}
     \end{figure}
 
-%    \begin{figure}[tbh]
-%        \begin{center}
-%        \includegraphics[width=6.9in]{best_wcn/select_variables/boxplot_wcn_two_in_one.pdf}
-%        \end{center}
-%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
-The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-%        \label{fig:best_wcn}
-%    \end{figure}
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_ER.pdf}
+        \end{center}
+        \caption{A comparison of the prediction power of five structural variables about site-specific evolutionary rates (ER).
+All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively.
+For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure.
+Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$.
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+        \label{fig:best_predictorER}
+    \end{figure}
 
-\bibliographystyle{mbe}
-\bibliography{mnv}
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_SE.pdf}
+        \end{center}
+        \caption{A comparison of the prediction power of five structural variables about site-specific Sequence Entropy (SE).
+All structural quantities correlate positively with SE, with the exception of Weighted Contact Number (WCN) which correlates negatively.
+For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure.
+Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$.
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+        \label{fig:best_predictorSE}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{fractal_dim_max_extent_volume.pdf}
+        \end{center}
+        \caption{The scaling behavior of protein maximum extent as defined by Eqn.
+\ref{eqn:max_extent} with protein volume for $209$ monomeric enzymes in the dataset.
+The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.47\pm0.06$.}
+        \label{fig:maxextent}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{fractal_dim_rad_gyration_nres.pdf}
+        \end{center}
+        \caption{The scaling behavior of protein's radius  as defined by Eqn.
+\ref{eqn:rg} with protein length for $209$ monomeric enzymes in the dataset.
+The mean \& median length of the proteins are $362$ \& $315$ respectively.
+The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.60\pm0.08$.}
+        \label{fig:rg}
+    \end{figure}
 

### 2015-04-17 analysis/src/adjacent_averaging_screen_only_nonvoro.r

@@ -9,7 +9,6 @@ library('ggplot2')
 library('fields') # used for function fudgeit
 library('zoo')
 
-
 # setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
 # source('get_res_data.r')
@@ -25,9 +24,6 @@ res_prop_concise = data.frame( zr4s_JC       = res_prop_jec$zr4s_JC
                              , hbe           = res_prop_dssp$hbe
                              )
 
-res_prop_concise_closed = res_prop_concise[res_prop_voroSC$VSCvolume_change_diff == 0,]
-res_prop_concise_open = res_prop_concise[res_prop_voroSC$VSCvolume_change_diff != 0,]
-
 #qplot(wcnSC, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
 #qplot(rsa, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
 #qplot(varea, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
@@ -37,7 +33,7 @@ res_prop_concise_open = res_prop_concise[res_prop_voroSC$VSCvolume_change_diff !
 
 # The following is an ordered list, in agreement with the column names of the above data frame.
 varnames_long = c('Evolutionary Rates (r4sJC)' , 'Sequence Entropy' , 'ddG Rate (Stability upon Substitution)' , 'Relative Solvent Accessibility (RSA)' ,
-				  'Hydrophobicity Scale' , 'Side-Chain Weighted Contact Number' , 'Average Side-Chain B factor' , 'Hydrogen Bond Energy (HBE)')
+				  'Side-Chain Weighted Contact Number' , 'Average Side-Chain B factor' , 'Hydrogen Bond Energy (HBE)')
 
 
 varnames_short = colnames(res_prop_concise)
@@ -45,8 +41,6 @@ varnames_short = colnames(res_prop_concise)
 
 # In order to reduce runtime, I am going to save the ordered data frames in a list once, which will be then used in the loops wherever needed.
 list_temp = list()
-list_temp_closed = list()
-list_temp_open   = list()
 for (i in 1:length(varnames_short))
 {
   cat('generating list element ', i, '\n')
@@ -54,16 +48,6 @@ for (i in 1:length(varnames_short))
   temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
[…]

### 2015-04-17 analysis/src/best_structural_predictors_of_ER.r

@@ -77,7 +77,7 @@ plot(  hist.vareaSC$x
     #,   main = 'Correlations with Evolutionary Rates'
     #,   xlab = expression(paste('Spearman Cor. with Evolutionary Rates ',rho))
     ,   xlab = 'Spearman Cor. with Evolutionary Rates'
-    ,   ylab = 'frequency'
+    ,   ylab = 'Frequency'
     )
 lines( -hist.wcnSC$x
      , hist.wcnSC$y
@@ -130,3 +130,8 @@ legend( 'topleft'
       )
 
 graphics.off()
+
+max(best_structural_predictors_of_ER$r.wcnSC.r4sJC)-min(best_structural_predictors_of_ER$r.wcnSC.r4sJC)
+max(best_structural_predictors_of_ER$r.vareaSC.r4sJC)-min(best_structural_predictors_of_ER$r.vareaSC.r4sJC)
+sd(best_structural_predictors_of_ER$r.vareaSC.r4sJC)
+sd(best_structural_predictors_of_ER$r.wcnSC.r4sJC)

### 2015-04-17 analysis/src/best_structural_predictors_of_SE.r

@@ -71,8 +71,8 @@ plot(  hist.vareaSC$x
        ,   lwd  = 2 
        #,   main = 'Correlations with Evolutionary Rates'
        #,   xlab = expression(paste('Spearman Cor. with Evolutionary Rates ',rho))
-       ,   xlab = 'Spearman Cor. with Evolutionary Rates'
-       ,   ylab = 'frequency'
+       ,   xlab = 'Spearman Cor. with Sequence Entropy'
+       ,   ylab = 'Frequency'
 )
 lines( -hist.wcnSC$x
        , hist.wcnSC$y

### 2015-04-17 analysis/src/pdb_prop_figures.r

@@ -3,11 +3,13 @@
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
-# source('input_ASAP_data')
+#source('input_ASAP_data')
 
 library('ppcor')
 
 ASAP_pdb_data = read.csv("../tables/ASAP_pdb_prop.csv", header = T)
+#ASAP_pdb_data = read.csv("../tables/all_pdb_prop_select_wide_asap.csv", header = T)
+#ASAP_pdb_data = read.csv("../tables/ASAP_pdb_prop_manually_corrected.csv", header = T)
 
 all_pdb_prop_select_wide = read.csv('../tables/all_pdb_prop_select_wide.csv', header = T)
 all_pdb_prop_select_wide_asap = read.csv('../tables/all_pdb_prop_select_wide_asap.csv', header = T)
@@ -20,17 +22,17 @@ n.screen.cols = 2   # number of columns in the screen plots
 
 filename = paste0('../figures/cordiv_similarities_seqent.pdf')
 pdf( filename, width=4.5*n.screen.cols, height=4*n.screen.rows, useDingbats=FALSE )
-column_list = c('r.rsa.seqent'
-                ,'r.ddgent.seqent'
+column_list = c( 'r.seqent.varea'
+                ,'r.rsa.seqent'
                 ,'r.bfSC.seqent'
-                ,'r.seqent.varea'
+                ,'r.ddgent.seqent'
                 ,'r.seqent.vsphericity'
                 ,'r.seqent.veccentricity'
                 )
-name_list = c('( Seq. Entropy - RSA )'
-              ,'( Seq. Entropy - ddG Entropy )'
+name_list = c('( Seq. Entropy - Voronoi Cell Surface Area )'
+              ,'( Seq. Entropy - RSA )'
               ,'( Seq. Entropy - Mean Residue Bfactor )'
-              ,'( Seq. Entropy - Voronoi Cell Surface Area )'
+              ,'( Seq. Entropy - ddG Rate )'
               ,'( Seq. Entropy - Voronoi Cell Compactness )'
               ,'( Seq. Entropy - Voronoi Cell Symmetry )'
               )
@@ -50,33 +52,37 @@ for (counter in seq(1,n.screen.rows*n.screen.cols))
        #cex.lab = 1.4,
[…]

### 2015-04-17 wcn_best_definition/analysis/src/get_quantiles.r

@@ -39,6 +39,7 @@ dflist = list( wcneSC_bfSC   = exp_wcneSC_bfSC
              )
 
 counter = 0
+best_params = data.frame()
 for (dataframe in dflist)
 {
   counter = counter + 1
@@ -61,7 +62,7 @@ for (dataframe in dflist)
   }
   rownames(quantiles) = NULL
   write.csv( quantiles, file = paste0('../tables/get_quantiles/sp_raw/',names(dflist)[[counter]],'_quantiles.csv'), row.names=F)
-  
+
   # Now generate the plot:
   model = substr(names(dflist)[[counter]],start=4,stop=4)
   atom  = substr(names(dflist)[[counter]],start=5,stop=6)
@@ -88,7 +89,11 @@ for (dataframe in dflist)
   #lines(predict(uspl,quantiles$parameter),lty=2,lwd=2,col='red')
   graphics.off()
 
-
+  #counter = 0
+  #best_params = data.frame()
+  #for (dataframe in dflist)
+  #{
+  #  counter = counter + 1
   ###############################################################################
   ###############################################################################
   # Now write out the quantiles for the absolute values of Spearman correlation strengths:
@@ -108,6 +113,16 @@ for (dataframe in dflist)
   rownames(quantiles) = NULL
   write.csv( quantiles, file = paste0('../tables/get_quantiles/sp_abs/',names(dflist)[[counter]],'_quantiles.csv'), row.names=F)
 
+  # Now find the best performing paramneters of the four kernels and their corresponding median correltions
+  best_param = quantiles$parameter[which.max(quantiles$median_sp)]
+  best_median_sp = quantiles$median_sp[which.max(quantiles$median_sp)]
+  row_best_param = data.frame( relation = names(dflist)[[counter]]
+                             , best_param = best_param
+                             , best_median_sp_cor = best_median_sp
+                             )
+  best_params = rbind(best_params, row_best_param)
+  #}
+  
   # Now generate the plot:
   model = substr(names(dflist)[[counter]],start=4,stop=4)
   atom  = substr(names(dflist)[[counter]],start=5,stop=6)
@@ -150,6 +165,8 @@ for (dataframe in dflist)
   graphics.off()
 }
 
+write.csv( best_params, file = paste0('../tables/get_quantiles/best_params.csv'), row.names=F)
+
 
[…]

### 2015-04-18 manuscript/voronoi/mnv.tex
minor revision of MN

@@
 
-    \subsection{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
+    \subsection*{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
     \label{sec:voronoi_SC}
@@
 
-    \subsection{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+    \subsection*{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
 

### 2015-04-18 wcn_best_definition/analysis/src/bestFP_determinants.r

@@ -9,6 +9,15 @@ pdb_CO     = read.table('../../../properties/pdb_prop_CO.out',header=T)
 pdb_CO     = pdb_CO[!(pdb_CO$pdb %in% excluded_pdbs),]
 pdb_CO$pdb = factor(pdb_CO$pdb)
 
+pdb_CO_varied = read.table('../../CO_cutoff_varied/CO_cutoff_varied.out',header=F)
+free_param_values = pdb_CO_varied[pdb_CO_varied$V1=='free_parameter',]
+free_param_values = data.frame( t( subset(free_param_values, select=-c(V1)) ) )
+
+pdb_CO_varied = read.table('../../CO_cutoff_varied/CO_cutoff_varied.out',header=T)
+pdb_CO_varied = pdb_CO_varied[!(pdb_CO_varied$free_parameter %in% excluded_pdbs),]
+pdb_CO_varied$free_parameter = factor(pdb_CO_varied$free_parameter)
+
+
 sum_wcnpSC_bfSC     = read.table('../../bfactor/sum_wcnpSC_bfSC.out',header=T)
 sum_wcnpSC_bfSC     = sum_wcnpSC_bfSC[!(sum_wcnpSC_bfSC$pdb %in% excluded_pdbs),]
 sum_wcnpSC_bfSC$pdb = factor(sum_wcnpSC_bfSC$pdb)
@@ -25,6 +34,8 @@ sum_wcnpSC_ddgent     = read.table('../../ddgent/sum_wcnpSC_ddgent.out',header=T
 sum_wcnpSC_ddgent     = sum_wcnpSC_ddgent[!(sum_wcnpSC_ddgent$pdb %in% excluded_pdbs),]
 sum_wcnpSC_ddgent$pdb = factor(sum_wcnpSC_ddgent$pdb)
 
+cor(pdb_CO_varied$X12.000000, sum_wcnpSC_bfSC$free_param_best, method='sp' )
+
 cor(pdb_CO$contact_orderSC, sum_wcnpSC_bfSC$free_param_best, method='sp' )
 cor(pdb_CO$contact_orderSC, sum_wcnpSC_r4sJC$free_param_best, method='sp' )
 cor(pdb_CO$contact_orderSC, sum_wcnpSC_seqent$free_param_best, method='sp' )
@@ -32,15 +43,38 @@ cor(pdb_CO$contact_orderSC, sum_wcnpSC_ddgent$free_param_best, method='sp' )
 
[…]

### 2015-04-18 properties/src/get_res_prop_dist_from_COM.f90
New code for calcaulation of distances of residues from the geometrical centers of PDBs.

@@ -1,0 +1,98 @@
+! This Fortran program attempts to find the distance of individual residues from the geometrical center of the protein.
+! This code used the output data from get_pdb_prop_CO.f90 and get_res_crd_bf.py
+
+! INPUT:
+!       -- Path to the input file pdb_prop_CO.out containing pdb names and number of residues
+!       -- Path to the input representative atomic coordinates and Bfactors (../properties/res_crd_bf)
+!       -- Path and name of the output file containing distance from COM of each residue for all pdb structures.
+
+! Amir Shahmoradi, Saturday 1:27 PM, April 18 2015, iCMB, UT Austin
+
+program get_res_prop_dist_from_COM
+implicit none
+
+! pdb file variables:
+  integer, parameter                            :: npdb=213                ! number of pdbs in pdb_prop_CO.out file
+  integer                                       :: natoms, nres  
+  character(len=3), dimension(:,:), allocatable :: res_name
+  integer         , dimension(:,:), allocatable :: res_num
+  real*8          , dimension(:,:), allocatable :: crd
+  real*8          , dimension(:)  , allocatable :: bfactor,bfactor_min,bfactor_max,distance
+  real*8          , dimension(3)                :: COM  ! geometrical center of protein
+  character(len=6)                              :: pdb,pdb_name            ! The name of the pdb structure and the SINGLE chain contained in it.
+  
+  
+! input files and variables:
+  character(len=300) :: co_in                                              ! input pdb file. ATTN: Any non ATOM records will be ignored!
+  character(len=300) :: crd_in                                             ! input pdb file. ATTN: Any non ATOM records will be ignored!
+  character(len=300) :: output	                                           ! output file containing distances of residues from COM of proteins
+  
+! input files variables:
+  integer            :: iostat, counter
+  integer            :: co_in_unit=10,crd_in_unit=11                       ! input file units
+  integer            :: output_unit=21                                     ! output file units
+  integer            :: ios
+
+! other variables:
[…]

### 2015-04-18 properties/src/get_res_prop_dist_from_COM.f90
distance from COM calculated

@@ -14,8 +14,8 @@ implicit none
 ! pdb file variables:
   integer, parameter                            :: npdb=213                ! number of pdbs in pdb_prop_CO.out file
   integer                                       :: natoms, nres  
-  character(len=3), dimension(:,:), allocatable :: res_name
-  integer         , dimension(:,:), allocatable :: res_num
+  character(len=3), dimension(:)  , allocatable :: res_name
+  integer         , dimension(:)  , allocatable :: res_num
   real*8          , dimension(:,:), allocatable :: crd
   real*8          , dimension(:)  , allocatable :: bfactor,bfactor_min,bfactor_max,distance
   real*8          , dimension(3)                :: COM  ! geometrical center of protein
@@ -25,7 +25,7 @@ implicit none
 ! input files and variables:
   character(len=300) :: co_in                                              ! input pdb file. ATTN: Any non ATOM records will be ignored!
   character(len=300) :: crd_in                                             ! input pdb file. ATTN: Any non ATOM records will be ignored!
-  character(len=300) :: output	                                           ! output file containing distances of residues from COM of proteins
+  character(len=300) :: output                                             ! output file containing distances of residues from COM of proteins
   
 ! input files variables:
   integer            :: iostat, counter
@@ -56,7 +56,8 @@ implicit none
   read(co_in_unit,*)        ! read file header
   read(crd_in_unit,*)       ! read file header
 
-  write(output_unit,'7A25') 'pdb','resnam','resnum','distance_from_COM','bf','bfmin','bfmax'
+  open(unit=output_unit,file=trim(adjustl(output)),status='replace')
+  write(output_unit,'(7A25)') 'pdb','resnam','resnum','distance_from_COM','bf','bfmin','bfmax'
   
 call cpu_time(tstart)
 do ii = 1,npdb

### 2015-04-18 analysis/src/dist_from_COM.r
some more analysis for distance from COM

@@ -1,0 +1,164 @@
+# Amir Shahmoradi, Saturday 2:48 PM, April 18 2015, Wilke Lab, ICMB, UT Austin
+# This code reads in data for distances of residues from the geometrical centers of  proteins.
+# The main goal was to see how other residue properties, in particular Bfcator and WCN, behave as a function distance from COM.
+
+#install.packages('zoo')
+#install.packages('ggplot2')
+#install.packages("fields")
+library('ggplot2')
+library('fields') # used for function fudgeit
+library('zoo')
+
+setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+# source('get_res_data.r')
+
+# The following is short version of the most useful residue properties that I have found so far.
+# Now since the three Voronoi quantities vnvortices, vnedges and vnfaces happen to be exactly the same as seen in the cormat calculated above, I am going to remove them from the data set, in addition to resvol which is not as informative.
+res_prop_concise = data.frame( dfcsc         = res_prop_dfcSC$distance_from_COM
+                             , seqent        = res_prop_elj$seqent
+                             , ddgent        = res_prop_jec_ddg$rate.ddg.foldx
+                             , rsa           = res_prop_dssp$rsa
+                             , wcnSC         = res_prop_wcn_bf$wcnSC
+                             , bfSC          = res_prop_wcn_bf$bfSC
+                             , bfSC_min      = res_prop_dfcSC$bfmin
+                             , bfSC_min      = res_prop_dfcSC$bfmax
+                             , hbe           = res_prop_dssp$hbe
+                             )
+
+#qplot(wcnSC, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
+#qplot(rsa, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
+#qplot(varea, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
+#qplot(ddgent, zr4s_JC, data = res_prop_concise, geom = c("smooth"))
+#qplot(log10(bfSC), zr4s_JC, data = res_prop_concise, geom = c("point"))
+
+
[…]

### 2015-04-18 analysis/src/get_res_data.r

@@ -20,6 +20,11 @@
 # setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
 
   excluded_pdbs = c('1BBS_A','1BS0_A','1DIN_A','1HPL_A')   # These are the 4 PDBs that did not have complete r4s evolutionary rates and are omitted from the dataset to avoid NA values.
+
+  # distances of sites from the geometrical centers of proteins
+  res_prop_dfcSC         = read.table('../../properties/res_prop_dist_from_COM_SC.out', header=T)
+  res_prop_dfcSC         = res_prop_dfcSC[!(res_prop_dfcSC$pdb %in% excluded_pdbs),]
+  res_prop_dfcSC$pdb     = factor(res_prop_dfcSC$pdb)
   
   # site-specific entropies
   res_prop_elj         = read.table('../../elj_pdb_entropies.in', header=T)

### 2015-04-18 wcn_best_definition/analysis/src/bestFP_determinants.r

@@ -61,6 +61,9 @@ plot( free_param_values$X1
     )
 
 max(co_cormat$bfsc[-1:-4],na.rm=TRUE)
+which.max(co_cormat$bfsc[-1:-4])
+co_cormat$bfsc[66]
+free_param_values$X1[66]
 
 plot( free_param_values$X1
     , co_cormat$ddg[-1:-4]

### 2015-04-19 properties/src/get_res_prop_dist_from_COM.f90
PDB gyration radii calculated and added normalized distances from COM added.

@@ -1,10 +1,12 @@
 ! This Fortran program attempts to find the distance of individual residues from the geometrical center of the protein.
-! This code used the output data from get_pdb_prop_CO.f90 and get_res_crd_bf.py
+! This code now also outputs the radius of gyration of proteins ina separater file.
+! This code uses the output data from get_pdb_prop_CO.f90 and get_res_crd_bf.py
 
 ! INPUT:
 !       -- Path to the input file pdb_prop_CO.out containing pdb names and number of residues
 !       -- Path to the input representative atomic coordinates and Bfactors (../properties/res_crd_bf)
 !       -- Path and name of the output file containing distance from COM of each residue for all pdb structures.
+!       -- Path and name of the output file containing the radii of gyrations of PDBs.
 
 ! Amir Shahmoradi, Saturday 1:27 PM, April 18 2015, iCMB, UT Austin
 
@@ -18,7 +20,8 @@ implicit none
   integer         , dimension(:)  , allocatable :: res_num
   real*8          , dimension(:,:), allocatable :: crd
   real*8          , dimension(:)  , allocatable :: bfactor,bfactor_min,bfactor_max,distance
-  real*8          , dimension(3)                :: COM  ! geometrical center of protein
+  real*8          , dimension(3)                :: COM                     ! geometrical center of protein
+  real*8                                        :: radius_of_gyration      ! geometrical center of protein
   character(len=6)                              :: pdb,pdb_name            ! The name of the pdb structure and the SINGLE chain contained in it.
   
   
@@ -26,11 +29,12 @@ implicit none
   character(len=300) :: co_in                                              ! input pdb file. ATTN: Any non ATOM records will be ignored!
   character(len=300) :: crd_in                                             ! input pdb file. ATTN: Any non ATOM records will be ignored!
   character(len=300) :: output                                             ! output file containing distances of residues from COM of proteins
+  character(len=300) :: rogout                                             ! output file containing radii of gyrations
   
 ! input files variables:
[…]

### 2015-05-09 manuscript/cordiv/ej_manuscript_cordiv.tex
Creating a new version of the paper

--- 2015-01-29 manuscript/cordiv/manuscript_cordiv.tex
@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 \usepackage{hyperref}
@@
 \usepackage{rotating}
-\usepackage{slashbox}
+%\usepackage{slashbox}
 \usepackage{afterpage}
 \usepackage{float}
-\citestyle{apj}
+\usepackage{color}
+%\citestyle{apj}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%  address and funding acknowledgement data:
-\def\affiliation{Department of Physics \\
-The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
-%\def\support{\footnote{Work supported by the BEACON-NSF center for the study of evolution in action, under contract xxxxxx.}}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\ddg}{$\Delta\Delta G~$}
-%\newcommand{\se}{\it seqent}
-%\newcommand{\rs}{\it r4sJC}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\def\Title#1{\begin{center} {\Large #1 } \end{center}}
-\def\Author#1{\begin{center}{ \sc #1} \end{center}}
-\def\Address#1{\begin{center}{ \it #1} \end{center}}
-\def\andauth{\begin{center}{and} \end{center}}
-\def\submit#1{\begin{center}Submitted to {\sl #1} \end{center}}
-\newcommand\pubblock{\rightline{\begin{tabular}{l} \pubnumber\\
-         \pubdate  \end{tabular}}}
-\newenvironment{Abstract}{\begin{quotation}  }{\end{quotation}}
-%\newenvironment{Presented}{\begin{quotation} \begin{center}
-%             PRESENTED AT\end{center}\bigskip
-%      \begin{center}\begin{large}}{\end{large}\end{center} \end{quotation}}
-\def\Acknowledgements{\bigskip  \bigskip \begin{center} \begin{large}
-             \bf ACKNOWLEDGEMENTS \end{large}\end{center}}
 
@@
 
+\title{Dissecting the relationship between protein structure and sequence variation}
+\author{Amir Shahmoradi$^1$, Eleisha L. Jackson$^2$, Claus O. Wilke$^2$}%\support}
 \begin{document}
-\begin{titlepage}
-\pubblock
 
-\vfill
-%\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
-\Title{Dissecting the relationship between protein structure and sequence variation}
-\vfill
-\Author{Amir Shahmoradi, Eleisha Jackson, Claus Wilke}%\support}
-\Address{\affiliation}
-\vfill
-\begin{Abstract}
+\date{\today}
+\maketitle
+
+\noindent
+$^1$ Department of Physics, The University of Texas at Austin, TX 78712, USA \\
+$^2$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, USA\\
+
+\bigskip
+\noindent
+$^*$Corresponding author\\
+$\phantom{^*}$Email: amir@physics.utexas.edu\\
+$\phantom{^*}$Phone:{ \color{red} Need a phone} \\
+%$\phantom{^*}$Phone: +1 512 232 2459\\
+
+\bigskip
+\noindent
+Manuscript type: research article\\
+\bigskip
+\noindent  Keywords: protein evolution, relative solvent accessibility, site variability
+
+\begin{abstract}
     Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution.
@@
 
-    %Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations.
-The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
-In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes.
-We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths.
-In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations.
-In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
-\end{Abstract}
+\end{abstract}
 \vfill
 \vfill
-\end{titlepage}
 \def\thefootnote{\fnsymbol{footnote}}
@@
 This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein.
-According to stability threshold model of proteins (xx), some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation.
-Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins (xx).
+According to stability threshold model of proteins {\color{red}(xx)}, some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation.
+Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins {\color{red}(xx)}.
 One of the earliest discovered examples of such relations, is the correlation of the site-specific Relative Solvent Accessibility (RSA) with sequence variation, that is, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
@@
     Other structural properties have also been recently identified and proposed to influence or explain the site-specific evolutionary variations of proteins.
-Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest (xx).
-Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx).
-In a recent work Echave et al.
-xx presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time.
+Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest {\color{red}(xx)}.
+Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed {\color{red}(xx)} and have been shown to correlate better with sequence evolutionary rates {\color{red}(xx)}.
+In a recent work {\color{red}Echave et al.
+xx} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time.
 They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
@@
     Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures.
-By analyzing a data set of $216$ monomeric enzymes, Yeh et al.
-xx found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA).
-Similarly, Echave et al.
-xx recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
+By analyzing a data set of $216$ monomeric enzymes, {\color{red}Yeh et al.
+xx} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA).
+Similarly, {\color{red}Echave et al.
+xx} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
 It appears that sequence-structure correlations tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
@@
 In the following sections, we present evidence in support of these findings and discuss their implications.
+%
     \\
@@
 
-    \subsection*{Sequence Data, Alignments and Evolutionary Rates}
+    \subsection*{Sequence Data, Alignments and Evolutionary Rates} {\color{red} Change this!!!!}
 
-    The results presented in this work are based on a dataset of $209$ monomeric enzymes (Echave papers, Wilke ddg paper xx) randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al.
+    The results presented in this work are based on a dataset of $209$ monomeric enzymes {\color{red}(Echave papers, Wilke ddg paper xx)} randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
+2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes {\color{red}(Webb 1992)} and domains of all main SCOP structural classes (Murzin et al.
 1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by {\color{red}(Yeh et all xx)} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol {\color{red}(Goldenberg et al.
 2009; Ashkenazy et al.
-2010).
-Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al.
-2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
+2010)}.
+Sequence alignments were then constructed using amino-acid sequences with MAFFT {\color{red}(Katoh et al.
+2002, 2005)}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
 The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset.
@@
     where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
-Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software rate4site (xx).
-To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity (Stamatakis 2014).
-For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) (Mayrose et al.
-2004).
+Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(xx)}.
+To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity {\color{red}(Stamatakis 2014)}.
+For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) {\color{red}(Mayrose et al.
+2004)}.
 
@@
 {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the site-specific structural properties include RSA (Tien et al.
-2012 xx), WCN (shih?
-xx).
+Prominent examples of the site-specific structural properties include RSA {\color{red}(Tien et al.
+2012 xx),} WCN {\color{red}(shih?
+xx)}.
 2.
 {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include pdb Contact Order (CO) as defined by xx, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
+Examples include pdb Contact Order (CO) as defined by {\color{red}xx}, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
 In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
@@
 There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-The same definition degeneracy also exists for the set of atomic Bfactors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor (e.g., Halle 2001 xx).
+The same definition degeneracy also exists for the set of atomic Bfactors {\color{red}(xx)} that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor {\color{red}(e.g., Halle 2001 xx)}.
     \\
@@
 
-    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+    We use VORO++ software {\color{red}(xx)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP {\color{red}(xx)} for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al {\color{red}(20112 xx)} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
 In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
@@
 
-    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
+All data and analysis scripts required to reproduce the work are publicly available are to view and download at \url{https://github.com/shahmoradi/cordiv}.
 
@@
         To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.
-Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
+Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression {\color{red}(reference to R package xx)} on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
 Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
@@
 
-\section{Discussion and Concluding Remarks}
+\section{Discussion}
 \label{sec:dcr}
@@
 
-\Acknowledgements
-
+\section{Acknowledgements}
+The authors acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing High-performance computing resources.
+ELJ is is funded by a National Science Graduate Research Fellowship, grant number DGE-1110007.
+COW is funded by {\color{red} Which grants??}.
+AS is funded by {\color{red} Which grants??}.
 ....
@@
 
-%\section*{Figures}
+%\cleardoublepage
 
+\bibliographystyle{peerj} %"style
+\bibliography{ej_manuscript_bib} %expected file "my refs.bib"
+
+%\citep{Bloometal2006}
+\section*{Figures}
     \begin{figure}[tbh]
@@
     \end{figure}
-
-%    \begin{figure}[tbh]
-%        \begin{center}
-%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
-%        \end{center}
-%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
-The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-%        \label{fig:best_wcn}
-%    \end{figure}
 
@@
 The black circles represent $209$ proteins used in this work.
-For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
+For comparison and validation, the red circles represent data for $9$ viral proteins taken from {\color{red}shahmoradi2014xx}}.
         \label{fig:validation}
     \end{figure*}
-
-%\newpage
-
-%\section*{Supplementary Figures}
-
-    %\centerline{\includegraphics[width=6.5in]{PC_screen_entropy_CSrmsf.pdf}}
-    %\noindent \textbf{Fig.
-S1} Principal Component (PC) Regression of sequence entropy against the structural variables, including CS RMSF.
-{\bf (A)} Variance in entropy explained by each principal component.
-For most proteins, PC1 and either PC2 or PC3 show the strongest correlations with sequence entropy.
-Significant correlations ($P<0.05$) are shown as filled symbols, and insignificant correlations ($P\geq0.05$) are shown as open symbols.
-{\bf (B)} and {\bf (C)} Composition of the three leading components.
-Red arrows represent the loadings of each of the structural variables on the principal components; black dots represent the amino acid sites in the PC coordinate system.
-    %\customlabel{fig:cor_entropy_PC_screen_CSrmsf}{S1}
-
-%    \begin{figure}[tbh]
-%        \begin{center}
-%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
-%        \end{center}
-%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN.
-The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-%        \label{fig:best_wcn}
-%    \end{figure}
-%
-%    \begin{figure}[tbh]
-%        \begin{center}
-%        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
-%        \end{center}
-%        \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset.
-In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other.
-The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors.
-The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
-%        \label{fig:best_bf}
-%    \end{figure}
-%
-%    \begin{figure}[tbh]
-%        \begin{center}
-%        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
-%        \end{center}
-%        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
-%        \label{fig:best_voronoi}
-%    \end{figure}
 
@@
 The black circles represent $209$ proteins used in this work.
-For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
+For comparison and validation, the red circles represent data for $9$ viral proteins taken from {\color{red} shahmoradi2014xx}}.
         \label{fig:validation}
@@
 
-\bibliographystyle{apj}
-\bibliography{proc}
+%\bibliographystyle{apj}
+%\bibliography{proc}
 

### 2015-05-12 manuscript/cordiv/ej_manuscript_cordiv.tex
More changes

@@
 \usepackage{color}
-%\citestyle{apj}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%force setting figures at the top of the page:
-    \makeatletter
-    \setlength{\@fptop}{0pt}
-    \makeatother
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%  personal abbreviations and macros
-%    the following package contains macros used in this document:
-%\input econfmacros.tex
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\textwidth=6.9in  \textheight=9.6in
-%%  Adjust these for your printer:
-\leftmargin=-5in   \topmargin=-0.8in
-\hoffset = -0.95in  \voffset = 0pt
-\marginparwidth = 0pt
-%% preprint number data:
-%\newcommand\pubnumber{Article 14 in eConf C1304143}
-\newcommand\pubnumber{}
-\newcommand\pubdate{\today}
+
+\usepackage[margin = 1.0 in]{geometry}
+\usepackage{natbib}
+
+\renewcommand{\bottomfraction}{.9}
+\renewcommand{\topfraction}{.9}
+\renewcommand{\textfraction}{0.1}
+\renewcommand{\floatpagefraction}{.9}
 
@@
 
-\title{Dissecting the relationship between protein structure and sequence variation}
-\author{Amir Shahmoradi$^1$, Eleisha L. Jackson$^2$, Claus O. Wilke$^2$}%\support}
+\title{Sequence divergence as the main determinant of sequence-structure relationships}
+\author{Amir Shahmoradi$^{1*}$, Eleisha L. Jackson$^2$, Claus O. Wilke$^2$}
 \begin{document}
@@
 \noindent
-$^1$ Department of Physics, The University of Texas at Austin, TX 78712, USA \\
-$^2$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, USA\\
+$^1$ Department of Physics, The University of Texas at Austin, Austin, TX 78712, USA \\
+$^2$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, 78712 USA\\
 
@@
 $\phantom{^*}$Email: amir@physics.utexas.edu\\
-$\phantom{^*}$Phone:{ \color{red} Need a phone} \\
-%$\phantom{^*}$Phone: +1 512 232 2459\\
+%$\phantom{^*}$Phone:{ \color{red} Need a phone} \\
+$\phantom{^*}$Phone: +1 512 232 2459\\
 
@@
 \begin{abstract}
-    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution.
-The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
-Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes.
-We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations.
+Recent work has shown that structural properties are capable of predicting site-specific evolutionary rates for a given protein.
+However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
+Here we present the results from a comprehensive search for potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes.
+We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structural-evolution relationships, explaining $10-30 \%$ of observed variation in sequence-structure relations.
 In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations.
@@
 \setcounter{footnote}{0}
-%
 
@@
 
-    Patterns of amino acid sequence variation are known to be influenced by the function of proteins (xx).
-The general consensus, based on the flurry of research done over the past several decades, is that the amino-acid sequence determines the 3D structure of proteins, known as the native conformation.
-This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein.
-According to stability threshold model of proteins {\color{red}(xx)}, some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation.
-Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins {\color{red}(xx)}.
-One of the earliest discovered examples of such relations, is the correlation of the site-specific Relative Solvent Accessibility (RSA) with sequence variation, that is, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
+Proteins are subject to a number of biophysical constraints and functional constraints \citep{Scherreretal2012, Wilkeetal2010}.
+These constraints result site-specific patterns of sequence variability within a protein.
+Recently several site-specific structural properties that can explain the general patterns of sequence variability in proteins have been identified.
+One of the earliest examples is Relative Site Accessibility \citep{Ramseyetal2011, Fransozaetal2009}.
+It was found that residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.
+Another structural property that has been found to significantly correlate with with site-specific variation is local packing density, often measured by Contact Number (CN) or Weighted Contact Number (WCN) \citep{Liaoetal2005, Yehetal2014, Huangetal2014}\\
+\indent In a recent work, \cite{Echaveetal2015} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg)  to the rate at which mutations accumulate in the corresponding site over evolutionary time.
+They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
 \\
-
-    Other structural properties have also been recently identified and proposed to influence or explain the site-specific evolutionary variations of proteins.
-Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest {\color{red}(xx)}.
-Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed {\color{red}(xx)} and have been shown to correlate better with sequence evolutionary rates {\color{red}(xx)}.
-In a recent work {\color{red}Echave et al.
-xx} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time.
-They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
-    \\
-
-    Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures.
-By analyzing a data set of $216$ monomeric enzymes, {\color{red}Yeh et al.
-xx} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA).
-Similarly, {\color{red}Echave et al.
-xx} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
-It appears that sequence-structure correlations tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}.
+\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures.
+By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA.
+Similarly, \citep{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
+It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates(Figure \ref{fig:seqent_structure_cors}).
 This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
-    \\
-
-    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability, implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins.
-Motivated by these observations, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
-We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations.
+{\color{red} Where is the stress model, include it!!!}\\
+\indent The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins.
+Motivated by these observations, here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
+We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relationships.
 In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.
-In the following sections, we present evidence in support of these findings and discuss their implications.
-%
 \\
@@
 
-    \subsection*{Sequence Data, Alignments and Evolutionary Rates} {\color{red} Change this!!!!}
-
-    The results presented in this work are based on a dataset of $209$ monomeric enzymes {\color{red}(Echave papers, Wilke ddg paper xx)} randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al.
-2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes {\color{red}(Webb 1992)} and domains of all main SCOP structural classes (Murzin et al.
-1995).
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by {\color{red}(Yeh et all xx)} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol {\color{red}(Goldenberg et al.
-2009; Ashkenazy et al.
-2010)}.
-Sequence alignments were then constructed using amino-acid sequences with MAFFT {\color{red}(Katoh et al.
-2002, 2005)}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
+    \subsection*{Structures, sequence data, and measures of sequence properties } 
+    The results presented in this work are based on a dataset of $209$ monomeric enzymes \citep{Echaveetal2015} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep{Webb1992} and domains of all main SCOP structural classes \citep{Murzinetal1995}.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citep{Yehetal2014, Huangetal2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep{Goldenbergetal2008}.
+Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given data set.
 The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset.
 To do so, we relied on two independent methods of measuring sequence variability measure.
-First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
-
+First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$:
     \begin{equation}
@@
     where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment.
-Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(xx)}.
-To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity {\color{red}(Stamatakis 2014)}.
-For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) {\color{red}(Mayrose et al.
-2004)}.
+The sequence entropy is a measure of variability at each site.
+We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(Citation??)}.
+First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity cited{Stamatakis2006, Stamatakis2014}.
+For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) \citep{Mayroseetal2004}.
 
-    \subsection*{Structural Properties}
-
-    The goal of the presented work is to identify the prominent structural or evolutionary properties of proteins that modulate sequence-structure correlations.
+    \subsection*{Calculation of Structural Properties}
+    The goal of the presented work is to identify the prominent structural or sequence properties of proteins that modulate sequence-structure correlations.
 These potential modulators represent a unique characteristics of the protein as a whole.
@@
 {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the site-specific structural properties include RSA {\color{red}(Tien et al.
-2012 xx),} WCN {\color{red}(shih?
-xx)}.
+Prominent examples of the site-specific structural properties include RSA \cite{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.
 2.
 {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include pdb Contact Order (CO) as defined by {\color{red}xx}, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.
+Examples include protein Contact Order (CO) as defined by {\color{red}Need Citation!}, protein size and compactness, sequence length, and structural resolution of the protein in X-ray crystallography.
 In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
 Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.
-A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.
+A comprehensive list of protein properties and their definitions are given in Table
+ \ref{tab:definitions} {\color{red} Where is this table???!!!}. \\
+\indent A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other.
+These regions are called Voronoi cells.
+The structure of proteins can be considered as a set of 3D coordinates representing individual sites.
+We use VORO++ software {\color{red}(Need citation!)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP \citep{Kabschetal2005}  for the calculation of Accessible Surface Area (ASA) for each site.
+We normalized the ASA for each site by the theoretical maximum solvent accessibility values of \cite{Tienetal2013}  to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
+To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
+\\
 
     \subsection*{Eliminating Degeneracy in Structural Property Definitions}
-
     In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s).
@@
 There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-The same definition degeneracy also exists for the set of atomic Bfactors {\color{red}(xx)} that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor {\color{red}(e.g., Halle 2001 xx)}.
-    \\
-
-    Similar definition degeneracy also exists for the set of coordinates that can be used for Voronoi tessellation of proteins.
-A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other.
-These regions are called Voronoi cells.
-The structure of proteins can be considered as a set of 3D coordinates representing individual sites.
-Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.
-    \\
-
+The same definition degeneracy also exists for the set of atomic Bfactors {\color{red}(Citation???)} that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor \citep{Halle2002}.
 Thus, for the sake of comprehensiveness and in order to identify the best definitions of structural properties such as WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of properties depending on the choice of the representative set of atomic coordinates used.
@@
 The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
-    \\
+Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.\\
 
-    We use VORO++ software {\color{red}(xx)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP {\color{red}(xx)} for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al {\color{red}(20112 xx)} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns.
-To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
-    \\
-
-All data and analysis scripts required to reproduce the work are publicly available are to view and download at \url{https://github.com/shahmoradi/cordiv}.
+All data and analysis scripts required to reproduce the work are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.
 
@@
     \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}
-
+	{\color{red} CONDENSE THIS WHOLE SECTION. THIS IS COVERED IN HIS PAPER.}
         As explained in previous section, there is a high level of redundancy in the initial set of collected protein properties.
@@
 This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.
-Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
-        \\
-
-        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
-        \\
-
-        For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
+Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.m For the WCN we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).\\
+\indent For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
 Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
-        \\
-
 Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
-        \\
-
 All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites.
@@
 An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
-        \\
+Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
+The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.\\
 
-        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
-The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
-        \\
-
-    \subsection*{Sequence divergence as the main Determinant of Sequence-Structure Relation}
-
+    \subsection*{Sequence divergence as the main determinant of Sequence-Structure Relation}
+	{\color{red}COLOR Delta Delta G dor viral proteins???}
         In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
@@
 
-        A hierarchical clustering of the correlation matrix however, reveals two main independent factors that have the strongest influence on the strengths of sequence-structure correlations: 1.
-The sequence divergence as measured by the standard deviation of sequence entropy and evolutionary rates (denoted by {\it sd.seqent} \& {\it sd.r4sJC}) among all sites in each protein structure.
+        A hierarchical clustering of the correlation matrix reveals two main independent factors that have the strongest influence on the strengths of sequence-structure correlations: 1.
+The sequence {\color{red} divergence} as measured by the standard deviation of sequence entropy and evolutionary rates (denoted by {\it sd.seqent} \& {\it sd.r4sJC}) among all sites in each protein structure.
 2.
-The homogeneity of the hydrogen bond strengths among the back bone atoms of each protein structure, as measured by the standard deviation of hydrogen bond energies (denoted by {\it sd.hbe}) among all pdb sites.
+The homogeneity of the hydrogen bond strengths among the backbone atoms of each protein structure, as measured by the standard deviation of hydrogen bond energies (denoted by {\it sd.hbe}) among all protein sites.
 A reduced-size of the Spearman correlation matrix for the most influential factors on the two strongest sequence--structure (seqent$/$r4s -- wcnSC$/$varea) relations is illustrated in Figure \ref{fig:mainmods}.
@@
 
-%\citep{Bloometal2006}
 \section*{Figures}
@@
 
-    \begin{figure*}[tbh]
-        \centering
-        %\begin{tabular}{cc}
-        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
-        %\end{tabular}
-        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
+%    \begin{figure*}[tbh]
+  %      \centering
+ %       \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
+ %       \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy.
 The black circles represent $209$ proteins used in this work.
 For comparison and validation, the red circles represent data for $9$ viral proteins taken from {\color{red}shahmoradi2014xx}}.
-        \label{fig:validation}
-    \end{figure*}
+%        \label{fig:validation}
+%    \end{figure*}
+
+    \begin{figure}[tbh]
+        \centering
+        \includegraphics[width=7in]{/best_wcn/select_variables/boxplot_wcn_all_in_one.pdf} \\
+        \caption{NEEDS A CAPTION.}
+        \label{fig:best_wcn}
+    \end{figure}
 
@@
 
-%\bibliographystyle{apj}
-%\bibliography{proc}
-
 \end{document}

### 2015-05-14 manuscript/cordiv/ej_manuscript_cordiv.tex
Making more changes

@@
 \begin{abstract}
-Recent work has shown that structural properties are capable of predicting site-specific evolutionary rates for a given protein.
+Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein.
 However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.
@@
 In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations.
-In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relations.
-
+In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest correlations between structural properties and site variability.
 \end{abstract}
@@
 
-Proteins are subject to a number of biophysical constraints and functional constraints \citep{Scherreretal2012, Wilkeetal2010}.
+{\color{blue}
+The result in this paper: Site variabilty is the main determine of the difference in structure-sequence correlations.
+This is the main.
+It might be good to mention that we looked at 200 other predictors and found this as the more important one.
+The issue: There is a difference between the Huang et al paper vs.
+Sharamoradi et al.
+In Echave's paper with Huang or Yeh?, they say that the CN is the best predictor.
+In Amir's paper, we say that RSA is still the best predictor.
+Why is this different?
+The difference is in the datasets in terms of variability.
+This paper should have about 3 figures according to Claus possibly.
+So far I see 4 plots that need to be made: \\
+1. WCN-Rate4site vs Mean RateSite??? Does this make sense? (Also for RSA) \\
+2. WCN-Entropy vs Mean Entropy \\
+3. RSA-Entropy vs Mean Entropy \\
+Open Questions: Should I make the stress  and delta delta G correlations plots for the viral sequences that would complete the story?}
+
+Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}.
 These constraints result site-specific patterns of sequence variability within a protein.
-Recently several site-specific structural properties that can explain the general patterns of sequence variability in proteins have been identified.
+Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified.
 One of the earliest examples is Relative Site Accessibility \citep{Ramseyetal2011, Fransozaetal2009}.
-It was found that residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.
+Residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.
 Another structural property that has been found to significantly correlate with with site-specific variation is local packing density, often measured by Contact Number (CN) or Weighted Contact Number (WCN) \citep{Liaoetal2005, Yehetal2014, Huangetal2014}\\
 \indent In a recent work, \cite{Echaveetal2015} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg)  to the rate at which mutations accumulate in the corresponding site over evolutionary time.
-They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
+They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates comparable to the predictive powers of solvent accessibility and contact number.
 \\
-\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures.
-By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA.
-Similarly, \citep{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
-It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates(Figure \ref{fig:seqent_structure_cors}).
+\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different proteins.
+By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho (\sim0.1-0.8)$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA.
+Similarly, \cite{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
+{\color{blue} It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates.
+SHOULD THIS BE IN RESULTS??} (Figure \ref{fig:seqent_structure_cors}).
 This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
-{\color{red} Where is the stress model, include it!!!}\\
+{\color{red} Where is the stress model, include it!!!
+Need to mention Huang Paper somewhere!}\\
 \indent The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins.
-Motivated by these observations, here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
-We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relationships.
+The underlying principles behind the wide variation in correlation strengths among proteins in not well understood.
+Here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
+We show that among all properties considered, {\color{blue}sequence divergence.
+What??
+Can we discuss divergence when talking about entropy?} appears to be the primary determinant for the strength of sequence-structure relationships.
 In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.
@@
 
-    \subsection*{Structures, sequence data, and measures of sequence properties } 
+    \subsection*{Structures, sequences, and measures of sequence properties } 
     The results presented in this work are based on a dataset of $209$ monomeric enzymes \citep{Echaveetal2015} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep{Webb1992} and domains of all main SCOP structural classes \citep{Murzinetal1995}.
@@
 We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(Citation??)}.
-First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity cited{Stamatakis2006, Stamatakis2014}.
+First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity \citep{Stamatakis2006, Stamatakis2014}.
 For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) \citep{Mayroseetal2004}.
@@
 {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the site-specific structural properties include RSA \cite{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.
+Prominent examples of the site-specific structural properties include RSA \citep{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.
 2.
@@
 Examples include protein Contact Order (CO) as defined by {\color{red}Need Citation!}, protein size and compactness, sequence length, and structural resolution of the protein in X-ray crystallography.
-In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein.
+{\color{blue} RE-WRITE THIS In addition, the distribution of each residue-level property can be summarized by its statistical moments as protein-level property of the protein}.
 Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.

### 2015-05-14 manuscript/cordiv/ej_manuscript_cordiv.tex
Removing non-relevant

@@
 \section*{Figures}
-    \begin{figure}[tbh]
-            \begin{center}
-            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
-            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
-            \end{center}
-            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
+%    \begin{figure}[tbh]
+%            \begin{center}
+%            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
+%            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
+%            \end{center}
+%            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs.
 correlations of other structural properties with evolutionary rates.
@@
 For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
-            \label{fig:seqent_structure_cors}
-    \end{figure}
-
-    \begin{figure}[tbh]
-        \centering
-        \includegraphics[width=3.6in]{/main_modulators_cormat_r4s_wcn.pdf} \\
-        \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
+%            \label{fig:seqent_structure_cors}
+%    \end{figure}
+%
+%
+%    \begin{figure}[tbh]
+%        \centering
+%        \includegraphics[width=3.6in]{/main_modulators_cormat_r4s_wcn.pdf} \\
+%        \caption{{\bf The Spearman correlation matrix of the strongest sequence--structure correlations and the prominent determinants of the strengths of the corresponding relations}.
 The variables on the diagonal elements of the matrix from top to bottom represent respectively, the four strongest sequence--structure relations: r4sJC--wcnSC, r4sJC--varea, seqent--wcnSC, seqent--varea, followed by protein properties that modulate the strength of these relations: the correlation between r4sJC--seqent ({\it r.r4sJC.seqent}), variance of r4sJC ({\it var.r4sJC}), variance of sequence entropy ({\it var.seqent}), variance of back-bone hydrogen bond energies ({\it var.hbe}), and the fraction of amino acids in helical \& $\beta-$sheet secondary structures in the proteins ({\it mn.helix} \& {\it mn.betas} respectively).}
-        \label{fig:mainmods}
-    \end{figure}
+%        \label{fig:mainmods}
+%    \end{figure}
 
@@
 
-    \begin{figure}[tbh]
-        \centering
-        \includegraphics[width=7in]{/best_wcn/select_variables/boxplot_wcn_all_in_one.pdf} \\
-        \caption{NEEDS A CAPTION.}
-        \label{fig:best_wcn}
-    \end{figure}
-
-    \begin{figure*}[tbh]
-        \centering
-        \begin{tabular}{cc}
-        \includegraphics[width=3.4in]{/modulators_similarity_Sr4sJC_scors.pdf} & \includegraphics[width=3.4in]{/modulators_similarity_Sseqent_scors.pdf} \\
-        \end{tabular}
-        \caption{{\bf Spearman correlation matrices illustrating the similarity of the set of structure and sequence properties that contribute to the strength of sequence--structure relations}.
+%    \begin{figure}[tbh]
+%        \centering
+%        \includegraphics[width=7in]{/best_wcn/select_variables/boxplot_wcn_all_in_one.pdf} \\
+%        \caption{NEEDS A CAPTION.}
+%        \label{fig:best_wcn}
+%    \end{figure}
+%
+%    \begin{figure*}[tbh]
+%        \centering
+%        \begin{tabular}{cc}
+%        \includegraphics[width=3.4in]{/modulators_similarity_Sr4sJC_scors.pdf} & \includegraphics[width=3.4in]{/modulators_similarity_Sseqent_scors.pdf} \\
+%        \end{tabular}
+%        \caption{{\bf Spearman correlation matrices illustrating the similarity of the set of structure and sequence properties that contribute to the strength of sequence--structure relations}.
 {\bf Left:} The correlation matrix for the relation of evolutionary rates ({\it r4sJC}) with structural quantities on the diagonal of the matrix.
 {\bf Right:} The correlation matrix for the relation of sequence entropy ({\it seqent}) with structural quantities on the diagonal of the matrix.}
-        \label{fig:simcormats}
-    \end{figure*}
-
+%        \label{fig:simcormats}
+%    \end{figure*}
+%
     \begin{figure*}[tbh]

### 2015-05-15 analysis/src/wcnSC_bfSC_adjacent_avg_scaled.r
new figures added (hist of B factor for all proteins)

@@ -1,0 +1,191 @@
+# Amir Shahmoradi, Friday 8:26 PM, May 11 2015, Wilke Lab, ICMB, UT Austin
+
+#install.packages('zoo')
+#install.packages('ggplot2')
+#install.packages("fields")
+library('ggplot2')
+library('fields') # used for function fudgeit
+library('zoo')
+
+# setwd('C:/Users/Amir/Documents/GitHub/cordiv/analysis/src')
+
+# source('get_res_data.r')
+
+# The following is short version of the most useful residue properties that I have found so far.
+# Now since the three Voronoi quantities vnvortices, vnedges and vnfaces happen to be exactly the same as seen in the cormat calculated above, I am going to remove them from the data set, in addition to resvol which is not as informative.
+res_prop_wcnSC_bfSC = data.frame( wcnSC         = res_prop_wcn_bf$wcnSC
+                                , bfSC          = res_prop_wcn_bf$bfSC
+                                )
+
+# The following is an ordered list, in agreement with the column names of the above data frame.
+varnames_long = c('Side-Chain Weighted Contact Number' , 'Average Side-Chain B factor')
+
+
+varnames_short = colnames(res_prop_concise)
+
+
+# In order to reduce runtime, I am going to save the ordered data frames in a list once, which will be then used in the loops wherever needed.
+list_temp = list()
+for (i in 1:length(varnames_short))
+{
+  cat('generating list element ', i, '\n')
+  res_prop_ordered = res_prop_concise[with(res_prop_concise, order(res_prop_concise[[varnames_short[[i]][1]]])),]
+  temp = rollapply(res_prop_ordered, width = 3000, FUN = mean)
+  temp = data.frame(temp)
+  list_temp[[i]] = temp
+}
+
+
+fudgeit <- function(){
+  xm <- get('xm', envir = parent.frame(1))
+  ym <- get('ym', envir = parent.frame(1))
+  z  <- get('dens', envir = parent.frame(1))
+  colramp <- get('colramp', parent.frame(1))
+  image.plot(xm,ym,z, col = colramp(256)
+             , legend.only = T
+             , add = F
+  )
+}
+
+
[…]

### 2015-05-16 manuscript/voronoi/mnv.tex
MN updated

@@
 %\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
-\Title{Structural Determinants of Sequence Evolution in Enzymatic proteins}
+\Title{Dissecting the relationship between protein structure and sequence variation}
 \vfill
@@
 A number of site-specific structural properties have been proposed over the past decade to answer this question.
-The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition.
+Most importantly, the role of local packing density has been highlighted recently and shown to be the dominant factor in the observed patterns of site-specific sequence variability of proteins.
+A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which represents by definition the combined effects of local packing density and long-range amino acid interactions on sequence variability.
+Here we propose a methodology to segregate and quantify the extent to which long range interactions determine the general patterns of sequence variability in proteins, independently of local packing density.
+We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of local packing cell, defined as the inverse of the volume of the Voronoi cells corresponding to each site in protein.
+
+    The majority of these quantities however depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition.
 A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.

### 2015-05-16 manuscript/voronoi/mnv.tex
minor update to MN

--- 2015-04-18 manuscript/voronoi/mnv.tex
@@
     What are the best structural predictors of protein's sequence evolution?
-A number of site-specific structural properties have been proposed over the past decade to answer this question.
-The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition.
+A number of site-specific structural characteristics have been proposed over the past decade to answer this question.
+Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability of proteins.
+A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which by its definition represents the combined effects of local packing density and long-range amino acid interactions on sequence variability.
+Here we propose a methodology to segregate and quantify the extent to which long range amino acid interactions determine the general patterns of sequence variability in proteins, independently of local packing density.
+We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse of the volume of the Voronoi cells representing individual sites in protein.
+
+    %The majority of these quantities however depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition.
 A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.
@@
     Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
-It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
-\citet{franzosa_structural_2009} use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
-Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in sequence variability patterns in contrast to RSA and local flexibility measures.
+It is perceivable that quantities such as B factor, RSA, and CN, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility.
+\citet{franzosa_structural_2009} used a variety of the local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral impact.
+Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns of enzymatic proteins in contrast to RSA and local flexibility measures.
 \\

### 2015-05-17 manuscript/cordiv/ej_manuscript_cordiv.tex

@@
 \begin{abstract}
+NEED TO EDIT THE ABSTRACT\\
 Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein.
@@
 In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest correlations between structural properties and site variability.
+
 \end{abstract}
@@
 
-{\color{blue}
-The result in this paper: Site variabilty is the main determine of the difference in structure-sequence correlations.
-This is the main.
-It might be good to mention that we looked at 200 other predictors and found this as the more important one.
-The issue: There is a difference between the Huang et al paper vs.
-Sharamoradi et al.
-In Echave's paper with Huang or Yeh?, they say that the CN is the best predictor.
-In Amir's paper, we say that RSA is still the best predictor.
-Why is this different?
-The difference is in the datasets in terms of variability.
-This paper should have about 3 figures according to Claus possibly.
-So far I see 4 plots that need to be made: \\
-1. WCN-Rate4site vs Mean RateSite??? Does this make sense? (Also for RSA) \\
-2. WCN-Entropy vs Mean Entropy \\
-3. RSA-Entropy vs Mean Entropy \\
-Open Questions: Should I make the stress  and delta delta G correlations plots for the viral sequences that would complete the story?}
-
 Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}.
@@
 Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified.
-One of the earliest examples is Relative Site Accessibility \citep{Ramseyetal2011, Fransozaetal2009}.
-Residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.
-Another structural property that has been found to significantly correlate with with site-specific variation is local packing density, often measured by Contact Number (CN) or Weighted Contact Number (WCN) \citep{Liaoetal2005, Yehetal2014, Huangetal2014}\\
-\indent In a recent work, \cite{Echaveetal2015} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg)  to the rate at which mutations accumulate in the corresponding site over evolutionary time.
-They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates comparable to the predictive powers of solvent accessibility and contact number.
+One of the earliest examples is Relative Site Accessibility.
+\cite{Fransozaetal2009} identified RSA as the strongest predictor of evolutionary rate.
+They found that residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.
+In their analysis these concerned both RSA and various definitions of residue packing density to predict evolutionary rate.
+They found that RSA and evolutionary rate shared a significant linear relationship.
+Afterwards, several other works \citep{Ramseyetal2011, Scherreretal2012} also found that RSA as a very significant predictor of evolutionary rate and found this linear relationship as well.
+However in these papers all have the same flaw.
+During the course of their analysis they binned the protein sites and average over all sites within a bin when looking at the trend of RSA.
+This could produce artifacts when looking at the trend.
 \\
-\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different proteins.
-By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho (\sim0.1-0.8)$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA.
-Similarly, \cite{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates.
-{\color{blue} It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates.
-SHOULD THIS BE IN RESULTS??} (Figure \ref{fig:seqent_structure_cors}).
-This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
-{\color{red} Where is the stress model, include it!!!
-Need to mention Huang Paper somewhere!}\\
-\indent The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins.
-The underlying principles behind the wide variation in correlation strengths among proteins in not well understood.
-Here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.
-We show that among all properties considered, {\color{blue}sequence divergence.
-What??
-Can we discuss divergence when talking about entropy?} appears to be the primary determinant for the strength of sequence-structure relationships.
-In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.
+\indent Later, Yeh et al performed a similar analysis on a series of enzyme monomer proteins and found that packing density, as defined by CN and WCN \citep{Liaoetal2005, Yehetal2014, Huangetal2014}, was the strongest determinant of site variability.
+A year later, Sharamoradi et al also performed a site-wise analysis on a series of viral proteins.
+In this analysis they found that RSA had the strongest correlation with site variability.
+Additionally the effect seen between CN and WCN was of a much smaller magnitude as compared to \citep{Yehetal2014}.
+Here we attempt to reconcile the work done in this area.
+We find that site variability is the primary determinant of structure determinant of the strength of structure-sequence relationships and some differences in previous work can be explained in terms of differing levels of site variability.
 \\
+COPY EDIT THIS!!! \\
 
@@
     \subsection*{Structures, sequences, and measures of sequence properties } 
-    The results presented in this work are based on a dataset of $209$ monomeric enzymes \citep{Echaveetal2015} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep{Webb1992} and domains of all main SCOP structural classes \citep{Murzinetal1995}.
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citep{Yehetal2014, Huangetal2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep{Goldenbergetal2008}.
-Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given data set.
-The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset.
+    The results presented in this work are based on a two datasets.
+The first is a dataset of $209$ monomeric enzymes from \citep{Echaveetal2015} originally from Huang et al.
+(2014).The original dataset was comprised of $213$ but we removed four of the proteins (1BBS, 1BS0, 1DIN, 2HPL) that had did not have data at insertion sites.
+Briefly, these proteins are all enzyme monomers   randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$.
+For each structure we had a corresponding alignment of up to 300 homologous sequences.
+The second dataset was from taken from Sharamoradi et al.
+and is comprised of nine?
+viral proteins.
+The viral proteins range from 122 - 557 in length and each structure is accompanied by a sequence alignment of up to 2362 homologous sequences.
+Sequence alignments for both datasets constructed using amino-acid sequences with MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given data set.The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in both datasets.
 To do so, we relied on two independent methods of measuring sequence variability measure.
@@
 The sequence entropy is a measure of variability at each site.
-We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(Citation??)}.
+We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site.
 First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity \citep{Stamatakis2006, Stamatakis2014}.
@@
     \subsection*{Calculation of Structural Properties}
-    The goal of the presented work is to identify the prominent structural or sequence properties of proteins that modulate sequence-structure correlations.
-These potential modulators represent a unique characteristics of the protein as a whole.
-In general, the structural and evolutionary properties fall into two major categories.
-1.
-{\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence.
-Prominent examples of the site-specific structural properties include RSA \citep{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.
-2.
-{\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole.
-Examples include protein Contact Order (CO) as defined by {\color{red}Need Citation!}, protein size and compactness, sequence length, and structural resolution of the protein in X-ray crystallography.
-{\color{blue} RE-WRITE THIS In addition, the distribution of each residue-level property can be summarized by its statistical moments as protein-level property of the protein}.
-Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates.
-A comprehensive list of protein properties and their definitions are given in Table
- \ref{tab:definitions} {\color{red} Where is this table???!!!}. \\
-\indent A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other.
-These regions are called Voronoi cells.
-The structure of proteins can be considered as a set of 3D coordinates representing individual sites.
-We use VORO++ software {\color{red}(Need citation!)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP \citep{Kabschetal2005}  for the calculation of Accessible Surface Area (ASA) for each site.
+
+    HOW IS CN and WCN Defined and Calculated??? (Cite Shih et al for WCN) Possbily Lin Paper \\
+\indent We used DSSP \citep{Kabschetal2005}  for the calculation of Accessible Surface Area (ASA) for each site.
 We normalized the ASA for each site by the theoretical maximum solvent accessibility values of \cite{Tienetal2013}  to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
@@
 \\
-
-    \subsection*{Eliminating Degeneracy in Structural Property Definitions}
-    In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s).
-There are however degeneracies in the definition of the some site-specific variables.
-For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins.
-There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins.
-The same definition degeneracy also exists for the set of atomic Bfactors {\color{red}(Citation???)} that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor \citep{Halle2002}.
-Thus, for the sake of comprehensiveness and in order to identify the best definitions of structural properties such as WCN, Bfactor, and Voronoi cells, here we calculate and consider all possible definitions of properties depending on the choice of the representative set of atomic coordinates used.
-These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains.
-We also calculate a representative coordinate for each site that is an average over all heavy atom coordinates in the side chain and backbone of the amino acid.
-In rare cases where the side chain atoms are not resolved in the PDB file or the amino acid lacks the heavy atom needed (e.g., $C_\beta$ for Glycine).
-The coordinate for that specific site is replaced with the coordinate of the corresponding $C_\alpha$ atom in the amino acid backbone.
-Similar to WCN and Bfactor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the calculation of Voronoi cells.\\
 
@@
 
-    \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}
-	{\color{red} CONDENSE THIS WHOLE SECTION. THIS IS COVERED IN HIS PAPER.}
-        As explained in previous section, there is a high level of redundancy in the initial set of collected protein properties.
-In particular, depending on the set of atomic coordinates used, there are $7$ different measures for some residue characteristics such as the residue Weighted Contact Number, Bfactor and Voronoi cell properties.
-This in turn results in a large set of secondary variables at pdb-level that basically measure the same protein characteristics, but with different strengths.
-Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.m For the WCN we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).\\
-\indent For the measure of local flexibility in proteins (Bfactor) we similarly find that among all $7$ representative measures of site Bfactors, the average of Bfactor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties.
-Specifically, bfSC outperforms the commonly used $C_\alpha$ Bfactor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
-Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume and the cell compactness also correlate best with other structure and sequence properties, only if the average side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
-All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites.
-In particular, the strength of structure and sequence correlations decrease when moving from side chain to backbone atoms.
-An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
-Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics.
-The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.\\
-
-    \subsection*{Sequence divergence as the main determinant of Sequence-Structure Relation}
-	{\color{red}COLOR Delta Delta G dor viral proteins???}
-        In order to identify the potential contributing factors to the strength of sequence--structure correlations, we first employ one of the simplest nonparametric yet powerful tests of statistical dependence, that is, we construct the Spearman correlation matrix of all pdb-level structure and sequence properties.
-The choice of Spearman versus the popular Pearson's correlation measure is made in order to minimize the effects of any nonlinear variable relationships on the strengths of the correlations.
-The resulting correlation matrix reveals a myriad of pdb-level properties each having a small but nonzero contribution to the strength of the structure-sequence correlations.
-        \\
-
-        A hierarchical clustering of the correlation matrix reveals two main independent factors that have the strongest influence on the strengths of sequence-structure correlations: 1.
-The sequence {\color{red} divergence} as measured by the standard deviation of sequence entropy and evolutionary rates (denoted by {\it sd.seqent} \& {\it sd.r4sJC}) among all sites in each protein structure.
-2.
-The homogeneity of the hydrogen bond strengths among the backbone atoms of each protein structure, as measured by the standard deviation of hydrogen bond energies (denoted by {\it sd.hbe}) among all protein sites.
-A reduced-size of the Spearman correlation matrix for the most influential factors on the two strongest sequence--structure (seqent$/$r4s -- wcnSC$/$varea) relations is illustrated in Figure \ref{fig:mainmods}.
-        \\
-
-        For the other weaker sequence--structure relations, i.e.
-the correlations of seqent$/$r4s with RSA, \ddg entropy (ddgent), and Bfactor (bfSC) we find other pdb-level properties that also contribute to the correlation strengths, comparable to or even stronger than in sequence divergence and hydrogen bond homogeneity.
-In general, we observe that for the weaker the sequence-structure correlations, factors that determine the accuracy of the measured residue properties become more influential on the strength of the correlations.
-In particular, the X-ray crystallographic resolution of the structure and the definition of the \ddg entropy play dominant roles, with Spearman correlation coefficients of $\rho~0.3$, on the strengths of the corresponding sequence-structure relations.
-        \\
-
-        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.
-Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression {\color{red}(reference to R package xx)} on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data.
-Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest.
-In addition to regularized regression, we have also employed Principal Component Regression (PCR) on the original dataset and its rank transformation.
-Both regression methods, PCR \& regularized, point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.
+    \subsection*{}
+{\color{red}SOME RESULTS.....} \\
 
@@
 \label{sec:dcr}
-
-        Throughout this work, we have carried out a comprehensive analysis in search for the main determinants of the strength of sequence--structure correlations -- some of which are newly reported and discussed in this work.
-Examples of sequence--structure relations include the correlations of sequence entropy ({\it seqent}) and measures of evolutionary rates (such as {\it r4sJC} used in this work) with measures of residue Contact Number (e.g., {\it wcnSC}), Relative Solvent Accessibility (RSA), \ddg entropy ({\it ddgent}).
-In addition, we have derived new site-specific properties, based on Voronoi Tessellation of protein 3D structures, that are comparable to or better than several previously known structural properties in explaining site-specific sequence entropy or evolutionary rates (e.g., Figure \ref{fig:seqent_structure_cors}).
-Prime examples include Voronoi cell volume ({\it vvolume}), surface area ({\it varea}) and Voronoi cell sphericity defined as,
-
-        \begin{equation}
-        \label{eqn:sphericity}
-        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
-        \end{equation}
-
-        in which $V$ \& $A$ represent {\it vvolume} \& {\it varea} respectively.
-We have also shown that site-specific structural properties -- such as Weighted Contact Number, Bfactor and Voronoi Cell properties -- that are calculated from the average coordinates of side chain atoms, have the best explanatory powers for the sequence variability measures such as {\it seqent} and {\it r4sJC}.
-Compared to the common choice of back bone {\it CA} atomic coordinates, site-specific properties averaged over side chain atoms can outperform in predicting sequence evolutionary rates by as much as $0.12$ in terms of Spearman correlation strength.
-
-        In search for the determinants of the strength of sequence-structure relations, we compiled a set of more than $200$ protein properties for a dataset of $209$ monomeric enzymes.
-By employing several independent parametric and non-parametric statistics, such as Spearman rank test, regularized regression and Principal Regression methods, we identify sequence divergence as the dominant factor in the strength of sequence-structure correlations, capable of explaining $10-30\%$ of the observed correlation strengths alone, in both the original and rank-transformed data.
+{\color{red}SOME DISCUSSION.....} \\
 
@@
 \section*{Figures}
+
+1. WCN-Rate4site vs Variance of RateSite (Also for RSA) and CN-Rate4site vs Variance of RateSite \\
+2. WCN-Entropy vs Mean Entropy  and CN-Entropy vs Mean Entropy \\
+3. RSA-Entropy vs Mean Entropy \\
+4. RSA vs Rate4Site \\
+5. Should I make the stress  and delta delta G correlations plots for the viral sequences that would complete the story?
+
 %    \begin{figure}[tbh]

### 2015-05-18 manuscript/voronoi/mnv.tex
MN update

@@
 Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability of proteins.
-A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which by its definition represents the combined effects of local packing density and long-range amino acid interactions on sequence variability.
-Here we propose a methodology to segregate and quantify the extent to which long range amino acid interactions determine the general patterns of sequence variability in proteins, independently of local packing density.
-We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse of the volume of the Voronoi cells representing individual sites in protein.
+A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which represents by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability.
+Here we propose a methodology to segregate and quantify the extent to which long-range interactions determine the general patterns of sequence variability in proteins, independently of local packing density.
+We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein.
 
@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
 \\
 
-    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
-It is perceivable that quantities such as B factor, RSA, and CN, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility.
-\citet{franzosa_structural_2009} used a variety of the local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral impact.
+    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
+It is perceivable that quantities such as B factor and RSA, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
+\citet{franzosa_structural_2009} used a variety of local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
 Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns of enzymatic proteins in contrast to RSA and local flexibility measures.
@@
 A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
-An estimate of the accessible volume for each site in protein can be generally obtained through a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
+An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
 In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}.
@@
 A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN.
-There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009}.
+There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $5\AA$ to $18\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014}.
 \\
 
-    In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{lin_deriving_2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
+    In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{lin_deriving_2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, WCN$_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,
 
@@
             \label{eqn:wcn_pwrl}
-            WCN_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}},
+            \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}}.
         \end{equation}
 
-    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel, which is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
-Moreover, no physical model has been so far proposed to support the power-law kernel used in the definition of WCN and the specific value of exponent often used.
+    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel.
+The value of the exponent that results in the best predictions of B factors appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
+A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability.
+The specific value of $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, in addition to local packing density.
+Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on B factor and sequence variability, aside from local packing density.
+In this regard, the two commonly-used proxy measures of packing density may not be considered {\it local}.
+An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, separately from each other.
 \\
 
-    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, their definitions does not involve any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution.
-This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility.
-Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number (CN) and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level.
-\\
+    Motivated by the existing gaps in the current understanding of the role of flexibility, packing density and long-range interactions on sequence-structure relations in proteins, here we derive a new set of site-specific structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions.
+In particular, we propose a methodology to derive a minimally-biased measure of {\it local} packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.
+This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.
 
-    Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
+    %Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution.
 Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility.
@@
 
-    In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+    In addition to site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
 We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
-The {\it ddG rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
-In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
-Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
+
+    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins was also defined and obtained following the stability threshold model of Bloom et al.
+(2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al.
+(2014) \cite{echave_relationship_2014}.
+Based upon this model, the \ddg {\it rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
+In brief, the site-specific quantity, \ddg rate (also denoted by {\it ddG rate} hereafter in figures), is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
+Therefore, a low \ddg rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
 \\
 
-    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins can be defined and obtained following the stability threshold model of Bloom et al.
-(2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al.
-(2014) \cite{echave_relationship_2014}.
-Based upon this model which was extensively described in Chapter \ref{chap:sschar}, a quantity $\Delta\Delta G$ {\it rate} (or {it ddG rate}) was derived for each individual site in all proteins in dataset.
-A high ddG rate for the $i^{th}$ site in a protein indicates a high stability of the site and the overall conformation of the protein to perturbations caused by substitution of the amino acid residing the site.
-\\
-
-    As a measure of local flexibility or fluctuation in different parts of the protein structure the temperature, factors ({\it B factor}) for all atoms in PDB files were extracted (c.f., Chapter \ref{chap:sschar}).
+    As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted.
 Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.
@@
 
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{fractal_dim_max_extent_volume.pdf}
-        \end{center}
-        \caption{The scaling behavior of protein maximum extent as defined by Eqn.
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{fractal_dim_max_extent_volume.pdf}
+%        \end{center}
+%        \caption{The scaling behavior of protein maximum extent as defined by Eqn.
 \ref{eqn:max_extent} with protein volume for $209$ monomeric enzymes in the dataset.
 The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.47\pm0.06$.}
-        \label{fig:maxextent}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{fractal_dim_rad_gyration_nres.pdf}
-        \end{center}
-        \caption{The scaling behavior of protein's radius  as defined by Eqn.
+%        \label{fig:maxextent}
+%    \end{figure}
+%
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{fractal_dim_rad_gyration_nres.pdf}
+%        \end{center}
+%        \caption{The scaling behavior of protein's radius  as defined by Eqn.
 \ref{eqn:rg} with protein length for $209$ monomeric enzymes in the dataset.
@@
 The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.60\pm0.08$.}
-        \label{fig:rg}
-    \end{figure}
+%        \label{fig:rg}
+%    \end{figure}
 

### 2015-05-19 manuscript/voronoi/mnv.tex
MN update

@@
 An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
-In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}.
+In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009},
+
+    \begin{equation}
+        \label{eqn:cn}
+        \text{CN}_{i} = \sum^N_{j\neq i} H\big(r_0-r_{ij}\big),
+    \end{equation}
+
+    \noindent in which $r_{ij}$ represents the distance between sites $i$ \& $j$ and,
+
+    \begin{equation}
+        \label{eqn:heaviside}
+        H\big(r_0-r_{ij}\big) = \int^{r_0-r_{ij}}_{-\infty}\delta\big(x)\operatorname{x},
+    \end{equation}
+
+    \noindent is the Heaviside step function, with $\delta(x)$ standing for the Dirac delta function.
 Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN.
-A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN.
+A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r_0$ in the definition of CN.
 There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $5\AA$ to $18\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014}.
@@
         \label{eqn:wcn_pwrl}
-            \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}}.
+        \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{\alpha=2}}.
     \end{equation}
@@
 
+    \subsection*{Contact Number Definition and Long-Range Amino Acid Interactions}
+    \label{sec:cnwcn}
+
+    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigated the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and amino acid fluctuations (B factor), for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns.
+\ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).
+
+    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcn}.
+
     \subsection*{Voronoi Partitioning of Protein's Structure}
@@
 
-    There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
-The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
+    There is already an extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
+The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
 For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
@@
         \begin{center}
+        \begin{tabular}{cc}
+            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_bfSC.pdf}
+        \end{tabular}
+        \end{center}
+        \caption{}
+        \label{fig:wcnh}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \begin{tabular}{cc}
+            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
+        \end{tabular}
+        \end{center}
+        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN$_\operatorname{P}$) with power-law kernel as given by Eqn.
+\ref{eqn:wcn_pwrl}, with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$.
+The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution.
+The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.
+Note that for $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ for $\alpha<0$.
+In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot.
+The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
+        \label{fig:wcnp}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
         \includegraphics[width=6in]{voronoi_diagram.pdf}

### 2015-05-19 manuscript/voronoi/mnv.tex
MN update

@@
 
-    \subsection*{Contact Number Definition and Long-Range Amino Acid Interactions}
+    \subsection*{Packing Density Definitions and Long-Range Amino Acid Interactions}
     \label{sec:cnwcn}
 
-    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigated the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and amino acid fluctuations (B factor), for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns.
+    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigate the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and the average B factor of the amino acid side-chains, for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns.
 \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).
 
-    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcn}.
+    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcn}.
+The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:cnwcn}.
+Evidently, the best-performing free parameters of both model indicate the non-negligible contributions of long-range amino acid interactions in the strengths of the observed correlations.
+The extent and significance of the influences however, cannot be deciphered solely based on the two common definitions of packing density: CN \& WCN.
+In the following section we present a methodology to segregate and quantify the role of local packing density from long-range interactions in on site-specific sequence variability and flexibility.
 
@@
 
-    \subsection*{Voronoi Cell Area and Volume as Proxy Measures of Local Packing Density and Flexibility in Proteins}
+    \subsection*{Voronoi Cell Volume as Proxy Measure of Local Packing Density and Flexibility in Proteins}
 
@@
     In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
+We show in Appendix \ref{app:xx} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
+\\
+
     Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
 It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
-In general, it is observed that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
+In general, it is observed that the cell surface volume and area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
 \\
 
-    The cell properties are also strongly correlated with each other.
-Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in Figure \ref{fig:voronoi_ER_cond}.
+    The Voronoi cell characteristics are also strongly associated with each other.
+Although the Voronoi cell volume and area are the best correlating variables with ER, none of the two exhibit any significant independent correlation with ER once the other is controlled for control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in Figure \ref{fig:voronoi_ER_cond}.
 Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.

### 2015-05-20 manuscript/voronoi/mnv.tex
MN update + new figures and tables on best definition of WCN

@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
 \\
@@
         \label{eqn:wcn_pwrl}
-        \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{\alpha=2}}.
+        \text{WCN}_{i} = \sum^N_{j\neq i} {r_{ij}^{\alpha=-2}}.
     \end{equation}
@@
 We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.
+\\
 
@@
     As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted.
-Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.
 Alternatively, the site-specific fluctuation could be calculated from MD simulations.
 This was however impossible for this study for the large dataset of $209$ proteins considered in this work.
+Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.
+Here, we use instead the average over all B factors of the side chain atoms if individual amino acids.
+We show in Appendix \ref{app:best_bf} that the average side chain B factor is the best representation of site-specific flexibility among all possible atomic B factors in an amino acid.
 \\
@@
 \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).
+\\
 
-    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcn}.
+    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcnp}.
 The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:cnwcn}.
@@
 
-    \subsection*{Voronoi Cell Volume as Proxy Measure of Local Packing Density and Flexibility in Proteins}
+    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility in Proteins}
 
@@
     In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
-We show in Appendix \ref{app:xx} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
+We show in Appendix \ref{app:best_voronoi} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
 \\
@@
 Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}.
-More recently, Marcos \& Echave (2014)\cite{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
+More recently, Marcos \& Echave (2015) \cite{marcos_too_2015} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
 \\
@@
         \begin{tabular}{cc}
-            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_bfSC.pdf}
+            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_bfSC.pdf}
         \end{tabular}
         \end{center}
-        \caption{}
-        \label{fig:wcnh}
+        \caption{Average absolute Spearman's correlation strengths of Contact Number (CN) -- as defined by Eqn.
+\ref{eqn:wcnh} using Heaviside kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $r_0$.
+The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $r_0$, and the dashed black line indicates the median of the distribution.
+The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.
+The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
+        \label{fig:cn}
     \end{figure}
@@
         \begin{tabular}{cc}
-            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
+            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
         \end{tabular}
         \end{center}
-        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN$_\operatorname{P}$) with power-law kernel as given by Eqn.
-\ref{eqn:wcn_pwrl}, with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$.
+        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN) -- as defined by Eqn.
+\ref{eqn:wcn_pwrl} using a power-law kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$.
 The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution.
@@
 In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot.
-The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
+The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:xx}.}
         \label{fig:wcnp}
@@
 
+\newpage
+
+% Tables
+
+    \begin{table}[htbp]
+    \caption{Best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of WCN with four site-specific quantities (average side-chain B factor, evolutionary rates (r4sJC), sequence entropy, and \ddg rate) for the entire dataset of 209 proteins.
+The corresponding median correlation coefficients ($\rho$) are reported  inside parenthesis next to each parameter value in the table.
+\label{tab:best_params}}
+    \bigskip
+    \centerline{
+    \begin{tabular}{|l|c|c|c|c|}
+      \hline
+      \multirow{3}{*}{Correlation} & \multicolumn{4}{|c|}{Best Performing Free Parameter of the Kernel} \\ \cline{2-5}
+                        & Power-law & Exponential & Gaussian & Cutoff \\
+                        & $\alpha$ & $\lambda~[\AA]$ & $\sigma~[\AA]$ & $r~[\AA]$ \\ \hline
+      WCN--B factor     & -2.2 ($0.70$) & 3.8 ($0.70$) & 9.6 ($0.70$) & 13.0 ($0.69$) \\
+      WCN--r4sJC        & -2.3 ($0.64$) & 3.4 ($0.64$) & 9.8 ($0.63$) & 14.8 ($0.62$) \\
+      WCN--Seq. Entropy & -2.2 ($0.54$) & 3.4 ($0.53$) & 8.0 ($0.53$) & 15.2 ($0.52$) \\
+      WCN--\ddg Rate    & -3.4 ($0.76$) & 2.0 ($0.78$) & 5.4 ($0.79$) & 9.8 ($0.75$) \\
+      \hline
+    \end{tabular}}
+    \end{table}
+
 \end{document}

### 2015-05-20 wcn_best_definition/analysis/src/get_sum_quantiles.r

@@ -1,49 +1,51 @@
-# This R script generates the quantile tables and the corresponding plots for the free parameters of different models of WCN
-# Amir Shahmoradi, Sunday 9:57 PM, March 29, 2015, Wilke Lab, iCMB, UT Austin
+# This R script generates the quartiles of the best free parameters of different models of WCN from the summary files.
+# What this means is this: Each pdb file has a set of best performing free parameters of the specific WCN model, for which WCN correlates the strongest with a given structure or sequence variable (e.g., B factor, r4sJC, or sequence entropy). This code finds the 0%, 5%, 25%, 50%, 75%, 95% best free parameters among the 209 best free parameter values in the summary files.
+
+# Amir Shahmoradi, Tuesday 6:30 PM, May 19, 2015, Wilke Lab, iCMB, UT Austin
 # Note that R script input_data.r must be first sourced in order to use this script.
 
 setwd('C:/Users/Amir/Documents/GitHub/cordiv/wcn_best_definition/analysis/src')
 
-dflist = list( wcneSC_bfSC     = exp_wcneSC_bfSC  
-             , wcneSC_r4sJC    = exp_wcneSC_r4sJC 
-             , wcneSC_seqent   = exp_wcneSC_seqent
-             , wcneSC_ddgent   = exp_wcneSC_ddgent
-             , wcneSC_distance = exp_wcneSC_distance
-             , wcngSC_bfSC     = exp_wcngSC_bfSC  
-             , wcngSC_r4sJC    = exp_wcngSC_r4sJC 
-             , wcngSC_seqent   = exp_wcngSC_seqent
-             , wcngSC_ddgent   = exp_wcngSC_ddgent
-             , wcngSC_distance = exp_wcngSC_distance
-             , wcnhSC_bfSC     = exp_wcnhSC_bfSC  
-             , wcnhSC_r4sJC    = exp_wcnhSC_r4sJC 
-             , wcnhSC_seqent   = exp_wcnhSC_seqent
-             , wcnhSC_ddgent   = exp_wcnhSC_ddgent
-             , wcnhSC_distance = exp_wcnhSC_distance
-             , wcnpSC_bfSC     = exp_wcnpSC_bfSC  
-             , wcnpSC_r4sJC    = exp_wcnpSC_r4sJC 
-             , wcnpSC_seqent   = exp_wcnpSC_seqent
-             , wcnpSC_ddgent   = exp_wcnpSC_ddgent
[…]

### 2015-05-20 manuscript/voronoi/mnv.tex
MN update

@@
 \begin{Abstract}
+
     What are the best structural predictors of protein's sequence evolution?
@@
     In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
-We show in Appendix \ref{app:best_voronoi} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
+We show in Appendix \ref{app:best_crd} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
 \\
@@
 
-    \subsection*{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
-    \label{sec:voronoi_SC}
+    \subsection*{Effects of Long-Range Amino Acid Interactions on Sequence Variability}
 
-    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables.
-For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
-The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins.
-Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}.
-More recently, Marcos \& Echave (2015) \cite{marcos_too_2015} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
+    Throughout the previous sections, a variety of site-specific structural characteristics of proteins were calculated and discussed, using different sets of atomic coordinates representing individual sites in proteins.
+These include the weighted contact number, the Voronoi cell characteristics, and representative site-specific B factor.
+To make comprehensive comparison of all site-specific structural properties of proteins, we also calculate the Relative Solvent Accessibility (RSA) and \ddg rate \citep[e.g.,][]{echave_relationship_2014} of individual amino acids in their sites in proteins.
 \\
 
-    Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \cite{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
-In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra.
-\\
+    Figure \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE} compare the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) and sequence entropy.
+It is notable that once WCN is recalculated using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific sequence variability.
+The better performance of WCN compared to local packing density as measured from Voronoi cell volume and area may not be surprising, since WCN by its definition takes into account the potential long-range interactions among amino acids in different regions of protein.
+Indeed, the fractal dimension of proteins very much resembles that of lattice percolation models \cite{stauffer_introduction_1994} and similarly the random packing of hard spheres near percolation threshold \cite{lorenz_universality_1993, liang_are_2001}.
+To expand on this, define the average maximum extent of a protein as,
 
-    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
-Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties.
-These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains.
-Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
-In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid.
-The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations.
-The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}.
-Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all.
+    % Figure goes here
+
+    % Figure goes here
+
+    \begin{equation}
+        \label{eqn:max_extent}
+        R_m = \frac{1}{2d}\sum_{i=1}^d \big( x_{i,max} - x_{i,min} \big),
+    \end{equation}
+
+    \noindent in which $d=3$ is the dimension of the Euclidean space $x_{i,max} - x_{i,min}$ is the maximum physical extent of the protein, as represented by the geometric center of the side-chain coordinates, in each of the three spatial dimensions.
+Alternatively, the radius of gyration of a protein of length $N$ can be defined as (similar to that of a finite size cluster: Egn.
+45a, Sec.
+3.2, in \cite{stauffer_introduction_1994}),
+
+    \begin{equation}
+        \label{eqn:rg}
+        R_g = \sqrt{\sum_{i=1}^N \frac{|\vec r_i-\vec r_0|^2}{N}},
+    \end{equation}
+
+    \noindent where,
+
+    \begin{equation}
+        \label{eqn:com}
+        \vec r_0 = \sum_{i=1}^N \frac{\vec r_i}{N},
+    \end{equation}
+
+    \noindent is the geometric center of the protein, and $r_i$ is the position of the geometric center of the side-chain of the $i^{th}$ amino acid in protein.
+This definition is such that the kinetic energy and the angular momentum of the protein about the $\vec r_0$ is equivalent to the kinetic energy and the angular momentum of all amino acids residing on a ring of radius $R_g$ centered at $\vec r_0$.
+Figure \ref{fig:maxextent} \& \ref{fig:rg} depict the behavior of the maximum extent ($R_m$) versus protein volume ($V$) and the radius of gyration ($R_g$) versus the protein length ($N$), respectively.
 \\
@@
 
-    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
-Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
-\\
+    The protein volumes and surface areas calculated using {\it 3V} software \cite{voss_3v:_2010}.
+A symmetric linear fit (i.e., Deming regression) to the plot of $\log R_m$ vs.
+$\log V$ and $\log R_g$ vs.
+$\log N$ results in regression slopes of $D\simeq2.47\pm0.06$ \& $D\simeq2.60\pm0.08$ respectively.
+The observed exponents are very similar to those of the scaling relation,
 
-    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells.
-Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}).
-\\
+    \begin{equation}
+        V \propto R^{D}, ~~~~~ (p\simeq p_c)
+    \end {equation}
 
-    It is notable that the standard deviations of the difference distributions for both quantities: WCN, and Voronoi cell area, are an order of magnitude smaller than the observed differences, implying that the correlation coefficients for all proteins in dataset uniformly translate to higher values by moving from C$_\alpha$ atomic coordinates to the geometric centers of the side chains, regardless of the strength of the correlation coefficients.
-\\
+    \noindent with $D\simeq2.51\pm0.01$ in numerical simulations of hard-sphere packing \cite{lorenz_universality_1993} in three dimensions near percolation threshold ($p\simeq p_c$) or the exponents derived from lattice models $D\simeq2.54\pm0.05$ \cite{adler_series_1990} \& $D\simeq2.5$ \cite{stauffer_introduction_1994}.
+It is notable that far from percolation threshold (i.e., $p\rightarrow 0$) in three dimensions, $D=2$ \cite{parisi_critical_1981}, while above the threshold ($p>p_c$), $D=d=3$, in 3D space.
 
-    \subsection*{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+\section{Discussion}
+\label{sec:discussion}
+
+    Throughout the previous sections,  a comprehensive analysis and comparison of the main structural determinants of sequence variability was carried out, using a dataset of $209$ monomeric enzymes.
+Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg rate as defined in Chapter \ref{chap:sschar} (see also Echave et al.
+(2014) \cite{echave_relationship_2014}), which is essentially a proxy measure of the stability of protein's native conformation upon substitution of amino acids in individual sites in proteins.
+In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg rate, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figures \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE}).
+
+\appendix{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
 
@@
 
-    \subsection*{Effects of Long-Range Amino Acid Interactions on Sequence Variability}
+    \appendix{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
+    \label{app:best_crd}
 
-    Throughout the previous sections, a variety of site-specific structural characteristics of proteins were calculated and discussed, using different sets of atomic coordinates representing individual sites in proteins.
-These include the weighted contact number, the Voronoi cell characteristics, and representative site-specific B factor.
-To make comprehensive comparison of all site-specific structural properties of proteins, we also calculate the Relative Solvent Accessibility (RSA) and \ddg rate \citep[e.g.,][]{echave_relationship_2014} of individual amino acids in their sites in proteins.
+    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables.
+For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein.
+The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins.
+Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}.
+More recently, Marcos \& Echave (2015) \cite{marcos_too_2015} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.
 \\
 
-    Figure \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE} compare the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) and sequence entropy.
-It is notable that once WCN is recalculated using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific sequence variability.
-The better performance of WCN compared to local packing density as measured from Voronoi cell volume and area may not be surprising, since WCN by its definition takes into account the potential long-range interactions among amino acids in different regions of protein.
-Indeed, the fractal dimension of proteins very much resembles that of lattice percolation models \cite{stauffer_introduction_1994} and similarly the random packing of hard spheres near percolation threshold \cite{lorenz_universality_1993, liang_are_2001}.
-To expand on this, define the average maximum extent of a protein as,
+    Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \cite{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility.
+In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra.
+\\
 
-    % Figure goes here
-
-    % Figure goes here
-
-    \begin{equation}
-        \label{eqn:max_extent}
-        R_m = \frac{1}{2d}\sum_{i=1}^d \big( x_{i,max} - x_{i,min} \big),
-    \end{equation}
-
-    \noindent in which $d=3$ is the dimension of the Euclidean space $x_{i,max} - x_{i,min}$ is the maximum physical extent of the protein, as represented by the geometric center of the side-chain coordinates, in each of the three spatial dimensions.
-Alternatively, the radius of gyration of a protein of length $N$ can be defined as (similar to that of a finite size cluster: Egn.
-45a, Sec.
-3.2, in \cite{stauffer_introduction_1994}),
-
-    \begin{equation}
-        \label{eqn:rg}
-        R_g = \sqrt{\sum_{i=1}^N \frac{|\vec r_i-\vec r_0|^2}{N}},
-    \end{equation}
-
-    \noindent where,
-
-    \begin{equation}
-        \label{eqn:com}
-        \vec r_0 = \sum_{i=1}^N \frac{\vec r_i}{N},
-    \end{equation}
-
-    \noindent is the geometric center of the protein, and $r_i$ is the position of the geometric center of the side-chain of the $i^{th}$ amino acid in protein.
-This definition is such that the kinetic energy and the angular momentum of the protein about the $\vec r_0$ is equivalent to the kinetic energy and the angular momentum of all amino acids residing on a ring of radius $R_g$ centered at $\vec r_0$.
-Figure \ref{fig:maxextent} \& \ref{fig:rg} depict the behavior of the maximum extent ($R_m$) versus protein volume ($V$) and the radius of gyration ($R_g$) versus the protein length ($N$), respectively.
+    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
+Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties.
+These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
+In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains.
+Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
+In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid.
+The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations.
+The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}.
+Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all.
 \\
@@
 
-    The protein volumes and surface areas calculated using {\it 3V} software \cite{voss_3v:_2010}.
-A symmetric linear fit (i.e., Deming regression) to the plot of $\log R_m$ vs.
-$\log V$ and $\log R_g$ vs.
-$\log N$ results in regression slopes of $D\simeq2.47\pm0.06$ \& $D\simeq2.60\pm0.08$ respectively.
-The observed exponents are very similar to those of the scaling relation,
+    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
+\\
 
-    \begin{equation}
-        V \propto R^{D}, ~~~~~ (p\simeq p_c)
-    \end {equation}
+    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells.
+Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}).
+\\
 
-    \noindent with $D\simeq2.51\pm0.01$ in numerical simulations of hard-sphere packing \cite{lorenz_universality_1993} in three dimensions near percolation threshold ($p\simeq p_c$) or the exponents derived from lattice models $D\simeq2.54\pm0.05$ \cite{adler_series_1990} \& $D\simeq2.5$ \cite{stauffer_introduction_1994}.
-It is notable that far from percolation threshold (i.e., $p\rightarrow 0$) in three dimensions, $D=2$ \cite{parisi_critical_1981}, while above the threshold ($p>p_c$), $D=d=3$, in 3D space.
-
-\section{Discussion}
-\label{sec:discussion}
-
-    Throughout the previous sections,  a comprehensive analysis and comparison of the main structural determinants of sequence variability was carried out, using a dataset of $209$ monomeric enzymes.
-Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg rate as defined in Chapter \ref{chap:sschar} (see also Echave et al.
-(2014) \cite{echave_relationship_2014}), which is essentially a proxy measure of the stability of protein's native conformation upon substitution of amino acids in individual sites in proteins.
-In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg rate, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figures \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE}).
+    It is notable that the standard deviations of the difference distributions for both quantities: WCN, and Voronoi cell area, are an order of magnitude smaller than the observed differences, implying that the correlation coefficients for all proteins in dataset uniformly translate to higher values by moving from C$_\alpha$ atomic coordinates to the geometric centers of the side chains, regardless of the strength of the correlation coefficients.
+\\
 

### 2015-05-21 manuscript/voronoi/mnv.tex
MN updates. Some Figures regenrated

@@
 \usepackage{float}
+\usepackage{rotating}
+\usepackage{url}
+\usepackage{float}
+\usepackage{multirow}
+\usepackage{rotating}
+
 \citestyle{mbe}
@@
 It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
-In general, it is observed that the cell surface volume and area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
+In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
 \\
@@
 
-    Not shown here for brevity, almost identical results to the above are obtained if sequence entropy as defined by Eqn.
-\ref{eqn:shannon} were used in place of sequence evolutionary rates.
+    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy as defined by Eqn.
+\ref{eqn:shannon} used in place of sequence evolutionary rates.
 The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy.
+A potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure.
+We show in appendix \ref{app:edge_effects} that such effects are generally negligible in the results presented in this section.
 \\
-
-    One potential caveat with Voronoi tessellation of finite structures in Euclidean space is the {\it edge effects}.
-Sites that are close to the surface of protein are often associated with Voronoi cells that are bounded by the cubic box containing the protein (Figure \ref{fig:voronoi}).
-Here to ensure that these edge effects do not influence the observed sequence-structure correlations, the open cells -- i.e., cells that are partially bounded and closed by the cubic box containing the protein -- are identified in all proteins by examining the variations in individual cell volumes upon changing the size of the cubic box containing the protein to a given extreme value.
-The open cells in individual proteins are then ranked by the fraction of volume changes observed upon changing the box size and then normalized to the the largest volume observed among closed cells.
-It should be noted that the specific extreme value chosen for the box sizes of the proteins or the rank ordering of the open cells does not have any influence on the resulting correlation strengths, since the Spearman's $\rho$ by its definition is a rank correlation coefficient.
-\\
-
-    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work.
-Similar conclusions are reached if the open cells were alternatively ranked by different criteria such as the fractional changes in cell area (vs.
-cell volume) upon changing the box size.
-The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for edge effects.
-An exception however is cell sphericity as defined in Eqn.
-\ref{eqn:sphericity}, which turns out to behave differently for open and closed cells.
-This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates.
-For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths.
-\\
-
-    % Figure goes here xx
-
-    % Figure goes here xx
 
@@
 
-    Throughout the previous sections, a variety of site-specific structural characteristics of proteins were calculated and discussed, using different sets of atomic coordinates representing individual sites in proteins.
-These include the weighted contact number, the Voronoi cell characteristics, and representative site-specific B factor.
-To make comprehensive comparison of all site-specific structural properties of proteins, we also calculate the Relative Solvent Accessibility (RSA) and \ddg rate \citep[e.g.,][]{echave_relationship_2014} of individual amino acids in their sites in proteins.
+    Figure \ref{fig:best_predictorER} compares the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes.
+Not shown here for brevity, similar results are also obtained for sequence entropy used as the measure of sequence variability.
+For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the geometric center of the side chains as the representative coordinates of individual sites.
+Notably, the quantity WCN outperforms all other structural quantities -- including those derived from Voronoi tessellation -- in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}.
+The better performance of WCN compared to local packing density as measured from Voronoi cell volume may not be surprising, knowing that WCN by its definition (with a power-law weighting function of exponent $\alpha=-2$) takes into account the potential long-range interactions among amino acids in different regions of protein.
 \\
 
-    Figure \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE} compare the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) and sequence entropy.
-It is notable that once WCN is recalculated using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific sequence variability.
-The better performance of WCN compared to local packing density as measured from Voronoi cell volume and area may not be surprising, since WCN by its definition takes into account the potential long-range interactions among amino acids in different regions of protein.
-Indeed, the fractal dimension of proteins very much resembles that of lattice percolation models \cite{stauffer_introduction_1994} and similarly the random packing of hard spheres near percolation threshold \cite{lorenz_universality_1993, liang_are_2001}.
-To expand on this, define the average maximum extent of a protein as,
-
-    % Figure goes here
-
-    % Figure goes here
-
-    \begin{equation}
-        \label{eqn:max_extent}
-        R_m = \frac{1}{2d}\sum_{i=1}^d \big( x_{i,max} - x_{i,min} \big),
-    \end{equation}
-
-    \noindent in which $d=3$ is the dimension of the Euclidean space $x_{i,max} - x_{i,min}$ is the maximum physical extent of the protein, as represented by the geometric center of the side-chain coordinates, in each of the three spatial dimensions.
-Alternatively, the radius of gyration of a protein of length $N$ can be defined as (similar to that of a finite size cluster: Egn.
-45a, Sec.
-3.2, in \cite{stauffer_introduction_1994}),
-
-    \begin{equation}
-        \label{eqn:rg}
-        R_g = \sqrt{\sum_{i=1}^N \frac{|\vec r_i-\vec r_0|^2}{N}},
-    \end{equation}
-
-    \noindent where,
-
-    \begin{equation}
-        \label{eqn:com}
-        \vec r_0 = \sum_{i=1}^N \frac{\vec r_i}{N},
-    \end{equation}
-
-    \noindent is the geometric center of the protein, and $r_i$ is the position of the geometric center of the side-chain of the $i^{th}$ amino acid in protein.
-This definition is such that the kinetic energy and the angular momentum of the protein about the $\vec r_0$ is equivalent to the kinetic energy and the angular momentum of all amino acids residing on a ring of radius $R_g$ centered at $\vec r_0$.
-Figure \ref{fig:maxextent} \& \ref{fig:rg} depict the behavior of the maximum extent ($R_m$) versus protein volume ($V$) and the radius of gyration ($R_g$) versus the protein length ($N$), respectively.
-\\
-
-    % Figure goes here xx
-
-    % Figure goes here xx
-
-    The protein volumes and surface areas calculated using {\it 3V} software \cite{voss_3v:_2010}.
-A symmetric linear fit (i.e., Deming regression) to the plot of $\log R_m$ vs.
-$\log V$ and $\log R_g$ vs.
-$\log N$ results in regression slopes of $D\simeq2.47\pm0.06$ \& $D\simeq2.60\pm0.08$ respectively.
-The observed exponents are very similar to those of the scaling relation,
-
-    \begin{equation}
-        V \propto R^{D}, ~~~~~ (p\simeq p_c)
-    \end {equation}
-
-    \noindent with $D\simeq2.51\pm0.01$ in numerical simulations of hard-sphere packing \cite{lorenz_universality_1993} in three dimensions near percolation threshold ($p\simeq p_c$) or the exponents derived from lattice models $D\simeq2.54\pm0.05$ \cite{adler_series_1990} \& $D\simeq2.5$ \cite{stauffer_introduction_1994}.
-It is notable that far from percolation threshold (i.e., $p\rightarrow 0$) in three dimensions, $D=2$ \cite{parisi_critical_1981}, while above the threshold ($p>p_c$), $D=d=3$, in 3D space.
+    In order to segregate the combined effects of long-range interactions from local packing density in the relationship between WCN and sequence variability (e.g., as measured by r4sJC evolutionary rates), the inverse of Voronoi cell volume can be used as a maximally local measure of packing density.
+By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the the effects long-range amino acid interactions beyond the immediate neighbors -- on sequence variability.
+Figure
 
@@
 
-\appendix{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+\appendix
+
+\section*{Appendix A: Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+\label{app:best_crd}
 
@@
 
-    \appendix{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
+    \section*{Appendix B: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
     \label{app:best_crd}
@@
 \\
+
+\section{Appendix C: Edge-Effects in Voronoi Partitioning of Protein Structures}
+\label{app:edge_effects}
+
+    One potential caveat with Voronoi tessellation of finite structures in Euclidean space is the {\it edge effects}.
+Sites that are close to the surface of protein are often associated with Voronoi cells that are bounded by the cubic box containing the protein (Figure \ref{fig:voronoi}).
+Here to ensure that these edge effects do not influence the observed sequence-structure correlations, the open cells -- i.e., cells that are partially bounded and closed by the cubic box containing the protein -- are identified in all proteins by examining the variations in individual cell volumes upon changing the size of the cubic box containing the protein to a given extreme value.
+The open cells in individual proteins are then ranked by the fraction of volume changes observed upon changing the box size and then normalized to the the largest volume observed among closed cells.
+It should be noted that the specific extreme value chosen for the box sizes of the proteins or the rank ordering of the open cells does not have any influence on the resulting correlation strengths, since the Spearman's $\rho$ by its definition is a rank correlation coefficient.
+\\
+
+    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work.
+Similar conclusions are reached if the open cells were alternatively ranked by other criteria such as the fractional changes in cell area (vs.
+cell volume) upon changing the box size.
+The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for the edge effects.
+An exception however is cell sphericity as defined in Eqn.
+\ref{eqn:sphericity}, which turns out to behave differently for open and closed cells.
+This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates.
+For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths.
+
+    % Figure goes here xx
+
+    % Figure goes here xx
 
@@
         \begin{center}
-        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_voro/zr4s_JC_voro.png}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_ER.pdf}
         \end{center}
-        \caption{General behavior of Voronoi cell characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
-The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
-The black \& orange curves represent respectively the general behaviors of closed \& open Voronoi cell characteristics.
-The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
-        \label{fig:zr4s_JC_voro}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
-        \end{center}
-        \caption{General behavior of site-specific structural characteristics versus site-specific evolutionary rates among all sites in all 209 proteins in dataset.
-The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
-The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
-        \label{fig:zr4s_JC_nonvoro}
+        \caption{A comparison of the prediction power of five structural variables about site-specific evolutionary rates (ER).
+All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively.
+For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure.
+Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$.
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+        \label{fig:best_predictorER}
     \end{figure}
@@
         \begin{center}
-        \includegraphics[width=5.5in]{best_structural_predictors_of_ER.pdf}
+        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_voro/zr4s_JC_voro.png}
         \end{center}
-        \caption{A comparison of the prediction power of five structural variables about site-specific evolutionary rates (ER).
-All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively.
-For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure.
-Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$.
-The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
-        \label{fig:best_predictorER}
+        \caption{General behavior of Voronoi cell characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The black \& orange curves represent respectively the general behaviors of closed \& open Voronoi cell characteristics.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+        \label{fig:zr4s_JC_voro}
     \end{figure}
@@
         \begin{center}
-        \includegraphics[width=5.5in]{best_structural_predictors_of_SE.pdf}
+        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
         \end{center}
-        \caption{A comparison of the prediction power of five structural variables about site-specific Sequence Entropy (SE).
+        \caption{General behavior of site-specific structural characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+        \label{fig:zr4s_JC_nonvoro}
+    \end{figure}
+
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{best_structural_predictors_of_SE.pdf}
+%        \end{center}
+%          \caption{A comparison of the prediction power of five structural variables about site-specific Sequence Entropy (SE).
 All structural quantities correlate positively with SE, with the exception of Weighted Contact Number (WCN) which correlates negatively.
@@
 The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
-        \label{fig:best_predictorSE}
-    \end{figure}
+%        \label{fig:best_predictorSE}
+%    \end{figure}
 
@@
     \begin{table}[htbp]
-    \caption{Best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of WCN with four site-specific quantities (average side-chain B factor, evolutionary rates (r4sJC), sequence entropy, and \ddg rate) for the entire dataset of 209 proteins.
-The corresponding median correlation coefficients ($\rho$) are reported  inside parenthesis next to each parameter value in the table.
+    \caption{Median best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of CN \& WCN with three site-specific quantities (evolutionary rates (r4sJC), sequence entropy, and average side-chain B factor) in the entire dataset of 209 proteins.
+The $25\%$ \& $75\%$ quartile distances in the distribution of parameters from the median are given as superscripts and subscripts to the median values.
 \label{tab:best_params}}
@@
     \centerline{
-    \begin{tabular}{|l|c|c|c|c|}
+    \begin{tabular}{|l|c|c|}
       \hline
-      \multirow{3}{*}{Correlation} & \multicolumn{4}{|c|}{Best Performing Free Parameter of the Kernel} \\ \cline{2-5}
-                        & Power-law & Exponential & Gaussian & Cutoff \\
-                        & $\alpha$ & $\lambda~[\AA]$ & $\sigma~[\AA]$ & $r~[\AA]$ \\ \hline
-      WCN--B factor     & -2.2 ($0.70$) & 3.8 ($0.70$) & 9.6 ($0.70$) & 13.0 ($0.69$) \\
-      WCN--r4sJC        & -2.3 ($0.64$) & 3.4 ($0.64$) & 9.8 ($0.63$) & 14.8 ($0.62$) \\
-      WCN--Seq. Entropy & -2.2 ($0.54$) & 3.4 ($0.53$) & 8.0 ($0.53$) & 15.2 ($0.52$) \\
-      WCN--\ddg Rate    & -3.4 ($0.76$) & 2.0 ($0.78$) & 5.4 ($0.79$) & 9.8 ($0.75$) \\
+      Correlation with  & $r_0[\AA]$ (CN)      & $\alpha$ (WCN)        \\ \hline
+      r4sJC             & $14.3^{+5.3}_{-4.0}$ & $-2.3^{+0.8}_{-0.4}$  \\
+      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ & $-2.2^{+0.8}_{-0.4}$  \\
+      B factor          & $13.1^{+4.0}_{-2.2}$ & $-2.2^{+0.5}_{-0.4}$  \\
       \hline

### 2015-05-21 analysis/src/best_structural_predictors_of_ER_given_VSCvolume.r

@@ -1,4 +1,4 @@
-# This code is aimed at finding which structural property best predicts the Evolutionary Rates.
+# This R script is very similar to best_structural_predictors_of_ER.r , with the exception that here the PARTIAL correlations are measured while controlling for Voronoi cell volume.
 
 # Last updated by Amir Shahmoradi, Friday 4:25 PM, February 13 2015, Wilke Lab, ICMB, UT Austin
 

### 2015-05-21 analysis/src/best_voronoi_predictors_of_ER.r

@@ -51,8 +51,20 @@ for(pdb in levels(res_prop_elj$pdb))
   #pcor.test( pdb_temp$zr4s_JC, pdb_temp$VSCeccentricity, pdb_temp$VSCsphericity, method='sp')
   #row = data.frame( pdb = pdb, rsa = r.rsa.r4sJC, wcn = r.wcn.r4sJC, vareaSC = r.vareaSC.r4sJC, ddgent = r.ddgent.r4sJC )
   #row = data.frame( pdb, r.nvertices.r4sJC, r.nedges.r4sJC, r.nfaces.r4sJC, r.edgelength.r4sJC, r.area.r4sJC , r.volume.r4sJC, r.eccentricity.r4sJC, r.sphericity.r4sJC )
+  
+  # You know what? Also test for partial correlation of cell variables with ER, controlling for cell volume (needed for the manuscript):
+  x = pcor.test( pdb_temp$zr4s_JC, pdb_temp$VSCedge, pdb_temp$VSCvolume, method='sp')
+  r.edg_given_volume = x$estimate
+  x = pcor.test( pdb_temp$zr4s_JC, pdb_temp$VSCarea, pdb_temp$VSCvolume, method='sp')
+  r.area_given_volume = x$estimate
+  x = pcor.test( pdb_temp$zr4s_JC, pdb_temp$VSCeccentricity, pdb_temp$VSCvolume, method='sp')
+  r.ecc_given_volume = x$estimate
+  x = pcor.test( pdb_temp$zr4s_JC, pdb_temp$VSCsphericity, pdb_temp$VSCvolume, method='sp')
+  r.sph_given_volume = x$estimate
+  
   row = data.frame( pdb, edge, area , volume, eccentricity, sphericity
-                  , r.edg_given_area, r.vol_given_area, r.ecc_given_area, r.sph_given_area )
+                  , r.edg_given_area, r.vol_given_area, r.ecc_given_area, r.sph_given_area
+                  , r.edg_given_volume, r.area_given_volume, r.ecc_given_volume, r.sph_given_volume )
   best_voronoi_predictors_of_ER = rbind( best_voronoi_predictors_of_ER, row )
 }
 
@@ -85,6 +97,10 @@ hist.r.edg_given_area = density(best_voronoi_predictors_of_ER$r.edg_given_area)
 hist.r.vol_given_area = density(best_voronoi_predictors_of_ER$r.vol_given_area)
[…]

### 2015-05-27 manuscript/voronoi/mnv.tex
Further MN update, + new tables and figures

@@
 \usepackage{rotating}
+\usepackage[titletoc,title]{appendix}
 
@@
 \vfill
-\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
+\Author{Amir Shahmoradi$^{1,2}$, Claus O. Wilke$^2$}
 \Address{\affiliation}
@@
 
-    What are the best structural predictors of protein's sequence evolution?
+    What are the best structural determinants of protein sequence evolution?
 A number of site-specific structural characteristics have been proposed over the past decade to answer this question.
-Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability of proteins.
-A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which represents by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability.
+Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability in proteins.
+The most commonly used measures of local packing density such as the Contact Number and the Weighted Contact Number represent by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability.
 Here we propose a methodology to segregate and quantify the extent to which long-range interactions determine the general patterns of sequence variability in proteins, independently of local packing density.
 We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein.
+Using a dataset of 209 monomeric enzymes, we show that the long-range amino acid interactions can explain $\sim000\%$ of variability observed in protein sequence, whereas only $\sim000\%$ of sequence variability can be attributed to site-specific packing density.
 
@@
 
-    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently.
+    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each characteristics influences the sequence evolution independently of the others.
 It is perceivable that quantities such as B factor and RSA, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
 \citet{franzosa_structural_2009} used a variety of local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
-Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns of enzymatic proteins in contrast to RSA and local flexibility measures.
+Recently however, \citet{huang_mechanistic_2014} argued through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns among enzymatic proteins in contrast to RSA and local flexibility measures.
 \\
 
-    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein.
+    It is notable that site-specific flexibility is often represented by the $C_\alpha$ atomic B factors, a quantity that is not necessarily an unbiased measure of the amino acid flexibility in a given site in protein.
 A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
-An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
-In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009},
+An estimate of the accessible volume to each site, and therefore a proxy measure of site-specific flexibility in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
+In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around the site \citep[e.g.,][]{franzosa_structural_2009},
 
@@
 
-    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel.
+    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable parameter: the exponent of the power-law kernel ($\alpha$).
 The value of the exponent that results in the best predictions of B factors appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
@@
 The specific value of $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, in addition to local packing density.
-Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on B factor and sequence variability, aside from local packing density.
+Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009}, aside from local packing density.
 In this regard, the two commonly-used proxy measures of packing density may not be considered {\it local}.
@@
     The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes \citep[e.g.,][]{yeh_site-specific_2014, echave_relationship_2014} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$ amino acids, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}.
-To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citet{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}.
+To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected \citep{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}.
 Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep[][]{katoh_mafft_2005}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data.
@@
     A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins was also defined and obtained following the stability threshold model of Bloom et al.
-(2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al.
-(2014) \cite{echave_relationship_2014}.
-Based upon this model, the \ddg {\it rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}.
-In brief, the site-specific quantity, \ddg rate (also denoted by {\it ddG rate} hereafter in figures), is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
-Therefore, a low \ddg rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
+(2006) \cite{bloom_structural_2006}.
+Based upon this model, the \ddg {\it rate} estimates for all structures in the dataset were calculated using FoldX program and the methodology proposed by \citet{echave_relationship_2014}.
+In brief, the site-specific quantity \ddg rate (denoted by {\it ddG rate} hereafter in figures), is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids.
+Therefore, a low \ddg rate for a given site would imply a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.
 \\
@@
 Alternatively, the site-specific fluctuation could be calculated from MD simulations.
-This was however impossible for this study for the large dataset of $209$ proteins considered in this work.
+This was however impossible in this study for the large dataset of $209$ proteins considered in this work.
 Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.
-Here, we use instead the average over all B factors of the side chain atoms if individual amino acids.
+Here, we use instead the average over all B factors of the side chain atoms of individual amino acids.
 We show in Appendix \ref{app:best_bf} that the average side chain B factor is the best representation of site-specific flexibility among all possible atomic B factors in an amino acid.
@@
 
-    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcnp}.
-The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:cnwcn}.
-Evidently, the best-performing free parameters of both model indicate the non-negligible contributions of long-range amino acid interactions in the strengths of the observed correlations.
-The extent and significance of the influences however, cannot be deciphered solely based on the two common definitions of packing density: CN \& WCN.
-In the following section we present a methodology to segregate and quantify the role of local packing density from long-range interactions in on site-specific sequence variability and flexibility.
+    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp}.
+The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:best_params}.
+Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of long-range amino acid interactions to the strengths of the observed correlations.
+The extent and significance of the influences however, cannot be deciphered solely from the two common definitions of packing density: CN \& WCN.
+In the following section we present a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability and flexibility measures.
 
@@
 
-    There is already an extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
+    There is already an extensive body of literature on applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
 The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
@@
 
-    In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation.
+    In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of proteins.
 An example of Voronoi tessellation of protein structure in two dimensions (PDB ID: {\it 1LBA}) is shown in Figure \ref{fig:voronoi}.
-The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions.
+The properties of individual cells resulting from tessellation can be then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions.
 \\
 
-    Here in this work, the simplest and most widely used definition of Voronoi tessellation described above is applied on a dataset of $209$  monomeric enzymes.
+    Here in this work, the simplest and most widely used definition of Voronoi tessellation is applied on the dataset of $209$  monomeric enzymes.
 We use VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
-Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
-A measure of the cell {\it eccentricity} can be also obtained by finding the distance between the cell seed and geometrical center of the cell.
+Among the most important properties are the length of the cell edges, cell surface area and volume, number of faces of each cell, and the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
 In addition, the cell {\it sphericity} can be calculated as a measure of the cell's {\it compactness} defined as,
@@
 
-    \noindent in which $V$ \& $A$ stand for the volume \& area of the cell respectively.
+    \noindent in which $V$ \& $A$ stand for the volume \& area of the cell, respectively.
 For a perfectly spherical cell, $\Psi=1$, while it becomes zero for a 2-dimensional object that has no volume but only surface area.
@@
 
-    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
-We show in Appendix \ref{app:best_crd} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
+    In order to assess the prediction power of the site-specific characteristics derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
+We show in Appendix \ref{app:best_crd} that the choice of the geometric centers of the side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
 \\
@@
     Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER).
-It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties.
-In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy.
+It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and with other Voronoi cell properties.
+In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity, the total edge length, and the cell sphericitiy.
 \\
@@
     The Voronoi cell characteristics are also strongly associated with each other.
-Although the Voronoi cell volume and area are the best correlating variables with ER, none of the two exhibit any significant independent correlation with ER once the other is controlled for control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in Figure \ref{fig:voronoi_ER_cond}.
+Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for.
+The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (c.f., Figure \ref{fig:voronoi_ER_cond}).
 Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
-In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell.
+In conclusion, the cell area, volume, and edge length appear to represent almost the same property of the Voronoi cell.
 Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates.
-These cell characteristics are however, discrete (integer) quantities and in general have a limited range.
+However, these cell characteristics are discrete (integer) quantities and in general have a limited range.
 \\
 
-    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy as defined by Eqn.
-\ref{eqn:shannon} used in place of sequence evolutionary rates.
+    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy used in place of evolutionary rates, as defined by Eqn.
+\ref{eqn:shannon}.
 The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy.
-A potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure.
-We show in appendix \ref{app:edge_effects} that such effects are generally negligible in the results presented in this section.
+One potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure.
+We show in appendix \ref{app:edge_effects} that such effects are generally negligible in our results presented in this section.
 \\
 
-    \subsection*{Effects of Long-Range Amino Acid Interactions on Sequence Variability}
-
-    Figure \ref{fig:best_predictorER} compares the predictive power of each of site-specific structural quantities about sequence variability as measured by evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes.
-Not shown here for brevity, similar results are also obtained for sequence entropy used as the measure of sequence variability.
-For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the geometric center of the side chains as the representative coordinates of individual sites.
-Notably, the quantity WCN outperforms all other structural quantities -- including those derived from Voronoi tessellation -- in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}.
-The better performance of WCN compared to local packing density as measured from Voronoi cell volume may not be surprising, knowing that WCN by its definition (with a power-law weighting function of exponent $\alpha=-2$) takes into account the potential long-range interactions among amino acids in different regions of protein.
+    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Variability}
+    \label{sec:long_range}
+    Figure \ref{fig:best_predictorER} compares the prediction power of each of the site-specific structural quantities about sequence evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes.
+Not shown here for brevity, similar results are also obtained for sequence entropy as the measure of sequence variability.
+For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains.
+Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}.
+The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn.
+\ref{eqn:wcn_pwrl} also takes into account the potential long-range interactions among amino acids in different regions of protein.
 \\
 
-    In order to segregate the combined effects of long-range interactions from local packing density in the relationship between WCN and sequence variability (e.g., as measured by r4sJC evolutionary rates), the inverse of Voronoi cell volume can be used as a maximally local measure of packing density.
-By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the the effects long-range amino acid interactions beyond the immediate neighbors -- on sequence variability.
-Figure
+    In order to segregate the combined effects of long-range interactions from local packing density, the inverse of the Voronoi cell volume can be used as a maximally local measure of packing density.
+By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the effects long-range amino acid interactions beyond immediate neighbors -- on sequence variability.
+Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume.
+It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a rank correlation coefficient.
 
@@
 
-\appendix
+\begin{appendices}
 
-\section*{Appendix A: Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
-\label{app:best_crd}
+\section{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+\label{app:best_bf}
 
@@
 C$_\alpha$ B factor) with other structural properties also merit further attention.
-It was discussed in Section \ref{sec:voronoi_SC} and depicted in the plots of Figure \ref{fig:best_bf} that in general, as one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve.
+It was discussed in Section \ref{sec:voro} and depicted in the plots of Figure \ref{fig:best_bf} that in general, as one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve.
 In particular, the use of average side-chain B factor turned out to result in the highest correlation strengths with other site-specific properties, implying that this average B factor is likely the best representation of the overall amino acid fluctuations and flexibility in a given site in protein.
@@
 
-    \section*{Appendix B: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
+\section{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
 \label{app:best_crd}
@@
 
-\section{Appendix C: Edge-Effects in Voronoi Partitioning of Protein Structures}
+\section{Edge-Effects in Voronoi Partitioning of Protein Structures}
 \label{app:edge_effects}
@@
 
+\end{appendices}
+
 \Acknowledgements
@@
         \begin{tabular}{cc}
-            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_bfSC.pdf}
+            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf}
         \end{tabular}
         \end{center}
-        \caption{Average absolute Spearman's correlation strengths of Contact Number (CN) -- as defined by Eqn.
-\ref{eqn:wcnh} using Heaviside kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $r_0$.
-The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $r_0$, and the dashed black line indicates the median of the distribution.
-The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.
+        \caption{Average absolute Spearman's correlation strengths of Contact Number (CN, as defined by Eqn.
+\ref{eqn:wcnh} using Heaviside kernel) and the Weighted Contact Number (WCN, as defined by Eqn.
+\ref{eqn:wcn_pwrl} using a power-law kernel) with site-specific evolutionary rates, for different values of the free parameters of the two kernels ($r_0$ \& $\alpha$ respectively).
+On each plot, the solid black lines represent the mean correlation strength in the entire dataset of 209 proteins at each value of the free parameter, and the dashed black lines indicate the median of the distribution.
+The green-shaded region together with the red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.
+Note that for the case of WCN with $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ with $\alpha<0$.
+In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot.
 The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
-        \label{fig:cn}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \begin{tabular}{cc}
-            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
-        \end{tabular}
-        \end{center}
-        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN) -- as defined by Eqn.
-\ref{eqn:wcn_pwrl} using a power-law kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$.
-The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution.
-The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.
-Note that for $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ for $\alpha<0$.
-In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot.
-The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:xx}.}
-        \label{fig:wcnp}
+        \label{fig:cnwcnp}
     \end{figure}
@@
         \label{fig:best_predictorER}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_given_VSCvolume.pdf}
+        \end{center}
+        \caption{A comparison of the prediction power of four structural variables (as in Figure \ref{fig:best_predictorER}) about site-specific evolutionary rates (ER), while controlling for the voronoi cell volume.
+All structural quantities correlate positively with ER on average, with the exception of Weighted Contact Number (WCN) which correlates negatively.
+For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure.
+Note that the Spearman's $\rho$ is insensitive to the use of inverse WCN in place of WCN.
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+        \label{fig:best_predictorER_given_volume}
     \end{figure}
@@
     \begin{table}[htbp]
-    \caption{Median best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of CN \& WCN with three site-specific quantities (evolutionary rates (r4sJC), sequence entropy, and average side-chain B factor) in the entire dataset of 209 proteins.
-The $25\%$ \& $75\%$ quartile distances in the distribution of parameters from the median are given as superscripts and subscripts to the median values.
+    \caption{Median best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of CN \& WCN with site-specific sequence variability measures (evolutionary rates (r4sJC) and sequence entropy) in the entire dataset of 209 proteins.
+Given in parentheses are the corresponding median Spearman correlation coefficients at the best parameter values.
+The subscripts and superscripts to each value represent the $25\%$ percentile range below and above the median value of the distribution.
 \label{tab:best_params}}
@@
       Correlation with  & $r_0[\AA]$ (CN)      & $\alpha$ (WCN)        \\ \hline
-      r4sJC             & $14.3^{+5.3}_{-4.0}$ & $-2.3^{+0.8}_{-0.4}$  \\
-      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ & $-2.2^{+0.8}_{-0.4}$  \\
-      B factor          & $13.1^{+4.0}_{-2.2}$ & $-2.2^{+0.5}_{-0.4}$  \\
+      r4sJC             & $14.3^{+5.3}_{-4.0}$ ($\rho\sim0.64^{+0.06}_{-0.06}$) & $-2.3^{+0.8}_{-0.4}$ ($\rho\sim0.65^{+0.05}_{-0.07}$)  \\
+      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ ($\rho\sim0.55^{+0.06}_{-0.06}$) & $-2.2^{+0.8}_{-0.4}$ ($\rho\sim0.55^{+0.07}_{-0.06}$)  \\
+      %B factor          & $13.1^{+4.0}_{-2.2}$ ($\rho\sim0.70^{+0.05}_{-0.06}$) & $-2.2^{+0.5}_{-0.4}$ ($\rho\sim0.71^{+0.05}_{-0.06}$)  \\
+      \hline
+    \end{tabular}}
+    \end{table}
+
+    \begin{table}[htbp]
+    \caption{The proportion of variance explained by local packing density (as measured by the inverse volume of Voronoi cells) and long-range amino acid interactions (as described in Sec.
+\ref{sec:long_range}, for site-specific sequence variability measures.
+The subscripts and superscripts to each value represent the $25\%$ percentile range below and above the median value of the distribution.
+\label{tab:best_params}}
+    \bigskip
+    \centerline{
+    \begin{tabular}{|l|c|c|}
+      \hline
+      Correlation with  & local packing density & long-range interactions   \\ \hline
+      r4sJC             & $14.3^{+5.3}_{-4.0}$ ($\rho\sim0.64^{+0.06}_{-0.06}$) & $-2.3^{+0.8}_{-0.4}$ ($\rho\sim0.65^{+0.05}_{-0.07}$)  \\
+      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ ($\rho\sim0.55^{+0.06}_{-0.06}$) & $-2.2^{+0.8}_{-0.4}$ ($\rho\sim0.55^{+0.07}_{-0.06}$)  \\
+      %B factor          & $13.1^{+4.0}_{-2.2}$ ($\rho\sim0.70^{+0.05}_{-0.06}$) & $-2.2^{+0.5}_{-0.4}$ ($\rho\sim0.71^{+0.05}_{-0.06}$)  \\
       \hline

### 2015-05-27 analysis/src/best_structural_predictors_of_ER_given_VSCvolume.r

@@ -154,8 +154,28 @@ legend( 'topleft'
 graphics.off()
 
 
+# Now summarize the partial correlations data over the entire dataset:
 
+best_structural_predictors_of_ER_given_VSCvolume = read.csv(file = "../tables/best_structural_predictors_of_ER_given_VSCvolume.csv", header=TRUE )
 
+data_subset = best_structural_predictors_of_ER_given_VSCvolume[,-1]  # remove pdb column from the data set to loop over the columns below
+best_structural_predictors_of_ER_given_VSCvol_summary = data.frame()
+for (correlation in names(data_subset))
+{
+  row = data.frame( correlation = correlation
+                  , mean = mean(best_structural_predictors_of_ER_given_VSCvolume[[correlation]])
+                  , median = median(best_structural_predictors_of_ER_given_VSCvolume[[correlation]])
+                  , min = min(best_structural_predictors_of_ER_given_VSCvolume[[correlation]])
+                  , quantile05 = quantile(best_structural_predictors_of_ER_given_VSCvolume[[correlation]], probs=0.05)
+                  , quantile25 = quantile(best_structural_predictors_of_ER_given_VSCvolume[[correlation]], probs=0.25)
+                  , quantile75 = quantile(best_structural_predictors_of_ER_given_VSCvolume[[correlation]], probs=0.75)
+                  , quantile95 = quantile(best_structural_predictors_of_ER_given_VSCvolume[[correlation]], probs=0.95)
+                  , max = max(best_structural_predictors_of_ER_given_VSCvolume[[correlation]])
+                  , sd = sd(best_structural_predictors_of_ER_given_VSCvolume[[correlation]])
+                  )
+  best_structural_predictors_of_ER_given_VSCvol_summary = rbind(best_structural_predictors_of_ER_given_VSCvol_summary,row)  
+}
+write.csv(best_structural_predictors_of_ER_given_VSCvol_summary, file = "../tables/best_structural_predictors_of_ER_given_VSCvol_summary.csv", row.names=F )
 
 max(best_structural_predictors_of_ER_given_VSCvolume$r.wcnSC.r4sJC)-min(best_structural_predictors_of_ER_given_VSCvolume$r.wcnSC.r4sJC)
 max(best_structural_predictors_of_ER_given_VSCvolume$r.vareaSC.r4sJC)-min(best_structural_predictors_of_ER_given_VSCvolume$r.vareaSC.r4sJC)

### 2015-05-27 wcn_best_definition/analysis/src/get_sum_quantiles.r

@@ -46,6 +46,7 @@ dflist = list( wcneSC_bfSC     = sum_wcneSC_bfSC
 
 counter = 0
 quantiles = data.frame()
+quantiles_rho = data.frame()
 for (dataframe in dflist)
 {
   counter = counter + 1
@@ -65,9 +66,24 @@ for (dataframe in dflist)
                     max_best_param        = max(as.vector(dataframe$free_param_best))
                     )
   quantiles = rbind(quantiles,row)
+  # Now get the quantiles for the correlation coefficients corresponding to the best parametr values
+  row_rho = data.frame(model      = names(dflist)[[counter]],
+                   mean_rho       = mean(abs(dataframe$sp_cor_best)),
+                   median_rho     = quantile(as.vector(abs(dataframe$sp_cor_best)), probs = 0.50),
+                   min_rho        = min(as.vector(abs(dataframe$sp_cor_best))),
+                   quantile05_rho = quantile(as.vector(abs(dataframe$sp_cor_best)), probs = 0.05),
+                   quantile25_rho = quantile(as.vector(abs(dataframe$sp_cor_best)), probs = 0.25),
+                   quantile75_rho = quantile(as.vector(abs(dataframe$sp_cor_best)), probs = 0.75),
+                   quantile95_rho = quantile(as.vector(abs(dataframe$sp_cor_best)), probs = 0.95),
+                   stdev_rho      = sd(as.vector(dataframe$sp_cor_best)),
+                   max_rho        = max(as.vector(dataframe$sp_cor_best))
+                   )
+  quantiles_rho = rbind(quantiles_rho,row_rho)
 }
 rownames(quantiles) = NULL
-write.csv( quantiles, file = paste0('../tables/get_quantiles/SPbest_quantiles.csv'), row.names=F)
+rownames(quantiles_rho) = NULL
+write.csv( quantiles, file = paste0('../tables/get_quantiles/SPbest_param_quantiles.csv'), row.names=F)
+write.csv( quantiles_rho, file = paste0('../tables/get_quantiles/SPbest_rho_quantiles.csv'), row.names=F)
 
   # Now generate the plot:
   model = substr(names(dflist)[[counter]],start=4,stop=4)

### 2015-06-01 manuscript/voronoi/mnv.tex
MN update

@@
 
-    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each characteristics influences the sequence evolution independently of the others.
-It is perceivable that quantities such as B factor and RSA, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids.
+    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each has a unique independent influence on the sequence variability patterns in proteins.
+It is perceivable that quantities such as B factor and RSA, all serve as different proxy measures of local packing density of individual sites in proteins.
 \citet{franzosa_structural_2009} used a variety of local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence.
@@
 
-    It is notable that site-specific flexibility is often represented by the $C_\alpha$ atomic B factors, a quantity that is not necessarily an unbiased measure of the amino acid flexibility in a given site in protein.
-A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure.
-An estimate of the accessible volume to each site, and therefore a proxy measure of site-specific flexibility in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}.
+    It is notable that \citet{halle_flexibility_2002} laid a theoretical foundation for the existence of a inverse relationship between the site-specific flexibility -- often represented by the $C_\alpha$ atomic B factors -- and the local packing density in proteins.
+Such an inverse relationship could be intuitively understood knowing that the inverse of local packing density can serve as an estimate of the accessible free volume to each site in the protein's 3-dimensional structure, and thus a proxy measure of site-specific flexibility in protein.
+A crude measure of local packing density for each site can be obtained through a quantity widely known as Contact Number \citep[e.g.,][]{liao_protein_2005}.
 In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around the site \citep[e.g.,][]{franzosa_structural_2009},
@@
 A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability.
-The specific value of $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, in addition to local packing density.
-Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009}, aside from local packing density.
-In this regard, the two commonly-used proxy measures of packing density may not be considered {\it local}.
-An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, separately from each other.
+The specific value $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, independently of the local packing density.
+Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009, shahmoradi_predicting_2014}, aside from local packing density.
+In this regard, the two commonly-used definitions of packing density (CN \& WCN) may not be considered {\it local}.
+An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, independently of each other.
 \\
@@
     Motivated by the existing gaps in the current understanding of the role of flexibility, packing density and long-range interactions on sequence-structure relations in proteins, here we derive a new set of site-specific structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions.
-In particular, we propose a methodology to derive a minimally-biased measure of {\it local} packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.
+In particular, we propose a methodology to derive a minimally-biased measure of local packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.
 This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.
@@
 
-\section{Protein Dataset and Structure$/$Sequence Variability Measures}
+\section{Protein Dataset and Site-Specific Structure$/$Sequence Variability Measures}
 \label{sec:methods}
@@
 
-    In addition to site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
+    In addition to site-specific evolutionary rates, we also calculate the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:
 
@@
 
-    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins was also defined and obtained following the stability threshold model of Bloom et al.
+    %A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins was also defined and obtained following the stability threshold model of Bloom et al.
 (2006) \cite{bloom_structural_2006}.
@@
 
-    As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted.
+    %As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted.
 Alternatively, the site-specific fluctuation could be calculated from MD simulations.
@@
 
-    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigate the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and the average B factor of the amino acid side-chains, for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns.
+    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability in proteins, first we investigate the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and sequence entropy for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30,30]$ as in Eqns.
 \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).
@@
 
-    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp}.
-The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:best_params}.
-Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of long-range amino acid interactions to the strengths of the observed correlations.
-The extent and significance of the influences however, cannot be deciphered solely from the two common definitions of packing density: CN \& WCN.
-In the following section we present a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability and flexibility measures.
+    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}.
+Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations.
+The significance and impact of these non-local interactions however, cannot be deciphered solely from the two common definitions of packing density: CN \& WCN.
+In the following section we present a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.
 
@@
 The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
-For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}).
+For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points.
 \\
@@
 An example of Voronoi tessellation of protein structure in two dimensions (PDB ID: {\it 1LBA}) is shown in Figure \ref{fig:voronoi}.
-The properties of individual cells resulting from tessellation can be then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions.
+The properties of individual cells resulting from tessellation can be then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions, also about sequence evolution as will be shown in the following section.
 \\
 
-    Here in this work, the simplest and most widely used definition of Voronoi tessellation is applied on the dataset of $209$  monomeric enzymes.
+    Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation on the dataset of $209$  monomeric enzymes.
 We use VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
-Among the most important properties are the length of the cell edges, cell surface area and volume, number of faces of each cell, and the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
+Among the most important properties are the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell.
 In addition, the cell {\it sphericity} can be calculated as a measure of the cell's {\it compactness} defined as,

### 2015-06-03 manuscript/voronoi/mnv.tex
MN update + new figures

@@
     The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}.
+(XX a comparison of CN cutoff distance with average amino acid neighbor distance is needed here to quantify evidence for the following sentence.
+An additional figure may also be necessary plotting cutoff distance vs.
+average NN distance for each protein XX).
 Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations.
-The significance and impact of these non-local interactions however, cannot be deciphered solely from the two common definitions of packing density: CN \& WCN.
-In the following section we present a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.
+The significance and impact of these non-local interactions on sequence evolution cannot be deciphered solely from the two common definitions of packing density: CN \& WCN.
+Therefore, we present, in the following section, a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.
 
@@
 
-    There is already an extensive body of literature on applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
+    There is already an extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}.
 The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}.
-For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points.
+For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the Euclidean space into regions, called {\it cells}, such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points.
 \\
@@
     In order to assess the prediction power of the site-specific characteristics derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra.
-We show in Appendix \ref{app:best_crd} that the choice of the geometric centers of the side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.
+We show in Appendix \ref{app:best_crd} that the choice of the geometric centers of the side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in strongest correlations of Voronoi cell properties with site-specific sequence variability of proteins.
 \\
@@
 It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and with other Voronoi cell properties.
-In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity, the total edge length, and the cell sphericitiy.
+In general, it is observed that the cell volume and surface area have the best predictive power compared to other cell characteristics, followed by the cell eccentricity, total edge length, and the cell's sphericitiy.
 \\
@@
 Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates.
-However, these cell characteristics are discrete (integer) quantities and in general have a limited range.
+However, these cell characteristics are discrete (integer) quantities and in general have limited ranges.
 \\
@@
 \ref{eqn:shannon}.
-The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy.
-One potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure.
+The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range that is inherent in the definition of sequence entropy.
+One potential caveat with the Voronoi tessellation of finite structures, such as proteins, is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure.
 We show in appendix \ref{app:edge_effects} that such effects are generally negligible in our results presented in this section.
@@
 
-    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Variability}
+    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Evolution}
     \label{sec:long_range}
@@
 Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume.
-It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a rank correlation coefficient.
+It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a non-parametric rank correlation coefficient.
 
@@
         \begin{center}
-        \includegraphics[width=5.5in]{best_structural_predictors_of_ER.pdf}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_limited.pdf}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_given_VSCvolume.pdf}
+        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_given_vol.pdf}
         \end{center}

### 2015-06-03 analysis/src/best_structural_predictors_of_ER.r

@@ -155,6 +155,59 @@ legend( 'topleft'
 graphics.off()
 
 
+### The following plot is specifically generated for the manuscript:
+# Now plot histograms in a single plot
+#colors = c('green', 'blue', 'red', 'black', 'gray', 'cyan2')
+pdf( "../figures/best_structural_predictors_of_ER_limited.pdf", width=4.5, height=4, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.05, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(  hist.vvolumeSC$x
+       ,  hist.vvolumeSC$y
+       ,   col = 'red'
+       #,   xlim = c(-0.5,0.85)
+       ,   xlim = c(0.1,0.9)
+       ,   ylim = c(0,5.5)
+       #,   col=colors[1]
+       #,   ylim=c(0,7)
+       #,   border = colors[1]
+       #,   lty = 0
+       ,   type = 'l'
+       ,   lwd  = 2 
+       #,   main = 'Correlations with Evolutionary Rates'
+       #,   xlab = expression(paste('Spearman Cor. with Evolutionary Rates ',rho))
+       ,   xlab = expression( paste('Correlation with Evolutionary Rates: Spearman ', rho ) )
+       ,   ylab = 'Relative Frequency'
+)
+lines( -hist.wcnSC$x
+       , hist.wcnSC$y
+       , col = 'black'
+       , lwd = 2
+)
+lines( -hist.wcnCA$x
+       , hist.wcnCA$y
+       , col = 'black'
+       , lwd = 2
+       , lty = 2
+)
+lines( hist.rsa$x
+       , hist.rsa$y
+       , col = 'blue'
+       , lwd  = 2
+)
+legend( 'topleft'
+        , c("Voronoi Cell Volume (SC)","1 / WCN (SC)", "1 / WCN (CA)", "RSA")
+        , col = c('red','black','black','blue')
+        , lty = c(1,1,2,1)
+        , lwd = 2
+        , bty = 'n'
+        , cex = 0.9
+)
+
+graphics.off()
+
+
+
+
+
 
 
 

### 2015-06-03 analysis/src/best_structural_predictors_of_ER_given_VSCvolume.r

@@ -182,6 +182,59 @@ max(best_structural_predictors_of_ER_given_VSCvolume$r.vareaSC.r4sJC)-min(best_s
 sd(best_structural_predictors_of_ER_given_VSCvolume$r.vareaSC.r4sJC)
 sd(best_structural_predictors_of_ER_given_VSCvolume$r.wcnSC.r4sJC)
 
+
+
+### The following plot is specifically generated for the manuscript:
+
+# Now plot histograms in a single plot
+#colors = c('green', 'blue', 'red', 'black', 'gray', 'cyan2')
+pdf( "../figures/best_structural_predictors_of_ER_given_vol.pdf", width=4.5, height=4, useDingbats=FALSE )
+par( mai=c(0.65, 0.65, 0.05, 0.05), mgp=c(2, 0.5, 0), tck=-0.03 )
+plot(  -5
+       ,  -5
+       ,   col = 'red'
+       #,   xlim = c(-0.5,0.85)
+       ,   xlim = c(-0.3,0.7)
+       ,   ylim = c(0,4.)
+       #,   col=colors[1]
+       #,   ylim=c(0,7)
+       #,   border = colors[1]
+       #,   lty = 0
+       ,   type = 'l'
+       ,   lwd  = 2 
+       #,   main = 'Correlations with Evolutionary Rates'
+       #,   xlab = expression(paste('Spearman Cor. with Evolutionary Rates ',rho))
+       ,   xlab = expression( paste('Correlation with Evolutionary Rates: Spearman ', rho ) )
+       ,   ylab = 'Relative Frequency'
+)
+lines( -hist.wcnSC$x
+       , hist.wcnSC$y
+       , col = 'black'
+       , lwd = 2
+)
+lines( -hist.wcnCA$x
+       , hist.wcnCA$y
+       , col = 'black'
+       , lwd = 2
+       , lty = 2
+)
+lines( hist.rsa$x
+       , hist.rsa$y
+       , col = 'blue'
+       , lwd  = 2
+)
+
+legend( 'topleft'
+        , c("1 / WCN (SC)", "1 / WCN (CA)", "RSA")
+        , col = c('black','black','blue')
+        , lty = c(1,2,1)
+        , lwd = 2
+        , bty = 'n'
+        , cex = 0.9
+)
+
+graphics.off()
+
 # Now do some paired t-tests:
 
 #temp1 = data.frame(var = 'wcnSC', r = best_structural_predictors_of_ER_given_VSCvolume$r.wcnSC.r4sJC)

### 2015-06-04 manuscript/voronoi/mnv.tex
MN update

@@
 Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume.
-It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a non-parametric rank correlation coefficient.
+It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman $\rho$ is a non-parametric rank correlation coefficient.
+
+    The resulting distribution of the Spearman's partial correlation coefficients of evolutionary rates with wcn (calculated using side-chain coordinates, wcnSC) has an absolute median value of 0.32 with a $50\%$ quartile range of $[0.23,0.40]$ about the median of the distribution.
 

### 2015-06-05 manuscript/voronoi/mnv.tex
MN update

@@
 It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman $\rho$ is a non-parametric rank correlation coefficient.
+\\
 
-    The resulting distribution of the Spearman's partial correlation coefficients of evolutionary rates with wcn (calculated using side-chain coordinates, wcnSC) has an absolute median value of 0.32 with a $50\%$ quartile range of $[0.23,0.40]$ about the median of the distribution.
+    The resulting distribution of the Spearman's partial correlation coefficients of evolutionary rates with wcn (calculated using side-chain coordinates, wcnSC) has an absolute median value of $0.32$ with a $50\%$ quartile range of $[0.23,0.40]$ about the median of the distribution.
+Therefore, the long-range amino acid interactions appear to explain approximately $10\%$ of the site-specific sequence evolutionary rates.
+The local packing density as measured by the inverse of Voronoi cell volume is alone capable of explaining a median $35\%$ of sequence evolutionary rates, corresponding to a median Spearman's correlation strength of $0.59$ with $50\%$ quartile range of $[0.52,0.64]$.
+By contrast, the partial correlation distribution of RSA while controlling for the Voronoi cell volume for the same dataset, indicates a negligible median contribution of $\sim0.006$ in explaining the observed variability in evolutionary rates of individual proteins.
 
@@
 
-    Throughout the previous sections,  a comprehensive analysis and comparison of the main structural determinants of sequence variability was carried out, using a dataset of $209$ monomeric enzymes.
-Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg rate as defined in Chapter \ref{chap:sschar} (see also Echave et al.
-(2014) \cite{echave_relationship_2014}), which is essentially a proxy measure of the stability of protein's native conformation upon substitution of amino acids in individual sites in proteins.
-In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg rate, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figures \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE}).
+    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most important structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes.
+Examples of important sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA).
+In addition, we have derived a new set of site-specific characteristics from the Voronoi Tessellation of protein 3D structures, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).
+
+    We also argued that the commonly used definitions of local packing density as measured by the Contact Number (CN) and the Weighted Contact Number (WCN), involve a free adjustable parameter in the definition.
+In order to get the best predictions of sequence variability, we fine-tuned the free parameters of the models for each proteins individually, and showed that the best performing parameter values of CN and WCN indicate a non-negligible influence of the long-range amino acid interactions, independently of the local packing density.
 

### 2015-06-05 analysis/src/best_structural_predictors_of_ER.r

@@ -205,6 +205,28 @@ legend( 'topleft'
 graphics.off()
 
 
+# Now summarize the partial data over the entire dataset:
+
+best_structural_predictors_of_ER = read.csv(file = "../tables/best_structural_predictors_of_ER.csv", header=TRUE )
+
+data_subset = best_structural_predictors_of_ER[,-1]  # remove pdb column from the data set to loop over the columns below
+best_structural_predictors_of_ER_summary = data.frame()
+for (correlation in names(data_subset))
+{
+  row = data.frame( correlation = correlation
+                    , mean = mean(best_structural_predictors_of_ER[[correlation]])
+                    , median = median(best_structural_predictors_of_ER[[correlation]])
+                    , min = min(best_structural_predictors_of_ER[[correlation]])
+                    , quantile05 = quantile(best_structural_predictors_of_ER[[correlation]], probs=0.05)
+                    , quantile25 = quantile(best_structural_predictors_of_ER[[correlation]], probs=0.25)
+                    , quantile75 = quantile(best_structural_predictors_of_ER[[correlation]], probs=0.75)
+                    , quantile95 = quantile(best_structural_predictors_of_ER[[correlation]], probs=0.95)
+                    , max = max(best_structural_predictors_of_ER[[correlation]])
+                    , sd = sd(best_structural_predictors_of_ER[[correlation]])
+  )
+  best_structural_predictors_of_ER_summary = rbind(best_structural_predictors_of_ER_summary,row)  
+}
+write.csv(best_structural_predictors_of_ER_summary, file = "../tables/best_structural_predictors_of_ER_summary.csv", row.names=F )
 
 
 

### 2015-06-06 manuscript/voronoi/mnv.tex
MN update

@@
 
-    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most important structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes.
+    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most widely studied structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes.
 Examples of important sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA).
-In addition, we have derived a new set of site-specific characteristics from the Voronoi Tessellation of protein 3D structures, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).
+In addition, we have derived a new set of site-specific characteristics from Voronoi Tessellation of protein 3D structure, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).
 
-    We also argued that the commonly used definitions of local packing density as measured by the Contact Number (CN) and the Weighted Contact Number (WCN), involve a free adjustable parameter in the definition.
-In order to get the best predictions of sequence variability, we fine-tuned the free parameters of the models for each proteins individually, and showed that the best performing parameter values of CN and WCN indicate a non-negligible influence of the long-range amino acid interactions, independently of the local packing density.
+    We also argued that the commonly used measures of local packing density, most importantly the Contact Number (CN) and the Weighted Contact Number (WCN), involve free adjustable parameters in the definitions which can be fine-tuned for each individual protein to obtain the best correlations between CN$/$WCN and sequence evolutionary rates.
+Although the best parameter values for CN \& WCN vary from one protein structure to another, the median values over the entire dataset indicate a non-negligible selection pressure due to long-range amino acid interactions on sequence evolution, independently of the local packing density.
 

### 2017-01-19 manuscript/voronoi/dissecting-roles-local/main-submission1.tex
final article source files added

\documentclass[12pt]{article}

\usepackage[super,comma,sort&compress]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{fancyhdr}

% allow floats to take up as much space as they want on a page
\renewcommand{\topfraction}{1.}	% max fraction of floats at top
\renewcommand{\bottomfraction}{1.}
\renewcommand{\textfraction}{0.}
\renewcommand*{\thefootnote}{\fnsymbol{\star}}

%   document style macros
\newcommand{\ddg}{$\Delta\Delta G~$}

\pagestyle{fancy}
\thispagestyle{empty}
\fancyhead[L]{Roles of packing density and longer-range effects in protein evolution}
%\fancyhead[R]{2015 Shahmoradi and Wilke}

\begin{document}

\title{Dissecting the roles of local packing density and longer-range effects in protein sequence evolution}

\author{Amir Shahmoradi$^{1,2,3}$\footnote{Correspondence to: amir@physics.utexas.edu}~, Claus O. Wilke$^{2,3,4}$\footnote{Correspondence to: wilke@austin.utexas.edu}}
\date{}
\maketitle

\noindent
$^1$ Department of Physics, The University of Texas at Austin\\
$^2$ Center for Computational Biology and Bioinformatics, The University of Texas at Austin\\ 
$^3$ Institute for Cellular and Molecular Biology, The University of Texas at Austin\\
$^4$ Department of Integrative Biology, The University of Texas at Austin\\

\begin{abstract}
What are the structural determinants of protein sequence evolution? A number of site-specific structural characteristics have been proposed, most of which are broadly related to either the density of contacts or the solvent accessibility of individual residues. Most importantly, there has been disagreement in the literature over the relative importance of solvent accessibility and local packing density for explaining site-specific sequence variability in proteins. We show here that this discussion has been confounded by the definition of local packing density. The most commonly used measures of local packing, such as the contact number and the weighted contact number, represent by definition the combined effects of local packing density and longer-range effects. As an alternative, we here propose a truly local measure of packing density around a single residue, based on the Voronoi cell volume. We show that the Voronoi cell volume, when calculated relative to the geometric center of amino-acid side chains, behaves nearly identically to the relative solvent accessibility, and both can explain, on average, approximately 34\% of the site-specific variation in evolutionary rate in a data set of 209 enzymes. An additional 10\% of variation can be explained by non-local effects that are captured in the weighted contact number. Consequently, evolutionary variation at a site is determined by the combined action of the immediate amino-acid neighbors of that site and of effects mediated by more distant amino acids. We conclude that instead of contrasting solvent accessibility and local packing density, future research should emphasize the relative importance of immediate contacts and longer-range effects on evolutionary variation.
\end{abstract}

%\smallskip
\noindent Keywords: protein structure; protein evolution; packing density; solvent accessibility; contact number.

\section*{Introduction}

A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \cite{Overingtonetal1992, goldman_assessing_1998, MirnyShakhnovich1999, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross_species_2013, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number (CN) \cite{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \cite{MirnyShakhnovich1999, DokholyanShakhnovich2001, Liberlesetal2012, echave_relationship_2014, Dasmehetal2014}, protein designability \cite{england_structural_2003, Jacksonetal2013}, and measures of local flexibility, such as the B factor \cite{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014}, or flexibility measures based on elastic network models \cite{liu_sequence_2012, huang_mechanistic_2014} and Molecular Dynamics (MD) simulations \cite{shahmoradi_predicting_2014}.

It is conceivable that the majority of these site-specific quantities predicting evolutionary rate (ER) simply act as different proxy measures of an underlying biophysical constraint, such as the strength of amino acid interactions at individual sites. Indeed, Huang et al. \cite{huang_mechanistic_2014} recently proposed that the underlying constraint is mechanistic stress, which can be estimated by calculating the weighted contact number (WCN) at each site. Quantities such as relative solvent accessibility and local flexibility are expected to correlate with WCN, because all these measures depend to some extent on the density of contacts around an amino acid. Huang and coworkers thus argued that the \emph{local packing density} is the primary constraint of protein sequence evolution \cite{huang_mechanistic_2014, yeh_site-specific_2014, yeh_local_2014}.

However, a priori it is not clear whether quantities based on the contact-number concept (either CN or WCN) are truly measures of \emph{local} packing density. All these measures involve adjustable free parameters in their definitions, and depending on the parameter choices, the measures can incorporate substantial long-range effects. In particular, WCN by default uses an inverse-square weighting, which assumes that amino acids at an arbitrary distance away from the focal residue still exert some measurable effect on that residue. Thus, some of the best predictors of evolutionary variation seem to confound local and longer-range effects, and the relative importance of the immediate neighborhood of an amino acid vs.\ longer range effects is not known.

To separate out the effects of direct-neighbor and longer-range effects, we here derive a new set of structural characteristics that, unlike CN and WCN, depend only on the immediate contacts of a given amino acid and do not involve freely adjustable parameters in their definitions. We achieve this goal by employing tessellation methods developed in the field of computational geometry. We find that quantities based on tessellation methods explain much but not all of the variance in ER that WCN captures. Therefore, we conclude that longer-range effects beyond the first coordination shell play a significant role in protein sequence evolution.

\section*{Methods}

\subsection*{Protein data set and site-specific sequence variability}

We analyzed the same data set of 209 monomeric enzymes we have previously analyzed \cite{echave_relationship_2014} and that was originally published with four additional proteins \cite{yeh_site-specific_2014}. The enzymes were originally randomly picked from the Catalytic Site Atlas $2.2.11$ \cite{porter_catalytic_2004}, range from 95 to 1287 amino acids in length, and include representatives from all six main EC functional classes \cite{webb_enzyme_1992} and domains of all main SCOP structural classes \cite{murzin_scop:_1995}. Evolutionary rates at all sites in all proteins were calculated as described \cite{echave_relationship_2014}. In brief, we employed the software Rate4Site \cite{mayrose_comparison_2004}, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model.

\subsection*{Packing density and relative solvent accessibility}

Local Packing Density (LPD) for individual sites in all proteins was calculated according to the two most-commonly used definitions, Contact Number (CN) and Weighted Contact Number (WCN) (defined in Eqs.~\ref{eqn:cn} and \ref{eqn:wcn_pwrl} in Results). Both CN and WCN generally require each residue to be represented by a single point in space 
(but see the work by Marcos and Echave \cite{marcos_too_2015}, who used a two-point representation). Here, we examined seven different ways to choose the reference point for each amino acid: We considered all backbone atomic coordinates, i.e., the coordinates of N, C, O, or C{$_\alpha$}, the coordinates of the first heavy atom in the side-chains, C{$_\beta$}, and two more representations obtained by averaging over the coordinates of all heavy atoms in the amino acid including the backbone atoms (denoted as AA coordinates) or of all side-chain heavy atoms (denoted as SC coordinates). For each of these seven representations, we calculated CN and WCN for each residue in each protein. Appendix~A shows that the SC representation results in the highest correlations between structural measures and sequence evolution, and therefore we used SC coordinates for all results reported in the main body of the manuscript.

We used the DSSP software \cite{kabsch_dictionary_1983} to calculate  the Accessible Surface Area (ASA) for each residue in each protein, and we normalized these ASA values by their theoretical maximum \cite{tien_maximum_2013} to obtain Relative Solvent Accessibility (RSA).

\subsection*{Voronoi tessellation}

We used the VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset. Among the most important properties are the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity, defined as the distance between the cell's seed and the geometrical center of the cell. In addition, the cell {\it sphericity}, which is a measure of the compactness of the cell, can be defined as
    \begin{equation}
        \label{eqn:sphericity}
        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
    \end{equation}
Here, $V$ and $A$ represent the volume and the area of the cell, respectively.  By definition, $\Psi$ falls between 0 and 1. The limit of $\Psi=1$ indicates that the cell is perfectly spherical, while the limit of $\Psi=0$ indicates that the cell is a 2-dimensional object that has no volume but only surface area.

Each Voronoi cell is a polyhedron consisting of a set of polygons joined together at their edges. The total edge length of a cell was calculated as the sum of the edge lengths of all such polygons. The resulting sum was divided by two to obtain the total edge length, since each edge is shared between two polygons. Similarly, the cell surface area was calculated as the sum of the surface areas of all polygons that bounded it. Finally, the cell volume was defined as the volume of the Voronoi polyhedron.

As for CN and WCN, we considered seven different coordinate representations, and we found that the SC coordinates performed best (Appendix A). Therefore, unless otherwise noted, all results reported in the main body of the text are based on SC coordinates.

\subsection*{Edge effects in Voronoi partitioning of protein structures}

    One potential caveat with Voronoi tessellation of finite structures is the existence of {\it edge effects}. Sites that are close to the surface of the protein can be associated with Voronoi cells that are open, only bounded by the cubic box containing the protein. We performed the following analysis to verify that these edge effects did not influence the observed sequence-structure correlations. We first identified all open cells in all proteins by examining the fraction of change in the cell volumes upon changing the size of the cubic box that contained the protein. For each protein, once the open cells were identified, we multiplied their volumes by a normalizing factor such that the smallest-volume open cell had a larger volume than largest closed cell. We then ordered the open cells by the fraction of volume change observed upon changing the box size. In other words, the more the cell volume increased upon increasing the box size, the larger the volume of that cell was assumed to be, compared to all other open cells in the protein. Finally, we calculated Spearman's correlation coefficient between this ordered set of cell volumes (that included both open and closed cells) and the sequence evolutionary rates. We then compared the computed correlation coefficients to those obtained by using the original cell volumes without corrections for edge effects. The resulting distributions of correlation coefficients for the original and corrected cell volumes were nearly identical (mean difference $<0.001$).

For our dataset of 209 monomeric proteins, we concluded that edge effects due to Voronoi tessellation appear to have virtually no influence on the observed sequence-structure correlations. We reached similar conclusions when open cells were alternatively ranked by other criteria, such as the fractional changes in cell area upon changing the box size. Thus, the Voronoi cell characteristics, in particular cell volume and cell area, can be safely used in predicting sequence variability, as long as a sufficiently large enclosing box is used to calculate the Voronoi cell characteristics of open cells.

One exception, however, is cell sphericity, as defined in Eqn. \ref{eqn:sphericity}. This quantity behaves differently for open and closed cells. Unlike the case for closed cells in the protein's interior,  the sphericity of open cells on the surface of the protein is generally {\it positively} correlated with site-specific evolutionary rates. This is an artifact of the edge effects in the tessellation of finite structure of protein. A possible correction to the sphericity of open cells could be therefore applied by negating the value of sphericity for open cells. This correction resulted on average in a 0.05 increase in the correlation coefficients of sphericity with ER. However, since this correction has no obvious justification based on first principles, we did not employ it for further analyses. More importantly, the inclusion or exclusion of edge-effect corrections to cell sphericity did not substantively affect any of the major findings of this work.

\section*{Statistical analysis and data availability}

Statistical analyses were carried out with the software package R \cite{ihaka1998r}. We used Spearman correlations and partial Spearman correlations throughout this work. Partial Spearman correlations were calculated using the R package {\it ppcor} \cite{kim_understanding_2006}.

All data and analysis scripts necessary to reproduce this work are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.

\section*{Results}

\subsection*{Optimal definition of contact number incorporates longer-range effects}

Prior work on the same enzyme data set that we have analyzed here has shown that contact number generally outperforms all other structural predictors of site variability \cite{yeh_site-specific_2014, yeh_local_2014, huang_mechanistic_2014}. This finding has been interpreted as indicating that local packing density is the primary evolutionary constraint imposed by protein structure. However, to what extent the contact number (CN) is actually a truly local measure is unclear. The contact number at a site $i$, CN$_i$, is defined as the number of amino acids within a fixed radius $r_0$ from that site,
    \begin{equation}
        \label{eqn:cn}
        \text{CN}_{i} = \sum_{j\neq i} H\big(r_0-r_{ij}\big),
    \end{equation}
where $r_{ij}$ is the distance between sites $i$ and $j$ and $H(r)$ is the Heaviside step function
    \begin{equation}
        \label{eqn:heaviside}
        H(r) = \begin{cases}
        			1 \quad\text{for $r\leq 0$,}\\
                    0 \quad\text{otherwise.}
               \end{cases}
    \end{equation}
The location of individual sites is commonly chosen as the position of the residues' C$_\alpha$ backbone atoms, but recent work has shown that the geometric center of the side chain may be a better choice \cite{marcos_too_2015}. Irrespective of the particular choice of reference point for each site, how local of a measure CN$_i$ actually is depends on the arbitrary cutoff parameter $r_0$. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range of 5\AA\ to 18\AA\ \cite{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014, yeh_local_2014}. While a cutoff of 5\AA\ likely captures only immediate neighbors of the focal amino acid, a cutoff of 18\AA\ will capture many amino acids that do not directly interact with the focal amino acid, and thus captures both local and non-local packing-density information.

To determine at what cutoff parameter $r_0$ the contact number had the most predictive power for evolutionary variation, we calculated, for each protein in our data set, the Spearman correlation between evolutionary rate (ER) and contact number CN across the protein as a function of $r_0$ (Figure \ref{fig:cnwcnp}A). (A similar analysis has previously been published\cite{yeh_local_2014}.) We found that, on average, the correlation started to rise rapidly around 5\AA, reached a maximum around $14.3$\AA, and then started to decline slowly for larger $r_0$ (see also Table~\ref{tab:best_params}). Since the maximum correlation was located at a cutoff value that captured at least two complete shells of amino acids around the focal residue, this analysis suggested that non-local effects play some role in constraining amino-acid variability.

The arbitrariness of the cutoff $r_0$ in the CN definition has been long recognized, and therefore some authors \cite{lin_deriving_2008} have suggested an alternative quantity known as the Weighted Contact Number (WCN). For a given site $i$ in a protein of length $N$, WCN$_i$ is defined as the sum of the inverse-squared  distances to all other sites in protein,
    \begin{equation}
        \label{eqn:wcn_pwrl}
        \text{WCN}_{i} = \sum^N_{j\neq i} r_{ij}^{-2}.
    \end{equation}
It turns out that WCN generally performs better than CN at predicting ER \cite{yeh_site-specific_2014, yeh_local_2014, huang_mechanistic_2014}. However, WCN is clearly \emph{not} a local measure. By virtue of the exponent $-2$, the aggregated influence of all amino acids at distance $r$ on WCN is the same, irrespective of $r$. In other words, the factor $1/r^2$ exactly compensates the geometrical increase in the number of residues in each shell, for arbitrarily large radii $r$. Thus, rather than just measuring local packing density, WCN also measures where the focal amino acid falls relative to the overall shape and distribution of amino acids in the protein. For example, in a perfectly spherical protein of uniform density, WCN would be a direct proxy of the distance to the geometric center of the protein.

Moreover, just as was the case with CN, the definition of WCN contains a somewhat arbitrary parameter \cite{yang_protein_2009}, the exponent $-2$. If we chose a more negative exponent, then WCN would be a more local measure of packing density, putting less weight on increasingly distant amino acids. Likewise, if we chose a less negative exponent, WCN would put more weight on more distant amino acids and less weight on amino acids close in. Therefore, just as we had done with CN, we  calculated the correlations between WCN and ER as a function of the exponent $\alpha$ (Figure \ref{fig:cnwcnp}B). We found that the correlations peaked at $\alpha\sim -2.3$, almost exactly the canonical value of $-2$.

Evidently, both CN and WCN perform best at describing ER variation when they capture a non-negligible contribution of longer-range effects, beyond the immediate neighborhood of each focal site. To disentangle the contributions of local and longer-range effects to sequence evolution, we next proceeded to develop a measure of packing density that by definition depended on only the immediate neighbors around a single amino acid.

\subsection*{Measures of local packing density based on Voronoi tessellations}

To describe the local surroundings of a residue inside a protein in an unambiguous manner, we need to partition the space inside a protein into regions that uniquely belong to specific residues. Such structural partitioning has a long history in the analysis of protein structures \cite{richards_interpretation_1974, gerstein_volume_1994}. In particular, the Voronoi tessellation and its dual graph, the Delaunay triangulation, have been used in several studies analyzing the internal structure of proteins and/or developing empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. 
The Voronoi tessellation divides the Euclidean space into regions, called {\it cells}, and each belonging to exactly one centroid point (seed), such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid point (Figure~\ref{fig:voronoi}).

When applying a Voronoi tessellation to a protein structure, we usually choose one seed point per residue in the structure. As is the case with CN and WCN, there are multiple possible choices for which point best represents each residue. While many authors simply use the locations of the C$_\alpha$ backbone atoms, we can also use the geometric center of the amino-acid side-chain, the geometric center of the entire amino acid, or any particular heavy atom in the amino-acid backbone or side-chain. We show in Appendix A that we obtain the strongest correlations of Voronoi cell properties with site-specific sequence variability of proteins if we use the geometric centers of the residue side chains as Voronoi seeds. Thus, we used these specific seeds for the remainder of this work.

How can the Voronoi tessellation be used to describe the local packing density in a protein? Most importantly, the volume of a Voronoi cell is a direct measure of the amount of space surrounding a particular residue. More densely packed regions of the protein will yield smaller cell volumes than less densely packed regions. However, in addition to cell volume, several other cell properties also measure the local geometric arrangement of residues surrounding the focal one, such as cell edge length, cell area, cell eccentricity, and cell sphericity (see Methods for definitions of these quantities). We asked which of these quantities could be used to predict ER, and we found that they all did (Figure \ref{fig:voronoi_ER_screen}A). Specifically, we found that cell volume and surface area were most strongly correlated with ER, followed by the cell eccentricity, total edge length, and the cell's sphericity (Figure \ref{fig:voronoi_ER_screen}A and Table \ref{tab:ttest-Voronoi}). All of the Voronoi cell characteristics except sphericity correlated positively with ER, while sphericity displayed negative correlations.

In this context, we would like to emphasize that most Voronoi cell characteristics increase with decreasing local packing density. For example, the larger the cell volume or cell surface area, the fewer amino acids are located in the immediate neighborhood of the focal residue. By contrast, increasing sphericity implies increasing packing density. In general, a polyhedron with a higher number of faces corresponds to higher sphericity, the limiting case of which is a perfect sphere with an infinite number of faces. In the context of Voronoi polyhedra in proteins, a higher number of faces of a cell indicates more amino-acid neighbors around the site of interest, and therefore more interactions among them. Consequently, correlations with cell sphericity tend to have the opposite sign from correlations with other Voronoi cell characteristics.

Not unexpectedly, we found that the Voronoi cell characteristics correlated strongly with each other. Therefore, we next used cell volume (measuring inverse local packing density) as the reference and calculated the partial correlations of ER with Voronoi cell properties controlling for cell volume. We found that these partial correlations were generally quite low, with median absolute values of approximately 0.1  (Figure \ref{fig:voronoi_ER_screen}B). Cell edge length and cell sphericity displayed negative partial correlations with ER while cell eccentricity and cell area displayed positive partial correlations. In conclusion, with regards to ER variation, the cell area, volume, and edge length appear to represent largely the same underlying property of a Voronoi cell, its size.

\subsection*{Influence of longer-range effects on sequence evolution}

We next asked how Voronoi cell volume compared as predictor of ER relative to the more commonly studied quantities WCN and RSA. As seen in Figure \ref{fig:best_predictorER}A, Voronoi cell volume, RSA, and C$_{\alpha}$ WCN all displayed comparable correlations with ER, with median (absolute) correlation coefficients of 0.59, 0.57, and 0.56, respectively. By contrast, side-chain WCN out-performed these predictors by a substantial margin, with median (absolute) correlation coefficient of 0.64 (see Table~\ref{tab:ttest-best} for pairwise paired t-tests among all distributions). The superior performance of side-chain WCN on this data set is consistent with previous independent results in the literature \cite{marcos_too_2015}.

Note that both Voronoi cell volume and RSA are purely local quantities, determined entirely by the immediate neighborhood of an amino acid. In fact, they correlate strongly with each other, and measure approximately the same quantity (Appendix A). By contrast, as mentioned previously, WCN takes into account the amino-acid arrangement in the entire protein, and it equally weighs each shell with radius $r$ around the focal residue. Thus, to disentangle local from non-local effects, we calculated partial correlations of ER with WCN while controlling for Voronoi cell volume. In this way, we could quantify the residual influence of WCN---that is, the influence of longer-range effects beyond immediate neighbors---on sequence variability. As a control, we performed the same calculation for RSA instead of WCN.

We found that side-chain WCN displayed substantial residual correlations with ER after controlling for Voronoi cell volume (Figure \ref{fig:best_predictorER}B). The distribution of (absolute) partial correlation coefficients had a median value of 0.32, with an interquartile range of 0.17 (from 0.23 to 0.40). Thus, after controlling for local packing, the longer-range effects appeared to explain approximately $10\%$ of the site-specific sequence evolutionary rates. Similarly, there were significant residual correlations of C$_\alpha$ WCN with ER after controlling for Voronoi cell volume, with a median (absolute) correlation coefficient of $0.21$. By contrast, the residual correlations of RSA with ER were by-and-large negligible. The median was only $0.09$, with an interquartile range of 0.17 (from $-0.02$ to $0.15$).

Thus, in conclusion, either WCN definition (side-chain or C$_\alpha$) provided a significant, independent contribution to explaining ER when purely local effects of packing were controlled for by Voronoi cell volume. By contrast, RSA did not make an independent contribution once Voronoi cell volume was controlled for, as was expected given these quantities' strong correlations with each other.

\section*{Discussion}

In this work, we have shown that the commonly used measures of packing density, contact density (CN) and weighted contact density (WCN), capture non-local effects of amino-acid packing in a protein. To quantify the magnitude of these effects, we have defined purely local measures of packing density via the Voronoi tessellation, and we have shown that Voronoi cell volume encapsulates most of the relevant information about packing contained in various Voronoi cell characteristics. We have then shown that the variation in site-specific evolutionary rates (ER) can be partitioned into two independent components: variation explained by local, site-specific packing density (as measured by Voronoi cell volume) and variation explained by more long-range effects that extend beyond the first coordination shell around each site in a given protein. The longer-range effects play a non-negligible role in protein sequence evolution and explain on average approximately 10\% of the observed variability in ER in the data set studied here. These 10\% are in addition to and independent of the proportion of variance explained by the site-specific packing density, which explains on average another 34\%. By contrast, the relative solvent accessibility (RSA) has negligible explanatory power for ER once site-specific packing is taken into account via Voronoi cell volume.

   The quantities CN and WCN involve freely adjustable parameters in their definitions, and these parameters control to what extent CN and WCN assess the density of both immediate-neighbor and longer-range amino-acid packing. In principle, one could fine-tune these parameters for each individual protein to obtain the strongest correlations between ER and either CN or WCN \cite{yeh_local_2014}. Although the best parameterization of CN and WCN varies from one protein structure to another, we can take the values that maximize the median correlation for our dataset as the best overall parameterization of CN and WCN for our dataset. For both quantities, we find that the optimum parametrization is not entirely local. In particular, for WCN, which is the overall best predictor of ER in this dataset, the optimum exponent is $-2.3$, very close to the canonical value of $-2$ that is commonly used in the definition of WCN  \cite{lin_deriving_2008, yang_protein_2009, huang_mechanistic_2014, yeh_site-specific_2014, yeh_local_2014, marcos_too_2015}. Since an exponent of $-2$ gives nearly equal weight to all shells of any radius around the focal residue, WCN is not really a site-specific measure of packing density. Instead it measures both the local packing density around the residue and the position of the residue relative to the global shape of the entire protein. By contrast, the Voronoi cell volume introduced here as an alternative measure of site-specific packing density depends exclusively on contributions from the nearest neighbors of each site in protein, that is, the first coordination shell around each focal residue.

    Several earlier works \cite{yeh_local_2014, huang_mechanistic_2014} have argued that local packing density is the best predictor of ER, while others \cite{franzosa_structural_2009, shahmoradi_predicting_2014} have implicated RSA as the main determinant of sequence evolution, with local packing density playing a more  peripheral role. Importantly, all these works have defined local packing density as CN or WCN. We have shown here that if we define local packing density as the inverse of the Voronoi cell volume, which by definition excludes the effects of longer-range effects beyond the first coordination shell, then it performs comparable to RSA.  Our conclusion is that RSA and local packing density---when measured in a way that excludes longer-range effects---in principle represent the same characteristics of the local environment in a protein, that is, both quantities are proxy measures of the number of neighboring amino acids in the first coordination shell. Additional evolutionary variation is explained by non-local effects and/or the distribution of amino acids at larger distances, and these effects are captured by the most commonly used definition of WCN with inverse $r^2$ weighting.

    What is the origin of these non-local effects? There are two potential causes. First, longer-range effects can originate from steric interactions of local clusters of amino acids that extend beyond the first coordination shell. Evidence for such clusters comes from the discovery of protein sectors, regions of proteins with shared co-variation but little co-variation with other regions \cite{Halabietal2009}. Further evidence comes from the observation that correlations between fluctuations of pairs of sites in proteins decay inversely with distances of the sites from each other \cite{dehouck_effective_2013}. Second, evolutionary variation may be influenced by shape and finite-size effects. If proteins have approximately uniform amino-acid density (as our work suggests, see next paragraph), then WCN measures primarily where in the three-dimensional protein structure a given residue is located. Thus, if there are systematic trends of evolutionary variation related to shape and finite size, for example conserved active sites that are located near the center of the protein, then WCN can partially capture these effects and provide further explanatory power for ER.

The finding that RSA and Voronoi cell volume are highly correlated was unexpected. According to their definitions, these two quantities measure different aspects of a residue's local environment. In particular, RSA can only vary to the extent to which a residue is actually near the surface. For all completely buried residues $\text{RSA}=0$, with no variability. By contrast, Voronoi cell volume depends on the exact location and number of amino-acid neighbors, and hence can, at least in principle, vary substantially inside the protein. The congruence of RSA and cell volume suggests that cell volume is approximately constant for buried residues and is determined primarily by solvent exposure on the protein surface. These conditions can only be satisfied if the amino-acid density is approximately constant throughout the protein. Thus, the strong correlations we have found between RSA and cell volume imply that amino-acid density is, to first order, uniform across protein structures.

One limitation of our present study is the choice of data set, 209 monomeric enzymes that have been previously studied by multiple groups. All enzymes in this data set are relatively small, globular, and have a well-defined active site, usually close to the center of the protein \cite{shih_evolutionary_2012}. It is possible that the weighted contact number, with its radial weighting, performs particularly well on such proteins. For example, in the large and elongated influenza protein hemagglutinin, RSA explains over twice the variation in evolutionary rate than WCN does  \cite{meyer_geometric_2015}. Moreover, active sites and protein--protein interfaces generally induce additional evolutionary constraints, and how these constraints interact with local packing density and longer-range steric interactions remains largely unexplored \cite{Deanetal2002, franzosa_structural_2009, meyer_geometric_2015,Abriataetal2015}. Therefore, future work will have to disentangle the contributions of local packing and longer-range effects in a wider set of protein structures, and it will also have to disentangle these contributions from the evolutionary constraints imposed by active sites and protein--protein interfaces.

\section*{Appendix A: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
\label{app:best_crd}

The calculations of the quantities CN and WCN, as well as the Voronoi tessellation, all require us to represent each amino acid by a reference point in 3D space. Most commonly, that reference point is chosen to be the C$_\alpha$ atom of the residue. However, this choice is mainly driven by convenience and convention, and \emph{a priori} there is no particular reason to assume that this set of atomic coordinates is the best representation of individual sites in a protein. Indeed, earlier works have already suggested to use the center-of-mass of side chain coordinates to represent residues \cite{soyer_voronoi_2000}. More recently, it was shown that WCN when calculated from side-chain center-of-mass coordinates generally results in significantly better correlations with ER than WCN calculated from $C_\alpha$ atoms does \cite{marcos_too_2015}.

    To identify the best reference point in a more principled manner, we here considered seven different choices of atomic coordinates for the calculation of CN, WCN, and Voronoi cells. These include the set of coordinates of all backbone atoms (N, C, C$_\alpha$, O) and the first heavy atom in the amino acid side chains (C$_\beta$). In addition, representative coordinates for each site in the protein can be also calculated by averaging over the coordinates of all heavy atoms in the side chains (referred to as SC). Finally, we calculated a representative coordinate by averaging over all heavy atom coordinates in both the side chain and the backbone of the amino acid (referred to as AA). For all calculations that required C$_\beta$ coordinates, if the side chain $C_\beta$ atom was not been resolved in the PDB file or the amino acid lacked $C_\beta$ (i.e., Glycine), the $C_\alpha$ coordinate for the corresponding amino acid was used instead.

    For all seven coordinate choices, we correlated both ER and RSA with both WCN and with Voronoi cell volume for each protein (Figure \ref{fig:best_wcn_vvol}). In all cases, we found a systematic trend of declining correlation strengths as we moved the reference point further away from the geometric center of the side chain. In fact, the two atoms furthest away from the side chain, the nitrogen in the amino-group and the oxygen in the carboxyl group, generally produced the weakest correlations both with ER and with RSA. We also found that Voronoi cell volume and RSA measured approximately the same quantity, as long as the Voronoi tessellation was carried out on either SC or AA coordinates. The median correlation coefficient between RSA and cell volume for these coordinates was 0.92, and there was very little variation around that value (Figure \ref{fig:best_wcn_vvol}D). (In comparison, the distribution of the correlation coefficients of side-chain WCN with cell volume had a median of 0.89, and on average, the correlations between cell volume and RSA were 0.05 higher than the correlations between cell volume and side-chain WCN.)

    In combination, these results show that the biophysically relevant reference point for amino-acid position is the side chain, not the commonly used C$_\alpha$ atom. Further, they highlight that RSA properly accounts for the position of the side chain, and behaves nearly identically to side-chain Voronoi cell volume.

\section*{Acknowledgements}

We thank Julian Echave, Benjamin Jack, Austin Meyer, Stephanie Spielman, and Eleisha Jackson for helpful discussions and comments. This work was supported in part by NIH Grant R01 GM088344, DTRA Grant HDTRA1-12-C-0007, and the BEACON Center for the Study of Evolution in Action (NSF Cooperative Agreement DBI-0939454). The Texas Advanced Computing Center at UT Austin provided high-performance computing resources.

\bibliographystyle{unsrt}
\bibliography{main}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figures

\cleardoublepage
\section*{Figures}
    \begin{figure}[h]
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
        \end{tabular}
        \end{center}
        \caption{Absolute correlation of evolutionary rate with (A) Contact Number and (B) Weighted Contact Number for varying degrees of locality of these quantities. In (A), we vary the cutoff parameter $r_0$ in Eq.~\ref{eqn:cn} from 0 to 50\AA. In (B), we vary the exponent $\alpha$ in Eq.~\ref{eqn:wcn_pwrl} from $-30$ to 30. In each plot, the solid black line represents the mean correlation strength in the entire dataset of 209 proteins and the dashed black line indicates the median of the distribution. The green-shaded region together with the red-dashed lines represent the $25\%$ and $75\%$ quartiles of the correlation-strength distribution. Note that for the case of WCN with $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ with $\alpha<0$. In addition, $\rho$ is undefined at $\alpha=0$ and not shown in this plot. The parameter values at which the correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
        \label{fig:cnwcnp}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6in]{voronoi_diagram.pdf}
        \end{center}
        \caption{Example of a Voronoi tessellation in two dimensions. The red dots represent the seed points, and the black lines delineate the Voronoi cells. For protein structures, the tessellation is carried out in three dimensions.}
        \label{fig:voronoi}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
        \end{center}
        \caption{Correlations and partial correlations of ER with various Voronoi cell properties. (A) Distributions of ER correlations with Voronoi cell properties sphericity, edge length, eccentricity, volume, and area. Note that all cell characteristics correlate positively with ER, except sphericity which correlates negatively with ER. We show here the correlations of ER with $(-1)\times \text{cell sphericity}$ so that the correlation distributions all appear on the same positive scale. The correlations for cell area and cell volume are not significantly different from each other, and are higher than the correlations for all other Voronoi cell properties (Table~\ref{tab:ttest-Voronoi}). (B) Distributions of partial correlations of ER with Voronoi cell properties, controlling for cell volume. All partial correlations are relatively small, with medians of approximately 0.1. Therefore, none of these cell properties provide much independent information about ER once cell volume is accounted for.}
        \label{fig:voronoi_ER_screen}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
        \end{center}
        \caption{Correlations and partial correlations of ER with Voronoi cell volume, WCN, and RSA. (A) Distributions of ER correlations with Voronoi cell volume, WCN (using side-chain and C$_\alpha$  coordinates, denoted by SC and CA, respectively), and RSA. Note that the WCN measures correlate negatively with ER. We show here correlations of ER with $(-1)\times \text{WCN}$ so that the correlation distributions all appear on the same positive scale. The correlations of ER with WCN (SC) are significantly higher than the correlations of ER with all other quantities shown (Table~\ref{tab:ttest-best}). (B) Distributions of partial correlations of ER with WCN and RSA, controlling for cell volume. The partial correlation of ER with $-\text{WCN}$ are substantial, with median values of 0.32 (side-chain WCN) and 0.21 (C$_\alpha$ WCN). By contrast, the partial correlations of ER with RSA largely vanish, with a median value of 0.09.} \label{fig:best_predictorER}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
        \end{center}
        \caption{Side-chain centers provide most informative reference points for both WCN and Voronoi tessellation. (A) Distribution of correlations of WCN with ER, for seven different coordinate sets according to which WCN was calculated: SC, AA, CB, CA, N, C, O. Each coordinate set represents a different way of identifying the reference location of each residue. For SC (Side Chain) and AA (entire Amino Acid), the reference point is given, respectively, by the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms. The latter include the backbone but exclude any Hydrogen. The coordinate sets CB, CA, N, C, and O use the respective atom in the amino acid as the reference point. (B) As in (A), but using Voronoi Cell Volume instead of WCN. (C) As in (A), but the correlations are calculated with RSA instead of with ER. (D) As in (B), but the correlations are calculated with RSA instead of with ER.}
        \label{fig:best_wcn_vvol}
    \end{figure}

\cleardoublepage

\section*{Tables}

    \begin{table}[htbp]
    \caption{The first, second (median), and the third quartiles of the distribution of the best free parameters of the Contact Number (CN) and the Weighted Contact Number (WCN) resulting in the strongest correlations with sequence evolutionary rates. The best parameter distributions are obtained by tuning the free parameters of CN and WCN for individual proteins in the dataset of 209 monomeric enzymes such that the strongest correlation with evolutionary rates is obtained. \label{tab:best_params}}
    \bigskip
    \centerline{
    \begin{tabular}{lccc}
      \hline
      \multirow{2}{*}{Local Packing Density Measure} & \multicolumn{3}{c}{Best free-parameter} \\ \cline{2-4}
                                           & $25\%$ quartile & median & $75\%$ quartile  \\ \hline
      CN    & $r_0=9.3~[\text{\AA}]$ & $r_0=14.3~[\text{\AA}]$ & $r_0=19.6~[\text{\AA}]$  \\
      WCN & $\alpha=-2.7$ & $\alpha=-2.3$ & $\alpha=-1.5$  \\
      \hline
    \end{tabular}}
    \end{table}

    \begin{table}[htbp]
    \caption{The percentage of variance of site-specific evolutionary rates explained by Voronoi cell volume (representing the inverse of  site-specific packing density) and Weighted Contact Number controlling for cell volume (representing longer-range effects). In each case, the first, second (median) and the third quartiles of the percentage distribution are reported.  \label{tab:explained_var}}
    \bigskip
    \centerline{
    \begin{tabular}{lccc}
      \hline
      \multirow{2}{*}{Structural Quantity} & \multicolumn{3}{c}{Percentage of ER Variance Explained} \\ \cline{2-4}
                                           & $25\%$ quartile & median & $75\%$ quartile  \\ \hline
      Voronoi Cell Volume        & $27\%$          & $34\%$ & $41\%$  \\
      WCN, controlling for Cell Volume   & $5\%$           & $10\%$ & $16\%$  \\
      \hline
    \end{tabular}}
    \end{table}

% pairwise t-test table for the similarity of the correlation of ER with different Voronoi cell characteristics

    \begin{table}[htbp]
    \caption{\label{tab:ttest-Voronoi}$P$-values of pairwise paired t-tests of the absolute correlation strengths of evolutionary rates (ER) with different Voronoi cell characteristics: volume, area, total edge length, sphericity, eccentricity. Cell area and cell volume perform equally well in predicting ER, while all other cell characteristics perform significantly worse (Fig.~\ref{fig:voronoi_ER_screen}). The $P$-values were corrected for multiple testing using the Hochberg correction \citep{hochberg_sharper_1988}.}
    \bigskip
    \centerline{
    \begin{tabular}{lcccc}
      \hline
                         & ER -- Area        & ER -- Eccentricity & ER -- Edge        & ER -- Sphericity \\ \hline
      ER -- Volume       & $6\times10^{-20}$ & $10^{-26}$         & $2\times10^{-21}$ & $1\times10^{-60}$  \\
      ER -- Sphericity   & $2\times10^{-70}$ & $2\times10^{-43}$  & $4\times10^{-88}$ & --  \\
      ER -- Edge         & $5\times10^{-86}$ & $8\times10^{-15}$  & --                & -- \\
      ER -- Eccentricity & $6\times10^{-37}$ & --                 & --                & --  \\
      \hline
    \end{tabular}}
    \end{table}

    \begin{table}[htbp]
    \caption{\label{tab:ttest-best}$P$-values of pairwise paired t-tests of the absolute correlation strengths of evolutionary rates (ER) with four structural characteristics depicted in Figure \ref{fig:best_predictorER}A: $\text{WCN (SC)}$, $\text{WCN (CA)}$, cell volume, and RSA. The $P$-values were corrected for multiple testing using the Hochberg correction \citep{hochberg_sharper_1988}.}
    \bigskip
    \centerline{
    \begin{tabular}{lccc}
      \hline
         & ER -- WCN(SC) & ER -- WCN(CA)  & ER -- Volume\\ \hline
     ER -- RSA     & $2\times10^{-12}$  & $0.083$ & $6\times10^{-17}$  \\
     ER --  Volume  & $7\times10^{-6}$ & $5\times10^{-4}$ & --  \\
     ER -- WCN(CA) & $5\times10^{-11}$ & -- & -- \\
      \hline
    \end{tabular}}
    \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figures

\cleardoublepage
\section*{Figures}
    \begin{figure}[h]
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
        \end{tabular}
        \end{center}
        \caption{Absolute correlation of evolutionary rate with (A) Contact Number and (B) Weighted Contact Number for varying degrees of locality of these quantities. In (A), we vary the cutoff parameter $r_0$ in Eq.~\ref{eqn:cn} from 0 to 50\AA. In (B), we vary the exponent $\alpha$ in Eq.~\ref{eqn:wcn_pwrl} from $-30$ to 30. In each plot, the solid black line represents the mean correlation strength in the entire dataset of 209 proteins and the dashed black line indicates the median of the distribution. The green-shaded region together with the red-dashed lines represent the $25\%$ and $75\%$ quartiles of the correlation-strength distribution. Note that for the case of WCN with $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ with $\alpha<0$. In addition, $\rho$ is undefined at $\alpha=0$ and not shown in this plot. The parameter values at which the correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
        \label{fig:cnwcnp}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6in]{voronoi_diagram.pdf}
        \end{center}
        \caption{Example of a Voronoi tessellation in two dimensions. The red dots represent the seed points, and the black lines delineate the Voronoi cells. For protein structures, the tessellation is carried out in three dimensions.}
        \label{fig:voronoi}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
        \end{center}
        \caption{Correlations and partial correlations of ER with various Voronoi cell properties. (A) Distributions of ER correlations with Voronoi cell properties sphericity, edge length, eccentricity, volume, and area. Note that all cell characteristics correlate positively with ER, except sphericity which correlates negatively with ER. We show here the correlations of ER with $(-1)\times \text{cell sphericity}$ so that the correlation distributions all appear on the same positive scale. The correlations for cell area and cell volume are not significantly different from each other, and are higher than the correlations for all other Voronoi cell properties (Table~\ref{tab:ttest-Voronoi}). (B) Distributions of partial correlations of ER with Voronoi cell properties, controlling for cell volume. All partial correlations are relatively small, with medians of approximately 0.1. Therefore, none of these cell properties provide much independent information about ER once cell volume is accounted for.}
        \label{fig:voronoi_ER_screen}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
        \end{center}
        \caption{Correlations and partial correlations of ER with Voronoi cell volume, WCN, and RSA. (A) Distributions of ER correlations with Voronoi cell volume, WCN (using side-chain and C$_\alpha$  coordinates, denoted by SC and CA, respectively), and RSA. Note that the WCN measures correlate negatively with ER. We show here correlations of ER with $(-1)\times \text{WCN}$ so that the correlation distributions all appear on the same positive scale. The correlations of ER with WCN (SC) are significantly higher than the correlations of ER with all other quantities shown (Table~\ref{tab:ttest-best}). (B) Distributions of partial correlations of ER with WCN and RSA, controlling for cell volume. The partial correlation of ER with $-\text{WCN}$ are substantial, with median values of 0.32 (side-chain WCN) and 0.21 (C$_\alpha$ WCN). By contrast, the partial correlations of ER with RSA largely vanish, with a median value of 0.09.} \label{fig:best_predictorER}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
        \end{center}
        \caption{Side-chain centers provide most informative reference points for both WCN and Voronoi tessellation. (A) Distribution of correlations of WCN with ER, for seven different coordinate sets according to which WCN was calculated: SC, AA, CB, CA, N, C, O. Each coordinate set represents a different way of identifying the reference location of each residue. For SC (Side Chain) and AA (entire Amino Acid), the reference point is given, respectively, by the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms. The latter include the backbone but exclude any Hydrogen. The coordinate sets CB, CA, N, C, and O use the respective atom in the amino acid as the reference point. (B) As in (A), but using Voronoi Cell Volume instead of WCN. (C) As in (A), but the correlations are calculated with RSA instead of with ER. (D) As in (B), but the correlations are calculated with RSA instead of with ER.}
        \label{fig:best_wcn_vvol}
    \end{figure}

\end{document}

### 2017-01-19 manuscript/voronoi/dissecting-roles-local/main.tex
final article source files added

--- 2017-01-19 manuscript/voronoi/dissecting-roles-local/main-submission1.tex
@@
 \usepackage{fancyhdr}
+\usepackage{setspace}
+\doublespacing
 
@@
 \renewcommand*{\thefootnote}{\fnsymbol{\star}}
+\newcommand{\angstrom}{\text{\normalfont\AA}}
 
@@
 Most importantly, there has been disagreement in the literature over the relative importance of solvent accessibility and local packing density for explaining site-specific sequence variability in proteins.
-We show here that this discussion has been confounded by the definition of local packing density.
-The most commonly used measures of local packing, such as the contact number and the weighted contact number, represent by definition the combined effects of local packing density and longer-range effects.
-As an alternative, we here propose a truly local measure of packing density around a single residue, based on the Voronoi cell volume.
-We show that the Voronoi cell volume, when calculated relative to the geometric center of amino-acid side chains, behaves nearly identically to the relative solvent accessibility, and both can explain, on average, approximately 34\% of the site-specific variation in evolutionary rate in a data set of 209 enzymes.
+We show that this discussion has been confounded by the definition of local packing density.
+The most commonly used measures of local packing, such as contact number and the weighted contact number, represent the combined effects of local packing density and longer-range effects.
+As an alternative, we propose a truly local measure of packing density around a single residue, based on the Voronoi cell volume.
+We show that the Voronoi cell volume, when calculated relative to the geometric center of amino-acid side chains, behaves nearly identically to the relative solvent accessibility, and each individually can explain, on average, approximately 34\% of the site-specific variation in evolutionary rate in a data set of 209 enzymes.
 An additional 10\% of variation can be explained by non-local effects that are captured in the weighted contact number.
-Consequently, evolutionary variation at a site is determined by the combined action of the immediate amino-acid neighbors of that site and of effects mediated by more distant amino acids.
-We conclude that instead of contrasting solvent accessibility and local packing density, future research should emphasize the relative importance of immediate contacts and longer-range effects on evolutionary variation.
+Consequently, evolutionary variation at a site is determined by the combined effects of the immediate amino-acid neighbors of that site and effects mediated by more distant amino acids.
+We conclude that instead of contrasting solvent accessibility and local packing density, future research should emphasize on the relative importance of immediate contacts and longer-range effects on evolutionary variation.
 \end{abstract}
@@
 A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \cite{Overingtonetal1992, goldman_assessing_1998, MirnyShakhnovich1999, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross_species_2013, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number (CN) \cite{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \cite{MirnyShakhnovich1999, DokholyanShakhnovich2001, Liberlesetal2012, echave_relationship_2014, Dasmehetal2014}, protein designability \cite{england_structural_2003, Jacksonetal2013}, and measures of local flexibility, such as the B factor \cite{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014}, or flexibility measures based on elastic network models \cite{liu_sequence_2012, huang_mechanistic_2014} and Molecular Dynamics (MD) simulations \cite{shahmoradi_predicting_2014}.
-
-It is conceivable that the majority of these site-specific quantities predicting evolutionary rate (ER) simply act as different proxy measures of an underlying biophysical constraint, such as the strength of amino acid interactions at individual sites.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \cite{Overingtonetal1992, goldman_assessing_1998, MirnyShakhnovich1999, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross_species_2013, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number (CN) \cite{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \cite{MirnyShakhnovich1999, DokholyanShakhnovich2001, Liberlesetal2012, echave_relationship_2014, Dasmehetal2014}, protein designability \cite{england_structural_2003, Jacksonetal2013}, and measures of local flexibility, such as the B factor \cite{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014}, or flexibility measures based on elastic network models \cite{liu_sequence_2012, huang_mechanistic_2014} and Molecular Dynamics (MD) simulations \cite{shahmoradi_predicting_2014}, or direct comparison of fluctuations among static structures of homologous proteins \cite{fuglebakk_measuring_2012,shahmoradi_predicting_2014, carvalho_comparison_2015}.
+(For a recent review, see Ref.~\cite{echave_causes_2016}.) It is conceivable that the majority of these site-specific quantities predicting evolutionary rate (ER) simply act as different proxy measures of an underlying biophysical constraint, such as the strength of amino acid interactions at individual sites.
 Indeed, Huang et al.
 \cite{huang_mechanistic_2014} recently proposed that the underlying constraint is mechanistic stress, which can be estimated by calculating the weighted contact number (WCN) at each site.
-Quantities such as relative solvent accessibility and local flexibility are expected to correlate with WCN, because all these measures depend to some extent on the density of contacts around an amino acid.
+Quantities such as relative solvent accessibility and local flexibility are expected to correlate with WCN, because all these measures depend to some extent on the density of contacts around an amino acid\cite{echave_causes_2016}.
 Huang and coworkers thus argued that the \emph{local packing density} is the primary constraint of protein sequence evolution \cite{huang_mechanistic_2014, yeh_site-specific_2014, yeh_local_2014}.
@@
 Thus, some of the best predictors of evolutionary variation seem to confound local and longer-range effects, and the relative importance of the immediate neighborhood of an amino acid vs.\ longer range effects is not known.
+Importantly, by longer-range effects we mean here effects between amino acids that are distant in 3D space.
+Amino acids that are distant in sequence but close in 3D space, as are frequently observed\cite{plaxco_contact_1998,lai_sequence_2015}, would contribute to local packing but not to longer-range effects by our definition.
 
@@
 Evolutionary rates at all sites in all proteins were calculated as described \cite{echave_relationship_2014}.
-In brief, we employed the software Rate4Site \cite{mayrose_comparison_2004}, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model.
+In brief, we employed the software Rate4Site \cite{mayrose_comparison_2004}, using the empirical Bayesian method and the amino-acid Jukes-Cantor-like (JC-like) mutational model.
 
@@
 We used the DSSP software \cite{kabsch_dictionary_1983} to calculate  the Accessible Surface Area (ASA) for each residue in each protein, and we normalized these ASA values by their theoretical maximum \cite{tien_maximum_2013} to obtain Relative Solvent Accessibility (RSA).
+Throughout this work, we considered RSA rather than ASA, consistent with common practice in the literature.
+RSA is preferred in evolutionary analyses because ASA confounds residue size and residue exposure.
+The maximum ASA varies by a factor of three from the smallest amino acid (glycine) to the largest (tryptophan) \cite{tien_maximum_2013}.
+As a consequence, a fully exposed glycine may have the same ASA as a largely but not completely buried tryptophan.
+By contrast, an RSA of 1 indicates that the residue is fully exposed, regardless of the nature of the residue.
+Thus, we generally expect RSA to be less variable under amino-acid substitution than ASA.
+Fully exposed residues tend to remain fully exposed, partially exposed residues tend to remain partially exposed, and so on.
 
@@
 We used the VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset.
-Among the most important properties are the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity, defined as the distance between the cell's seed and the geometrical center of the cell.
-In addition, the cell {\it sphericity}, which is a measure of the compactness of the cell, can be defined as
+Specifically, we calculated the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity, defined as the distance between the cell's seed and the geometrical center of the cell.
+We also calculated the cell {\it sphericity}, which is a measure of the compactness of the cell, defined as
     \begin{equation}
@@
 
-    One potential caveat with Voronoi tessellation of finite structures is the existence of {\it edge effects}.
-Sites that are close to the surface of the protein can be associated with Voronoi cells that are open, only bounded by the cubic box containing the protein.
-We performed the following analysis to verify that these edge effects did not influence the observed sequence-structure correlations.
-We first identified all open cells in all proteins by examining the fraction of change in the cell volumes upon changing the size of the cubic box that contained the protein.
-For each protein, once the open cells were identified, we multiplied their volumes by a normalizing factor such that the smallest-volume open cell had a larger volume than largest closed cell.
-We then ordered the open cells by the fraction of volume change observed upon changing the box size.
-In other words, the more the cell volume increased upon increasing the box size, the larger the volume of that cell was assumed to be, compared to all other open cells in the protein.
-Finally, we calculated Spearman's correlation coefficient between this ordered set of cell volumes (that included both open and closed cells) and the sequence evolutionary rates.
+When performing a Voronoi tessellation of a protein, which has a finite size, we have to be aware of the existence of {\it edge effects} on the protein surface.
+Most Voronoi cells on the protein surface are not completely enclosed by other neighboring cells.
+Therefore, to ensure that all cells are closed in all directions, we need to place the protein structure into a hypothetical box of arbitrary size.
+This box provides the missing boundaries for the otherwise open cells on the protein surface.
+Voronoi cells that are bounded by this box tend to have artifactual properties, such as much larger volumes and areas compared to the cells that are fully surrounded by other neighboring cells.
+In addition, the volumes and areas of these cells depend on the size of the box, which is arbitrary.
+
+To verify that these edge effects did not influence the observed sequence-structure correlations, we investigated how our results changed as a function of the size of the enclosing box.
+To do so, we first identified all open cells in all proteins by examining the fraction of change in the cell volumes upon changing the size of the containing box.
+Initially, each wall of the box was set to minimum distance of $30.0$\AA\ from all atoms in the protein, and then the size of the box was changed, such that each wall would be at least $60.0$\AA\ away from any atom in the protein.
+Note that changing the size of the box surrounding the protein will only affect the properties of those cells that are inherently open in the absence of the box.
+
+For each protein, once open cells were identified, we sorted these cells according to their rates of volume change upon changing the box size, such that the slowest changing cell was assumed to have the smallest volume among open cells, and the fastest changing cell was assumed to have the largest volume among open cells.
+Note that this method of ranking the open cells is very similar to ranking the cells by the solid opening angle of open cells, without the requirement of explicitly measuring the solid opening angles for all open cells.
+The underlying assumption here is that cells with larger solid opening angles will show larger rate of cell volume change upon changing the box size surrounding the protein.
+The wider the solid opening angle of a cell is, the more space is assumed to be available to the amino acid residing the open cell.
+In essence, the rate of volume change provides a method of ranking the open cells based on their solid opening angles, without direct calculation of the angles.
+
+All open cells were assumed to have a larger volume than any closed cell.
+Finally, we calculated Spearman's correlation coefficient between this ordered set of cell volumes (that included both open and closed cells) and the evolutionary rates.
 We then compared the computed correlation coefficients to those obtained by using the original cell volumes without corrections for edge effects.
@@
 
-For our dataset of 209 monomeric proteins, we concluded that edge effects due to Voronoi tessellation appear to have virtually no influence on the observed sequence-structure correlations.
+For our dataset of 209 monomeric proteins, we thus concluded that edge effects due to Voronoi tessellation appear to have virtually no influence on the observed sequence-structure correlations.
 We reached similar conclusions when open cells were alternatively ranked by other criteria, such as the fractional changes in cell area upon changing the box size.
@@
 
+Throughout this study, we ask to what extent various measures of local packing in proteins predict evolutionary variation.
+To address this question, we carry out a number of analyses that all consist of the same conceptual steps illustrated in Figure \ref{fig:pipeline}.
+We first correlate the evolutionary rate (ER) at each site with a measure of packing at the same site (Figure \ref{fig:pipeline}A, B).
+We then calculate the corresponding correlation coefficients separately for all proteins in our data set (Figure \ref{fig:pipeline}C).
+Finally, we compare the distributions of these correlation coefficients for different measures of packing (Figure \ref{fig:pipeline}D).
+
 \subsection*{Optimal definition of contact number incorporates longer-range effects}
@@
         H(r) = \begin{cases}
-        			1 \quad\text{for $r\leq 0$,}\\
+        			1 \quad\text{for $r\geq 0$,}\\
                     0 \quad\text{otherwise.}
@@
 
-Evidently, both CN and WCN perform best at describing ER variation when they capture a non-negligible contribution of longer-range effects, beyond the immediate neighborhood of each focal site.
+Next we asked whether the improved performance of WCN over CN was caused specifically by the power-law weighting of close contacts or by the inclusion of long-range interactions.
+To do so, we considered a hybrid WCN model with power-law weighting and a hard cutoff beyond which all other residues were excluded.
+More specifically, we fixed the exponent of the power-law weighting function to $\alpha=-2$ and further included only those sites in the calculation of WCN that were within a specific radius of neighborhood $r_0$ from the site of interest,
+    \begin{equation}
+        \label{eqn:wcn_pwrl_cutoff}
+        \text{WCN}_{i}^\text{cutoff} = \sum^N_{j\neq i} r_{ij}^{-2} H\big(r_0-r_{ij}\big),
+    \end{equation}
+\noindent with $H$ defined by Eqn. \ref{eqn:heaviside}.
+We then varied the cutoff $r_0$ and searched for the value that would result in the strongest correlation of WCN with site-specific evolutionary rates, separately for each individual protein in our dataset.
+Finally, we measured the mean fraction of neighboring sites that were required in the calculation of $\text{WCN}^\text{cutoff}$  to obtain the highest correlation strengths.
+We found for the vast majority of proteins that the highest correlation strengths were obtained when over 90\% of the sites in the protein were included in the calculation of $\text{WCN}^\text{cutoff}$ (Figure \ref{fig:wcn_pwrl_cutoff}).
+This large fraction is a clear indication of the importance of distant sites in their evolution.
+
+In combination, these results demonstrate that both CN and WCN perform best at describing ER variation when they capture a non-negligible contribution of longer-range effects, beyond the immediate neighborhood of each focal site.
 To disentangle the contributions of local and longer-range effects to sequence evolution, we next proceeded to develop a measure of packing density that by definition depended on only the immediate neighbors around a single amino acid.
@@
 
-In this context, we would like to emphasize that most Voronoi cell characteristics increase with decreasing local packing density.
+In this context, we emphasize that most Voronoi cell characteristics increase with decreasing local packing density.
 For example, the larger the cell volume or cell surface area, the fewer amino acids are located in the immediate neighborhood of the focal residue.
@@
 Cell edge length and cell sphericity displayed negative partial correlations with ER while cell eccentricity and cell area displayed positive partial correlations.
-In conclusion, with regards to ER variation, the cell area, volume, and edge length appear to represent largely the same underlying property of a Voronoi cell, its size.
 
@@
 
-We found that side-chain WCN displayed substantial residual correlations with ER after controlling for Voronoi cell volume (Figure \ref{fig:best_predictorER}B).
+We found that side-chain WCN displayed substantial residual correlations with ER after controlling for Voronoi cell volume (Figure \ref{fig:best_predictorER}B, Table \ref{tab:explained_var}).
 The distribution of (absolute) partial correlation coefficients had a median value of 0.32, with an interquartile range of 0.17 (from 0.23 to 0.40).
@@
 
-Thus, in conclusion, either WCN definition (side-chain or C$_\alpha$) provided a significant, independent contribution to explaining ER when purely local effects of packing were controlled for by Voronoi cell volume.
+In conclusion, either WCN definition (side-chain or C$_\alpha$) provided a significant, independent contribution to explaining ER when purely local effects of packing were controlled for by Voronoi cell volume.
 By contrast, RSA did not make an independent contribution once Voronoi cell volume was controlled for, as was expected given these quantities' strong correlations with each other.
+
+\subsection*{Sensitivity of results to evolutionary-rate model matrix}
+
+The evolutionary rates we have used throughout this work were calculated based on a Jukes-Cantor-like (JC-like) model, in which all amino-acid substitutions are assumed to be equally likely and no corrections are made for mutation biases and/or the structure of the genetic code \cite{echave_relationship_2014}.
+More sophisticated evolutionary models, such as the JTT model\cite{jones_rapid_1992}, the WAG model\cite{WhelanGoldman2001}, or the LG model\cite{LeGascuel2008}, attempt to correct for these factors, and generally yield improved model fit compared to simpler models.
+To what extent these various models affect rate inferences is not well understood.
+In our analysis here, we used the same rates that had been used in previous analyses \cite{echave_relationship_2014,marcos_too_2015}, so that our results were directly comparable to these previous results.
+However, the same enzyme data set we analyzed here has also been studied previously with rates calculated under the JTT model\cite{yeh_site-specific_2014,huang_mechanistic_2014,yeh_local_2014}, so we asked to what extent our conclusions differed among these two sets of evolutionary rates.
+
+We found that, first, for each individual protein in the dataset, the evolutionary rates calculated under the JTT model correlated strongly with the evolutionary rates calculated under the JC-like model.
+The average Spearman correlation coefficient was $0.97$ with the standard deviation of distribution $\sigma=0.02$ over the entire dataset.
+Second, the results that we report in Table \ref{tab:explained_var} (percent variance explained by Voronoi cell volume and by WCN controlling for Voronoi cell volume) change only by approximately 2 percentage points when JTT rates are used in place of JC-like rates.
+Similarly, all other analyses also showed only minor variations upon using JTT rates.
+Therefore, we conclude that our findings are not strongly affected by the specific model with which evolutionary rates are calculated.
 
@@
 
-In this work, we have shown that the commonly used measures of packing density, contact density (CN) and weighted contact density (WCN), capture non-local effects of amino-acid packing in a protein.
+We have shown that the commonly used measures of packing density, contact density (CN) and weighted contact density (WCN), capture non-local effects of amino-acid packing in a protein.
 To quantify the magnitude of these effects, we have defined purely local measures of packing density via the Voronoi tessellation, and we have shown that Voronoi cell volume encapsulates most of the relevant information about packing contained in various Voronoi cell characteristics.
@@
 By contrast, the relative solvent accessibility (RSA) has negligible explanatory power for ER once site-specific packing is taken into account via Voronoi cell volume.
+
+One limitation of our study is the choice of the data set, 209 monomeric enzymes that have been previously studied by multiple groups.
+All enzymes in this data set are relatively small, globular, and have a well-defined active site, usually close to the center of the protein \cite{shih_evolutionary_2012}.
+It is possible that the weighted contact number, with its radial weighting, performs particularly well on such proteins.
+For example, in the large and elongated influenza protein hemagglutinin, RSA explains over twice the variation in evolutionary rate than WCN does  \cite{meyer_geometric_2015}.
+Moreover, active sites and protein--protein interfaces generally induce additional evolutionary constraints, and how these constraints interact with local packing density and longer-range steric interactions remains largely unexplored \cite{Deanetal2002, franzosa_structural_2009, meyer_geometric_2015,Abriataetal2015}.
+Therefore, future work will have to disentangle the contributions of local packing and longer-range effects in a wider set of protein structures, and it will also have to disentangle these contributions from the evolutionary constraints imposed by active sites and protein--protein interfaces.
 
@@
 By contrast, the Voronoi cell volume introduced here as an alternative measure of site-specific packing density depends exclusively on contributions from the nearest neighbors of each site in protein, that is, the first coordination shell around each focal residue.
+
+When searching for both the optimal cutoff for CN and the optimal exponent for WCN, we found a single global optimum in either case (Figure \ref{fig:cnwcnp}).
+This result was not unexpected.
+First, the weighting functions (Heaviside step function for CN, power-law for WCN) are monotonically decreasing functions of pairwise distance.
+Second, even though the pairwise interaction strengths of amino acids in a protein are complex functions of inter-residue distance, virtually unique to each pair of residues in each protein, the overall ``averaged over protein-medium'' behavior of interaction strength vs.\ inter-residue distance decays monotonically with distance \cite{dehouck_effective_2013,xia_correlation_2015,xia_multiscale_2015}.
+Therefore, the observed behavior in Figure \ref{fig:cnwcnp} merely assert that there is a single optimal parameter value that represents the best relative weighting of local and longer-range interactions.
+Either increasing or decreasing this parameter value will yield a sub-optimal relative weighting of local and longer-range interactions.
 
@@
 
+Other residue--residue interactions besides steric interactions are less likely to cause the long-range effects we have observed here.
+Covalent bonds and van der Waals' interactions hardly reach beyond the first coordination shells of amino acids, limiting the effective ranges of these interactions to the amino acid's immediate neighborhood $\lesssim10$\AA.
+And disulfide bonds and ionic interactions are infrequently observed in our sample of enzymatic proteins.
+
 The finding that RSA and Voronoi cell volume are highly correlated was unexpected.
@@
 
-One limitation of our present study is the choice of data set, 209 monomeric enzymes that have been previously studied by multiple groups.
-All enzymes in this data set are relatively small, globular, and have a well-defined active site, usually close to the center of the protein \cite{shih_evolutionary_2012}.
-It is possible that the weighted contact number, with its radial weighting, performs particularly well on such proteins.
-For example, in the large and elongated influenza protein hemagglutinin, RSA explains over twice the variation in evolutionary rate than WCN does  \cite{meyer_geometric_2015}.
-Moreover, active sites and protein--protein interfaces generally induce additional evolutionary constraints, and how these constraints interact with local packing density and longer-range steric interactions remains largely unexplored \cite{Deanetal2002, franzosa_structural_2009, meyer_geometric_2015,Abriataetal2015}.
-Therefore, future work will have to disentangle the contributions of local packing and longer-range effects in a wider set of protein structures, and it will also have to disentangle these contributions from the evolutionary constraints imposed by active sites and protein--protein interfaces.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%    
 \section*{Appendix A: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
@@
 In fact, the two atoms furthest away from the side chain, the nitrogen in the amino-group and the oxygen in the carboxyl group, generally produced the weakest correlations both with ER and with RSA.
-We also found that Voronoi cell volume and RSA measured approximately the same quantity, as long as the Voronoi tessellation was carried out on either SC or AA coordinates.
-The median correlation coefficient between RSA and cell volume for these coordinates was 0.92, and there was very little variation around that value (Figure \ref{fig:best_wcn_vvol}D).
-(In comparison, the distribution of the correlation coefficients of side-chain WCN with cell volume had a median of 0.89, and on average, the correlations between cell volume and RSA were 0.05 higher than the correlations between cell volume and side-chain WCN.)
 
-    In combination, these results show that the biophysically relevant reference point for amino-acid position is the side chain, not the commonly used C$_\alpha$ atom.
-Further, they highlight that RSA properly accounts for the position of the side chain, and behaves nearly identically to side-chain Voronoi cell volume.
+Finally, we investigated how Voronoi cell volume related to RSA and WCN.
+In this analysis, we focused on SC coordinates and considered different exponents for WCN.
+First, we found that RSA and cell volume were strongly correlated, with a mean correlation coefficient of 0.92 and a standard deviation of $0.02$ (see also Figure \ref{fig:best_wcn_vvol}D).
+Under the default definition of WCN, with power-law exponent $\alpha=-2$, we found that the mean correlation coefficient between WCN and cell volume was 0.89, with a standard deviation of 0.04.
+Thus, on average, the correlations between cell volume and RSA were 0.03 higher than the correlations between cell volume and side-chain WCN.
+We next identified the power-law exponents that resulted in the strongest correlations between WCN and cell volume for each protein.
+We found a mean exponent of $\alpha=-3.3$ (standard deviation $\sigma=0.3$).
+Using WCN with the best exponent value for each protein instead of fixed value $\alpha=-2$ for all, the correlation between WCN and cell volume increased to $0.94$ on average (standard deviation $0.01$).
+These results show that the default WCN definition, with exponent $-2$, is less similar to Voronoi cell volume than a more local definition with an exponent near $-3.3$.
+
+To summarize, our results show that the biophysically relevant reference point for amino-acid position is the side chain, not the commonly used C$_\alpha$ atom.
+Further, they highlight that RSA properly accounts for the position of the side chain, and behaves nearly identically to side-chain Voronoi cell volume, whereas WCN additional contains information about longer-range interactions not accounted for in the Voronoi cell volume.
 
@@
 \section*{Figures}
+
     \begin{figure}[h]
-        \begin{center}
-        \begin{tabular}{cc}
-            \includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
-        \end{tabular}
-        \end{center}
+        %\begin{center}
+        %\includegraphics[width=6in]{pipeline.png}
+        %\end{center}
+        \caption{Schematic of our analysis approach.
+(A, B) Evolutionary rates at each site in a protein correlate with measures of packing density, such as WCN (A) or Voronoi cell volume (B).
+Results are shown here for all amino acids in one representative protein (PDB ID: 1ONR, chain A) (C) For all proteins in the data set, the Spearman's correlation coefficients of ER--WCN and ER--Cell Volume relations are calculated and compared to each other.
+Each black point represents the two correlation values for one protein.
+The red dashed line represents the equality line for the absolute values of the correlation strengths.
+(D) Finally, we convert the set of correlation coefficients into distributions and compare their relative means.
+On average, WCN correlates more strongly with ER than cell volume does.}
+        \label{fig:pipeline}
+    \end{figure}
+
+    \begin{figure}
+        %\begin{center}
+        %\begin{tabular}{cc}
+            %\includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
+        %\end{tabular}
+        %\end{center}
         \caption{Absolute correlation of evolutionary rate with (A) Contact Number and (B) Weighted Contact Number for varying degrees of locality of these quantities.
@@
     \begin{figure}
-        \begin{center}
-        \includegraphics[width=6in]{voronoi_diagram.pdf}
-        \end{center}
+        %\begin{center}
+        %\includegraphics[width=6in]{WCN_fraction_of_residues_included.pdf}
+        %\end{center}
+        \caption{Fraction of residues that are included in the calculation of $\text{WCN}^\text{cutoff}$ when its correlation with site-specific evolutionary rate is maximized.
+For the vast majority of proteins, the highest correlation strengths are obtained when over 90\% of the residues in the protein are included in the calculation of $\text{WCN}^\text{cutoff}$.}
+        \label{fig:wcn_pwrl_cutoff}
+    \end{figure}
+
+    \begin{figure}
+        %\begin{center}
+        %\includegraphics[width=6in]{voronoi_diagram.pdf}
+        %\end{center}
         \caption{Example of a Voronoi tessellation in two dimensions.
@@
     \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
-        \end{center}
+        %\begin{center}
+        %\includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
+        %\end{center}
         \caption{Correlations and partial correlations of ER with various Voronoi cell properties.
@@
     \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
-        \end{center}
+        %\begin{center}
+        %\includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
+        %\end{center}
         \caption{Correlations and partial correlations of ER with Voronoi cell volume, WCN, and RSA.
@@
     \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
-        \end{center}
+        %\begin{center}
+        %\includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
+        %\end{center}
         \caption{Side-chain centers provide most informative reference points for both WCN and Voronoi tessellation.
@@
 (D) As in (B), but the correlations are calculated with RSA instead of with ER.}
+        \label{fig:best_wcn_vvol}
+    \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%This is only for figure plots without captions, to conform with Proteins: SFB manuscript style
+
+    \begin{figure}[h]
+        \begin{center}
+        \includegraphics[width=6in]{pipeline.png}
+        \end{center}
+        \caption{THIS IS FIGURE 1 ILLUSTRATION. THE CAPTIONS AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:pipeline}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \begin{tabular}{cc}
+            \includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
+        \end{tabular}
+        \end{center}
+        \caption{THIS IS FIGURE 2 ILLUSTRATION. THE CAPTIONS AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:cnwcnp}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=6in]{WCN_fraction_of_residues_included.pdf}
+        \end{center}
+        \caption{THIS IS FIGURE 3 ILLUSTRATION. THE CAPTION AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:wcn_pwrl_cutoff}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=6in]{voronoi_diagram.pdf}
+        \end{center}
+        \caption{THIS IS FIGURE 4 ILLUSTRATION. THE CAPTION AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:voronoi}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
+        \end{center}
+        \caption{THIS IS FIGURE 5 ILLUSTRATION. THE CAPTION AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:voronoi_ER_screen}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
+        \end{center}
+        \caption{THIS IS FIGURE 6 ILLUSTRATION. THE CAPTION AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
+        \label{fig:best_predictorER}
+    \end{figure}
+
+    \begin{figure}
+        \begin{center}
+        \includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
+        \end{center}
+        \caption{THIS IS FIGURE 7 ILLUSTRATION. THE CAPTION AND FIGURES HAVE BEEN SEPARATED TO COMPLY WITH THE RULES OF PROTEINS: SFB JOURNAL.}
         \label{fig:best_wcn_vvol}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figures
-
-\cleardoublepage
-\section*{Figures}
-    \begin{figure}[h]
-        \begin{center}
-        \begin{tabular}{cc}
-            \includegraphics[width=3.1in]{spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.1in]{spcor_cnwSC_r4sJC.pdf}
-        \end{tabular}
-        \end{center}
-        \caption{Absolute correlation of evolutionary rate with (A) Contact Number and (B) Weighted Contact Number for varying degrees of locality of these quantities.
-In (A), we vary the cutoff parameter $r_0$ in Eq.~\ref{eqn:cn} from 0 to 50\AA.
-In (B), we vary the exponent $\alpha$ in Eq.~\ref{eqn:wcn_pwrl} from $-30$ to 30.
-In each plot, the solid black line represents the mean correlation strength in the entire dataset of 209 proteins and the dashed black line indicates the median of the distribution.
-The green-shaded region together with the red-dashed lines represent the $25\%$ and $75\%$ quartiles of the correlation-strength distribution.
-Note that for the case of WCN with $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ with $\alpha<0$.
-In addition, $\rho$ is undefined at $\alpha=0$ and not shown in this plot.
-The parameter values at which the correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
-        \label{fig:cnwcnp}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=6in]{voronoi_diagram.pdf}
-        \end{center}
-        \caption{Example of a Voronoi tessellation in two dimensions.
-The red dots represent the seed points, and the black lines delineate the Voronoi cells.
-For protein structures, the tessellation is carried out in three dimensions.}
-        \label{fig:voronoi}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_voronoi_predictors_of_ER_screen_V.pdf}
-        \end{center}
-        \caption{Correlations and partial correlations of ER with various Voronoi cell properties.
-(A) Distributions of ER correlations with Voronoi cell properties sphericity, edge length, eccentricity, volume, and area.
-Note that all cell characteristics correlate positively with ER, except sphericity which correlates negatively with ER.
-We show here the correlations of ER with $(-1)\times \text{cell sphericity}$ so that the correlation distributions all appear on the same positive scale.
-The correlations for cell area and cell volume are not significantly different from each other, and are higher than the correlations for all other Voronoi cell properties (Table~\ref{tab:ttest-Voronoi}).
-(B) Distributions of partial correlations of ER with Voronoi cell properties, controlling for cell volume.
-All partial correlations are relatively small, with medians of approximately 0.1.
-Therefore, none of these cell properties provide much independent information about ER once cell volume is accounted for.}
-        \label{fig:voronoi_ER_screen}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_structural_predictors_of_ER_screen.pdf}
-        \end{center}
-        \caption{Correlations and partial correlations of ER with Voronoi cell volume, WCN, and RSA.
-(A) Distributions of ER correlations with Voronoi cell volume, WCN (using side-chain and C$_\alpha$  coordinates, denoted by SC and CA, respectively), and RSA.
-Note that the WCN measures correlate negatively with ER.
-We show here correlations of ER with $(-1)\times \text{WCN}$ so that the correlation distributions all appear on the same positive scale.
-The correlations of ER with WCN (SC) are significantly higher than the correlations of ER with all other quantities shown (Table~\ref{tab:ttest-best}).
-(B) Distributions of partial correlations of ER with WCN and RSA, controlling for cell volume.
-The partial correlation of ER with $-\text{WCN}$ are substantial, with median values of 0.32 (side-chain WCN) and 0.21 (C$_\alpha$ WCN).
-By contrast, the partial correlations of ER with RSA largely vanish, with a median value of 0.09.} \label{fig:best_predictorER}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=6.5in]{best_wcn_vvol_boxplot.pdf}
-        \end{center}
-        \caption{Side-chain centers provide most informative reference points for both WCN and Voronoi tessellation.
-(A) Distribution of correlations of WCN with ER, for seven different coordinate sets according to which WCN was calculated: SC, AA, CB, CA, N, C, O.
-Each coordinate set represents a different way of identifying the reference location of each residue.
-For SC (Side Chain) and AA (entire Amino Acid), the reference point is given, respectively, by the geometric average coordinates of the Side Chain (SC) atoms and the entire Amino Acid (AA) atoms.
-The latter include the backbone but exclude any Hydrogen.
-The coordinate sets CB, CA, N, C, and O use the respective atom in the amino acid as the reference point.
-(B) As in (A), but using Voronoi Cell Volume instead of WCN.
-(C) As in (A), but the correlations are calculated with RSA instead of with ER.
-(D) As in (B), but the correlations are calculated with RSA instead of with ER.}
-        \label{fig:best_wcn_vvol}
-    \end{figure}
-
 \end{document}

### 2017-01-19 manuscript/voronoi/trash/mnv.tex
final article source files added

--- 2015-06-06 manuscript/voronoi/mnv.tex
@@
 %\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
-\Title{Dissecting the relationship between protein structure and sequence variation}
+\Title{Dissecting the Role of Long-Range Amino Acid Interactions and Local Packing Density in Protein Sequence Evolution}
 \vfill
@@
 We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein.
-Using a dataset of 209 monomeric enzymes, we show that the long-range amino acid interactions can explain $\sim000\%$ of variability observed in protein sequence, whereas only $\sim000\%$ of sequence variability can be attributed to site-specific packing density.
+Using a dataset of 209 monomeric enzymes, we show that the long-range amino acid interactions can explain $\sim10\%$ of variability observed in protein sequence, whereas $\sim30\%$ of sequence variability can be attributed to site-specific packing density.
 
@@
     A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties.
-Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
+Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}.
 \\
@@
 
-    It is notable that \citet{halle_flexibility_2002} laid a theoretical foundation for the existence of a inverse relationship between the site-specific flexibility -- often represented by the $C_\alpha$ atomic B factors -- and the local packing density in proteins.
-Such an inverse relationship could be intuitively understood knowing that the inverse of local packing density can serve as an estimate of the accessible free volume to each site in the protein's 3-dimensional structure, and thus a proxy measure of site-specific flexibility in protein.
-A crude measure of local packing density for each site can be obtained through a quantity widely known as Contact Number \citep[e.g.,][]{liao_protein_2005}.
+    A crude measure of site-specific packing density for each site in protein is often obtained through a quantity widely known as Contact Number \citep[e.g.,][]{liao_protein_2005}.
+Alternative packing density measures have also been proposed leading to better predictions of sequence evolution and site-specific flexibility, a prime example of which is the Weighted Contact Number (WCN) proposed by \citet{lin_deriving_2008}.
+These proxy measures of packing density however, involve adjustable free parameters in their definitions.
+Moreover, the packing density as defined by these quantities is more representative of the local environment around the site of interest, and does necessarily correspond to the site-specific packing density.
+\\
+
+    Motivated by the existing gaps in the current understanding of the role of site-specific packing density and long-range amino acid interactions in protein sequence evolution, here we derive a new set of structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions.
+In particular, we propose a methodology to derive a minimally-biased measure of local packing density based upon which the role of long-range amino-acid interactions on sequence variability can be segregated and isolated from the effects of site-specific packing density.
+This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.
+In this regard, the two commonly-used definitions of local packing density (CN \& WCN) may not be considered {\it site-specific}.
+An important question therefore remains unanswered as to what extent the site-specific packing density and long-range interactions influence sequence variability, independently of each other.
+\\
+
+\section{Methods}
+\label{sec:methods}
+
+    \subsection*{Local Packing Density Definition and Measurement}
+
+    The local packing density is commonly measured by a quantity known as Contact Number (CN).
 In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around the site \citep[e.g.,][]{franzosa_structural_2009},
@@
 
-    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable parameter: the exponent of the power-law kernel ($\alpha$).
-The value of the exponent that results in the best predictions of B factors appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
+    Although WCN is in general a better predictor of site-specific sequence variability, the proposed definition of WCN still involves an adjustable parameter: the exponent of the power-law kernel ($\alpha$).
+The value of the exponent that results in the best predictions appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.
 A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability.
 The specific value $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, independently of the local packing density.
-Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009, shahmoradi_predicting_2014}, aside from local packing density.
-In this regard, the two commonly-used definitions of packing density (CN \& WCN) may not be considered {\it local}.
-An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, independently of each other.
+Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009, shahmoradi_predicting_2014}, aside from site-specific packing density.
 \\
-
-    Motivated by the existing gaps in the current understanding of the role of flexibility, packing density and long-range interactions on sequence-structure relations in proteins, here we derive a new set of site-specific structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions.
-In particular, we propose a methodology to derive a minimally-biased measure of local packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.
-This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.
 
@@
 
-\section{Protein Dataset and Site-Specific Structure$/$Sequence Variability Measures}
+    \subsection*{Protein Dataset and Site-Specific Structure$/$Sequence Variability Measures}
     \label{sec:methods}
@@
     The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}.
-(XX a comparison of CN cutoff distance with average amino acid neighbor distance is needed here to quantify evidence for the following sentence.
-An additional figure may also be necessary plotting cutoff distance vs.
-average NN distance for each protein XX).
 Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations.
@@
 
-    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility in Proteins}
+    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility}
 
@@
 Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for.
-The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (c.f., Figure \ref{fig:voronoi_ER_cond}).
+The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (Figure \ref{fig:voronoi_ER_cond}).
 Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled.
@@
 For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains.
-Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}.
+Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability.
+In particular, a pairwise t-test between the correlation strengths of r4sJC-WCN and the correlation of r4sJC with the inverse of Voronoi cell volume yields a p-value of $<10^{-16}$.
 The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn.
@@
     Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most widely studied structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes.
-Examples of important sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA).
-In addition, we have derived a new set of site-specific characteristics from Voronoi Tessellation of protein 3D structure, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).
+Examples of important sequence--structure relations include the correlations of measures of sequence variability -- such as evolutionary rates (e.g., {\it r4sJC}) and sequence entropy -- with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA).
+In addition, we derived a new set of site-specific characteristics from the Voronoi partitioning of protein's 3D structure, some of which are capable of explaining sequence variability equally well or better than several structural quantities that were previously considered in the literature -- including B factor, RSA, and the traditional definitions of CN and WCN using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).
+\\
 
-    We also argued that the commonly used measures of local packing density, most importantly the Contact Number (CN) and the Weighted Contact Number (WCN), involve free adjustable parameters in the definitions which can be fine-tuned for each individual protein to obtain the best correlations between CN$/$WCN and sequence evolutionary rates.
-Although the best parameter values for CN \& WCN vary from one protein structure to another, the median values over the entire dataset indicate a non-negligible selection pressure due to long-range amino acid interactions on sequence evolution, independently of the local packing density.
+    The commonly used measures of local packing density, most importantly CN and WCN, involve free adjustable parameters in their definitions that can be fine-tuned for each individual protein to obtain the strongest correlations between CN$/$WCN and sequence variability.
+Although the best parameter values for CN \& WCN vary from one protein structure to another, the median of the values over the entire dataset clearly indicate a non-negligible selection pressure on protein sequence solely due to long-range amino acid interactions, independently of the local packing density.
+Unlike CN and WCN, the site-specific packing density as defined by the inverse of Voronoi cell volume in this work, represents only the contributions from the nearest neighbors of each site in protein, that is, the first coordination shell.
+\\
+
+    Although the site-specific packing density as defined by the inverse of the Voronoi cell volume excludes the effects of long-range interactions beyond the first coordinations shell, it remains as best predictor of sequence variability in our dataset among all other structural determinants, most importantly RSA.
+Contrary to the findings of this work however, some other studies \citep[e.g., ][]{franzosa_structural_2009, scherrer_modeling_2012, shahmoradi_predicting_2014} have argued for RSA as the main determinant of sequence evolution with local packing density having a peripheral role.
+Our conclusion is that RSA and the local packing density -- once corrected for long-range interactions -- in principle represent the same characteristics of the local environment in protein, that is, both quantities are proxy measures of the number of neighboring amino acids in the first coordination shell.
+This argument is also further supported by running a pairwise t-test between the correlation strengths of RSA with evolutionary rates and the correlation strengths of the inverse Voronoi cell volume with evolutionary rates for the entire protein dataset, with a resulting p-value $\sim0.29$.
+An important question however remains unanswered in this work and merits further study, as to what extent the findings of the presented analysis -- based on a dataset of 209 monomeric enzymes -- can be generalized to other types of proteins in nature.
 
 \begin{appendices}
-
-\section{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
-\label{app:best_bf}
-
-    For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain results in the best correlations with other structural and sequence properties.
-Specifically, the average side chain B factor outperforms the commonly used $C_\alpha$ B factor in predicting RSA, \ddg rate, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
-\\
-
-    % Figure goes here xx
-
-    The observed improvements in correlations of average side-chain B factor (vs.
-C$_\alpha$ B factor) with other structural properties also merit further attention.
-It was discussed in Section \ref{sec:voro} and depicted in the plots of Figure \ref{fig:best_bf} that in general, as one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve.
-In particular, the use of average side-chain B factor turned out to result in the highest correlation strengths with other site-specific properties, implying that this average B factor is likely the best representation of the overall amino acid fluctuations and flexibility in a given site in protein.
-The definition of B factor and its derivation from Debye-Waller factor has been already discussed in Chapter \ref{chap:sschar}, Eqns.
-\ref{eqn:dwf_def}--\ref{eqn:bf_def}.
-\\
-
-    The mean-square-displacement $\langle u^2 \rangle$ in Eqn. \ref{eqn:bf_def} can be decomposed into four contributing components \cite{frauenfelder_temperature-dependent_1979},
-
-    \begin{equation}
-        \label{eqn:bf_comp}
-        \langle u^2 \rangle = \langle u^2 \rangle_c + \langle u^2 \rangle_d + \langle u^2 \rangle_{ld} + \langle u^2 \rangle_v,
-    \end {equation}
-
-    \noindent in which subscripts $c, d, ld, v$ refer to fluctuations due to conformational substates, diffusion, lattice disorder, and thermal vibrations respectively.
-The second term $\langle u^2 \rangle_d$ is generally negligible and can be ignored in Eqn.
-\ref{eqn:bf_comp}.
-Of particular interest to this study is the first term, which is also typically the major contributor to the overall value of the atomic B factor, specially in high-resolution X-ray crystallography of proteins.
-This term represents the positional displacements of the atom of interest together with other atoms in the amino acid between many different conformational substates of the protein, with the transition probability between the substates governed by the Boltzmann distribution.
-Compared to atomic coordinates, there are comparatively fewer restraints on the atomic B factors during X-ray crystallography refinement process, and thus in this regard B factor is generally considered as the {\it error sinks} for static and dynamic disorder and various kinds of model errors in the refinement process \cite{read_structure-factor_1990}.
-The noise and model uncertainty contributions to the atomic B factors in particular increase with decreasing the resolution of the X-ray crystallography.
-Better resolution in general corresponds to lower average B factors for the entire structure of the protein \cite{read_structure-factor_1990}.
-\\
-
-    Although the extraction of conformational fluctuations from noise in B factors seems a daunting task \cite{read_structure-factor_1990}, the effects of noise, model error and uncertainties due to limited X-ray crystallography resolution can be minimized by averaging B factors over the entire amino acid in a given site: To expand on this, consider the contribution of conformational fluctuations between different substates to be approximately the same for all atoms in the amino acid.
-The conformational fluctuations can be regarded as the collective motion of all atoms in the amino acid, on top of which there are noise fluctuations in each of the atoms.
-These collective motions are the type of fluctuations in B factors that are expected to reflect the biologically relevant and important factors for the proper functioning of the protein.
-The stochastic noise in the fluctuations is often assumed to have an isotropic Gaussian origin.
-Therefore, averaging over the atomic B factors in each individual amino acid essentially results in higher Signal-to-Noise Ratio (SNR) in the measurement of the amino acid conformational fluctuations.
-Figure \ref{fig:bfC_bfAA_ratio} illustrates how this averaging over all atomic B factors increases the SNR in measuring the fluctuations due to conformational substate transitions of the amino acid.
-\\
-
-    % Figure goes here
-
-    To expand further on this, a simple argument may be given to explain the observed strongly-positive approximately-linear correlation between the two parameters in the plot of Figure \ref{fig:bfC_bfAA_ratio}.
-The contributions to the atomic B factor values of the $i^{th}$ atom in the amino acid in the $j^{th}$ site in a given protein can be assumed to originate from two major sources: conformational substates and stochastic noise due to model uncertainties in refinement process and limited resolution of the X-ray crystallography,
-
-    \begin{equation}
-        \label{eqn:bf_comp}
-        \langle u^2 \rangle_{ij} = \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij}.
-    \end {equation}
-
-    For simplicity and without loss of generality, one can assume that the contribution of fluctuations due to conformational substate transitions is approximately the same for all atomic B factors in a given amino acid residing the $j^{th}$ site.
-In other words, the term $\langle u^2 \rangle_{\text{substates},ij}$ in the above equation has almost the same value $\langle u^2 \rangle_{\text{substates},j}$ for all atoms in the amino acid in the $j^{th}$ site in protein.
-Thus, the average B factor for the entire amino acid molecule of size $N_j$ atoms would be,
-
-    \begin{eqnarray}
-        \label{eqn:bf_avg}
-        \langle u^2 \rangle_{j} &=& \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
-                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
-                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \mu_{\text{noise},j}
-    \end {eqnarray}
-
-    \noindent in which $\mu_{\text{noise},j}$ is the average noise in the $j^{th}$ amino acid.
-The ratio of the B factor of the $ij^{th}$ atom to the average B factor of the $j^{th}$ site in protein can be approximated as,
-
-    \begin{eqnarray}
-        \label{eqn:bf_ratio}
-        \frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} &\simeq& \frac{\langle u^2 \rangle_{substates,j} + \langle u^2 \rangle_{noise,ij}}{\langle u^2 \rangle_{substates,j} + \mu_{noise,j}} \\
-                                 &=& \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \nonumber \\
-                                 &+& \bigg(\frac{\langle u^2 \rangle_{\text{noise},ij}}{\langle u^2 \rangle_{\text{substates},j}} \bigg)
-                                 \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \label{eqn:bf_ratio1} \\
-                                 &\simeq& 1 - \frac{\mu_{noise,j}}{\langle u^2 \rangle_{\text{substates},j}}, \label{eqn:bf_ratio2}
-    \end {eqnarray}
-
-    \noindent where from line \ref{eqn:bf_ratio1} to \ref{eqn:bf_ratio2}, an assumption was made that the second term in line \ref{eqn:bf_ratio1} could be neglected compared to the first term and that the noise compared to conformational fluctuation is small, that is, $\mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j} < 1$ (an error of $0.2\AA$ corresponds approximately to $1\AA$ increase in B factor \cite{read_structure-factor_1990}).
-Knowing that the average noise across different amino acids is approximately the same \cite{frauenfelder_temperature-dependent_1979}, that is $\mu_{noise,j}\sim \mu_{noise}$, and that the noise due to X-ray crystallography almost negatively linearly correlates with crystallography resolution in the range $\sim1-3~[\AA]$ \cite{read_structure-factor_1990}, that is $\mu_{noise}\propto -\text{resolution}$, a positive approximately-linear relationship between the average of the B factor ratios over the entire amino acids in the protein structure and the X-ray crystallography resolution would be obtained,
-
-    \begin{eqnarray}
-        \frac{\text{BF}_C}{\text{BF}_{AA}} &=& \frac{1}{L}\sum_{j=1}^{L}\frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} \label{eqn:avg_bf_ratio} \\
-                                           &\propto& -\mu_{noise} \sum_{j=1}^{L}\frac{1}{\langle u^2 \rangle_{substates,j}} \label{eqn:avg_bf_ratio1} \\
-                                           &\propto& \text{resolution}
-    \end {eqnarray}
-
-    \noindent in which $L$ represents the length of the protein sequence.
-The summation term in line \ref{eqn:avg_bf_ratio1} would not influence this linear relationship, causing only scatter in the relation, so long as the length of the protein not does impose limitations on the resolution of X-ray crystallography of proteins.
-In general, however this may not be the case.
-For the sample of 209 proteins considered here, there exists indeed a weak Spearman's correlation coefficient of $\rho\sim0.2$ between protein length ($L$) and resolution.
-Figure \ref{fig:bfC_bfAA_ratio} illustrates the relationship between the average B factors ratio and the resolution in the dataset, using atom $C$ in the backbone of all amino acids in proteins representing the $i^{th}$ atom in the notation of Eqn.
-\ref{eqn:avg_bf_ratio}.
-It is also notable that the the atomic fluctuations due to conformational substates may not be exactly the same for all atoms in an amino acid in a given site in protein.
-Indeed, one may expect the conformational fluctuations in the backbone atoms would be less significant compared to conformational fluctuations of side-chain atoms.
-\\
-
-    Although averaging B factor over the entire amino acid atoms would reduce the noise further than averaging over side-chain atoms, the functionally important conformational fluctuations that are better captured by the side-chain atomic B factors would compensate for the increase in the noise, such that overall, the B factors averaged over side-chain atoms results in slightly better correlations with sequence variability and other relevant structural characteristics depicted in Figure \ref{fig:best_bf}.
-\\
 
@@
 
-    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
+    Here in this work, all possible choices of the representative set of atomic coordinates were considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells.
 Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties.
 These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$).
-In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains.
-Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
+In addition, representative coordinates for each site in protein can be also calculated by averaging over the coordinates of all heavy atoms in the side chains.
+Also calculated was a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together.
 In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid.
-The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations.
+The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate as defined by \citet{echave_relationship_2014}, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and the average backbone Hydrogen bond energy with different measures of WCN and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn} and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations.
 The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}.
 Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all.
+The Hydrogen bonds were identified and the corresponding energies were calculated according to the prescription of \citet{kabsch_dictionary_1983}.
 \\
@@
 
-    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
+    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible sets of coordinates, the average over coordinates of all heavy atoms of each individual side-chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates.
 Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
@@
     Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells.
-Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}).
+For brevity, only the results for the cell area are given here.
+Specifically, the cell area from average side-chain coordinates outperforms cell area calculated based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}).
 \\
@@
 \ref{eqn:sphericity}, which turns out to behave differently for open and closed cells.
-This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates.
+%This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates.
 For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths.
@@
     % Figure goes here xx
+
+%\section{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
+%\label{app:best_bf}
+%
+%    For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side-chain results in the best correlations with other structural and sequence properties (Figure \ref{fig:best_bf}).
+Specifically, the average side chain B factor outperforms the commonly used $C_\alpha$ B factor in predicting RSA, \ddg rate, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
+\\
+%
+%    % Figure goes here xx
+%
+%    The observed improvements in correlations of average side-chain B factor (vs.
+C$_\alpha$ B factor) with other structural properties also merit further attention.
+As one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve.
+In particular, the use of average side-chain B factor turned out to result in the highest correlation strengths with other site-specific properties, implying that this average B factor is likely the best representation of the overall amino acid fluctuations and flexibility in a given site in protein.
+The mean-square-displacement of an atom as measured by B factor can be decomposed into four contributing components \cite{frauenfelder_temperature-dependent_1979},
+%
+%    \begin{equation}
+%        \label{eqn:bf_comp}
+%        \langle u^2 \rangle = \langle u^2 \rangle_c + \langle u^2 \rangle_d + \langle u^2 \rangle_{ld} + \langle u^2 \rangle_v,
+%    \end {equation}
+%
+%    \noindent in which subscripts $c, d, ld, v$ refer to fluctuations due to conformational substates, diffusion, lattice disorder, and thermal vibrations respectively.
+The second term $\langle u^2 \rangle_d$ is generally negligible and can be ignored in Eqn.
+\ref{eqn:bf_comp}.
+Of particular interest to this study is the first term, which is also typically the major contributor to the overall value of the atomic B factor, specially in high-resolution X-ray crystallography of proteins.
+This term represents the positional displacements of the atom of interest together with other atoms in the amino acid between many different conformational substates of the protein, with the transition probability between the substates governed by the Boltzmann distribution.
+Compared to atomic coordinates, there are comparatively fewer restraints on the atomic B factors during X-ray crystallography refinement process, and thus in this regard B factor is generally considered as the {\it error sinks} for static and dynamic disorder and various kinds of model errors in the refinement process \cite{read_structure-factor_1990}.
+The noise and model uncertainty contributions to the atomic B factors in particular increase with decreasing the resolution of the X-ray crystallography.
+Better resolution in general corresponds to lower average B factors for the entire structure of the protein \cite{read_structure-factor_1990}.
+\\
+%
+%    Although the extraction of conformational fluctuations from noise in B factors seems a daunting task \cite{read_structure-factor_1990}, the effects of noise, model error and uncertainties due to limited X-ray crystallography resolution can be minimized by averaging B factors over the entire atoms of an amino acid in a given site: To expand on this, consider the contribution of conformational fluctuations between different substates to be approximately the same for all atoms in the amino acid.
+The conformational fluctuations can be regarded as the collective motion of all atoms in the amino acid, on top of which there are noise fluctuations in each of the atoms.
+These collective motions are the type of fluctuations in B factors that are expected to reflect the biologically relevant and important factors for the proper functioning of the protein.
+The stochastic noise in the fluctuations is often assumed to have an isotropic Gaussian origin.
+Therefore, averaging over the atomic B factors in each individual amino acid essentially results in higher Signal-to-Noise Ratio (SNR) in the measurement of the amino acid conformational fluctuations.
+Figure \ref{fig:bfC_bfAA_ratio} illustrates how this averaging over all atomic B factors increases the SNR in measuring the fluctuations due to conformational substate transitions of the amino acid.
+\\
+%
+%    % Figure goes here
+%
+%    To expand further on this, a simple argument may be given to explain the observed strongly-positive approximately-linear correlation between the two parameters in the plot of Figure \ref{fig:bfC_bfAA_ratio}.
+The contributions to the atomic B factor values of the $i^{th}$ atom in the amino acid in the $j^{th}$ site in a given protein can be assumed to originate from two major sources: conformational substates and stochastic noise due to model uncertainties in refinement process and limited resolution of the X-ray crystallography,
+%
+%    \begin{equation}
+%        \label{eqn:bf_comp}
+%        \langle u^2 \rangle_{ij} = \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij}.
+%    \end {equation}
+%
+%    For simplicity and without loss of generality, one can assume that the contribution of fluctuations due to conformational substate transitions is approximately the same for all atomic B factors in a given amino acid residing the $j^{th}$ site.
+In other words, the term $\langle u^2 \rangle_{\text{substates},ij}$ in the above equation has almost the same value $\langle u^2 \rangle_{\text{substates},j}$ for all atoms in the amino acid in the $j^{th}$ site in protein.
+Thus, the average B factor for the entire amino acid molecule of size $N_j$ atoms would be,
+%
+%    \begin{eqnarray}
+%        \label{eqn:bf_avg}
+%        \langle u^2 \rangle_{j} &=& \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
+%                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
+%                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \mu_{\text{noise},j}
+%    \end {eqnarray}
+%
+%    \noindent in which $\mu_{\text{noise},j}$ is the average noise in the $j^{th}$ amino acid.
+The ratio of the B factor of the $ij^{th}$ atom to the average B factor of the $j^{th}$ site in protein can be approximated as,
+%
+%    \begin{eqnarray}
+%        \label{eqn:bf_ratio}
+%        \frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} &\simeq& \frac{\langle u^2 \rangle_{substates,j} + \langle u^2 \rangle_{noise,ij}}{\langle u^2 \rangle_{substates,j} + \mu_{noise,j}} \\
+%                                 &=& \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \nonumber \\
+%                                 &+& \bigg(\frac{\langle u^2 \rangle_{\text{noise},ij}}{\langle u^2 \rangle_{\text{substates},j}} \bigg)
+%                                 \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \label{eqn:bf_ratio1} \\
+%                                 &\simeq& 1 - \frac{\mu_{noise,j}}{\langle u^2 \rangle_{\text{substates},j}}, \label{eqn:bf_ratio2}
+%    \end {eqnarray}
+%
+%    \noindent where from line \ref{eqn:bf_ratio1} to \ref{eqn:bf_ratio2}, an assumption was made that the second term in line \ref{eqn:bf_ratio1} could be neglected compared to the first term and that the noise compared to conformational fluctuation is small, that is, $\mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j} < 1$ (an error of $0.2\AA$ corresponds approximately to $1\AA$ increase in B factor \cite{read_structure-factor_1990}).
+Knowing that the average noise across different amino acids is approximately the same \cite{frauenfelder_temperature-dependent_1979}, that is $\mu_{noise,j}\sim \mu_{noise}$, and that the noise due to X-ray crystallography almost negatively linearly correlates with crystallography resolution in the range $\sim1-3~[\AA]$ \cite{read_structure-factor_1990}, that is $\mu_{noise}\propto -\text{resolution}$, a positive approximately-linear relationship between the average of the B factor ratios over the entire amino acids in the protein structure and the X-ray crystallography resolution would be obtained,
+%
+%    \begin{eqnarray}
+%        \frac{\text{BF}_C}{\text{BF}_{AA}} &=& \frac{1}{L}\sum_{j=1}^{L}\frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} \label{eqn:avg_bf_ratio} \\
+%                                           &\propto& -\mu_{noise} \sum_{j=1}^{L}\frac{1}{\langle u^2 \rangle_{substates,j}} \label{eqn:avg_bf_ratio1} \\
+%                                           &\propto& \text{resolution}
+%    \end {eqnarray}
+%
+%    \noindent in which $L$ represents the length of the protein sequence.
+The summation term in line \ref{eqn:avg_bf_ratio1} would not influence this linear relationship, causing only scatter in the relation, so long as the length of the protein not does impose limitations on the resolution of X-ray crystallography of proteins.
+In general, however this may not be the case.
+For the sample of 209 proteins considered here, there exists indeed a weak Spearman's correlation coefficient of $\rho\sim0.2$ between protein length ($L$) and resolution.
+Figure \ref{fig:bfC_bfAA_ratio} illustrates the relationship between the average B factors ratio and the resolution in the dataset, using atom $C$ in the backbone of all amino acids in proteins representing the $i^{th}$ atom in the notation of Eqn.
+\ref{eqn:avg_bf_ratio}.
+It is also notable that the the atomic fluctuations due to conformational substates may not be exactly the same for all atoms in an amino acid in a given site in protein.
+Indeed, one may expect the conformational fluctuations in the backbone atoms would be less significant compared to conformational fluctuations of side-chain atoms.
+\\
+%
+%    Although averaging B factor over the entire amino acid atoms would reduce the noise further than averaging over side-chain atoms, the functionally important conformational fluctuations that are better captured by the side-chain atomic B factors would compensate for the increase in the noise, such that overall, the B factors averaged over side-chain atoms results in slightly better correlations with sequence variability and other relevant structural characteristics depicted in Figure \ref{fig:best_bf}.
+\\
 
@@
         \caption{Average absolute Spearman's correlation strengths of Contact Number (CN, as defined by Eqn.
-\ref{eqn:wcnh} using Heaviside kernel) and the Weighted Contact Number (WCN, as defined by Eqn.
+\ref{eqn:cn} using Heaviside kernel in Eqn.
+\ref{eqn:heaviside}) and the Weighted Contact Number (WCN, as defined by Eqn.
 \ref{eqn:wcn_pwrl} using a power-law kernel) with site-specific evolutionary rates, for different values of the free parameters of the two kernels ($r_0$ \& $\alpha$ respectively).
@@
 Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$.
-The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
         \label{fig:best_predictorER}
@@
 Note that the Spearman's $\rho$ is insensitive to the use of inverse WCN in place of WCN.
-The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
+The abbreviation {\it SC} refers to the use of average Side-Chain coordinates wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
         \label{fig:best_predictorER_given_volume}
@@
 
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{best_bf/select_variables/boxplot_bf_all_in_one.pdf}
-        \end{center}
-        \caption{A comparison of the correlation strength of 6 different measures of B factor with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_voro/zr4s_JC_voro.png}
+%        \end{center}
+%        \caption{General behavior of Voronoi cell characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The black \& orange curves represent respectively the general behaviors of closed \& open Voronoi cell characteristics.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+%        \label{fig:zr4s_JC_voro}
+%    \end{figure}
+%
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
+%        \end{center}
+%        \caption{General behavior of site-specific structural characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
+The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
+The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
+%        \label{fig:zr4s_JC_nonvoro}
+%    \end{figure}
+
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{best_bf/select_variables/boxplot_bf_all_in_one.pdf}
+%        \end{center}
+%        \caption{A comparison of the correlation strength of 6 different measures of B factor with 6 coordinate-independent structural or sequence properties for 209 proteins in dataset.
 Shown on the horizontal axes, are the 6 representative atomic B factors: {\it SC, AA, CB, CA, N, C, O} used as flexibility measures of individual sites in proteins.
 The two variables {\it SC} \& {\it AA} stand respectively for the average B factor of all Side Chain (SC) atoms and the entire Amino Acid (AA) atoms, excluding hydrogens.}
-        \label{fig:best_bf}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{mean_bfC_bfAA_ratio_resol.pdf}
-        \end{center}
-        \caption{An illustration of the strong positive correlation of X-ray crystallography resolution with the ratio of the backbone $C$ atomic B factor to the average amino acid B factor ($\text{BF}_C/\text{BF}_{AA}$), averaged over all sites in individual proteins, highlighting the significant contributions of noise and model errors to atomic B factor values.
+%        \label{fig:best_bf}
+%    \end{figure}
+%
+%    \begin{figure}
+%        \begin{center}
+%        \includegraphics[width=5.5in]{mean_bfC_bfAA_ratio_resol.pdf}
+%        \end{center}
+%        \caption{An illustration of the strong positive correlation of X-ray crystallography resolution with the ratio of the backbone $C$ atomic B factor to the average amino acid B factor ($\text{BF}_C/\text{BF}_{AA}$), averaged over all sites in individual proteins, highlighting the significant contributions of noise and model errors to atomic B factor values.
 The Spearman's correlation coefficient between the two quantities is $\rho\sim0.76$.
@@
 Each filled circle in the plot represents one protein in the dataset of 209 enzymes used in this work.}
-        \label{fig:bfC_bfAA_ratio}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_voro/zr4s_JC_voro.png}
-        \end{center}
-        \caption{General behavior of Voronoi cell characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
-The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
-The black \& orange curves represent respectively the general behaviors of closed \& open Voronoi cell characteristics.
-The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
-        \label{fig:zr4s_JC_voro}
-    \end{figure}
-
-    \begin{figure}
-        \begin{center}
-        \includegraphics[width=5.5in]{adjacent_averaging_screen/only_nonvoro/zr4s_JC_nonvoro.png}
-        \end{center}
-        \caption{General behavior of site-specific structural characteristics versus normalized site-specific evolutionary rates among all sites in all 209 proteins in dataset.
-The red curves in each plot is obtained by adjacent-averaging of every $3000$ sites.
-The background heat map in each plot is a 2D density plot of all 75755 amino acid sites in all 209 proteins, showing the overall distribution of sites about the average curve.}
-        \label{fig:zr4s_JC_nonvoro}
-    \end{figure}
+%        \label{fig:bfC_bfAA_ratio}
+%    \end{figure}
 
@@
     \begin{table}[htbp]
-    \caption{The proportion of variance explained by local packing density (as measured by the inverse volume of Voronoi cells) and long-range amino acid interactions (as described in Sec.
-\ref{sec:long_range}, for site-specific sequence variability measures.
-The subscripts and superscripts to each value represent the $25\%$ percentile range below and above the median value of the distribution.
+    \caption{The percentage of variance of site-specific evolutionary rates (r4sJC) explained by site-specific packing density (represented by the inverse volume of Voronoi cells) and long-range amino acid interactions (as described in Sec.
+\ref{sec:long_range}).
+For each structural quantity, the first, second (median) and the third quartiles of the percentage distribution of the explained variance are reported on each row, for the entire dataset of 209 proteins.
 \label{tab:best_params}}
@@
     \centerline{
-    \begin{tabular}{|l|c|c|}
+    \begin{tabular}{|l|c|c|c}
       \hline
-      Correlation with  & local packing density & long-range interactions   \\ \hline
-      r4sJC             & $14.3^{+5.3}_{-4.0}$ ($\rho\sim0.64^{+0.06}_{-0.06}$) & $-2.3^{+0.8}_{-0.4}$ ($\rho\sim0.65^{+0.05}_{-0.07}$)  \\
-      Seq. Entropy      & $12.4^{+5.5}_{-2.6}$ ($\rho\sim0.55^{+0.06}_{-0.06}$) & $-2.2^{+0.8}_{-0.4}$ ($\rho\sim0.55^{+0.07}_{-0.06}$)  \\
-      %B factor          & $13.1^{+4.0}_{-2.2}$ ($\rho\sim0.70^{+0.05}_{-0.06}$) & $-2.2^{+0.5}_{-0.4}$ ($\rho\sim0.71^{+0.05}_{-0.06}$)  \\
+      \multirow{2}{*}{structural quantity} & variance explained ($25\%$ quartile) & variance explained (median) & variance explained ($75\%$ quartile)  \\ \hline
+      site-specific packing density      & $27\%$ & $34\%$ & $41\%$  \\
+      long-range amino-acid interactions & $5\%$  & $10\%$ & $16\%$  \\
       \hline
